### 2013-10-24 paper/paper.tex
creating folder for the paper

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{natbib}

\oddsidemargin 0in
\evensidemargin 0in
\textwidth= 6.5in
\topmargin -0.50in
\textheight 9in

\usepackage{latexsym,amssymb,amsmath,color}
\newcommand{\revised}[1]{\textcolor{blue}{{#1}}}
\newcommand{\comment}[1]{\textcolor{red}{{#1}}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\xxi}{\vec{\xi}}
\newcommand{\xxii}{\xxi^{(\scriptscriptstyle{-}\scriptstyle{i})}}
\newcommand{\var}[1]{{\mathrm{Var}}\left[ {#1} \right]}
\newcommand{\ip}[2]{\left( {#1}, {#2} \right)}
\newcommand{\U}{\mathcal{U}}
\newcommand{\normim}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega^{*})}}
\newcommand{\avemu}[1]{\mathrm{E}\left({#1}\right)}
\newcommand{\ave}[1]{\left\langle {#1} \right\rangle}
\newcommand{\prob}[1]{\mathrm{Prob}\left\{ {#1} \right\}}
\newcommand{\ind}[1]{\mathrm{\chi}_{\scriptscriptstyle {#1} }}
\newcommand{\NISP}{\mathcal{S}}
\newcommand{\ipmu}[2]{\left( {#1}, {#2} \right)_\mu}
\newcommand{\norm}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega)}}
\newcommand{\pard}[2]{\frac{\partial{#1}}{\partial{#2}}}
\def\xbold{{\vec{x}}}
\renewcommand{\vec}[1]{{\mathchoice
                     {\mbox{\boldmath$\displaystyle{#1}$}}
                     {\mbox{\boldmath$\textstyle{#1}$}}
                     {\mbox{\boldmath$\scriptstyle{#1}$}}
                     {\mbox{\boldmath$\scriptscriptstyle{#1}$}}}}
                     %

\begin{document}
\begin{center}{\Large\bf Tsunami Uncertainty Quantification}\vskip 5pt

\bf Preliminary results
\end{center}
\vskip 15pt

\input{intro}

\input{tsunami}

\clearpage
%\input{inference}

\input{uqpce}
\input{results}

\clearpage

\clearpage
\bibliographystyle{plainnat}
\bibliography{biblio}
\end{document}

### 2013-10-24 paper/analysis.tex
adding tex files

\subsection{Statistical and sensitivity analysis}
\label{sec:forward}
We now exploit the PC surrogate to study the statistics of the 
water surface elevation, and to quantify its sensitivity to the
uncertain input parameters and thus to anticipate their impact on
the inverse problem.  In doing so, it is emphasized that 
no additional Geoclaw simulations were needed to obtain
the information presented below, rather it is obtained either directly 
from the coefficients of the PC surrogate or by sampling the corresponding
representations for different values of $\xxi$.

Figure~\ref{fig:ave} shows the evolution of
PC mean water surface elevation and its two standard deviations
bounds at the four gauge locations as indicated in each panel.  
The standard deviation is insignificant during the two hours
of the simulations and therefore was not included in the plots.
This standard deviation however increases as the tsunami increase
and its effect approaches the gauge locations. 
To quantity the contribution of each
uncertain parameter to the variance in water surface elevation, we calculate the total sensitivity index 
using the PC coefficients ~\citep{Alexanderian2012,Sudret,Crestaux}. The total sensitivity index
of  each of the uncertain parameters is shown in
Figure~\ref{fig:sens} for the four gauges. The Manning's roughness coefficient
at the shore $N_2$ is clearly dominant and contributes
most to the  water surface elevation variance compared to the other two 
Manning's roughness coefficients;
this is true throughout the simulation time.  The Manning's roughness coefficient
in the ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
the during the second hour of simulation and Manning's roughness coefficient
in the land $N_1$ appears to be an insignificant contributor
to the variance.
\begin{figure}[h]
\begin{tabular}{clc}

\includegraphics[width=0.475\textwidth]{../figures/musigma1.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/musigma2.pdf} \\
\includegraphics[width=0.475\textwidth]{../figures/musigma3.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/musigma4.pdf}
\end{tabular}
\caption{Evolution of PC mean water surface elevation at different gauge locations.}
\label{fig:ave}
\end{figure}
\begin{figure}[h]
\begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{../figures/sens1.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/sens2.pdf} \\
\includegraphics[width=0.475\textwidth]{../figures/sens3.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/sens4.pdf}
\end{tabular}
\caption{Total sensitivity index of different input parameters.}
\label{fig:sens}
\end{figure}

%The impact of $\alpha$ and $V_{max}$ for fixed $m=0$ is illustrated in Figure \ref{fig:response}
%for the area-averaged SST within the $42~km\times42~km$ analysis region; the
%different panels illustrate the time evolution of the SST response
%surfaces. Their most striking features are the relatively flat
%horizontal contours on Sep~17 and Sep~18 indicating that SST depends
%only mildly on $V_{\max}$ even during peak winds;
% unsurprisingly these contours turn completely flat on Sep
%19 (and afterwards) when the winds dip below 10~m/s. The multiplicative
%factor $\alpha$, however, exerts a strong influence on SST even
%under mild wind conditions.  This influence is relatively
%weak for the low $\alpha$ range and increases substantially for the
%higher $\alpha$ as evidenced by the packed contours.  The temperature
%response surfaces at 50-m and 200-m depths (not shown) exhibit
%roughly the same structure as the surface, with milder
%dependence on $V_{\max}$ even during the peak winds of Sep~18.

The same statistical analysis can be performed for the
entire domain in 2D. Figure~\ref{fig:mean2d}(top row) shows
the PC mean water surface elevation for the considered computational
domain at three different times as indicated in the title of each panel.
The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row).
\begin{figure}[h]
        \begin{tabular}{ccc}
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/mean2d1.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/mean2d3.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/mean2d4.pdf} \\
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/sigma2d1.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/sigma2d3.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/sigma2d4.pdf}
\end{tabular}
\caption{PC mean (top row) and standard deviation (bottom row) of the water surface elevation at different times.}
\label{fig:mean2d}
\end{figure}

\begin{figure}[h]
\begin{tabular}{clc}
 \hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T12d1.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T12d3.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T12d4.pdf} \\
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T22d1.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T22d3.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T22d4.pdf} \\
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T32d1.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T32d3.pdf} &
\hspace*{-65pt}
\includegraphics[width=0.45\textwidth]{../figures/T32d4.pdf}
\end{tabular}
\caption{Total sensitivity index for $N_1$ (top row) $N_2$ (center row) and $N_3$ (bottom row)
 at different times as indicated.}
\end{figure}

  \clearpage

### 2013-10-24 paper/inference.tex
adding tex files

\section{Bayesian Inference}
 \label{sec:inference}

%Inferring optimal values for the uncertain parameters is performed by solving an inverse problem 
%using observations. Specifically, the AXBT temperature data obtained during typhoon Fanapi,
%which were 
%shown to be correlated to the typhoon wind speed, are used to sharpen 
%initial estimates of the uncertain drag parameters indicated in 
%Equations (\ref{eq:prior1}-\ref{eq:prior3}).

The key idea in the Bayesian approach is to express the 
forward problem (response surface/model)
as a product of conditional probability densities. Bayes theorem 
is then used to reverse the conditioning as explained below.  
Let the observation data (AXBTs) be denoted by \{$T_i = T_i(x_i,y_i,z_i,t_i)\}_{i=1}^N$;
their model predicted counterparts by \{$M_i= M_i(x_i,y_i,z_i,t_i)\}_{i=1}^N$
($N$ is the number of observations) and $H$ the set of control parameters ; 
according to the Bayes' theorem we have:
\begin{equation}
 p(H| \{T_i\}_{i=1}^N) \propto 
 p(\{T_i\}_{i=1}^N | H) \ p(H)  
\label{eq:bayes}
\end{equation}
where $p(H)$ is the prior of $H$, representing the \emph{a priori} knowledge
about $H$ (before considering the observations data); 
$p(\{T_i\}_{i=1}^N | H)$ is the likelihood function representing
the probability of obtaining the data given the set of parameters $H$;
and finally $p(H| \{ T_i \}_{i=1}^N)$ is the posterior,
representing the probability that $H$ is true given the data $( \{ T_i \}_{i=1}^N )$.
In our current inference problem the set of parameters will
be taken as $H = \{ \alpha, V_{\max}, m\}$. Thus, Equation~(\ref{eq:bayes}) can be rewritten as:
\begin{equation}
p(\{ \alpha, V_{\max}, m\} | \{ T_i \}_{i=1}^N)
\propto 
p(\{T_i\}_{i=1}^N | \{ \alpha, V_{\max}, m\}) \ p(\alpha)p(V_{\max}) p(m)  
\end{equation}
The key point in the Bayesian inference is to define the likelihood function and 
prior \citep{sivia}. For this purpose, we define
\begin{equation} 
	 \epsilon_i = T_i - M_i \ ,
\end{equation}
where $\epsilon_i$ represents the discrepancy between the model and observations. 
We assume
that the observations are independent
and that $\epsilon_i$ is normally distributed with mean
zero and variance $\sigma^2$, i.e. 
$\epsilon_i \sim N(0,\sigma^2)$. In this 
case the likelihood function can be written as:
\begin{equation} 
p(\{T_i\}_{i=1}^N | \{ \alpha, V_{\max}, m\}) 
= 
\prod_{i=1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} 
\exp \left( \frac{-(T_i - M_i)^2}{2 \sigma^2} \right)  	
\label{eq:likelihood}
\end{equation}

The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
While in general $\sigma^2$ depends on the observations, in cases where the error
amplitude is generally small and does not change throughout space and time, one may use
a simplified model and assume single hyper-parameter $\sigma^2$. In our problem, 
the typhoon wind forcing conditions vary significantly from one day to another
and AXBT data collected on each day are exposed to different measurement errors.
We thus consider different variance $\sigma^2$ for every set of AXBT data collected
on different days as indicated in Section~\ref{sec:results}\ref{sec:inverse}.

The next step is choosing the prior that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for the model parameters is assumed:
\begin{equation} 
p(\{ \alpha, V_{\max}, m\}) = \begin{cases}
		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i < \{ \alpha, V_{\max}, m\} \leq b_i ,  \\
		0 &\text{otherwise}  , 
\end{cases}
\end{equation}
where $ [a_i,b_i]$ denote the parameter ranges defined in 
Equations~(\ref{eq:prior1}--\ref{eq:prior3}).
Regarding the variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia}, expressed as:
\begin{equation} 
p(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}
\label{eq:var_pr}
\end{equation}
Consequently, Bayes' theorem gives:
\begin{equation} 
p(\{ \alpha, V_{\max}, m\},\sigma^2 | \{T_i\}_{i=1}^N) 
\propto
\left[ \prod_{i=1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} 
\exp \left( \frac{-(T_i - M_i)^2}{2 \sigma^2} \right) \right] 
\ p(\sigma^2)p(\alpha) p(V_{\max})p(m)
\end{equation}
where $p(\{ \alpha, V_{\max}, m\},\sigma^2 | \{T_i\}_{i=1}^N)$  is the joint posterior.

Inferring the drag coefficient parameters requires 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
sample the posterior distribution accurately and efficiently.
This MCMC phase requires repeated (tens of
thousands of)
HYCOM simulations initialized with different values of the
uncertain parameters; this step is prohibitively expensive. An
alternative is to construct a surrogate model that requires a much
smaller ensemble of HYCOM runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
PCEs to build the surrogate, which, in addition
also provide statistical properties,  such as the mean, variance and sensitivities,
efficiently.

PCE methods \citep{LeMaitreKnio2010} are based on expressing the dependence 
of a quantity of
interest, $R$ say, as a truncated expansion of the form:
\begin{equation}
  R(\xbold,t,\xxi) \doteq \sum_{k = 0}^P \hat{R}_k(\xbold,t) \Psi_k(\xxi),
\mbox{ with }
 \hat{R}_k \approx \frac{\left< R, \Psi_k \right>_Q}{\left< \Psi_k, \Psi_k \right>} .
\label{eq:stochseries}
\end{equation}
where $\hat{R}_k$ are unknown coefficients, $P$ is finite and depends on
the truncation strategy adopted, and the functions $\Psi_k(\xxi)$ form
an orthogonal basis of an underlying probability space, and $\left<
R, \Psi_k \right>_Q$ denotes an appropriate numerical quadrature.
The coefficients in the series are computed via a Non-Intrusive
Spectral Projection (NISP) as it does not require any modification to
the HYCOM code but only an ensemble of simulations. We have adopted an adaptive quadrature ~\citep{Constantine:2012,winokur:2012,conrad:2012}
to minimize the errors incurred in evaluating the
coefficients and to minimize the size of the HYCOM ensemble; furthermore
it has the virtue of automatically identifying the important stochastic
variables and steering the quadrature sampling along their directions.
Appendix C provides more details about PCEs and the adaptive quadrature.
Once the coefficients are known, the stochastic series in Equation~(\ref{eq:stochseries}) 
becomes an attractive and
faithfull surrogate for the model as it can be summed very efficiently
and at tiny fraction of the cost of a full HYCOM simulation. We will
use it to estimate the model counterparts $M_i$ of the AXBT temperatures.

### 2013-10-24 paper/intro.tex
adding tex files

\section{Introduction}

In this work, we aim to perform an uncertainty quantification
using Polynomial Chaos Expansion. Quantity of interest is the surface elevation.
The uncertainty in the manning coefficient is investigated.
We assume that the Manning coefficient vary by region. Specifically,
we set three values for the manning coefficient namely, at the land (denoted by N1)
and at the shore (denoted by N2) and below the water (denoted by N3).
Uniform distribution is assumed as the prior for the three manning coefficients.
such that UN = [0.005-0.2]. The three regions are shown in Figure~\ref{fig:setup}(right)

### 2013-10-24 paper/mcmc.tex
adding tex files

\subsection{Inverse Problem} 
\label{sec:inverse}

Before attempting solving the inverse problem,
we need to make sure that the observed data are
comparable to the simulated one.
To this end, Figure~\ref{fig:compare}
compares PC mean water surface elevation profiles
with their counterparts from the observed gauge locations.
The plots show good agreement between the simulations and the observations at different
locations and at different times. 
%
\begin{figure}[h]
\begin{tabular}{clc}
%        
\includegraphics[width=0.475\textwidth]{../figures/compare1.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/compare2.pdf} \\
\includegraphics[width=0.475\textwidth]{../figures/compare3.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/compare4.pdf}
\end{tabular}
\caption{Comparison of PC mean 
and observed data of water surface elevation with time at the four gauges.}
\label{fig:compare}
\end{figure}

Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
Figure~\ref{fig:scatter}.
\begin{figure}[h]
\centering
\includegraphics[width=0.475\textwidth]{../figures/scatter.pdf} 
\caption{Scatter plot of PC mean water surface elevation vs. observed ones.}
\label{fig:scatter}

\end{figure}

We now can exploit the surrogate model in the Bayesian inference of the Manning's 
roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's roughness coefficient distributions in light of the 
observed data. This sampling, demanding tens of thousands of forward simulations, 
would have been prohibitively expensive in the
absence of the surrogate, as the generation of each sample would have required an
independent Geoclaw realization. The surrogate provides a computationally
efficient alternative, and requires only evaluating the PCE for different values 
of the seed $\xxi$. The setup of the Bayesian inference problem,
MCMC chains for the input drag parameters and the corresponding posterior distributions are 
presented and discussed in this section.

The observed data collected at different locations 
in the likelihood function (Equation~\ref{eq:likelihood}) to update the input parameters.
MCMCs of $40,000$ iterations are obtained for the Manning roughness coefficients: 
$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
shows the sample chains for
the input parameters for different iterates of the MCMC algorithms. 
The different panels
shows well-mixed chains for all parameters.
However, all chains span the entire range
of the prior, and so it appears that the observations are not informative 
concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
parameter range, with values between 0.005 and 0.1.
The chain for the variance is also shown in 
Figure \ref{fig:mcmc} and appears to be well mixed.

\begin{figure}[h]
\begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{../figures/chain_p1.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/chain_p2.pdf} \\
\includegraphics[width=0.475\textwidth]{../figures/chain_p3.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/chain_s1.pdf}
\end{tabular}
\caption{Chain samples for the three Manning's roughness coefficients $N_1,N_2,N_3$ and $\sigma^2$
the variance between simulations and observations.}
\label{fig:mcmc} 
\end{figure}

The computed MCMC chains can be readily used to determine the posterior 
distribution; kernel density estimation (KDE) is used for this purpose
~\citep{Parzen1962,Silverman1986}(The first 6000 iterates, associated 
with the burn-in period, are discarded). The resulting posterior pdfs 
of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
with a Maximum A Posteriori (MAP) estimate of around 0.013; an extended tail 
towards the higher Manning's roughness coefficient values is also observed.
In contrast, the posterior pdfs of $N_1$ and $N_3$ appear to be fairly flat, 
and similar to the uniform prior. This is an indication that 
the observed data were not useful to refine our prior knowledge for $N_1$ and $N_3$.

The posterior distribution of the variance is also shown in Figure~\ref{fig:mcmc}. 
Taking the square root of the MAP values yields the water surface elevation standard 
deviation, and the latter is a reflection of the mismatch between the model and 
observed data estimated to be 0.144~$m$.

 \begin{figure}[h]
        \begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{../figures/pdf_p1.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/pdf_p2.pdf} \\
\includegraphics[width=0.475\textwidth]{../figures/pdf_p3.pdf} &
\includegraphics[width=0.475\textwidth]{../figures/pdf_s1.pdf}
        \end{tabular}
        \caption{Posterior distributions for the three Manning's roughness coefficients $N_1,N_2,N_3$ 
and $\sigma^2$ the variance between simulations and observations.}
\label{fig:pdfs} 
        \end{figure}

### 2013-10-24 paper/results.tex
adding tex files

\section{Results}
\label{sec:results}
The results of the forward and inverse problems are presented and 
discussed in three sections. First, we perform an error analysis 
of the PC expansion construction, namely  to establish the 
validity of the corresponding surrogate.  Second, we perform a 
sensitivity analysis and rank the impact of the different Manning's 
roughness coefficient on the predicted surface elevation. Finally, 
posterior distributions for the input parameters are determined 
and analyzed in light of the available gauge data.
\input{pcstudy}
\input{analysis}
\input{mcmc}

### 2013-10-24 paper/tsunami.tex
adding tex files

\section{T\={o}hoku Tsunami}
Our study focuses on quantifying ocean wave uncertainties associated with 
manning friction coefficient,

The setting chosen for our experiment is the tsunami in the Pacific ocean 
that occurred in 11 March 2011. The 2011 earthquake off the Pacific coast of Tōhoku lasting approximately six minutes

%The earthquake, caused by 6 to 8 meters upthrust on a 180-km wide seabed at 60 km offshore from the east coast of Tōhoku, resulted in a major tsunami that brought destruction along the Pacific coastline of Japan's northern islands. Thousands of lives were lost when entire towns were devastated. The tsunami propagated throughout the Pacific Ocean region reaching the entire Pacific coast of North and South America from Alaska to Chile

\subsection{Computational Model}
Figure~\ref{fig:setup}(left) shows the topography of the considered domain.
Land topography is shown using while the bathymetry is shown using a lighter colormap.

During this tsunami, different gauges were installed to observed the surface elevation
The locations of theses gauges are shown in Figure~\ref{setup} .
The gauges are used in the inverse problem to infer the manning coefficients.

\begin{figure}[h]
\centering
\begin{tabular}{clc}
\includegraphics[width=0.45\textwidth]{../figures/topo.pdf}  &
\includegraphics[width=0.45\textwidth]{../figures/coef.pdf} 
\label{setup}
\end{tabular}
\caption{Topography and gauge locations.}
\label{fig:setup}
\end{figure}

### 2013-10-24 paper/uqpce.tex
adding tex files

\section{Polynomial Chaos Expansion}
\label{sec:uqpce}

Polynomial Chaos (PC) is a spectral method that have been developed in the engineering community to represent uncertainties in the output of numerical simulations \citep{Villegas2012,Lin2009,Xiu2004}
due to the uncertainties in a model's input.
The PC method expresses the dependencies of a quantity of 
interest on the uncertain input variables as a truncated expansion.

Let $U=U(\bm{x},t,\xxi)$ 
denote an output quantity of 
interest normally function of space $\bm{x}$ and time $t$, and also on the uncertain variables $\xxi$ that went into
specifying the model's input data (such as uncertain initial and boundary conditions, and parameters). PC expansion express $U$ in the form:
\begin{equation}
  U(\xbold,t,\xxi) \doteq \sum_{k = 0}^P U_k(\xbold,t) \Psi_k(\xxi)
\label{eq:stochseries}
\end{equation}
where $U_k(\xbold,t)$ are the polynomial coefficients, $P$ is finite
and depends on the truncation strategy adopted, and the functions
$\Psi_k(\xxi)$ form an orthogonal basis of an underlying probability
space. This series representation can be viewed as a spectral expansion
of $U$ along the stochastic dimensions.  The existence and convergence of this series
is asserted by the Cameron-Martin theorem
\citep{Cameron:1947}  whenever $R$ has finite variance. The series rate of convergence, and
hence the number of terms to retain, depends on the smoothness of
$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
when $U$ is infinitely smooth; the convergence rate becomes algebraic
when $U$ has finite smoothness \citep{Canuto:2006}. 
%In practice the series convergence is monitored 
%via various error metrics; Section \ref{sec:results}\ref{sec:analysis} presents one possible error
%analysis for the present problem.

The choice of basis is dictated by the probability density
function of the stochastic variable $\xxi$, and which appears as a weight
function in the stochastic space's inner product:
\begin{equation}
 \left<\Psi_k,\Psi_m\right> = \int \Psi_k(\xxi) \;\Psi_m(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi.
\end{equation}
When the basis functions are orthonormal $\left<\Psi_k,\Psi_m\right>=\delta_{k,m}$, where
$\delta_{k,m}$ is the Kronecker delta.
For uniform
distributions the basis functions are hence scaled Legendre polynomials.
For multi-dimensional problems the basis functions are
tensor products of 1D basis functions. The identification of the inner product weight function
with the probability distribution of $\xxi$ simplifies the calculations of $U$'s statistical moments.
Noting that since $\Psi_0(\xxi)$ is a constant that can be normalized to satisfy 
$\left<\Psi_0,\Psi_0\right>=1$, the expectation and variance of $U$ can be computed as:
\begin{eqnarray}
 E[U]&=&\int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0 \\
 E[(U-E[U])^2]&=&\int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
\left<\Psi_k,\Psi_k\right>
\end{eqnarray}
The series representation (\ref{eq:stochseries}) can thus be seen as
combining approximation and probabilistic frameworks, a combination
that has proven extremely useful in solving UQ
problems.

A number of procedures have been devised to compute the expansion coefficients,
$U_k$. Here we rely on the Non-Intrusive Spectral Projection (NISP) 
since it does not require any modification to the deterministic model, and only
ensemble runs at specified values of the uncertain parameters are needed.
The NISP method
exploits the orthogonality of the basis, and replaces the
stochastic integrals by quadrature to obtain:
\begin{equation}
 U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>}
         \approx \frac{\left< U, \Psi_k \right>_Q}{\left< \Psi_k, \Psi_k \right>} .
\end{equation}
The subscript $Q$ refers to approximating the inner product integral with
quadrature, i.e.
\begin{equation}
  \left< U, \Psi_k \right> = \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
\approx \left< U, \Psi_k \right>_Q
= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q
\end{equation}
where $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
respectively.
The computation of the ${U}_k$ can thus be expressed as a matrix-vector product of the form:
\begin{equation} 
 U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
 \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>}
\end{equation} 
where $\Pi_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
from an ensemble of the deterministic model {\em realizations} with the uncertain parameters set at
the quadrature value $\xxi_q$. The NISP procedure can be carried out as a post-processing
operation using simple scripting languages like MATLAB, the only costs being those of the
ensemble calculation and its storage.

The calculation of the $U(\bm{x},t,\xxi_q)$ is the most expensive part of the
inference procedure; thus, reducing the number of sampling points
while maximizing their effectiveness is critical for the procedure's
efficiency.  The quadrature order should be commensurate with the
truncation order, and should be high enough to avoid aliasing artifacts.
The choice of quadrature rule is hence critical to the performance
of the PC (in its NISP version at least). In this work, we employ the 
tensorized Gaussian
quadratures as their computational costs are affordable.
For more details on PC exapnsions and other UQ methods,
refer to \citep{LeMaitreKnio2010}.

### 2013-10-25 paper/pcstudy.tex
fixed some links to figres

\subsection{Error and convergence study}
\label{sec:analysis}

Figure~\ref{fig:rlzs} shows the evolution of the
water surface elevation for the 125 realizations obtained 
to compute the PC expansions at the four different gauge 
locations. We notice that the variation in surface elevation 
is negligible during the first hour at gauge number 21418 
and during the first two hours for the remaining gauges
as the plots of the different realizations are seen to superimpose.
However, a noticeable uncertainty in the surface 
elevation can be seen  at all gauges as indicated from the thickness 
of the bands formed by the plots of the realizations. This uncertainty
occurs till the end of the simulations.
\begin{figure}[h]
\centering
\begin{tabular}{clc}        
\includegraphics[width=0.6\textwidth]{figures/rlzs_gauges.pdf} 
\end{tabular}
\caption{125 Geoclaw realizations at different gauge locations.}
\label{fig:rlzs}
\end{figure}

In order to check the consistency of the approximation, we compare
 water surface elevation from the realizations 
with those obtained from the PC surrogate. The different curves
show an excellent agreement (not shown) for all times. We also define
an error metric that measures the relative normalized root mean-square error between the left hand side function 
in Equation~(\ref{eq:stochseries})
and its PC representation at the sampling points:
\begin{equation} 
   E = \frac{\displaystyle
         \left(\sum_{\xxi \in \NISP} \left|U(\xxi) - \sum_{k = 0}^{P}
U_k\Psi_k(\xxi)\right|^2
         \right)^{1/2}}
        {\displaystyle
          \left(\sum_{\xxi \in \NISP} \left|U(\xxi)\right|^2\right)^{1/2} 
          },
\label{eq:error}
\end{equation}
where $\NISP$ is the 125-member ensemble obtained using the PC algorithm. 
This error metric calculated at the different gauge locations is shown in Figure~\ref{fig:error};
the largest relative normalized error for 
water surface elevation is about 0.1\% . 
\begin{figure}[h]
\centering
\begin{tabular}{clc}        
\includegraphics[width=0.5\textwidth]{figures/error_gauge1.pdf} &
\includegraphics[width=0.5\textwidth]{figures/error_gauge2.pdf} \\
\includegraphics[width=0.5\textwidth]{figures/error_gauge3.pdf} &
\includegraphics[width=0.5\textwidth]{figures/error_gauge4.pdf}

\end{tabular}
\caption{125 Geoclaw realizations at different gauge locations.}
\label{fig:error}
\end{figure}

%Contour maps of the relative normalized error for the entire simulation region are 
%shown in Figure~\ref{fig:error2D} for various depths and dates to confirm the error trends of the
%analysis box.  The error is largest after the wind
%intensifies (bottom row) for all depths. For either day, the maximum
%error is located at 50~m which coincides with the depth of the
%original mixed layer measured by the AXBT. The maximum magnitude
%recorded is about 1\% and occurs on Sep~18. The
%elevated error region is located to the right of the storm and
%extends from the surface down to 50~m, after which the impact of the
%input uncertainty decreases substantially.  For the purpose of our
%current study, and given that the majority of AXBT data are at
%depth, the surrogate's errors are considered acceptable and small.

A final check consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC representation.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3}

\begin{figure}[h]
\centering

\begin{tabular}{clcl}
\includegraphics[width=0.5\textwidth]{figures/pdfs1_2.pdf} &
\includegraphics[width=0.5\textwidth]{figures/pdfs2_2.pdf} \\
\includegraphics[width=0.5\textwidth]{figures/pdfs3_2.pdf} &
\includegraphics[width=0.5\textwidth]{figures/pdfs4_2.pdf}
\end{tabular}
\caption{pdf of water surface elevation at the different gauge locations at t = 7200 s.}
\label{fig:pdfs2}
\end{figure}

\begin{figure}[h]
\centering
\begin{tabular}{clc}

\includegraphics[width=0.5\textwidth]{figures/pdfs1_3.pdf} &
\includegraphics[width=0.5\textwidth]{figures/pdfs2_3.pdf} \\
\includegraphics[width=0.5\textwidth]{figures/pdfs3_3.pdf} &
\includegraphics[width=0.5\textwidth]{figures/pdfs4_3.pdf}
\end{tabular}
\caption{pdf of water surface elevation at the different gauge locations at t = 10800 s.}
\label{fig:pdfs3}
\end{figure}
The different curves
correspond to increased order of PC, from 1 to 5; 
The plots indicate that the double peaked distributions are
well-resolved with order  4 but becomes weakly insenstive to further refinement 
as shown with order 5.

The various error metrics presented above 
provide confidence that the PC expansion is a faithful 
model surrogate.

\clearpage

### 2013-10-30 paper/cover.tex
adding inference section

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\begin{center}
\begin{Large}
{\bf Polynomial Chaos for the Estimation of Manning's Friction}\\

\end{Large}
\bigskip
\bigskip
Ihab Sraj$^1$, Kyle Mandli$^2$ \& Ibrahim Hoteit$^1$\\
\bigskip
$^1$Department of Physical Sciences and Engineering,
King Abdullah University for Science and Technology, Thuwal, Saudi Arabia \\
$^2$Department of Engineering, Austin, TX 
\end{center}

\vspace{5.5cm}

\begin{tabbing}
Corresponding Author: \hspace{5mm} \=  \\
       \> Department of Physical Sciences and Engineering \\
       \> King Abdullah University for Science and Technology \\
       \> Thuwal, Saudi Arabia\\
       \> \\
Phone: \>  \\
Fax:   \>  \\
Email: \>    \\
\\
Submitted to: \> {\it TBD} \\
\> \today \\
\\
Revised: \>

\end{tabbing}

### 2013-10-30 paper/formulation.tex
adding inference section

\section{Formulation}
\label{sec:formu}
We begin by introducing the essential building blocks of the present formulation: 
the Bayesian approach to inverse problems~\ref{sec:inference}
and the stochastic spectral methods for forward uncertainty propagation in ~\ref{sec:uqpce}
\input{inference}
\input{uqpce}

### 2013-10-30 paper/inference.tex
adding inference section

@@
-\section{Bayesian Inference}
+\subsection{Bayesian Inference}
  \label{sec:inference}
+Bayesian inference is a statistical approach to inverse problems
+that has recently gained much interest in different applications.
+The key idea in this approach is to express all forms of uncertainty
+in terms of probability. The first step in a Bayesian inference 
+is to formulate the forward problem (response surface/model) using 
+a suitable likelihood function and product of conditional probability densities. 
+We then formulate the prior distribution over the unknown parameters of the model
+using our best belief. Lastly, given some observation data, Bayes rule 
+is used to obtain a posterior distribution for these unknown
+\citep{sivia}. We review this approach briefly below.
 
-%Inferring optimal values for the uncertain parameters is performed by solving an inverse problem 
-%using observations. Specifically, the AXBT temperature data obtained during typhoon Fanapi,
-%which were 
-%shown to be correlated to the typhoon wind speed, are used to sharpen 
-%initial estimates of the uncertain drag parameters indicated in 
-%Equations (\ref{eq:prior1}-\ref{eq:prior3}).
-
-The key idea in the Bayesian approach is to express the 
-forward problem (response surface/model)
-as a product of conditional probability densities. Bayes theorem 
-is then used to reverse the conditioning as explained below.  
-Let the observation data (AXBTs) be denoted by \{$T_i = T_i(x_i,y_i,z_i,t_i)\}_{i=1}^N$;
-their model predicted counterparts by \{$M_i= M_i(x_i,y_i,z_i,t_i)\}_{i=1}^N$
-($N$ is the number of observations) and $H$ the set of control parameters ; 
-according to the Bayes' theorem we have:
+Let the observation data (water surface elevation at the gauges) be denoted by \{$S_i = S_i(x_i,y_i,t_i)\}_{i=1}^m$;
+their model predicted counterparts by \{$M_i= M_i(x_i,y_i,t_i)\}_{i=1}^m$
+($m$ is the number of observations) and $\{N_i\}_{i=1}^3$ the set of 
+three Manning Friction coefficients treated as the unknown parameters. 
+Bayes' rule yields:
 \begin{equation}
- p(H| \{T_i\}_{i=1}^N) \propto 
- p(\{T_i\}_{i=1}^N | H) \ p(H)  
+ p(\{N_i\}_{i=1}^3| \{S_i\}_{i=1}^m) \propto 
+ p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3) \ p(\{N_i\}_{i=1}^3)  
 \label{eq:bayes}
 \end{equation}
-where $p(H)$ is the prior of $H$, representing the \emph{a priori} knowledge
-about $H$ (before considering the observations data); 
-$p(\{T_i\}_{i=1}^N | H)$ is the likelihood function representing
-the probability of obtaining the data given the set of parameters $H$;
-and finally $p(H| \{ T_i \}_{i=1}^N)$ is the posterior,
-representing the probability that $H$ is true given the data $( \{ T_i \}_{i=1}^N )$.
-In our current inference problem the set of parameters will
-be taken as $H = \{ \alpha, V_{\max}, m\}$. Thus, Equation~(\ref{eq:bayes}) can be rewritten as:
+where $p(\{N_i\}_{i=1}^3)$ is the prior of $\{N_i\}_{i=1}^3$, representing the \emph{a priori} knowledge
+about the manning coefficient; 
+$p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3)$ is the likelihood function representing
+the probability of obtaining the data given the set of parameters $\{N_i\}_{i=1}^3$;
+and finally $p(\{N_i\}_{i=1}^3| \{ S_i \}_{i=1}^m)$ is the posterior,
+representing the probability that $\{N_i\}_{i=1}^3$ is true given the data $( \{ S_i \}_{i=1}^m )$.
+
 \begin{equation}
-p(\{ \alpha, V_{\max}, m\} | \{ T_i \}_{i=1}^N)
+p(\{N_i\}_{i=1}^3 | \{ S_i \}_{i=1}^m)
 \propto 
-p(\{T_i\}_{i=1}^N | \{ \alpha, V_{\max}, m\}) \ p(\alpha)p(V_{\max}) p(m)  
+p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3) \ p(N_1)p(N_2) p(N_3)  
 \end{equation}
-The key point in the Bayesian inference is to define the likelihood function and 
-prior \citep{sivia}. For this purpose, we define
-\begin{equation} 
-	 \epsilon_i = T_i - M_i \ ,
-\end{equation}
-where $\epsilon_i$ represents the discrepancy between the model and observations. 
-We assume
-that the observations are independent
-and that $\epsilon_i$ is normally distributed with mean
+
+To formulate the likelihood function, we let 
+	 $\epsilon_i = S_i - M_i $
+represents the discrepancy between the model and observations 
+and assume that the observations are independent
+and normally distributed with mean
 zero and variance $\sigma^2$, i.e. 
@@
 \begin{equation} 
-p(\{T_i\}_{i=1}^N | \{ \alpha, V_{\max}, m\}) 
+p(\{S_i\}_{i=1}^m |  \{N_i\}_{i=1}^3) 
 = 
-\prod_{i=1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} 
-\exp \left( \frac{-(T_i - M_i)^2}{2 \sigma^2} \right)  	
+\prod_{i=1}^m  \frac{1}{\sqrt{2 \pi \sigma^2}} 
+\exp \left( \frac{-(S_i - M_i)^2}{2 \sigma^2} \right)  	
 \label{eq:likelihood}
@@
 amplitude is generally small and does not change throughout space and time, one may use
-a simplified model and assume single hyper-parameter $\sigma^2$. In our problem, 
-the typhoon wind forcing conditions vary significantly from one day to another
-and AXBT data collected on each day are exposed to different measurement errors.
-We thus consider different variance $\sigma^2$ for every set of AXBT data collected
-on different days as indicated in Section~\ref{sec:results}\ref{sec:inverse}.
+a simplified model and assume single hyper-parameter $\sigma^2$.
 
@@
 \begin{equation} 
-p(\{ \alpha, V_{\max}, m\}) = \begin{cases}
-		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i < \{ \alpha, V_{\max}, m\} \leq b_i ,  \\
+p(N_i) = \begin{cases}
+		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
 		0 &\text{otherwise}  , 
@@
 \begin{equation} 
-p(\{ \alpha, V_{\max}, m\},\sigma^2 | \{T_i\}_{i=1}^N) 
+p(\{N_i\}_{i=1}^3,\sigma^2 | \{S_i\}_{i=1}^m) 
 \propto
-\left[ \prod_{i=1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} 
-\exp \left( \frac{-(T_i - M_i)^2}{2 \sigma^2} \right) \right] 
-\ p(\sigma^2)p(\alpha) p(V_{\max})p(m)
+\left[ \prod_{i=1}^m  \frac{1}{\sqrt{2 \pi \sigma^2}} 
+\exp \left( \frac{-(S_i - M_i)^2}{2 \sigma^2} \right) \right] 
+\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
 \end{equation}
-where $p(\{ \alpha, V_{\max}, m\},\sigma^2 | \{T_i\}_{i=1}^N)$  is the joint posterior.
+where $p(\{N_i\}_{i=1}^3,\sigma^2 | \{S_i\}_{i=1}^m)$  is the joint posterior.
 
@@
 This MCMC phase requires repeated (tens of
-thousands of)
-HYCOM simulations initialized with different values of the
+thousands of) Geoclaw simulations initialized with different values of the
 uncertain parameters; this step is prohibitively expensive. An
 alternative is to construct a surrogate model that requires a much
-smaller ensemble of HYCOM runs, and that can be used instead
+smaller ensemble of GeoClaw runs, and that can be used instead
 at a significantly reduced computational cost.  Here, we rely on
@@
 efficiently.
-
-PCE methods \citep{LeMaitreKnio2010} are based on expressing the dependence 
-of a quantity of
-interest, $R$ say, as a truncated expansion of the form:
-\begin{equation}
-  R(\xbold,t,\xxi) \doteq \sum_{k = 0}^P \hat{R}_k(\xbold,t) \Psi_k(\xxi),
-\mbox{ with }
- \hat{R}_k \approx \frac{\left< R, \Psi_k \right>_Q}{\left< \Psi_k, \Psi_k \right>} .
-\label{eq:stochseries}
-\end{equation}
-where $\hat{R}_k$ are unknown coefficients, $P$ is finite and depends on
-the truncation strategy adopted, and the functions $\Psi_k(\xxi)$ form
-an orthogonal basis of an underlying probability space, and $\left<
-R, \Psi_k \right>_Q$ denotes an appropriate numerical quadrature.
-The coefficients in the series are computed via a Non-Intrusive
-Spectral Projection (NISP) as it does not require any modification to
-the HYCOM code but only an ensemble of simulations. We have adopted an adaptive quadrature ~\citep{Constantine:2012,winokur:2012,conrad:2012}
-to minimize the errors incurred in evaluating the
-coefficients and to minimize the size of the HYCOM ensemble; furthermore
-it has the virtue of automatically identifying the important stochastic
-variables and steering the quadrature sampling along their directions.
-Appendix C provides more details about PCEs and the adaptive quadrature.
-Once the coefficients are known, the stochastic series in Equation~(\ref{eq:stochseries}) 
-becomes an attractive and
-faithfull surrogate for the model as it can be summed very efficiently
-and at tiny fraction of the cost of a full HYCOM simulation. We will
-use it to estimate the model counterparts $M_i$ of the AXBT temperatures.

### 2013-10-30 paper/mcmc.tex
adding inference section

@@
 in the likelihood function (Equation~\ref{eq:likelihood}) to update the input parameters.
-MCMCs of $40,000$ iterations are obtained for the Manning roughness coefficients: 
+MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
 $N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
@@
 distribution; kernel density estimation (KDE) is used for this purpose
-~\citep{Parzen1962,Silverman1986}(The first 6000 iterates, associated 
+~\citep{Parzen1962,Silverman1986}(The first 5000 iterates, associated 
 with the burn-in period, are discarded). The resulting posterior pdfs 
@@
 
+   0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835
+
  \begin{figure}[h]

### 2013-10-30 paper/tsunami.tex
fixed some links to figres
adding inference section

@@
 
-\subsection{Computational Model}
+\subsection{GeoClaw}
+2D Shallow water equations
+
+Physical and computational parameters
 Figure~\ref{fig:setup}(left) shows the topography of the considered domain.
 Land topography is shown using while the bathymetry is shown using a lighter colormap.
-
-During this tsunami, different gauges were installed to observed the surface elevation
-The locations of theses gauges are shown in Figure~\ref{setup} .
-The gauges are used in the inverse problem to infer the manning coefficients.
 
@@
 \end{figure}
+\subsection{Parametric Uncertainty}
+\subsection{Observations}
+During this tsunami, different gauges were installed to observed the surface elevation
+The locations of theses gauges are shown in Figure~\ref{setup} .
+The gauges are used in the inverse problem to infer the manning coefficients.
+
+\begin{figure}[h]      
+\centering
+\includegraphics[width=0.6\textwidth]{../figures/obs.pdf}
+\caption{Observed data of water surface elevation with time at four different gauges.}
+\label{fig:obs}
+\end{figure}

### 2013-11-04 setplot.py
Changes to simulation code setup

@@ -11,6 +11,7 @@ import os
 import glob
 
 import numpy
+import matplotlib.pyplot as plt
 
 import clawpack.visclaw.colormaps as colormaps
 import clawpack.visclaw.geoplot as geoplot
@@ -279,9 +280,17 @@ def setplot(plotdata):
     plotaxes.title = r"Manning's $N$ Coefficient"
     plotaxes.scaled = True
 
-    surge.plot.add_friction(plotaxes, bounds=[0.01,0.04], shrink=0.9)
-    plotaxes.plotitem_dict['friction'].amr_patchedges_show = [0,0,0,0,0,0,0]
-    plotaxes.plotitem_dict['friction'].colorbar_label = ""
+    plotitem = plotaxes.new_plotitem(name='friction',plot_type='2d_pcolor')
+    plotitem.plot_var = lambda cd: cd.aux[1,:,:]
+    plotitem.pcolor_cmap = plt.get_cmap('YlOrRd')
+    plotitem.colorbar_shrink = 0.9
+    plotitem.pcolor_cmin = 0.04
+    plotitem.pcolor_cmax = 0.1    
+    plotitem.add_colorbar = True
+    plotitem.colorbar_label = "Manning's-$n$ Coefficient"
+    plotitem.amr_celledges_show = [0,0,0]
+    plotitem.amr_patchedges_show = [0,0,0,0,0,0,0]
+    plotitem.colorbar_label = r"$n$"
 
     #-----------------------------------------
     

### 2013-11-04 setrun.py

@@ -141,7 +141,7 @@ def setrun(claw_pkg='geoclaw'):
     clawdata.output_format == 'ascii'      # 'ascii' or 'netcdf' 
 
     clawdata.output_q_components = 'all'   # need all
-    clawdata.output_aux_components = 4  # eta=h+B is in q
+    clawdata.output_aux_components = [0,3]  # eta=h+B is in q
     clawdata.output_aux_onlyonce = False    # output aux arrays each frame
 
 

### 2013-11-07 paper/inference.tex
major changes to formulation part

@@
  \label{sec:inference}
+
 Bayesian inference is a statistical approach to inverse problems
-that has recently gained much interest in different applications.
+that has recently gained much interest in different applications including
+ocean~\citep{Alexanderian2011a,Zedler2012,sraj:2013a}
+climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
 The key idea in this approach is to express all forms of uncertainty
@@
 is to formulate the forward problem (response surface/model) using 
-a suitable likelihood function and product of conditional probability densities. 
+a suitable likelihood function and a product of conditional probability densities. 
 We then formulate the prior distribution over the unknown parameters of the model
@@
 
-Let the observation data (water surface elevation at the gauges) be denoted by \{$S_i = S_i(x_i,y_i,t_i)\}_{i=1}^m$;
-their model predicted counterparts by \{$M_i= M_i(x_i,y_i,t_i)\}_{i=1}^m$
-($m$ is the number of observations) and $\{N_i\}_{i=1}^3$ the set of 
-three Manning Friction coefficients treated as the unknown parameters. 
-Bayes' rule yields:
+Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs.
+We now consider a forward model $\vec G$ that predicts the data function of
+the parameters such that:
 \begin{equation}
- p(\{N_i\}_{i=1}^3| \{S_i\}_{i=1}^m) \propto 
- p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3) \ p(\{N_i\}_{i=1}^3)  
+\vec d \approx \vec{G}( \vec \theta)
+\end{equation}
+
+Applying the Bayes' rule yields:
+\begin{equation}
+ p(\vec{\theta}| \vec d) \propto 
+ p(\vec d | \vec{\theta}) \ p(\vec{\theta})  
 \label{eq:bayes}
 \end{equation}
-where $p(\{N_i\}_{i=1}^3)$ is the prior of $\{N_i\}_{i=1}^3$, representing the \emph{a priori} knowledge
-about the manning coefficient; 
-$p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3)$ is the likelihood function representing
-the probability of obtaining the data given the set of parameters $\{N_i\}_{i=1}^3$;
-and finally $p(\{N_i\}_{i=1}^3| \{ S_i \}_{i=1}^m)$ is the posterior,
-representing the probability that $\{N_i\}_{i=1}^3$ is true given the data $( \{ S_i \}_{i=1}^m )$.
+where $p(\vec{\theta})$ is the prior of $\vec{\theta}$, representing the \emph{a priori} knowledge
+about the parameters; 
+$p(\vec d| \vec{\theta})$ is the likelihood function representing
+the probability of obtaining the data given the set of parameters $\vec{\theta}$;
+and finally $p(\vec{\theta}| \vec d)$ is the posterior,
+representing the probability that $\vec{\theta}$ is true given the data $\vec d $.
 
+To formulate the likelihood function, we let $\vec \epsilon = \vec d - \vec{G}$
+represents the discrepancy between the model and observations.
+Here, the components of $\vec \epsilon $ are i.i.d. random variables with density $p_{\epsilon}$.
+In this case the likelihood function can be written as:
 \begin{equation} 
-p(\{N_i\}_{i=1}^3 | \{ S_i \}_{i=1}^m)
-\propto 
-p(\{S_i\}_{i=1}^m | \{N_i\}_{i=1}^3) \ p(N_1)p(N_2) p(N_3)  
-\end{equation}
-
-To formulate the likelihood function, we let 
-	 $\epsilon_i = S_i - M_i $
-represents the discrepancy between the model and observations 
-and assume that the observations are independent
-and normally distributed with mean
-zero and variance $\sigma^2$, i.e. 
-$\epsilon_i \sim N(0,\sigma^2)$. In this 
-case the likelihood function can be written as:
-\begin{equation} 
-p(\{S_i\}_{i=1}^m |  \{N_i\}_{i=1}^3) 
+p(\vec d |  \vec{\theta}) 
 = 
-\prod_{i=1}^m  \frac{1}{\sqrt{2 \pi \sigma^2}} 
-\exp \left( \frac{-(S_i - M_i)^2}{2 \sigma^2} \right)  	
+\prod_{i=1}^m  
+p_\epsilon (d_i - G_i(\vec \theta))  	
 \label{eq:likelihood}
@@
 
-The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
+In our application, we assume that the errors $\epsilon_i$ are independent
+and normally distributed with mean
+zero and variance $\sigma^2$, i.e. 
+$\epsilon_i \sim N(0,\sigma^2)$. 
 While in general $\sigma^2$ depends on the observations, in cases where the error
 amplitude is generally small and does not change throughout space and time, one may use
-a simplified model and assume single hyper-parameter $\sigma^2$.
-
-The next step is choosing the prior that should be based 
-on some \emph{a priori} knowledge about the parameters. In our case, a uniform
-prior for the model parameters is assumed:
+a simplified model and assume single $\sigma^2$ value.
+Consequently, Equation~\ref{eq:likelihood} becomes:
 \begin{equation} 
-p(N_i) = \begin{cases}
-		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
-		0 &\text{otherwise}  , 
-\end{cases}
+p(\vec d |  \vec{\theta}) 
+= 
+\frac{1}{\sqrt{2 \pi \sigma^2}}\prod_{i=1}^m   
+\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace 	
+\label{eq:likelihood2}
 \end{equation}
-where $ [a_i,b_i]$ denote the parameter ranges defined in 
-Equations~(\ref{eq:prior1}--\ref{eq:prior3}).
-Regarding the variance, the only information we know 
-is that $\sigma^2$ is always positive.
-We thus assume a Jeffreys prior \citep{sivia}, expressed as:
+The joint posterior from Bayes' rule can be given as:
 \begin{equation} 
-p(\sigma^2) =  \begin{cases}
-		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
-		0 &\text{otherwise}. 
-		\end{cases}
-\label{eq:var_pr}
+p(\vec{\theta}| \vec d)
+\propto
+\frac{1}{\sqrt{2 \pi \sigma^2}}   \prod_{i=1}^m  
+\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace  
+\prod_{i=1}^n p(\theta_i)
 \end{equation}
-Consequently, Bayes' theorem gives:
+The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
+In other words, $\sigma^2$ may become an additional object for Bayesian inference and may themselves be endowed with priors and estimated from data.
+In this case the joint posterior is finally written as follows:
 \begin{equation} 
-p(\{N_i\}_{i=1}^3,\sigma^2 | \{S_i\}_{i=1}^m) 
+p(\vec{\theta},\sigma^2 | \vec d)
 \propto
-\left[ \prod_{i=1}^m  \frac{1}{\sqrt{2 \pi \sigma^2}} 
-\exp \left( \frac{-(S_i - M_i)^2}{2 \sigma^2} \right) \right] 
-\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
+\frac{1}{\sqrt{2 \pi \sigma^2}}   \prod_{i=1}^m  
+\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace
+\ p(\sigma^2) \prod_{i=1}^n p(\theta_i)
+\label{eq:post}
 \end{equation}
-where $p(\{N_i\}_{i=1}^3,\sigma^2 | \{S_i\}_{i=1}^m)$  is the joint posterior.
-
-Inferring the drag coefficient parameters requires 
-sampling the posterior. In general, when the space of the unknown 
-parameters is multidimensional, a suitable computational strategy is 
-the Markov Chain Monte  Carlo (MCMC) method. 
-We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
-sample the posterior distribution accurately and efficiently.
-This MCMC phase requires repeated (tens of
-thousands of) Geoclaw simulations initialized with different values of the
-uncertain parameters; this step is prohibitively expensive. An
-alternative is to construct a surrogate model that requires a much
-smaller ensemble of GeoClaw runs, and that can be used instead
-at a significantly reduced computational cost.  Here, we rely on
-PCEs to build the surrogate, which, in addition
-also provide statistical properties,  such as the mean, variance and sensitivities,
-efficiently.

### 2013-11-07 paper/mcmc.tex
major changes to formulation part

@@
 %
-\begin{figure}[h]
-\begin{tabular}{clc}
-%        
-\includegraphics[width=0.475\textwidth]{../figures/compare1.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/compare2.pdf} \\
-\includegraphics[width=0.475\textwidth]{../figures/compare3.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/compare4.pdf}
-\end{tabular}
-\caption{Comparison of PC mean 
-and observed data of water surface elevation with time at the four gauges.}
-\label{fig:compare}
-\end{figure}
 
@@
 Figure~\ref{fig:scatter}.
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.475\textwidth]{../figures/scatter.pdf} 
-\caption{Scatter plot of PC mean water surface elevation vs. observed ones.}
-\label{fig:scatter}
-
-\end{figure}
 
@@
 
-\begin{figure}[h]
-\begin{tabular}{clc}
-\includegraphics[width=0.475\textwidth]{../figures/chain_p1.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/chain_p2.pdf} \\
-\includegraphics[width=0.475\textwidth]{../figures/chain_p3.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/chain_s1.pdf}
-\end{tabular}
-\caption{Chain samples for the three Manning's roughness coefficients $N_1,N_2,N_3$ and $\sigma^2$
-the variance between simulations and observations.}
-\label{fig:mcmc} 
-\end{figure}
-
 The computed MCMC chains can be readily used to determine the posterior 
@@
    0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835
-
- \begin{figure}[h]
-        \begin{tabular}{clc}
-\includegraphics[width=0.475\textwidth]{../figures/pdf_p1.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/pdf_p2.pdf} \\
-\includegraphics[width=0.475\textwidth]{../figures/pdf_p3.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/pdf_s1.pdf}
-        \end{tabular}
-        \caption{Posterior distributions for the three Manning's roughness coefficients $N_1,N_2,N_3$ 
-and $\sigma^2$ the variance between simulations and observations.}
-\label{fig:pdfs} 
-        \end{figure}

### 2013-11-07 paper/uqpce.tex
adding inference section
major changes to formulation part

@@
-\section{Polynomial Chaos Expansion}
+\subsection{Accelerating Bayesian Inference}
 \label{sec:uqpce}
 
-Polynomial Chaos (PC) is a spectral method that have been developed in the engineering community to represent uncertainties in the output of numerical simulations \citep{Villegas2012,Lin2009,Xiu2004}
-due to the uncertainties in a model's input.
-The PC method expresses the dependencies of a quantity of 
-interest on the uncertain input variables as a truncated expansion.
+MCMC exploration of the 4D posterior (Equation~\ref{eq:post_coef}) requires repeated simulations (tens of thousands) of the forward GeoClaw model, once for each proposed move of the Markov chain.
+While a single \emph{Geoclaw} simulation
+takes $\sim 15~$mins, depending on the details of the MCMC sampler, it is desirable to avoid repeated forward solutions altogether.
+An alternative is to construct a surrogate model that requires a much
+smaller ensemble of GeoClaw runs, and that can be used instead
+at a significantly reduced computational cost.  Here, we rely on
+Polynomial Chaos expansions for accelerating Bayesian inference in this context 
+by building a surrogate, which, in addition
+also provide statistical properties,  such as the mean, variance and sensitivities,
+efficiently.
 
-Let $U=U(\bm{x},t,\xxi)$ 
-denote an output quantity of 
-interest normally function of space $\bm{x}$ and time $t$, and also on the uncertain variables $\xxi$ that went into
-specifying the model's input data (such as uncertain initial and boundary conditions, and parameters). PC expansion express $U$ in the form:
+\subsubsection{Polynomial Chaos}
+
+Polynomial Chaos (PC) is a probabilistic methodology that expresses the 
+dependencies of model outputs on the uncertain model inputs
+as a polynomial truncated expansion. This method has been developed in 
+the engineering community to represent uncertainties in the output of 
+numerical simulations~\citep{Villegas2012,Lin2009,Xiu2004}
+due to the uncertainties in a model's input. We briefly describe the PC
+method below; for more details on this method, and other UQ methods
+refer to \citep{LeMaitreKnio2010}.
+
+Let $U=U(\bm{x},t,\xxi)$ denote a quantity of 
+interest (QoI) that is the output of certain model.
+$U$ is normally function of space $\bm{x}$ and time $t$, and 
+also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). PC expresses $U$ in the form:
 \begin{equation}
@@
 \end{equation} 
-where $U_k(\xbold,t)$ are the polynomial coefficients, $P$ is finite
-and depends on the truncation strategy adopted, and the functions
-$\Psi_k(\xxi)$ form an orthogonal basis of an underlying probability
-space. This series representation can be viewed as a spectral expansion
+where $U_k(\xbold,t)$ are the polynomial coefficients, and
+$\Psi_k(\xxi)$ are functions that form an orthogonal basis of an underlying probability
+space. The total number of terms in the PC expansion is
+$P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimension and $p$ is the highest order
+of polynomials. 
+This series representation can be viewed as a spectral expansion
 of $U$ along the stochastic dimensions.  The existence and convergence of this series
 is asserted by the Cameron-Martin theorem
-\citep{Cameron:1947}  whenever $R$ has finite variance. The series rate of convergence, and
+\citep{Cameron:1947}  whenever $U$ has finite variance. The series rate of convergence, and
 hence the number of terms to retain, depends on the smoothness of
@@
 \begin{equation}
- \left<\Psi_k,\Psi_m\right> = \int \Psi_k(\xxi) \;\Psi_m(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi.
+ \left<\Psi_i,\Psi_j\right> = \int \Psi_i(\xxi) \;\Psi_j(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi=\delta_{ij}\ave{\Psi_i^2}
+\label{eq:inner}
 \end{equation}
-When the basis functions are orthonormal $\left<\Psi_k,\Psi_m\right>=\delta_{k,m}$, where
-$\delta_{k,m}$ is the Kronecker delta.
+where $\delta_{ij}$ is the Kronecker delta.
 For uniform
@@
 \begin{eqnarray}
- E[U]&=&\int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0 \\
+ E[U]&=&\int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0  
+ \label{eq:mean} \\
  E[(U-E[U])^2]&=&\int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
+ \label{eq:sigma}
 \left<\Psi_k,\Psi_k\right>
 \end{eqnarray}
-The series representation (\ref{eq:stochseries}) can thus be seen as
-combining approximation and probabilistic frameworks, a combination
-that has proven extremely useful in solving UQ
-problems.
 
-A number of procedures have been devised to compute the expansion coefficients,
-$U_k$. Here we rely on the Non-Intrusive Spectral Projection (NISP) 
-since it does not require any modification to the deterministic model, and only
-ensemble runs at specified values of the uncertain parameters are needed.
-The NISP method
-exploits the orthogonality of the basis, and replaces the
-stochastic integrals by quadrature to obtain:
+PC representations of random variables also enable efficient
+approximation of the corresponding variance-based, global
+sensitivity indices. This can be done by computing the so-called {\it total} 
+sensitivity indices of random variables from the PC representations or Sobol decomposition~\citep{Sobol:1993,Homma:1996,Sobol:2001}. To get the total sensitivity corresponding to the uncertain
+input $\xi_i$ we compute the total index~\citep{LeMaitreKnio2010,Crestaux,Sudret}:
+\begin{equation} \label{equ:T-hard}
+   T_i = \displaystyle \sum_{u \ni i} S_u =
+         \frac{\displaystyle
+              \sum_{u \ni i} \sum_{k \in K_u} U_k^2 \ave{\Psi_k^2}}
+              {\displaystyle\sum_{k = 1}^P U_k^2 \ave{\Psi_k^2}} .
+\end{equation}
+where for each index set $u \subseteq \{1, \ldots, n\}$ we define:
+\[
+   K_u = \left\{ k \in \{1, \ldots, P\} :
+          \begin{cases}
+           \vec{\alpha}^k_i > 0, \quad \mbox{ if } i \in u,\\
+           \vec{\alpha}^k_i = 0, \quad \mbox{ if } i \notin u
+          \end{cases}\right\}.
+        \]
+        and $\vec{\alpha}^k$ is the
+multi-index associated with $k^{th}$ term in the
+PC expansion~\cite{LeMaitreKnio2010}.
+
+Using Equation~\eqref{equ:T-hard}, the computation of $T_i$ is straightforward.
+Note that the total sensitivity index $T_i$ measures the contribution of
+the $i^{th}$ random input to total model variability by
+computing the fraction of the total variance due to all the terms in the
+PC expansion which involve $\xi_i$. It is also worth noting that for random variables expanded in a PC basis, all the index sets
+above are dictated by the basis alone. Thus, we need to find the index sets $K_u$
+for computation of $S_u$ and $\mathcal{I}_i$ for computation
+of $T_i$ only once. Hence, the computation of $S_u$ or $T_i$ for a
+random variable is immediate once its PC coefficients are determined.
+
+%The series representation (\ref{eq:stochseries}) can thus be seen as
+%combining approximation and probabilistic frameworks, a combination
+%that has proven extremely useful in solving UQ
+%problems.
+
+\subsubsection{Non Intrusive Spectral Projection (NISP)}
+The computation of the coefficients of the PC expansions $U_k$
+can be done using a number of procedures. Here we adopt a non-intrusive
+approach that allows the use of the forward model \emph{Geoclaw} as a black box
+with no code modifications required. PC expansion coefficients are thus determined
+based on a set of response \emph{Geoclaw} simulations at specified set of the uncertain parameters. 
+In particular, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
 \begin{equation}
- U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>}
-         \approx \frac{\left< U, \Psi_k \right>_Q}{\left< \Psi_k, \Psi_k \right>} .
+ U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>} = 
+ \frac{1}{\left< \Psi_k, \Psi_k \right>} 
+ \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
 \end{equation}
-The subscript $Q$ refers to approximating the inner product integral with
-quadrature, i.e.
+This orthogonal projection minimizes the L2 error on the space spanned by the basis.
+with the inner-product indicated in Equation~\ref{eq:inner}.
+Using NISP the stochastic integrals are solved using a numerical quadrature to obtain:
+
 \begin{equation}
-  \left< U, \Psi_k \right> = \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
+  \left< U, \Psi_k \right> 
 \approx \left< U, \Psi_k \right>_Q
@@
 \end{equation}
-where $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
+where the subscript $Q$ refers to approximating the inner product integral with
+quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
 respectively.
@@
 from an ensemble of the deterministic model {\em realizations} with the uncertain parameters set at
-the quadrature value $\xxi_q$. The NISP procedure can be carried out as a post-processing
-operation using simple scripting languages like MATLAB, the only costs being those of the
-ensemble calculation and its storage.
+the quadrature value $\xxi_q$.
 
+The only cost of the NISP is that of the ensemble calculation and its storage.
 The calculation of the $U(\bm{x},t,\xxi_q)$ is the most expensive part of the
@@
 of the PC (in its NISP version at least). In this work, we employ the 
-tensorized Gaussian
-quadratures as their computational costs are affordable.
-For more details on PC exapnsions and other UQ methods,
-refer to \citep{LeMaitreKnio2010}.
+tensorized Gaussian quadratures as their computational costs are affordable.

### 2013-11-12 paper/mcmc.tex
Fix figure paths

@@
 %
+\begin{figure}[h]
+\begin{tabular}{clc}
+%        
+\includegraphics[width=0.475\textwidth]{./figures/compare1.pdf} &
+\includegraphics[width=0.475\textwidth]{./figures/compare2.pdf} \\
+\includegraphics[width=0.475\textwidth]{./figures/compare3.pdf} &
+\includegraphics[width=0.475\textwidth]{./figures/compare4.pdf}
+\end{tabular}
+\caption{Comparison of PC mean 
+and observed data of water surface elevation with time at the four gauges.}
+\label{fig:compare}
+\end{figure}
 
@@
 Figure~\ref{fig:scatter}.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.475\textwidth]{./figures/scatter.pdf} 
+\caption{Scatter plot of PC mean water surface elevation vs. observed ones.}
+\label{fig:scatter}
+
+\end{figure}
 
@@
 
+\begin{figure}[h]
+\begin{tabular}{clc}
+\includegraphics[width=0.475\textwidth]{./figures/chain_p1.pdf} &
+\includegraphics[width=0.475\textwidth]{./figures/chain_p2.pdf} \\
+\includegraphics[width=0.475\textwidth]{./figures/chain_p3.pdf} &
+\includegraphics[width=0.475\textwidth]{./figures/chain_s1.pdf}
+\end{tabular}
+\caption{Chain samples for the three Manning's roughness coefficients $N_1,N_2,N_3$ and $\sigma^2$
+the variance between simulations and observations.}
+\label{fig:mcmc} 
+\end{figure}
+
 The computed MCMC chains can be readily used to determine the posterior 
@@
    0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835
+
+ \begin{figure}[h]
+        \begin{tabular}{clc}
+\includegraphics[width=0.475\textwidth]{./figures/pdf_p1.pdf} &
+\includegraphics[width=0.475\textwidth]{./figures/pdf_p2.pdf} \\
+\includegraphics[width=0.475\textwidth]{./figures/pdf_p3.pdf} &
+\includegraphics[width=0.475\textwidth]{./figures/pdf_s1.pdf}
+        \end{tabular}
+        \caption{Posterior distributions for the three Manning's roughness coefficients $N_1,N_2,N_3$ 
+and $\sigma^2$ the variance between simulations and observations.}
+\label{fig:pdfs} 
+        \end{figure}

### 2013-11-12 paper/tsunami.tex
major changes to formulation part
removed figures from tsnumai.tex
Fix figure paths

@@
+%!TEX root = paper.tex
+
+Background literature so far
+\begin{itemize}
+    \item Myers and Baptista (2001) - Crude friction coefficient sensitivity, found it to be important but not fully explored
+    \item Jakeman et al (2010) - Again a crude (two coefficient) sensitivity analysis.
+    \item Dao and Tkalich (2007) - Really not a very good paper, also crude SA.
+\end{itemize}
+
+\section{Numerical Model}
+
 \section{T\={o}hoku Tsunami}
@@
 \begin{tabular}{clc}
-\includegraphics[width=0.45\textwidth]{../figures/topo.pdf}  &
-\includegraphics[width=0.45\textwidth]{../figures/coef.pdf} 
+\includegraphics[width=0.45\textwidth]{./figures/topo.pdf}  &
+\includegraphics[width=0.45\textwidth]{./figures/coef.pdf} 
 \label{setup}
@@
 \end{figure}
+
 \subsection{Parametric Uncertainty}
+
 \subsection{Observations}
+
+Observations used in th
+
 During this tsunami, different gauges were installed to observed the surface elevation
@@
 \centering
-\includegraphics[width=0.6\textwidth]{../figures/obs.pdf}
+\includegraphics[width=0.6\textwidth]{./figures/obs.pdf}
 \caption{Observed data of water surface elevation with time at four different gauges.}

### 2013-11-12 paper/analysis.tex
Fix figure paths

@@
-\subsection{Statistical and sensitivity analysis}
+\subsection{Forward Propagation of Uncertainty}
 \label{sec:forward}
-We now exploit the PC surrogate to study the statistics of the 
-water surface elevation, and to quantify its sensitivity to the
-uncertain input parameters and thus to anticipate their impact on
-the inverse problem.  In doing so, it is emphasized that 
-no additional Geoclaw simulations were needed to obtain
-the information presented below, rather it is obtained either directly 
-from the coefficients of the PC surrogate or by sampling the corresponding
-representations for different values of $\xxi$.
+We now use the surrogate created effectively using PC to propagate prior 
+uncertainty through the forward model. We mainly exploit the surrogate 
+to study the statistics of the water surface elevation, and to conduct 
+a global sensitivity analysis of the impact of the uncertain input parameters.
+We note that the statistical and sensitivity analysis presented below 
+is performed using the PC coefficients solely; these were calculated
+from an ensemble of 125 Geoclaw simulations.
+\subsubsection{Statistical analysis}
+The mean of a QoI and its standard deviation is computed
+from the PC coefficients as shown in Equation~\ref{eq:mean} and Equation~\ref{eq:sigma}. 
+In Figure~\ref{fig:ave} we plot the evolution of
+the calculated PC mean water surface elevation along with two standard deviations
+bounds at the four gauges as indicated in each panel.  
+Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
+reflected in a significant standard deviation as seen earlier from the realizations in Figure~\ref{fig:rlzs}.
+We observed that the standard deviation ......
 
-Figure~\ref{fig:ave} shows the evolution of
-PC mean water surface elevation and its two standard deviations
-bounds at the four gauge locations as indicated in each panel.  
-The standard deviation is insignificant during the two hours
-of the simulations and therefore was not included in the plots.
-This standard deviation however increases as the tsunami increase
-and its effect approaches the gauge locations. 
-To quantity the contribution of each
-uncertain parameter to the variance in water surface elevation, we calculate the total sensitivity index 
-using the PC coefficients ~\citep{Alexanderian2012,Sudret,Crestaux}. The total sensitivity index
-of  each of the uncertain parameters is shown in
-Figure~\ref{fig:sens} for the four gauges. The Manning's roughness coefficient
-at the shore $N_2$ is clearly dominant and contributes
-most to the  water surface elevation variance compared to the other two 
-Manning's roughness coefficients;
-this is true throughout the simulation time.  The Manning's roughness coefficient
-in the ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
-the during the second hour of simulation and Manning's roughness coefficient
-in the land $N_1$ appears to be an insignificant contributor
-to the variance.
-\begin{figure}[h]
-\begin{tabular}{clc}
-
-\includegraphics[width=0.475\textwidth]{../figures/musigma1.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/musigma2.pdf} \\
-\includegraphics[width=0.475\textwidth]{../figures/musigma3.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/musigma4.pdf}
-\end{tabular}
-\caption{Evolution of PC mean water surface elevation at different gauge locations.}
-\label{fig:ave}
-\end{figure}
-\begin{figure}[h]
-\begin{tabular}{clc}
-\includegraphics[width=0.475\textwidth]{../figures/sens1.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/sens2.pdf} \\
-\includegraphics[width=0.475\textwidth]{../figures/sens3.pdf} &
-\includegraphics[width=0.475\textwidth]{../figures/sens4.pdf}
-\end{tabular}
-\caption{Total sensitivity index of different input parameters.}
-\label{fig:sens}
-\end{figure}
-
-%The impact of $\alpha$ and $V_{max}$ for fixed $m=0$ is illustrated in Figure \ref{fig:response}
-%for the area-averaged SST within the $42~km\times42~km$ analysis region; the
-%different panels illustrate the time evolution of the SST response
-%surfaces. Their most striking features are the relatively flat
-%horizontal contours on Sep~17 and Sep~18 indicating that SST depends
-%only mildly on $V_{\max}$ even during peak winds;
-% unsurprisingly these contours turn completely flat on Sep
-%19 (and afterwards) when the winds dip below 10~m/s. The multiplicative
-%factor $\alpha$, however, exerts a strong influence on SST even
-%under mild wind conditions.  This influence is relatively
-%weak for the low $\alpha$ range and increases substantially for the
-%higher $\alpha$ as evidenced by the packed contours.  The temperature
-%response surfaces at 50-m and 200-m depths (not shown) exhibit
-%roughly the same structure as the surface, with milder
-%dependence on $V_{\max}$ even during the peak winds of Sep~18.
-
-The same statistical analysis can be performed for the
+\comment{The same statistical analysis can be performed for the
 entire domain in 2D. Figure~\ref{fig:mean2d}(top row) shows
@@
 domain at three different times as indicated in the title of each panel.
-The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row).
-\begin{figure}[h]
-        \begin{tabular}{ccc}
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/mean2d1.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/mean2d3.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/mean2d4.pdf} \\
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/sigma2d1.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/sigma2d3.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/sigma2d4.pdf}
-\end{tabular}
-\caption{PC mean (top row) and standard deviation (bottom row) of the water surface elevation at different times.}
-\label{fig:mean2d}
-\end{figure}
+The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row).}
 
-\begin{figure}[h]
-\begin{tabular}{clc}
- \hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T12d1.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T12d3.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T12d4.pdf} \\
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T22d1.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T22d3.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T22d4.pdf} \\
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T32d1.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T32d3.pdf} &
-\hspace*{-65pt}
-\includegraphics[width=0.45\textwidth]{../figures/T32d4.pdf}
-\end{tabular}
-\caption{Total sensitivity index for $N_1$ (top row) $N_2$ (center row) and $N_3$ (bottom row)
- at different times as indicated.}
-\end{figure}
+\subsubsection{Sensitivity analysis}
+A global sensitivity analysis is next conducted to quantity the contribution of each
+uncertain parameter to the variance in water surface elevation. To this end, we calculate 
+the total sensitivity index using the PC coefficients as shown in Equation~\ref{eq:sens}~\citep{Alexanderian2012,Sudret,Crestaux}. The evolution of the total sensitivity index
+of each of the uncertain parameters is shown in Figure~\ref{fig:sens} for the four gauges. 
+The Manning's roughness coefficient at the shore $N_2$ is clearly dominant and contributes
+the most to the variance in the water surface elevation compared to the other two 
+Manning's roughness coefficients $N_1$ and $N_3$; this is true for almost the entire simulation time
+and at the four gauges. The Manning's roughness coefficient
+in the bottom of ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
+during the second hour of simulation and Manning's roughness coefficient
+at the land $N_1$ appears to be an insignificant contributor
+to the variance.
+
+\comment{In 2D, the sensitivity analysis shows that .... Figure~\ref{fig:sens2d}(top row) shows
+the PC mean water surface elevation for the considered computational
+domain at three different times as indicated in the title of each panel.
+The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row)}.
+
+\subsubsection{Response surface}
+In addition to the static and analysis, the PC surrogate 
+can be used to construct a response surface for the uncertain input parameters.
+This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
+range.
+\comment{The impact of $N_2$ and $N_3$ for fixed $N_1=0.1025$ is illustrated in Figure~\ref{fig:response2}
+and Figure~\ref{fig:response3}
+for the area-averaged SST within the $42~km\times42~km$ analysis region; the
+different panels illustrate the time evolution of the SST response
+surfaces. 
+Their most striking features are the relatively flat
+horizontal contours on Sep~17 and Sep~18 indicating that SST depends
+only mildly on $V_{\max}$ even during peak winds;
+unsurprisingly these contours turn completely flat on Sep
+19 (and afterwards) when the winds dip below 10~m/s. The multiplicative
+factor $\alpha$, however, exerts a strong influence on SST even
+under mild wind conditions.  This influence is relatively
+weak for the low $\alpha$ range and increases substantially for the
+higher $\alpha$ as evidenced by the packed contours.  The temperature
+response surfaces at 50-m and 200-m depths (not shown) exhibit
+roughly the same structure as the surface, with milder
+dependence on $V_{\max}$ even during the peak winds of Sep~18.}
 

### 2013-11-12 paper/coefs.tex

\subsection{Inferring Manning Friction Coefficient}
 \label{sec:manning}

We seek to infer the Manning coefficient from water surface elevation
data measured at gauges as showm.
Bayesian inference can be applied directly to infer the uncertain parameters
using Equation~\ref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their \emph{GeoClaw} model predicted counterparts by $\vec G$.
The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the
three Manning Friction coefficients treated as the unknown parameters. 
In this case  Equation~\ref{eq:post} can be written as:
\begin{equation} 
p(N_1,N_2,N_3,\sigma^2 | \vec d) 
\propto \frac{1}{\sqrt{2 \pi \sigma^2}} 
 \prod_{i=1}^m  
\exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
\label{eq:post_coef}
\end{equation}

To complete deinfe the posterior equation, we need to choose a proper prior that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for the model parameters is assumed as indicated:
\begin{equation} 
p(N_i) = \begin{cases}
		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
		0 &\text{otherwise}  , 
\end{cases}
\end{equation}
where $ [a_i,b_i]$ denote the parameter ranges defined as.
Regarding the variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia}, expressed as:
\begin{equation} 
p(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}
\label{eq:var_pr}
\end{equation}

Inferring the coefficients requires 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
sample the posterior distribution accurately and efficiently.

### 2013-11-12 paper/formulation.tex

@@
 \label{sec:formu}
-We begin by introducing the essential building blocks of the present formulation: 
-the Bayesian approach to inverse problems~\ref{sec:inference}
-and the stochastic spectral methods for forward uncertainty propagation in ~\ref{sec:uqpce}
+We begin by introducing the methods employed in this work.
+The Bayesian approach to inverse problems is described in Section~\ref{sec:inference};
+and its application to our problem of inferring the Manning Friction Coefficients is described in Section~\ref{sec:manning}. Finally in Section~\ref{sec:uqpce}, we introduce
+a stochastic spectral method for forward uncertainty propagation and accelerating the Bayesian inference.
+%The two pieces are brought together in~\ref{sec:setup}.
+
 \input{inference}
+\input{coefs}
+
 \input{uqpce}

### 2013-11-12 paper/pcstudy.tex
adding inference section
Fix figure paths

@@
 occurs till the end of the simulations.
-\begin{figure}[h]
-\centering
-\begin{tabular}{clc}        
-\includegraphics[width=0.6\textwidth]{figures/rlzs_gauges.pdf} 
-\end{tabular}
-\caption{125 Geoclaw realizations at different gauge locations.}
-\label{fig:rlzs}
-\end{figure}
 
@@
 water surface elevation is about 0.1\% .
-\begin{figure}[h]
-\centering
-\begin{tabular}{clc}        
-\includegraphics[width=0.5\textwidth]{figures/error_gauge1.pdf} &
-\includegraphics[width=0.5\textwidth]{figures/error_gauge2.pdf} \\
-\includegraphics[width=0.5\textwidth]{figures/error_gauge3.pdf} &
-\includegraphics[width=0.5\textwidth]{figures/error_gauge4.pdf}
-
-\end{tabular}
-\caption{125 Geoclaw realizations at different gauge locations.}
-\label{fig:error}
-\end{figure}
 
@@
 
-\begin{figure}[h]
-\centering
-
-\begin{tabular}{clcl}
-\includegraphics[width=0.5\textwidth]{figures/pdfs1_2.pdf} &
-\includegraphics[width=0.5\textwidth]{figures/pdfs2_2.pdf} \\
-\includegraphics[width=0.5\textwidth]{figures/pdfs3_2.pdf} &
-\includegraphics[width=0.5\textwidth]{figures/pdfs4_2.pdf}
-\end{tabular}
-\caption{pdf of water surface elevation at the different gauge locations at t = 7200 s.}
-\label{fig:pdfs2}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\begin{tabular}{clc}
-
-\includegraphics[width=0.5\textwidth]{figures/pdfs1_3.pdf} &
-\includegraphics[width=0.5\textwidth]{figures/pdfs2_3.pdf} \\
-\includegraphics[width=0.5\textwidth]{figures/pdfs3_3.pdf} &
-\includegraphics[width=0.5\textwidth]{figures/pdfs4_3.pdf}
-\end{tabular}
-\caption{pdf of water surface elevation at the different gauge locations at t = 10800 s.}
-\label{fig:pdfs3}
-\end{figure}
 The different curves

### 2013-11-12 paper/tsunami.tex

@@
 
-Physical and computational parameters
-Figure~\ref{fig:setup}(left) shows the topography of the considered domain.
-Land topography is shown using while the bathymetry is shown using a lighter colormap.
-
-\begin{figure}[h]
-\centering
-\begin{tabular}{clc}
-\includegraphics[width=0.45\textwidth]{./figures/topo.pdf}  &
-\includegraphics[width=0.45\textwidth]{./figures/coef.pdf} 
-\label{setup}
-\end{tabular}
-\caption{Topography and gauge locations.}
-\label{fig:setup}
-\end{figure}
-
 \subsection{Parametric Uncertainty}
+Uniform distribution is assumed as the prior for the three manning coefficients.
+such that UN = [0.005-0.2]. The three regions are shown in Figure~\ref{fig:setup}(right)
 
 \subsection{Observations}
-
-Observations used in th
-
 During this tsunami, different gauges were installed to observed the surface elevation
@@
 The gauges are used in the inverse problem to infer the manning coefficients.
-
-\begin{figure}[h]      
-\centering
-\includegraphics[width=0.6\textwidth]{./figures/obs.pdf}
-\caption{Observed data of water surface elevation with time at four different gauges.}
-\label{fig:obs}
-\end{figure}

### 2013-11-12 paper/intro.tex
major changes to formulation part
Add more citations

@@
+%!TEX root = paper.tex
+
 \section{Introduction}
@@
 and at the shore (denoted by N2) and below the water (denoted by N3).
-Uniform distribution is assumed as the prior for the three manning coefficients.
-such that UN = [0.005-0.2]. The three regions are shown in Figure~\ref{fig:setup}(right)
+
+\cite{sraj:2013a}
+\cite{sraj:2013b}
+
+Background literature on friction coefficient sensitivity:
+\begin{itemize}
+    \item \cite{Myers:2001el} - Crude friction coefficient sensitivity, found it to be important but not fully explored
+    \item \cite{Jakeman:2010hk} - Again a crude (two coefficient) sensitivity analysis.
+    \item \cite{Dao:2007hr} - Really not a very good paper, also crude SA.
+\end{itemize}

### 2013-11-12 paper/tsunami.tex
Add more citations

@@
 
-Background literature so far
-\begin{itemize}
-    \item Myers and Baptista (2001) - Crude friction coefficient sensitivity, found it to be important but not fully explored
-    \item Jakeman et al (2010) - Again a crude (two coefficient) sensitivity analysis.
-    \item Dao and Tkalich (2007) - Really not a very good paper, also crude SA.
-\end{itemize}
+\section{Approach}
 
-\section{Numerical Model}
+% Our study focuses on quantifying the surface displacement uncertainties associated with the Manning's-N friction coefficient.  The analysis was applied to the \tohoku tsunami
 
-\section{T\={o}hoku Tsunami}
-Our study focuses on quantifying ocean wave uncertainties associated with 
-manning friction coefficient,
+% \tohoku
 
-The setting chosen for our experiment is the tsunami in the Pacific ocean 
-that occurred in 11 March 2011. The 2011 earthquake off the Pacific coast of Tōhoku lasting approximately six minutes
+% \subsection{\tohoku Tsunami}
 
-%The earthquake, caused by 6 to 8 meters upthrust on a 180-km wide seabed at 60 km offshore from the east coast of Tōhoku, resulted in a major tsunami that brought destruction along the Pacific coastline of Japan's northern islands.
-Thousands of lives were lost when entire towns were devastated.
-The tsunami propagated throughout the Pacific Ocean region reaching the entire Pacific coast of North and South America from Alaska to Chile
+\subsection{Numerical Model}
 
-\subsection{GeoClaw}
-2D Shallow water equations
+Based on work by \cite{MacInnes:2013cr}, \cite{Berger:2011du}, \cite{Berger:2011vi}
+
+\subsection{Observations}
 

### 2013-11-12 paper/paper.tex
adding inference section
Add formatting for Ocean Modelling

@@
-\documentclass[12pt]{article}
+%&pdflatex
+%
+%
+%
+%
+
+% \documentclass[12pt]{article}
+\documentclass[preprint,12pt]{elsarticle}
+
 \usepackage{graphicx}
@@
 
-\oddsidemargin 0in
-\evensidemargin 0in
-\textwidth= 6.5in
-\topmargin -0.50in
-\textheight 9in
+% Add line numbers
+\usepackage[mathlines]{lineno}
+
+% Multipart figures
+\usepackage{subcaption}
+
+% \oddsidemargin 0in
+% \evensidemargin 0in
+% \textwidth= 6.5in
+% \topmargin -0.50in
+% \textheight 9in
+
+\usepackage{ifpdf}
 
@@
                      {\mbox{\boldmath$\scriptscriptstyle{#1}$}}}}
-                     %
+
+\newcommand{\geoclaw}{{\sc GeoClaw}\xspace}
+\newcommand{\tohoku}{T\={o}hoku\xspace}
+
+\journal{Ocean Modelling}
 
 \begin{document}
-\begin{center}{\Large\bf Tsunami Uncertainty Quantification}\vskip 5pt
 
-\bf Preliminary results
-\end{center}
-\vskip 15pt
+\ifpdf
+\DeclareGraphicsExtensions{.pdf, .png, .jpg, .tif}
+\else
+\DeclareGraphicsExtensions{.png, .jpg, .tif, .eps}
+\fi
+
+\begin{frontmatter}
+
+\title{Polynomial Chaos for the Estimation of Manning's Friction}
+\date{\today}
+
+\author[kaust]{Ihab Sraj}
+\ead{ihab.sraj@kaust.edu.sa}
+\ead{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+\author[ut]{Kyle T. Mandli}
+% \ead{kyle@ices.utexas.edu}
+% \ead[http://users.ices.utexas.edu/~kyle]{http://users.ices.utexas.edu/~kyle}
+\author[kaust]{Ibrahim Hoteit}
+\author[ut]{Clint N. Dawson}
+
+\address[kaust]{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+\address[ut]{Institute for Computational Engineering and Science, University of Texas at Austin, 201 E 24th ST. Stop C0200, Austin, TX 78712-1229, USA}
+
+\date{\today}
+
+\begin{abstract}
+
+\end{abstract}
+
+\begin{keyword}
+tsunami \sep Manning's N friction \sep sensitivity analysis \sep polynomial chaos
+\end{keyword}
+
+\end{frontmatter}
+\linenumbers
+
+% \input{cover}
+% \clearpage
 
@@
 
-\clearpage
-%\input{inference}
+\input{formulation}
 
-\input{uqpce}
 \input{results}
 
+\input{figures} 
 \clearpage
@@
 \clearpage
-\bibliographystyle{plainnat}
+\bibliographystyle{elsarticle-num}
 \bibliography{biblio}

### 2013-11-13 paper/coefs.tex
minor edits and added class files

@@
 We seek to infer the Manning coefficient from water surface elevation
-data measured at gauges as showm.
+data measured at different gauges as shown in Figure~\ref{fig:setup}.
 Bayesian inference can be applied directly to infer the uncertain parameters
@@
 are the water surface elevation at the gauges;
-their \emph{GeoClaw} model predicted counterparts by $\vec G$.
+their model predicted counterparts $\vec G$ is predicted by \emph{GeoClaw}.
 The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the
@@
 
-To complete deinfe the posterior equation, we need to choose a proper prior that should be based 
+To complete the definition of the posterior, we need to choose a proper prior that should be based 
 on some \emph{a priori} knowledge about the parameters. In our case, a uniform

### 2013-11-13 paper/formulation.tex
minor edits and added class files

--- 2013-10-30 paper/formulation.tex
@@
 \label{sec:formu}
-We begin by introducing the essential building blocks of the present formulation: 
-the Bayesian approach to inverse problems~\ref{sec:inference}
-and the stochastic spectral methods for forward uncertainty propagation in ~\ref{sec:uqpce}
+In this section, we introduce the methods employed in this work.
+First, the Bayesian approach to inverse problems is described in Section~\ref{sec:inference};
+its application to our problem of inferring the Manning Friction Coefficients in Section~\ref{sec:manning}.
+Finally, in Section~\ref{sec:uqpce}, we introduce
+a stochastic spectral method for forward uncertainty propagation and accelerating the Bayesian inference.
+
 \input{inference}
+\input{coefs}
+
 \input{uqpce}

### 2013-11-13 paper/inference.tex
minor edits and added class files

@@
 The key idea in this approach is to express all forms of uncertainty
-in terms of probability. The first step in a Bayesian inference 
+in terms of probability. The first step in Bayesian inference 
 is to formulate the forward problem (response surface/model) using 
 a suitable likelihood function and a product of conditional probability densities. 
-We then formulate the prior distribution over the unknown parameters of the model
-using our best belief. Lastly, given some observation data, Bayes rule 
-is used to obtain a posterior distribution for these unknown
+The prior distributions over the unknown parameters of the model
+are then formulated using our best belief. Lastly, given some observation data, Bayes rule 
+is used to obtain a posterior distribution for these unknown parameters
 \citep{sivia}. We review this approach briefly below.
@@
 Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs.
-We now consider a forward model $\vec G$ that predicts the data function of
+We consider a forward model $\vec G$ that predicts the data function of
 the parameters such that:

### 2013-11-13 paper/uqpce.tex
minor edits and added class files

@@
 
-MCMC exploration of the 4D posterior (Equation~\ref{eq:post_coef}) requires repeated simulations (tens of thousands) of the forward GeoClaw model, once for each proposed move of the Markov chain.
+MCMC exploration of the four-dimensional posterior (Equation~\ref{eq:post_coef}) requires repeated simulations (tens of thousands) of the forward GeoClaw model, once for each proposed move of the Markov chain.
 While a single \emph{Geoclaw} simulation
@@
 of polynomials. 
-This series representation can be viewed as a spectral expansion
+\comment{This series representation can be viewed as a spectral expansion
 of $U$ along the stochastic dimensions.  The existence and convergence of this series
@@
 when $U$ is infinitely smooth; the convergence rate becomes algebraic
-when $U$ has finite smoothness \citep{Canuto:2006}. 
+when $U$ has finite smoothness \citep{Canuto:2006}. }
 %In practice the series convergence is monitored 
@@
 with the probability distribution of $\xxi$ simplifies the calculations of $U$'s statistical moments.
+
+\subsubsection{Non Intrusive Spectral Projection (NISP)}
+The computation of the coefficients of the PC expansions $U_k$
+can be done using a number of procedures. Here we adopt a non-intrusive
+approach that allows the use of the forward model \emph{Geoclaw} as a black box
+with no code modifications required. PC expansion coefficients are thus determined
+based on a set of response \emph{Geoclaw} simulations at specified set of the uncertain parameters. 
+In particular, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
+\begin{equation}
+ U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>} = 
+ \frac{1}{\left< \Psi_k, \Psi_k \right>} 
+ \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
+\end{equation}
+This orthogonal projection minimizes the L2 error on the space spanned by the basis.
+with the inner-product indicated in Equation~\ref{eq:inner}.
+Using NISP the stochastic integrals are solved using a numerical quadrature to obtain:
+
+\begin{equation}
+  \left< U, \Psi_k \right> 
+\approx \left< U, \Psi_k \right>_Q
+= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q
+\end{equation}
+where the subscript $Q$ refers to approximating the inner product integral with
+quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
+respectively.
+The computation of the ${U}_k$ can thus be expressed as a matrix-vector product of the form:
+\begin{equation} 
+ U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
+ \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>}
+\end{equation} 
+where $\Pi_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
+from an ensemble of the deterministic model {\em realizations} with the uncertain parameters set at
+the quadrature value $\xxi_q$.
+
+The only cost of the NISP is that of the ensemble calculation and its storage.
+The calculation of the $U(\bm{x},t,\xxi_q)$ is the most expensive part of the
+inference procedure; thus, reducing the number of sampling points
+while maximizing their effectiveness is critical for the procedure's
+efficiency.  The quadrature order should be commensurate with the
+truncation order, and should be high enough to avoid aliasing artifacts.
+The choice of quadrature rule is hence critical to the performance
+of the PC (in its NISP version at least). In this work, we employ the 
+tensorized Gaussian quadratures as their computational costs are affordable.
+
+\subsubsection{Use of Polynomial Chaos}
 Noting that since $\Psi_0(\xxi)$ is a constant that can be normalized to satisfy 
@@
 %problems.
-
-\subsubsection{Non Intrusive Spectral Projection (NISP)}
-The computation of the coefficients of the PC expansions $U_k$
-can be done using a number of procedures. Here we adopt a non-intrusive
-approach that allows the use of the forward model \emph{Geoclaw} as a black box
-with no code modifications required. PC expansion coefficients are thus determined
-based on a set of response \emph{Geoclaw} simulations at specified set of the uncertain parameters. 
-In particular, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
-\begin{equation}
- U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>} = 
- \frac{1}{\left< \Psi_k, \Psi_k \right>} 
- \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
-\end{equation}
-This orthogonal projection minimizes the L2 error on the space spanned by the basis.
-with the inner-product indicated in Equation~\ref{eq:inner}.
-Using NISP the stochastic integrals are solved using a numerical quadrature to obtain:
-
-\begin{equation}
-  \left< U, \Psi_k \right> 
-\approx \left< U, \Psi_k \right>_Q
-= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q
-\end{equation}
-where the subscript $Q$ refers to approximating the inner product integral with
-quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
-respectively.
-The computation of the ${U}_k$ can thus be expressed as a matrix-vector product of the form:
-\begin{equation} 
- U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
- \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>}
-\end{equation} 
-where $\Pi_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
-from an ensemble of the deterministic model {\em realizations} with the uncertain parameters set at
-the quadrature value $\xxi_q$.
-
-The only cost of the NISP is that of the ensemble calculation and its storage.
-The calculation of the $U(\bm{x},t,\xxi_q)$ is the most expensive part of the
-inference procedure; thus, reducing the number of sampling points
-while maximizing their effectiveness is critical for the procedure's
-efficiency.  The quadrature order should be commensurate with the
-truncation order, and should be high enough to avoid aliasing artifacts.
-The choice of quadrature rule is hence critical to the performance
-of the PC (in its NISP version at least). In this work, we employ the 
-tensorized Gaussian quadratures as their computational costs are affordable.

### 2013-11-13 paper/analysis.tex
minor edits and added class files
method section is completed for now

@@
 is performed using the PC coefficients solely; these were calculated
-from an ensemble of 125 Geoclaw simulations.
+from an ensemble of 125 \geoclaw simulations.
 \subsubsection{Statistical analysis}

### 2013-11-13 paper/coefs.tex
method section is completed for now

@@
 are the water surface elevation at the gauges;
-their model predicted counterparts $\vec G$ is predicted by \emph{GeoClaw}.
+their model predicted counterparts $\vec G$ is predicted by \geoclaw.
 The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the

### 2013-11-13 paper/uqpce.tex
method section is completed for now

--- 2013-11-07 paper/uqpce.tex
@@
 
-MCMC exploration of the 4D posterior (Equation~\ref{eq:post_coef}) requires repeated simulations (tens of thousands) of the forward GeoClaw model, once for each proposed move of the Markov chain.
-While a single \emph{Geoclaw} simulation
-takes $\sim 15~$mins, depending on the details of the MCMC sampler, it is desirable to avoid repeated forward solutions altogether.
-An alternative is to construct a surrogate model that requires a much
-smaller ensemble of GeoClaw runs, and that can be used instead
+The four-dimensional posterior in Equation~\ref{eq:post_coef} can be explored via MCMC;
+this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
+once for every proposed set of parameters of the Markov chain. While a single \geoclaw simulation
+takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
+to avoid forward model simulations altogether. This can be accomplished by constructing a 
+surrogate model that requires a much
+smaller ensemble of \geoclaw runs, and that can be used instead
 at a significantly reduced computational cost.  Here, we rely on
 Polynomial Chaos expansions for accelerating Bayesian inference in this context 
-by building a surrogate, which, in addition
-also provide statistical properties,  such as the mean, variance and sensitivities,
-efficiently.
+by building a surrogate, which, in addition can provide statistical properties, such as the mean, variance and sensitivities, efficiently.
 
@@
 $U$ is normally function of space $\bm{x}$ and time $t$, and 
-also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). PC expresses $U$ in the form:
+also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). 
+PC expresses $U$ in the form:
 \begin{equation}
@@
 of polynomials.
-This series representation can be viewed as a spectral expansion
-of $U$ along the stochastic dimensions.  The existence and convergence of this series
-is asserted by the Cameron-Martin theorem
-\citep{Cameron:1947}  whenever $U$ has finite variance. The series rate of convergence, and
-hence the number of terms to retain, depends on the smoothness of
-$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
-when $U$ is infinitely smooth; the convergence rate becomes algebraic
-when $U$ has finite smoothness \citep{Canuto:2006}. 
-%In practice the series convergence is monitored 
-%via various error metrics; Section \ref{sec:results}\ref{sec:analysis} presents one possible error
-%analysis for the present problem.
 
-The choice of basis is dictated by the probability density
+The choice of the basis is dictated by the probability density
 function of the stochastic variable $\xxi$, and which appears as a weight
@@
 For multi-dimensional problems the basis functions are
-tensor products of 1D basis functions. The identification of the inner product weight function
+tensor products of 1D basis functions.
+
+The series representation Equation~\ref{eq:stochseries} can be viewed as a spectral expansion
+of $U$ along the stochastic dimensions. The existence and convergence of this series
+is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
+The series rate of convergence, and hence the number of terms to retain, depends on the smoothness of
+$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
+when $U$ is infinitely smooth; the convergence rate becomes algebraic
+when $U$ has finite smoothness \citep{Canuto:2006}. In practice the series convergence is monitored 
+via various error metrics as discussed in the results section.
+
+\subsubsection{Non Intrusive Spectral Projection (NISP)}
+The computation of the coefficients of the PC expansions $U_k$
+can be done using a number of procedures. Here we adopt a non-intrusive
+approach that allows the use of the forward model \geoclaw as a black box
+with no code modifications required. PC expansion coefficients are thus determined
+based on a set of response \geoclaw simulations at specified set of the uncertain parameters. 
+In particular, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
+\begin{equation}
+ U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>} = 
+ \frac{1}{\left< \Psi_k, \Psi_k \right>} 
+ \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
+\end{equation}
+This orthogonal projection minimizes the $L_2$ error on the space spanned by the basis.
+Using NISP the stochastic integrals are solved using a numerical quadrature to obtain:
+
+\begin{equation}
+  \left< U, \Psi_k \right> 
+\approx \left< U, \Psi_k \right>_Q
+= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q
+\end{equation}
+where the subscript $Q$ refers to approximating the inner product integral with
+quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
+respectively. The quadrature order should be commensurate with the
+truncation order, and should be high enough to avoid aliasing artifacts.
+The choice of quadrature rule is hence critical to the performance
+of the PC (in its NISP version at least). In the current work, we employ the 
+tensorized Gaussian quadrature that yields 125 quadrature points for the set of three input parameters.
+The computational cost of this ensemble is low compared to the cost of the 
+tens of thousands of runs using the forward model.
+
+The computation of the ${U}_k$ can thus be expressed as a matrix-vector product of the form:
+\begin{equation} 
+ U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
+ \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>}
+\end{equation} 
+where $\Pi_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
+from an ensemble of the deterministic model realizations with the uncertain parameters set at
+the quadrature value $\xxi_q$.
+
+\subsubsection{Statistical moments and sensitivity analysis}
+The identification of the inner product weight function
 with the probability distribution of $\xxi$ simplifies the calculations of $U$'s statistical moments. 
@@
 \begin{equation} \label{equ:T-hard}
-   T_i = \displaystyle \sum_{u \ni i} S_u =
+   T_i =
          \frac{\displaystyle
-              \sum_{u \ni i} \sum_{k \in K_u} U_k^2 \ave{\Psi_k^2}}
+               \sum_{k \in K_i} U_k^2 \ave{\Psi_k^2}}
               {\displaystyle\sum_{k = 1}^P U_k^2 \ave{\Psi_k^2}} .
 \end{equation}
-where for each index set $u \subseteq \{1, \ldots, n\}$ we define:
-\[
-   K_u = \left\{ k \in \{1, \ldots, P\} :
-          \begin{cases}
-           \vec{\alpha}^k_i > 0, \quad \mbox{ if } i \in u,\\
-           \vec{\alpha}^k_i = 0, \quad \mbox{ if } i \notin u
-          \end{cases}\right\}.
+where we define \[
+   K_i = \left\{ k \in \{1, \ldots, P\} :
+           \vec{\alpha}^k_i > 0 \right\}
         \]
@@
 Using Equation~\eqref{equ:T-hard}, the computation of $T_i$ is straightforward.
-Note that the total sensitivity index $T_i$ measures the contribution of
+The total sensitivity index $T_i$ measures the contribution of
 the $i^{th}$ random input to total model variability by
 computing the fraction of the total variance due to all the terms in the
-PC expansion which involve $\xi_i$. It is also worth noting that for random variables expanded in a PC basis, all the index sets
-above are dictated by the basis alone. Thus, we need to find the index sets $K_u$
-for computation of $S_u$ and $\mathcal{I}_i$ for computation
-of $T_i$ only once. Hence, the computation of $S_u$ or $T_i$ for a
-random variable is immediate once its PC coefficients are determined.
+PC expansion which involve $\xi_i$.
+% It is also worth noting that for random variables expanded in a PC basis, all the index sets
+%above are dictated by the basis alone. Thus, we need to find the index sets $K_u$
+%for computation of $S_u$ and $\mathcal{I}_i$ for computation
+%of $T_i$ only once. Hence, the computation of $S_u$ or $T_i$ for a
+%random variable is immediate once its PC coefficients are determined.
 

### 2013-11-13 paper/analysis.tex
results section in progress

@@
 
-\comment{The same statistical analysis can be performed for the
+The same statistical analysis can be performed for the
 entire domain in 2D. Figure~\ref{fig:mean2d}(top row) shows
@@
 domain at three different times as indicated in the title of each panel.
-The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row).}
+The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row)
+at different times. We clearly notice the propagation of the variance
+due to the parametric uncertainty with time
+and along the tsunami's reflections.
 
@@
 uncertain parameter to the variance in water surface elevation. To this end, we calculate 
-the total sensitivity index using the PC coefficients as shown in Equation~\ref{eq:sens}~\citep{Alexanderian2012,Sudret,Crestaux}. The evolution of the total sensitivity index
-of each of the uncertain parameters is shown in Figure~\ref{fig:sens} for the four gauges. 
+the total sensitivity index using the PC coefficients as shown in Equation~\eqref{eq:T-hard}~\citep{Alexanderian2012,Sudret,Crestaux}. The evolution of the total sensitivity index
+of each of the uncertain parameters is shown in Figure~\ref{fig:sens} at the four gauges. 
 The Manning's roughness coefficient at the shore $N_2$ is clearly dominant and contributes
@@
 
-\comment{In 2D, the sensitivity analysis shows that .... Figure~\ref{fig:sens2d}(top row) shows
-the PC mean water surface elevation for the considered computational
-domain at three different times as indicated in the title of each panel.
-The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row)}.
+In 2D, the sensitivity analysis shows also that $N_2$ is dominant
+where we have variance in water surface elevation. This is
+indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
+for $N_1$(top row) $N_2$(center row) and $N_3$ (bottom row)
+at different times as indicated in each panel.
 
@@
 This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
-range.
-\comment{The impact of $N_2$ and $N_3$ for fixed $N_1=0.1025$ is illustrated in Figure~\ref{fig:response2}
+range. Since, $N_3$ shows insignificant contribution in the 
+uncertainty in the model output, we find the impact of $N_2$ and $N_3$ for 
+fixed $N_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
 and Figure~\ref{fig:response3}
-for the area-averaged SST within the $42~km\times42~km$ analysis region; the
-different panels illustrate the time evolution of the SST response
-surfaces. 
-Their most striking features are the relatively flat
-horizontal contours on Sep~17 and Sep~18 indicating that SST depends
-only mildly on $V_{\max}$ even during peak winds;
-unsurprisingly these contours turn completely flat on Sep
-19 (and afterwards) when the winds dip below 10~m/s. The multiplicative
-factor $\alpha$, however, exerts a strong influence on SST even
-under mild wind conditions.  This influence is relatively
-weak for the low $\alpha$ range and increases substantially for the
-higher $\alpha$ as evidenced by the packed contours.  The temperature
-response surfaces at 50-m and 200-m depths (not shown) exhibit
-roughly the same structure as the surface, with milder
-dependence on $V_{\max}$ even during the peak winds of Sep~18.}
-
-  \clearpage
+for different times as indicated. The most striking features are the relatively flat
+horizontal contours in the $N_3$ direction indicating that water surface elevation depends
+only mildly on $N_3$ even during peak tsunami events.

### 2013-11-13 paper/uqpce.tex
results section in progress

@@
 The series representation Equation~\ref{eq:stochseries} can be viewed as a spectral expansion
-of $U$ along the stochastic dimensions. The existence and convergence of this series
+of $U$ along the stochastic dimensions. It can be also seen as
+combining approximation and probabilistic frameworks, a combination
+that has proven extremely useful in solving UQ problems. The existence and convergence of this series
 is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
@@
 input $\xi_i$ we compute the total index~\citep{LeMaitreKnio2010,Crestaux,Sudret}:
-\begin{equation} \label{equ:T-hard}
+\begin{equation} \label{eq:T-hard}
    T_i =
@@
 
-Using Equation~\eqref{equ:T-hard}, the computation of $T_i$ is straightforward.
+Using Equation~\eqref{eq:T-hard}, the computation of $T_i$ is straightforward.
 The total sensitivity index $T_i$ measures the contribution of
@@
 PC expansion which involve $\xi_i$.
-% It is also worth noting that for random variables expanded in a PC basis, all the index sets
-%above are dictated by the basis alone. Thus, we need to find the index sets $K_u$
-%for computation of $S_u$ and $\mathcal{I}_i$ for computation
-%of $T_i$ only once. Hence, the computation of $S_u$ or $T_i$ for a
-%random variable is immediate once its PC coefficients are determined.
-
-%The series representation (\ref{eq:stochseries}) can thus be seen as
-%combining approximation and probabilistic frameworks, a combination
-%that has proven extremely useful in solving UQ
-%problems.

### 2013-11-13 paper/tsunami.tex
Minor modifications, mostly adding convenient packages

@@
 
-% Our study focuses on quantifying the surface displacement uncertainties associated with the Manning's-N friction coefficient.  The analysis was applied to the \tohoku tsunami
+Our study focuses on quantifying the surface displacement uncertainties associated with the Manning's-N friction coefficient using a polynomial chaos approach as applied to the \tohoku tsunami of 2011.
+The forward calculations were generated using \geoclaw, a finite volume wave-propagation numerical model that solves the two-dimensional nonlinear shallow water equations.
+In conjunction with
 
-% \tohoku
+\subsection{\tohoku Tsunami}
 
-% \subsection{\tohoku Tsunami}
+The \tohoku earthquake and tsunami of 2011 has been the subject of numerous studies due to the wealth of observational evidence and severity of the tsunami.
 
-\subsection{Numerical Model}
+The $\text{M}_\text{w}$ 9.0 great \tohoku earthquake of 2011 and the accompanying tsunami caused massive damage to the Japanese coastlines in the Sendai region and beyond including damage in Crescent City, CA.
+The earthquake magnitude was
 
-Based on work by \cite{MacInnes:2013cr}, \cite{Berger:2011du}, \cite{Berger:2011vi}
+A number of studies have looked at the
+
+\subsection{\geoclaw}
+
+\geoclaw is Based on work by \cite{MacInnes:2013cr}, \cite{Berger:2011du}, \cite{Berger:2011vi}, \cite{George:2008aa}
 
 \subsection{Observations}
+
+Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system
+
+gauges 21418, 21413, 21401, 21419 measured maximum wave amplitudes of 1.86, 0.77, 0.66 and 0.54 m respectively
 

### 2013-11-13 paper/mcmc.tex
minor edits and added class files

--- 2013-11-07 paper/mcmc.tex
@@
 The plots show good agreement between the simulations and the observations at different
-locations and at different times. 
-%
-
-Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
+locations and at different times. Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
 Figure~\ref{fig:scatter}.
@@
 absence of the surrogate, as the generation of each sample would have required an
-independent Geoclaw realization. The surrogate provides a computationally
+independent \geoclaw realization. The surrogate provides a computationally
 efficient alternative, and requires only evaluating the PCE for different values 
@@
 ~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
-with a Maximum A Posteriori (MAP) estimate of around 0.013; an extended tail 
+with a Maximum A Posteriori (MAP) estimate of around 0.011; an extended tail 
 towards the higher Manning's roughness coefficient values is also observed.
-In contrast, the posterior pdfs of $N_1$ and $N_3$ appear to be fairly flat, 
+In contrast, the posterior pdfs of $N_1$ appear to be fairly flat, 
 and similar to the uniform prior. This is an indication that 
-the observed data were not useful to refine our prior knowledge for $N_1$ and $N_3$.
-
+the observed data were not useful to refine our prior knowledge for $N_1$.
+While for $N_3$, we observed a posterior that has a well defined peak
+of 0.185 but no clear pdf shape.
 The posterior distribution of the variance is also shown in Figure~\ref{fig:mcmc}. 
@@
 deviation, and the latter is a reflection of the mismatch between the model and 
-observed data estimated to be 0.144~$m$.
+observed data estimated to be 0.145~$m$.
 
-   0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835
+   %MAPS 0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835

### 2013-11-13 paper/pcstudy.tex
minor edits and added class files

@@
 the largest relative normalized error for 
-water surface elevation is about 0.1\% .
+water surface elevation is about 0.1\%.
 
@@
 and Figure~\ref{fig:pdfs3}
-
 The different curves
-correspond to increased order of PC, from 1 to 5; 
+correspond to increased order of PC, from 1 to 5.
 The plots indicate that the double peaked distributions are
 well-resolved with order  4 but becomes weakly insenstive to further refinement 
-as shown with order 5.
-
-The various error metrics presented above 
+as shown with order 5. The various error metrics presented above 
 provide confidence that the PC expansion is a faithful 
 model surrogate.
-
-\clearpage

### 2013-11-13 paper/tsunami.tex
Fleshing out approach section

--- 2013-11-12 paper/tsunami.tex
@@
 
-% Our study focuses on quantifying the surface displacement uncertainties associated with the Manning's-N friction coefficient.  The analysis was applied to the \tohoku tsunami
+The approach taken to quantify the surface displacement uncertainties associated with the Manning's $n$ friction parameterization uses a polynomial chaos approach coupled with a shallow water forward model as applied to a real event, the \tohoku tsunami of 2011.
+Here we briefly described the setup of the forward model and the prior of the model used.
 
-% \tohoku
+\subsection{Forward Model}
 
-% \subsection{\tohoku Tsunami}
+The forward model we use is the \geoclaw package, a finite volume wave-propagation class that solves the two-dimensional nonlinear shallow water equations  
+\begin{equation} \label{eq:swe}
+    \begin{aligned}
+    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
+    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| u \\
+    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| v
+    \end{aligned}
+\end{equation}
+where $h$ is the depth of the water column, $u$ and $v$ the velocities in the longitudinal and latitudinal directions respectively, $g$ the acceleration due to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom friction coefficient.
+As is common, $C_f$ is calculated given a Manning's $n$ parameterization such that
+\begin{equation}
+    C_f = \frac{g n^2}{h^{5/3}}
+\end{equation}
+where $n$ is determined from empirical roughness studies.
 
-\subsection{Numerical Model}
+The input data
 
-Based on work by \cite{MacInnes:2013cr}, \cite{Berger:2011du}, \cite{Berger:2011vi}
+\subsubsection{Source Term Evaluation} \label{ssub:source_term_numerics}
+
+Since the effect of friction on the water column is the primary focus of this study, we give a brief description of the process by which friction and the other source terms are included in \geoclaw.
+We can write hyperbolic balance laws such as \eqref{eq:swe} commonly in the form
+\begin{equation}
+    q_t + f(q)_x + g(q)_y = \Psi(q)
+\end{equation}
+where $q$ are the conserved quantities, $f(q)$ and $g(q)$ the fluxes in the appropriate directions, and $\Psi(q)$ the source terms.  In the case of \eqref{eq:swe} these can be written
+\begin{align}
+    q = \begin{bmatrix}
+        h \\ hu \\ hv
+    \end{bmatrix} & &
+    f(q) = \begin{bmatrix}
+        hu \\ hu^2 + \frac{1}{2} gh^2 \\ huv
+    \end{bmatrix} & &
+    g(q) = \begin{bmatrix}
+        hv \\ huv \\ hv^2 + \frac{1}{2} gh^2
+    \end{bmatrix}
+    \Psi(q) = \begin{bmatrix}
+        0 \\ fhv - ghb_x - C_f |\vec{u}| u \\ -fhu - ghb_y - C_f |\vec{u}| v
+    \end{bmatrix}.
+\end{align}
+
+In \geoclaw, the bathymetry source term is handled directly in the Riemann solver step due to well-balancing concerns.
+The Coriolis and friction terms are handled via a operator-splitting approach that solves two simpler problems, $q_t + f(q)_x + g(q)_y = 0$ and $q_t = \Psi(q)$, combining the updates to $q$.
+The simplest splitting is Godunov splitting in which each system is solved alternately using the full time step $\dt$.
+Although this approach is only first-order accurate the observed splitting errors do no dominate the overall error in practice (see \cite{LevequeBook2002} for a more thorough discussion).
+
+Starting with the friction and Coriolis, the bottom friction terms can be evaluated as
+\[
+    (hu)_t = -C_f hu ~~~~\text{and}~~~~ (hv)_t = -C_f hv
+\]
+and are solved using a backwards Euler method for computing the loss of momentum so that
+\[
+    (hu)^{n+1}_{ij} = \frac{(hu)^n_{ij}}{1 + (C_f)_{ij} \dt} ~~~~\text{and}~~~~
+    (hv)^{n+1}_{ij} = \frac{(hv)^n_{ij}}{1 + (C_f)_{ij} \dt}
+\]
+where the drag coefficient is computed using the previous time step's state as
+\[
+    (C_f)_{ij} = \frac{gn_{ij}^2}{(h^n_{ij})^{-7/3}} \sqrt{[(hu)^n_{ij}]^2 + [(hv)^n_{ij}]^2} \left[1-\left(\frac{h_{\text{break}}}{h^n_{ij}}\right)^{\theta_f} \right]^{\gamma_f / \theta_f}.
+\]
+The values $h_{\text{break}}$, $\theta_f$, and $\gamma_f$ are constant parameters that define the hybridization between the Chezy and Manning's $n$ formulation from \eqref{eq:friction_term}.
+The Coriolis terms
+\[
+    (hu)_t = -f hv ~~~~\text{and}~~~~ (hv)_t = f hu
+\]
+are evaluated using a matrix exponential up to the 4th term in the series such that the update becomes
+\[
+    \begin{bmatrix}
+        hu \\ hv     
+    \end{bmatrix}^{n+1}_{ij} = 
+    \begin{bmatrix}
+        hu \\ hv     
+    \end{bmatrix}^{n}_{ij} \cdot 
+    \begin{bmatrix}
+        1 - \frac{1}{2} (f \dt)^2 + \frac{1}{24} (f \dt)^4 & f \dt - \frac{1}{6} (f \dt)^3 \\
+        - f \dt + \frac{1}{6} (f \dt)^3 & 1 - \frac{1}{2} (f \dt)^2 + \frac{1}{24} (f \dt)^4
+    \end{bmatrix}
+\]
+where $f = 2 \Omega \sin y$ with $\Omega = 2 \pi / 8.61642\times10^4$ and $y$ the longitudinal coordinate in radians.
+
+\begin{enumerate}
+    \item Finite volume wave-propagation approach
+    \item Adaptive mesh refinement
+    \item Riemann solver - included inundation and well-balanced properties
+    \item Implementation of friction as a source-term splitting
+\end{enumerate}
+
+\geoclaw is Based on work by \cite{MacInnes:2013cr}, \cite{Berger:2011du}, \cite{Berger:2011vi}, \cite{George:2008aa}
+
+\subsection{\tohoku Tsunami}
+
+The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
+The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
+In this study we have based our simulation on a recent study that attempted to ascertain which source earthquake model matched the observed data optimally \cite{MacInnes:2013cr}.
+
+\begin{enumerate}
+    \item Note that we do not have the bathymetry used in this paper
+    \item Figure out which source we are using
+\end{enumerate}
 
@@
 
+The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
+The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \cite{Percival:2011}.
+The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:dart_locations} and whose de-tided data for the event are show in figure~\ref{fig:dart_data}.
+
+% \begin{figure}[tb]    
+%     \centering
+%     \begin{subfigure}[b]{0.48\textwidth}
+%         \includegraphics[width=\textwidth]{./figures/DART_locations}
+%         \caption{}
+%         \label{fig:dart_locations}
+%     \end{subfigure}
+%     \begin{subfigure}[b]{0.48\textwidth}
+%         \includegraphics[width=\textwidth]{./figures/DART_data}
+%         \caption{}
+%         \label{fig:dart_data}
+%     \end{subfigure}
+%     \caption{Stuff}
+%   \end{figure}
+
 \subsection{Parametric Uncertainty}
+To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $n$ was sampled in the interval $[0.005-0.2]$.  The regions
+
 Uniform distribution is assumed as the prior for the three manning coefficients.

### 2013-11-14 paper/coefs.tex
results section in progress and minor edits

@@
 Bayesian inference can be applied directly to infer the uncertain parameters
-using Equation~\ref{eq:post}. In our case, the observation data $\vec d$ 
+using Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
 are the water surface elevation at the gauges;
@@
 three Manning Friction coefficients treated as the unknown parameters. 
-In this case  Equation~\ref{eq:post} can be written as:
+In this case  Equation~\eqref{eq:post} can be written as:
 \begin{equation} 

### 2013-11-14 paper/tsunami.tex
results section in progress and minor edits

@@
 The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \cite{Percival:2011}.
-The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:dart_locations} and whose de-tided data for the event are show in figure~\ref{fig:dart_data}.
+The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup} and whose de-tided data for the event are show in Figure~\ref{fig:obs}.
 

### 2013-12-11 paper/intro.tex
Paper update

@@
 \section{Introduction}
+
+Tsunamis in the past decade have been responsible for some of the largest natural disasters ever recorded.
+
+In the face of this hazard, risk assessment and forecasting systems have come under renewed scrutiny in order to build for future events.
+Work to improve the computational methods underlying these tasks has seen tremendous advances recently but work done to quantify the uncertainties in the input data such as bathymetry, source mechanisms, and friction field, has been lacking.
+
+Tsunami hazards have come under increasing interest in the past decade as massive tsunami events have turned the world's eyes to the utter destruction they can cause.
+Risk assessments and forecasting of these events have seen vast improvements over the past decades due to more readily available computational power and advances in the computational methods employed.
+These advances have been responsible for many increases in the veracity of predictive models but one area of tsunami modeling that has not been addressed fully is in the uncertainty of input data such as the bathymetry, source mechanism and friction field.
+
+Previous work to study uncertainty in the input
+
+In this work we focus on the uncertainty in the parameterization of the friction field.
+
+In this work, we aim to quantify the uncertainty in the friction field using polynomial chaos expansions.
+
+In this work we aim to quantify one aspect of the uncertainty involved in tsunami modeling, that of the friction field.
 
@@
 \end{itemize}
+
+\url{http://arxiv.org/pdf/1203.6297.pdf}
+\url{http://hal.inria.fr/docs/00/64/38/78/PDF/RR-7809.pdf}

### 2013-12-11 paper/tsunami.tex
Paper update

--- 2013-11-13 paper/tsunami.tex
@@
 
-\section{Approach}
+\section{Tsunami Computational Model and Case Study Descriptions} \label{sec:model_description}
 
-Our study focuses on quantifying the surface displacement uncertainties associated with the Manning's-N friction coefficient using a polynomial chaos approach as applied to the \tohoku tsunami of 2011.
-The forward calculations were generated using \geoclaw, a finite volume wave-propagation numerical model that solves the two-dimensional nonlinear shallow water equations.
-In conjunction with
+In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following UQ analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
 
-\subsection{\tohoku Tsunami}
+\subsection{\geoclaw} \label{ssub:geoclaw}
 
-The \tohoku earthquake and tsunami of 2011 has been the subject of numerous studies due to the wealth of observational evidence and severity of the tsunami.
+The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations
+\begin{equation} \label{eq:swe}
+    \begin{aligned}
+    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
+    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
+    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
+    \end{aligned}
+\end{equation}
+where $h$ is the depth of the water column, $u$ and $v$ the velocities in the longitudinal and latitudinal directions respectively, $g$ the acceleration due to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom friction coefficient.
+As is common in tsunami modeling, $C_f$ is calculated given a Manning's $n$ parameterization such that
+\begin{equation}
+    C_f = \frac{g n^2}{h^{5/3}}
+\end{equation}
+where $n$ is an empirically determined parameter.
 
-The $\text{M}_\text{w}$ 9.0 great \tohoku earthquake of 2011 and the accompanying tsunami caused massive damage to the Japanese coastlines in the Sendai region and beyond including damage in Crescent City, CA.
-The earthquake magnitude was
+The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.
+The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing even when the momentum is non-zero, and entropy corrections \cite{George:2008aa}.
 
-A number of studies have looked at the
+One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, a tsunami wave).
+\geoclaw implements these algorithms following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.
 
-\subsection{\geoclaw}
+\alert{Need to describe how we handled the AMR in the UQ calculation.}
 
-\geoclaw is Based on work by \cite{MacInnes:2013cr}, \cite{Berger:2011du}, \cite{Berger:2011vi}, \cite{George:2008aa}
+\subsection{The \tohoku Tsunami} \label{ssub:tohoku}
+
+The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
+The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
+In this study we have based our simulation on a recent study that attempted to ascertain which source earthquake model matched the observed data optimally \cite{MacInnes:2013cr}.
+From that study, we have used \alert{SOMETHING} source model and the publicly available bathymetry (the finest bathymetry located near the Japanese coastline was not included here and should not have a large impact on the deep-ocean observations we are utilizing).
 
@@
 
-Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system
+The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
+The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \cite{Percival:2011}.
+The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:dart_locations} and whose de-tided data for the event are show in figure~\ref{fig:dart_data}.
 
-gauges 21418, 21413, 21401, 21419 measured maximum wave amplitudes of 1.86, 0.77, 0.66 and 0.54 m respectively
+% \begin{figure}[tb]    
+%     \centering
+%     \begin{subfigure}[b]{0.48\textwidth}
+%         \includegraphics[width=\textwidth]{./figures/DART_locations}
+%         \caption{}
+%         \label{fig:dart_locations}
+%     \end{subfigure}
+%     \begin{subfigure}[b]{0.48\textwidth}
+%         \includegraphics[width=\textwidth]{./figures/DART_data}
+%         \caption{}
+%         \label{fig:dart_data}
+%     \end{subfigure}
+%     \caption{Stuff}
+%   \end{figure}
 
 \subsection{Parametric Uncertainty}
+To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $n$ was sampled in the interval $[0.005-0.2]$.  The regions
+
 Uniform distribution is assumed as the prior for the three manning coefficients.
 such that UN = [0.005-0.2]. The three regions are shown in Figure~\ref{fig:setup}(right)
-
-\subsection{Observations}
-During this tsunami, different gauges were installed to observed the surface elevation
-The locations of theses gauges are shown in Figure~\ref{setup} .
-The gauges are used in the inverse problem to infer the manning coefficients.

### 2013-12-11 setplot.py
Minor plotting fix

@@ -279,6 +279,7 @@ def setplot(plotdata):
     plotaxes.ylimits = 'auto'
     plotaxes.title = r"Manning's $N$ Coefficient"
     plotaxes.scaled = True
+    plotaxes.afteraxes = addgauges
 
     plotitem = plotaxes.new_plotitem(name='friction',plot_type='2d_pcolor')
     plotitem.plot_var = lambda cd: cd.aux[1,:,:]
@@ -290,7 +291,8 @@ def setplot(plotdata):
     plotitem.colorbar_label = "Manning's-$n$ Coefficient"
     plotitem.amr_celledges_show = [0,0,0]
     plotitem.amr_patchedges_show = [0,0,0,0,0,0,0]
-    plotitem.colorbar_label = r"$n$"
+    plotitem.colorbar_label = r"$n$")
+
 
     #-----------------------------------------
     

### 2013-12-11 plot_DART_data.py
Add base DART plotting facility

@@ -1,0 +1,16 @@
+#!/usr/bin/env python
+
+import numpy
+
+import matplotlib.pyplot as plt
+
+def read_gauge(path, time_interval=[0.0,1e10]):
+
+    data = numpy.loadtxt(path, skiprows=2)
+
+
+def plot_gauge_data():
+    fig = plt.figure()
+
+def plot_gauge_locations():
+    fig = plt.figure()

### 2013-12-11 paper/analysis.tex

@@
 The mean of a QoI and its standard deviation is computed
-from the PC coefficients as shown in Equation~\ref{eq:mean} and Equation~\ref{eq:sigma}. 
+from the PC coefficients as shown in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
 In Figure~\ref{fig:ave} we plot the evolution of

### 2013-12-11 paper/mcmc.tex

@@
 The observed data collected at different locations 
-in the likelihood function (Equation~\ref{eq:likelihood}) to update the input parameters.
+in the likelihood function (Equation~\eqref{eq:likelihood}) to update the input parameters.
 MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
@@
 The chain for the variance is also shown in 
-Figure \ref{fig:mcmc} and appears to be well mixed.
+Figure~\ref{fig:mcmc} and appears to be well mixed.
 

### 2013-12-11 paper/results.tex

@@
 \label{sec:results}
-The results of the forward and inverse problems are presented and 
-discussed in three sections. First, we perform an error analysis 
-of the PC expansion construction, namely  to establish the 
-validity of the corresponding surrogate.  Second, we perform a 
-sensitivity analysis and rank the impact of the different Manning's 
-roughness coefficient on the predicted surface elevation. Finally, 
-posterior distributions for the input parameters are determined 
-and analyzed in light of the available gauge data.
+We begin the results section by performing
+an error and convergence analysis of the
+constructed surrogate using the PC expansions
+to establish its validity in Section~\ref{sec:analysis}.
+Then in Section~\ref{sec:forward}, we present
+statistical analysis to quantify the uncertainty in the predicted surface elevation
+as well as a sensitivity analysis to rank the impact of the different Manning's 
+roughness coefficients on this uncertainty. 
+Finally, in Section~\ref{sec:inverse}, we present the results of the inverse
+problem where we determined the posterior distributions of the input parameters 
+and analyzed them in light of the available gauge data.
 \input{pcstudy}

### 2013-12-11 paper/uqpce.tex

@@
 
-The four-dimensional posterior in Equation~\ref{eq:post_coef} can be explored via MCMC;
+The four-dimensional posterior in Equation~\eqref{eq:post_coef} can be explored via MCMC;
 this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
@@
 
-The series representation Equation~\ref{eq:stochseries} can be viewed as a spectral expansion
+The series representation Equation~\eqref{eq:stochseries} can be viewed as a spectral expansion
 of $U$ along the stochastic dimensions. It can be also seen as

### 2013-12-13 paper/intro.tex
Reworked and filled in intro

--- 2013-10-24 paper/intro.tex
@@
+%!TEX root = paper.tex
+
 \section{Introduction}
 
-In this work, we aim to perform an uncertainty quantification
-using Polynomial Chaos Expansion. Quantity of interest is the surface elevation.
-The uncertainty in the manning coefficient is investigated.
-We assume that the Manning coefficient vary by region. Specifically,
-we set three values for the manning coefficient namely, at the land (denoted by N1)
-and at the shore (denoted by N2) and below the water (denoted by N3).
-Uniform distribution is assumed as the prior for the three manning coefficients.
-such that UN = [0.005-0.2]. The three regions are shown in Figure~\ref{fig:setup}(right)
+Tsunamis in the past decade have been responsible for some of the most deadly and costly natural disasters ever recorded.
+Coastal communities have faced this hazard by assessing the risk of these hazards, attempting to make informed decisions about the likelihood that such an event would occur with cost to life and property.
+To do this, computational models of tsunami events are often employed to explore multiple different scenarios.
+Unfortunately, since tsunami events are relatively rare, there is an abundance of uncertainty in the input data for these computational models ranging from effects of the domain such as bathymetry and friction parameterizations, to the earthquake source.
+In this article we present an approach to quantifying the uncertainty in the friction field and a method to find the best parameterized values of the friction field using the \tohoku tsunami as a case study.
+
+Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).
+In this work we concentrate on the uncertainty in the friction parameterization used in tsunami simulations.
+Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.
+
+In this work we have used the widely used Manning's-n law to provide a relationship between the friction coefficient and the roughness parameter $n$.
+The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
+To carry out the uncertainty quantification we have employed polynomial chaos expansions as used in \cite{sraj:2013a,sraj:2013b}.
+Here the quantity of interest is taken to be the surface elevation $\eta$ and will compared to available NOAA Deep-ocean Assessment and Reporting of Tsunamis (DART) buoys for the \tohoku tsunami.
+The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.
+
+\begin{itemize}
+    \item Possibly mention this \url{http://arxiv.org/pdf/1203.6297.pdf}?  Talks about creating a surrogate model for land-slide generated tsunamis.
+\end{itemize}

### 2013-12-13 paper/tsunami.tex
Minor changes to tsunami section

@@
 
-\section{Tsunami Computational Model and Case Study Descriptions} \label{sec:model_description}
+\section{Simulation Setup}
 
-In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following UQ analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
+In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following uncertainty analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
 
@@
 The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
-In this study we have based our simulation on a recent study that attempted to ascertain which source earthquake model matched the observed data optimally \cite{MacInnes:2013cr}.
-From that study, we have used \alert{SOMETHING} source model and the publicly available bathymetry (the finest bathymetry located near the Japanese coastline was not included here and should not have a large impact on the deep-ocean observations we are utilizing).
+In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.
 
-\subsection{Observations}
+\subsubsection{Observations}
 
 The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
-The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \cite{Percival:2011}.
+The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
 The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup} and whose de-tided data for the event are show in Figure~\ref{fig:obs}.
@@
 
+\subsubsection{Source Model}
+
+Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.
+After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and
+
+\begin{itemize}
+    \item Perhaps do one of the tsunami inversion fault models
+\end{itemize}
+
+Since the previous study found that the source model used did not impact the off-shore gauge stations, the ones used here from \cite{Ammon:2011dm}, was chose the source model that was commonly available.
+Access to the finest bathymetry located near the Japanese coastline was also not included as it primarily effected only the shore observations.
+
+\subsubsection{Bathymetry}
+
 \subsection{Parametric Uncertainty}

### 2013-12-16 paper/pcstudy.tex
revisiting results section

@@
 Figure~\ref{fig:rlzs} shows the evolution of the
-water surface elevation for the 125 realizations obtained 
-to compute the PC expansions at the four different gauge 
-locations. We notice that the variation in surface elevation 
-is negligible during the first hour at gauge number 21418 
-and during the first two hours for the remaining gauges
-as the plots of the different realizations are seen to superimpose.
-However, a noticeable uncertainty in the surface 
-elevation can be seen  at all gauges as indicated from the thickness 
-of the bands formed by the plots of the realizations. This uncertainty
-occurs till the end of the simulations.
+water surface elevation predicted at the four different gauges 
+using \geoclaw for the 125 different realizations 
+required to compute the PC expansions. We notice that the 
+variability in water surface elevation is insignificant for the first 
+hour at all gauges as the plots of the different realizations superimpose.
+Later, the variability starts to increase with time at the different gauges 
+as indicated from the thickness of the bands formed by the plots of the realizations
+and is significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
+for gauges 21401, 21413, 21418, 21419 respectively.
+This is consistent with the distance from the gauge to the source of the earthquake
+where gauge 21418 is the closest and gauge 21419 is the farthest.
+The uncertainty in the water surface  elevation continues till the end of the simulations
+at all gauges.
 
-In order to check the consistency of the approximation, we compare
+In order to check the consistency of the PC approximation, we compare
 water surface elevation from the realizations 
 with those obtained from the PC surrogate. The different curves
-show an excellent agreement (not shown) for all times. We also define
+show an excellent agreement (not shown) for all times and gauges locations.
+To quantify this agreement, we define
 an error metric that measures the relative normalized root mean-square error between the left hand side function 
-in Equation~(\ref{eq:stochseries})
-and its PC representation at the sampling points:
+in Equation~(\ref{eq:stochseries}) and its PC representation at the sampling points:
+
 \begin{equation} 
@@
 and Figure~\ref{fig:pdfs3}
-The different curves
-correspond to increased order of PC, from 1 to 5.
+The different curves correspond to increased order of PC, from 1 to 5.
 The plots indicate that the double peaked distributions are

### 2013-12-16 paper/results.tex
revisiting results section

@@
 \label{sec:results}
-We begin the results section by performing
-an error and convergence analysis of the
-constructed surrogate using the PC expansions
-to establish its validity in Section~\ref{sec:analysis}.
-Then in Section~\ref{sec:forward}, we present
-statistical analysis to quantify the uncertainty in the predicted surface elevation
-as well as a sensitivity analysis to rank the impact of the different Manning's 
-roughness coefficients on this uncertainty. 
-Finally, in Section~\ref{sec:inverse}, we present the results of the inverse
-problem where we determined the posterior distributions of the input parameters 
-and analyzed them in light of the available gauge data.
+We begin the results section by performing an error and convergence 
+analysis of the constructed surrogate using the PC expansions to 
+establish its validity in Section~\ref{sec:analysis}.  In Section
+~\ref{sec:forward}, we present a statistical analysis to quantify the 
+uncertainty in the predicted water surface elevation in addition to a 
+sensitivity analysis that enables us to rank the impact of the 
+different Manning's roughness coefficients on the uncertainty. Finally, 
+in Section~\ref{sec:inverse}, we present the results of the inverse problem 
+where we determine the posterior distributions of the input uncertain
+parameters and analyze them in light of the available gauge data.
 \input{pcstudy}

### 2013-12-16 paper/tsunami.tex
revisiting results section

@@
 
-The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations
+The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
+
 \begin{equation} \label{eq:swe}
@@
 \end{equation}
-where $h$ is the depth of the water column, $u$ and $v$ the velocities in the longitudinal and latitudinal directions respectively, $g$ the acceleration due to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom friction coefficient.
-As is common in tsunami modeling, $C_f$ is calculated given a Manning's $n$ parameterization such that
+
+where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
+longitudinal and latitudinal directions respectively, $g$ the acceleration due 
+to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
+friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
+given a Manning's $n$ parameterization such that:
+
 \begin{equation}

### 2013-12-16 paper/analysis.tex
more edits for results

--- 2013-11-13 paper/analysis.tex
@@
 \label{sec:forward}
-We now use the surrogate created effectively using PC to propagate prior 
-uncertainty through the forward model. We mainly exploit the surrogate 
-to study the statistics of the water surface elevation, and to conduct 
+The PC expansion created using an ensemble of 125 \geoclaw simulations
+is now used as a surrogate to propagate prior input parameter uncertainty 
+through the forward model. We here exploit the surrogate 
+to study the statistics of the water surface elevation and conduct 
 a global sensitivity analysis of the impact of the uncertain input parameters.
-We note that the statistical and sensitivity analysis presented below 
-is performed using the PC coefficients solely; these were calculated
-from an ensemble of 125 \geoclaw simulations.
+
 \subsubsection{Statistical analysis}
-The mean of a QoI and its standard deviation is computed
-from the PC coefficients as shown in Equation~\ref{eq:mean} and Equation~\ref{eq:sigma}. 
+The mean of the QoI and its standard deviation is computed
+from the PC coefficients as indicated in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
 In Figure~\ref{fig:ave} we plot the evolution of
@@
 Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
-reflected in a significant standard deviation as seen earlier from the realizations in Figure~\ref{fig:rlzs}.
-We observed that the standard deviation ......
+reflected in a significant standard deviation as seen earlier from the realizations shown 
+in Figure~\ref{fig:rlzs}. An interesting observations is that the
+standard deviation in water surface elevations vanishes at few times instances
+during the tsunami. This can be explained as ......
 
@@
 \subsubsection{Sensitivity analysis}
-A global sensitivity analysis is next conducted to quantity the contribution of each
+A global sensitivity analysis is next performed to quantity the contribution of each
 uncertain parameter to the variance in water surface elevation. To this end, we calculate 
@@
 In 2D, the sensitivity analysis shows also that $N_2$ is dominant
-where we have variance in water surface elevation. This is
+at the same regions where we observe variance in water surface elevation. This is
 indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
@@
 \subsubsection{Response surface}
-In addition to the static and analysis, the PC surrogate 
+In addition to the statistical and sensitivity analysis, the PC surrogate 
 can be used to construct a response surface for the uncertain input parameters.
 This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
-range. Since, $N_3$ shows insignificant contribution in the 
+range. Since $N_3$ shows insignificant contribution in the 
 uncertainty in the model output, we find the impact of $N_2$ and $N_3$ for 
-fixed $N_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
-and Figure~\ref{fig:response3}
-for different times as indicated. The most striking features are the relatively flat
+fixed value of $N_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
+and Figure~\ref{fig:response3} for different times as indicated. The most striking features are the relatively flat
 horizontal contours in the $N_3$ direction indicating that water surface elevation depends
-only mildly on $N_3$ even during peak tsunami events.
+only mildly on $N_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,
+at gauge 21419, its the opposite case where the horizontal contours are in the $N_2 $ direction
+and there water surface elevation depends mildly on $N_2$.

### 2013-12-16 paper/mcmc.tex
more edits for results

@@
 we need to make sure that the observed data are
-comparable to the simulated one.
-To this end, Figure~\ref{fig:compare}
-compares PC mean water surface elevation profiles
+comparable to the simulated one. To this end, Figure~\ref{fig:compare}
+compares PC mean water surface elevation profiles (with error bars)
 with their counterparts from the observed gauge locations.
-The plots show good agreement between the simulations and the observations at different
-locations and at different times. Figure~\ref{fig:compare} can be recast as a scatter plot shown in 
-Figure~\ref{fig:scatter}.
+The plots show a good agreement between the simulations and the 
+observations at different locations and at different times. 
+The same comparison can be recast as a scatter plot shown in 
+Figure~\ref{fig:scatter}. The difference between the observations
+and simulations is attributed to the uncertainty in the 
+Manning's N coefficient and to the measurement errors to be quantified
+in this section.
 
-We now can exploit the surrogate model in the Bayesian inference of the Manning's 
+The surrogate model is exploited in the Bayesian inference of the Manning's 
 roughness coefficient.  To this end, an adaptive MCMC method is used to sample 

### 2013-12-16 paper/pcstudy.tex
more edits for results

@@
 and Figure~\ref{fig:pdfs3}
-The different curves correspond to increased order of PC, from 1 to 5.
+The different curves correspond to increased order of PC (from 1 to 5).
 The plots indicate that the double peaked distributions are
-well-resolved with order 4 but becomes weakly insenstive to further refinement 
-as shown with order 5. The various error metrics presented above 
-provide confidence that the PC expansion is a faithful 
-model surrogate.
+well-resolved with order 3 at t = 7200 s as shown in Figure~\ref{fig:pdfs2}.
+However, at t = 10800 s further refinement is needed as the pdfs are
+sensitive to the refinement up to order 4 but then becomes weakly 
+insensitive with additional refinement  (order 5) as indicated in Figure~\ref{fig:pdfs3}. 
+We, therefore, used PC order 5 in all computations.
+This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 
+model surrogate that can be used in both the forward and inverse problems..

### 2013-12-16 paper/tsunami.tex
some edits to tsnunami section and adding authors

@@
 
-\subsubsection{Observations}
-
-The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
-The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
-The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup} and whose de-tided data for the event are show in Figure~\ref{fig:obs}.
-
-% \begin{figure}[tb]    
-%     \centering
-%     \begin{subfigure}[b]{0.48\textwidth}
-%         \includegraphics[width=\textwidth]{./figures/DART_locations}
-%         \caption{}
-%         \label{fig:dart_locations}
-%     \end{subfigure}
-%     \begin{subfigure}[b]{0.48\textwidth}
-%         \includegraphics[width=\textwidth]{./figures/DART_data}
-%         \caption{}
-%         \label{fig:dart_data}
-%     \end{subfigure}
-%     \caption{Stuff}
-%   \end{figure}
-
 \subsubsection{Source Model}
@@
 \subsubsection{Bathymetry}
+The bathymetry is shown in Figure~\ref{fig:setup} in addition to the topography
+of the Pacific Ocean.
+\subsubsection{Observations}
 
+The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
+The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
+The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
 \subsection{Parametric Uncertainty}
-To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $n$ was sampled in the interval $[0.005-0.2]$.  The regions
-
-Uniform distribution is assumed as the prior for the three manning coefficients.
-such that UN = [0.005-0.2]. The three regions are shown in Figure~\ref{fig:setup}(right)
+In this work, we seek to quantify parametric uncertainty namely due 
+to the uncertainty in the roughness parameter $n$. To this end,
+we characterize the roughness $N$ by three different parameters
+that are varied in three different regions but are assume
+constant in each region. The three regions are as follows: 
+on-shore who coefficnet is denoted by $N_1$, near-shore 
+denoted by $N_2$, and deep-water denoted by $N_3$.
+The three regions are shown in Figure~\ref{fig:ceofs}.
+To carry out the uncertainty quantification we have employed polynomial chaos expansions 
+as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
+to our problem is described in Section~\ref{sec:uqpce}.  
+To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $n$ was sampled in the interval $[0.005-0.2]$.

### 2013-12-17 paper/intro.tex
added two paragraphs to the intro but needs some cleaning

@@
 Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.
+Recently, Sarri \emph{et al.} built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}.
+The implemented emulator was the Gaussian process
+that requires using a combination of prior knowledge about the simulator and
+appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
+where they performed basic statistical and sensitivity analysis
 
-In this work we have used the widely used Manning's-n law to provide a relationship between the friction coefficient and the roughness parameter $n$.
+In this work, we take advantage of the recent developments in Uncertainty Quantification 
+(UQ) methods, which make it feasible to probe the sensitivity of complex and realistic tsunami models 
+to uncertain parameters, and to assess the amount of information that can
+be gained from specific measurements and observations. In particular, 
+we implement an inverse modeling approach towards the objective stated above.  
+Specifically, we will rely on Polynomial Chaos (PC) expansions to construct a 
+faithful surrogate of the response of the large and complex geophysical 
+model simulations.  The availability of the surrogate enables us to efficiently 
+implement a Bayesian inference formalism to the inverse problem, namely 
+because it dramatically reduces the cost of sampling the posterior distribution.
+
+PC expansions have been developed and applied in the engineering community to quantify uncertainties
+in numerical simulations; their principal use is in propagating input
+uncertainties through large, complex and non-linear models to compute the
+ensuing output uncertainties. Several researchers have been 
+investigating the applicability of PC expansions for UQ in oceanic simulations
+\citep{thacker2012,ashwanth2010,Alexanderian2012}. Recently,
+PC combined with Bayesian inference was used to estimate drag coefficient
+during Typhoon Fanapi 2010. The inference was done using
+adjoint method and MCMC \cite{sraj:2013a,sraj:2013b}.
+
+In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
+to estimate Manning N friction.
+We use the widely used Manning's-n law to provide a relationship between the friction coefficient and the roughness parameter $n$.
 The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
-To carry out the uncertainty quantification we have employed polynomial chaos expansions as used in \cite{sraj:2013a,sraj:2013b}.
+To carry out the uncertainty quantification we have employed polynomial chaos expansions.
 Here the quantity of interest is taken to be the surface elevation $\eta$ and will compared to available NOAA Deep-ocean Assessment and Reporting of Tsunamis (DART) buoys for the \tohoku tsunami.
 The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.
-
-\begin{itemize}
-    \item Possibly mention this \url{http://arxiv.org/pdf/1203.6297.pdf}?  Talks about creating a surrogate model for land-slide generated tsunamis.
-\end{itemize}

### 2013-12-17 paper/intro.tex
more changes and figs

@@
 appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
-where they performed basic statistical and sensitivity analysis
+where they performed basic statistical and sensitivity analysis.
+\comment{In this work we reformulate a statistical data assimilation method generally used in the estimation of model states to estimate this model parameter \cite{Mayo2013} . We show that
+low-dimensional representations of Manning’s n coefficients can be recovered by assimilating
+water elevation data. This is a promising approach to parameter estimation in coastal ocean
+modeling.}
 

### 2013-12-17 paper/pcstudy.tex
more changes and figs

--- 2013-12-16 paper/pcstudy.tex
@@
 hour at all gauges as the plots of the different realizations superimpose.
-Later, the variability starts to increase with time at the different gauges 
-as indicated from the thickness of the bands formed by the plots of the realizations
-and is significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
+Later in time, the variability starts to increase with time at the different gauges 
+as indicated from the thickness of the bands formed by the plots of the realizations.
+This variability appears to be significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
 for gauges 21401, 21413, 21418, 21419 respectively.
 This is consistent with the distance from the gauge to the source of the earthquake
-where gauge 21418 is the closest and gauge 21419 is the farthest.
-The uncertainty in the water surface  elevation continues till the end of the simulations
+(epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup}.
+The uncertainty in the prediction of water surface  elevation continues till the end of the simulations
 at all gauges.
@@
 water surface elevation from the realizations 
-with those obtained from the PC surrogate. The different curves
-show an excellent agreement (not shown) for all times and gauges locations.
+with those obtained from the PC surrogate. The different curves (not shown) 
+reveal an excellent agreement for all times and gauges locations.
 To quantify this agreement, we define
@@
 \end{equation}
-where $\NISP$ is the 125-member ensemble obtained using the PC algorithm. 
+where $\NISP$ is the 125-member ensemble obtained to construct the PC surrogate. 
 This error metric calculated at the different gauge locations is shown in Figure~\ref{fig:error};
@@
 
-A final check consists of verifying whether the probability density
+A final check for the validity of the PC approximation 
+consists of verifying whether the probability density
 functions (pdfs) of water surface elevation at the different gauge locations
-converges with increased order of the PC representation.  Sample
+converges with increased order of the PC expansion.  Sample
 water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
-and Figure~\ref{fig:pdfs3}
-The different curves correspond to increased order of PC, from 1 to 5.
-The plots indicate that the double peaked distributions are
-well-resolved with order 4 but becomes weakly insenstive to further refinement 
-as shown with order 5. The various error metrics presented above 
-provide confidence that the PC expansion is a faithful 
-model surrogate.
+and Figure~\ref{fig:pdfs3} where
+the different curves correspond to increased order of PC ($p= 1-5$).
+The plots indicate double peaked distributions that are
+well-resolved with PC order $p=3$ at t = 7200 s as shown in Figure~\ref{fig:pdfs2}.
+However, at t = 10800 s further refinement is needed as the pdfs are
+sensitive to the refinement up to order $p=4$ but then becomes weakly 
+insensitive with additional refinement  ($p = 5$) as indicated in Figure~\ref{fig:pdfs3}. 
+We, therefore, used PC order $p = 5$ in all computations below.
+This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 
+model surrogate that can be used in both the forward and inverse problems.

### 2013-12-19 paper/intro.tex
work in progress

--- 2013-12-17 paper/intro.tex
@@
 appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
-where they performed basic statistical and sensitivity analysis
+where they performed basic statistical and sensitivity analysis.
+In another recent work, Mayo et el. reformulated a statistical data assimilation method generally used in the estimation of model states to estimate Manning’s n coefficients~\cite{Mayo2013}. 
+They used a low-dimensional representations of Manning’s n coefficients and recovered it by assimilating
+water elevation data.
 

### 2013-12-19 paper/tsunami.tex
work in progress

@@
 
-\alert{Need to describe how we handled the AMR in the UQ calculation.}
+\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
+steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
+one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
+Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
+at the gauges ready.... Not sure if this explanation is worth including here ///}
 
@@
 to our problem is described in Section~\ref{sec:uqpce}.  
-To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $n$ was sampled in the interval $[0.005-0.2]$.
+To represent the uncertainty in the Manning's $N$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $N \in [0.005-0.2]$.

### 2013-12-19 paper/analysis.tex
last changes from my side... i think

@@
 standard deviation in water surface elevations vanishes at few times instances
-during the tsunami. This can be explained as ......
+during the tsunami. \alert{This can be explained as ...... }
+
+These statistics can be used to compare with the 
+(DART) buoys observations. Figure~\ref{fig:compare} 
+the PC mean water surface elevation with the observed 
+water surface elevation at the different gauges locations
+with error bars indicated at few time instances. 
+The plots show a good agreement between the PC simulations and the 
+observations at different locations and at different times.
 

### 2013-12-19 paper/mcmc.tex
last changes from my side... i think

@@
 
-Before attempting solving the inverse problem,
-we need to make sure that the observed data are
-comparable to the simulated one. To this end, Figure~\ref{fig:compare}
-compares PC mean water surface elevation profiles (with error bars)
-with their counterparts from the observed gauge locations.
+Before we attempt solving the inverse problem of parameter estimation, 
+we first check the ability of \geoclaw to simulate water surface elevation
+realistically. To this end, we present a comparison between the 
+(DART) buoys observations  and their \geoclaw model counterparts
+for a reference simulation in which the  the Manning's N coefficients were set to their optimal values $N_1=N_2=N_3=0.015$. Figure~\ref{fig:scatter} 
+shows a scatter plot that compares the observed 
+water surface elevation at the different gauges locations and the  \geoclaw model counterparts. 
 The plots show a good agreement between the simulations and the 
 observations at different locations and at different times. 
-The same comparison can be recast as a scatter plot shown in 
-Figure~\ref{fig:scatter}. The difference between the observations
-and simulations is attributed to the uncertainty in the 
-Manning's N coefficient and to the measurement errors to be quantified
-in this section.
+The difference between the observations and simulations is attributed to the uncertainty in the 
+Manning's N coefficient and to the measurement errors.
+\alert{An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
+the epicenter of the earthquake and therefore receiving more impact by the tsunami.}
 
-The surrogate model is exploited in the Bayesian inference of the Manning's 
-roughness coefficient.  To this end, an adaptive MCMC method is used to sample 
+Bayesian inference is now used to estimate the Manning's 
+roughness coefficient ranges that minimize this scatter between 
+observed and modeled water surface elevation, and to estimate the variance of the noise in the measured data.
+To this end, an adaptive MCMC method is used to sample 
 the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
 update the Manning's roughness coefficient distributions in light of the 
-observed data. This sampling, demanding tens of thousands of forward simulations, 
-would have been prohibitively expensive in the
-absence of the surrogate, as the generation of each sample would have required an
-independent \geoclaw realization. The surrogate provides a computationally
-efficient alternative, and requires only evaluating the PCE for different values 
-of the seed $\xxi$. The setup of the Bayesian inference problem,
-MCMC chains for the input drag parameters and the corresponding posterior distributions are 
-presented and discussed in this section.
+observed data. This sampling requires tens of thousands of 
+\geoclaw model runs that are prohibitively expensive as each MCMC 
+sample requires an independent \geoclaw realization. Instead,
+the surrogate model created using PC expansions provides a computationally
+efficient alternative that requires only evaluating the PC expansion
+for different values of the germ $\xxi$.
 
-The observed data collected at different locations 
-in the likelihood function (Equation~\eqref{eq:likelihood}) to update the input parameters.
-MCMCs of $50,000$ iterations are obtained for the Manning roughness coefficients: 
+The MCMC algorithm required $50,000$ iterations to guarantee convergence
+and to obtain the posteriors of the Manning roughness coefficients: 
 $N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
-shows the sample chains for
-the input parameters for different iterates of the MCMC algorithms. 
-The different panels
-shows well-mixed chains for all parameters.
-However, all chains span the entire range
-of the prior, and so it appears that the observations are not informative 
-concerning this uncertain input.  $N_{2}$ chain appears to be concentrated in the lower end of the
-parameter range, with values between 0.005 and 0.1.
-The chain for the variance is also shown in 
-Figure~\ref{fig:mcmc} and appears to be well mixed.
+shows the sample chains for the input parameters for different iterations of the MCMC algorithms. 
+The different panels shows well-mixed chains for all input parameters as well as the variance $\sigma^2$.
+The chain of  $N_{1}$ span the entire range of the prior meaning the observations are not informative 
+concerning this uncertain input.  On the contrary, the $N_{2}$ chain appears to be concentrated in the 
+lower end of the parameter range, with values between 0.005 and 0.1. 
+While for $N_{3}$, the chain appears to be concentrated in the 
+upper end of the parameter range. Finally, the chain for the variance 
+appears to be well mixed with a well defined posterior range between 0.02 and 0.022.
 
-The computed MCMC chains can be readily used to determine the posterior 
-distribution; kernel density estimation (KDE) is used for this purpose
-~\citep{Parzen1962,Silverman1986}(The first 5000 iterates, associated 
-with the burn-in period, are discarded). The resulting posterior pdfs 
+Next, the computed MCMC chains are used to determine the posterior 
+distributions using kernel density estimation (KDE)
+~\citep{Parzen1962,Silverman1986}.  The resulting posterior pdfs 
 of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
-in Figure~\ref{fig:pdfs}.  As expected from the chains shown in Figure
-~\ref{fig:mcmc}, the posterior pdf of $N_2$, exhibits well-defined peak, 
-with a Maximum A Posteriori (MAP) estimate of around 0.011; an extended tail 
-towards the higher Manning's roughness coefficient values is also observed.
-In contrast, the posterior pdfs of $N_1$ appear to be fairly flat, 
-and similar to the uniform prior. This is an indication that 
-the observed data were not useful to refine our prior knowledge for $N_1$.
+in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
+Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
+As expected from the chains shown in Figure
+~\ref{fig:mcmc}, the posterior pdfs of $N_1$ appear to be fairly flat, 
+and similar to the uniform prior; an indication that 
+the observed data were not useful to refine our prior knowledge for $N_1$.,
+In contrast, the posterior pdf of $N_2$ exhibits well-defined peak, 
+with a Maximum A Posteriori (MAP) estimate of around 0.011
+and an extended tail towards the higher Manning's roughness coefficient values.
 While for $N_3$, we observed a posterior that has a well defined peak
-of 0.185 but no clear pdf shape.
-The posterior distribution of the variance is also shown in Figure~\ref{fig:mcmc}. 
-Taking the square root of the MAP values yields the water surface elevation standard 
-deviation, and the latter is a reflection of the mismatch between the model and 
-observed data estimated to be 0.145~$m$.
+of 0.185 but no clear pdf shape. The posterior distribution of the variance 
+appears to be well-defined of Gaussian-like shape. 
+The MAP value of $\sigma^2=0.021$ can be used to estimate the water surface elevation standard 
+deviation estimated to be 0.145~$m$. 
+This value is a reflection of the mismatch between the model and 
+observed data.
 

### 2013-12-19 paper/tsunami.tex
last changes from my side... i think

@@
 The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
-The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21418, 21413, 21401, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
+The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
 \subsection{Parametric Uncertainty}

### 2013-12-23 paper/coefs.tex
revisiting results section
modified intro and added one section on manning

@@
 
-We seek to infer the Manning coefficient from water surface elevation
+We seek to infer the Manning's n coefficients from water surface elevation
 data measured at different gauges as shown in Figure~\ref{fig:setup}.
@@
 The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the
-three Manning Friction coefficients treated as the unknown parameters. 
+three Manning's coefficients treated as the unknown parameters. 
 In this case  Equation~\eqref{eq:post} can be written as:
+
 \begin{equation} 
@@
 prior for the model parameters is assumed as indicated:
+
 \begin{equation} 
@@
 We thus assume a Jeffreys prior \citep{sivia}, expressed as:
+
 \begin{equation} 

### 2013-12-23 paper/formulation.tex
modified intro and added one section on manning

@@
 First, the Bayesian approach to inverse problems is described in Section~\ref{sec:inference};
-its application to our problem of inferring the Manning Friction Coefficients in Section~\ref{sec:manning}.
+its application to our problem of inferring the Manning's n friction coefficients in Section~\ref{sec:manning}.
 Finally, in Section~\ref{sec:uqpce}, we introduce

### 2013-12-23 paper/intro.tex
last changes from my side... i think
modified intro and added one section on manning

--- 2013-12-17 paper/intro.tex
@@
 
-Tsunamis in the past decade have been responsible for some of the most deadly and costly natural disasters ever recorded.
-Coastal communities have faced this hazard by assessing the risk of these hazards, attempting to make informed decisions about the likelihood that such an event would occur with cost to life and property.
-To do this, computational models of tsunami events are often employed to explore multiple different scenarios.
-Unfortunately, since tsunami events are relatively rare, there is an abundance of uncertainty in the input data for these computational models ranging from effects of the domain such as bathymetry and friction parameterizations, to the earthquake source.
-In this article we present an approach to quantifying the uncertainty in the friction field and a method to find the best parameterized values of the friction field using the \tohoku tsunami as a case study.
+Tsunamis in the past decade have been responsible for some of the most 
+deadly and costly natural disasters ever recorded. Coastal communities 
+have faced this hazard by assessing the risk of these hazards, attempting 
+to make informed decisions about the likelihood that such an event would 
+occur with cost to life and property.  To do this, computational models of 
+tsunami events are often employed to explore multiple different scenarios
+to predict the water elevation caused by the tsunamis. The accurate prediction 
+of water elevations requires accurate estimation of many model parameters that 
+are either measured directly, defined empirically, or inferred from
+sparsely collected data. Unfortunately, since tsunami events are relatively rare, 
+there is an abundance of uncertainty in the input data for these computational 
+models ranging from effects of the domain such as bathymetry and friction parameterizations, 
+to the earthquake source.  In this article, we aim to estimate the Manning's n friction
+coefficient that plays a key role in predicting the water elevation. 
+Our approach is to quantify the uncertainty in the predicted water elevation 
+and employ an inverse modeling approach to estimate the Manning's n coefficient 
+using real observations of water elevation measured during the \tohoku tsunami as a case study.
 
 Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).
-In this work we concentrate on the uncertainty in the friction parameterization used in tsunami simulations.
+For instance, Sarri \emph{et al.} recently built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}.
+The implemented emulator was the Gaussian process that requires using a combination of prior knowledge about the simulator and
+appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
+where they performed basic statistical and sensitivity analysis.
+In this work we focus on the uncertainty in the friction parameterization used in tsunami simulations.
 Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.
-Recently, Sarri \emph{et al.} built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}.
-The implemented emulator was the Gaussian process
-that requires using a combination of prior knowledge about the simulator and
-appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
-where they performed basic statistical and sensitivity analysis
+In a recent work, Mayo et el.
+reformulated a statistical data assimilation method generally used in the estimation of model states to estimate Manning's n coefficient~\cite{Mayo2013}.
+They used a low-dimensional representations of Manning's n coefficients and recovered it by assimilating water elevation data.
 
-In this work, we take advantage of the recent developments in Uncertainty Quantification 
-(UQ) methods, which make it feasible to probe the sensitivity of complex and realistic tsunami models 
-to uncertain parameters, and to assess the amount of information that can
-be gained from specific measurements and observations. In particular, 
-we implement an inverse modeling approach towards the objective stated above.  
-Specifically, we will rely on Polynomial Chaos (PC) expansions to construct a 
+In this work, we aim at making use of the recent developments in Uncertainty Quantification 
+(UQ) methods that allows probing the sensitivity of realistic tsunami models 
+to uncertain parameters, and inferring those parameters from specific observations. 
+In particular, we implement an inverse modeling approach that relied on the Bayesian inference approach.  
+We will also rely on Polynomial Chaos (PC) expansions to construct a 
 faithful surrogate of the response of the large and complex geophysical 
-model simulations.  The availability of the surrogate enables us to efficiently 
-implement a Bayesian inference formalism to the inverse problem, namely 
-because it dramatically reduces the cost of sampling the posterior distribution.
+model simulations.  The construction of the surrogate enables us to efficiently 
+implement the Bayesian inference formalism as it dramatically reduces the cost of 
+sampling the posterior distribution.
 
@@
 PC combined with Bayesian inference was used to estimate drag coefficient
-during Typhoon Fanapi 2010. The inference was done using
-adjoint method and MCMC \cite{sraj:2013a,sraj:2013b}.
+during Typhoon Fanapi 2010. The same problem was solved using
+the adjoint method and MCMC \cite{sraj:2013a,sraj:2013b}.
 
 In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
-to estimate Manning N friction.
-We use the widely used Manning's-n law to provide a relationship between the friction coefficient and the roughness parameter $n$.
+to estimate Manning'n n coefficient of friction. 
+%We use the widely used Manning's n law to provide a relationship between the friction coefficient and the roughness parameter $n$.
 The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
@@
 The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.
+
+\alert{maybe an outline for this paper}

### 2013-12-23 paper/paper.tex
method section is completed for now
Minor modifications, mostly adding convenient packages
Fleshing out approach section
results section in progress and minor edits
Paper update
Minor changes to tsunami section
some edits to tsnunami section and adding authors
modified intro and added one section on manning

@@
 \usepackage{natbib}
+\usepackage{xspace} % Incorporates correct optional space after commands
+\usepackage{hyperref}
 
 % Add line numbers
-\usepackage[mathlines]{lineno}
+\usepackage[mathlines]{lineno} % Control line numbers
 
 % Multipart figures
-\usepackage{subcaption}
+\usepackage{subcaption}  % "Newest" subfigure like package
 
@@
 
-\usepackage{ifpdf}
+\usepackage{ifpdf} % For declaring rules for graphics below
 
@@
 \newcommand{\comment}[1]{\textcolor{red}{{#1}}}
+\newcommand{\alert}[1]{\textbf{\textcolor{red}{{#1}}}}
 \newcommand{\be}{\begin{equation}}
@@
 
+\newcommand{\pd}{\pard}
 \newcommand{\geoclaw}{{\sc GeoClaw}\xspace}
 \newcommand{\tohoku}{T\={o}hoku\xspace}
+\newcommand{\dx}{\ensuremath{\Delta x}}   % Delta x
+\newcommand{\dy}{\ensuremath{\Delta y}}   % Delta y
+\newcommand{\dt}{\ensuremath{\Delta t}}   % Delta t
 
@@
 
-\title{Polynomial Chaos for the Estimation of Manning's Friction}
+\title{Estimation of Manning's Friction Coefficient using Polynomial Chaos and Bayesian inference}
 \date{\today}
@@
 % \ead[http://users.ices.utexas.edu/~kyle]{http://users.ices.utexas.edu/~kyle}
+\author[duke]{Omar M. Knio}
+\author[ut]{Clint N. Dawson}
 \author[kaust]{Ibrahim Hoteit}
-\author[ut]{Clint N. Dawson}
 
@@
 \address[ut]{Institute for Computational Engineering and Science, University of Texas at Austin, 201 E 24th ST. Stop C0200, Austin, TX 78712-1229, USA}
-
+\address[duke]{Department of Mechanical Engineering and Materials Science, Duke University, 144
+Hudson Hall, Durham, North Carolina 27708, USA}
 \date{\today}
@@
 
+Tsunamis have been lately catastrophic due to the warnings
+issued late and to poorly predicted water elevations. Computational 
+models of tsunami events are often employed to predict water elevations 
+by exploring multiple different scenarios due to the abundance of uncertainty 
+in the input data. Accurate estimation of water elevations requires 
+accurate estimation of many model parameters. Our objective here is to 
+present an efficient method to estimate the Manning's n friction coefficient
+that plays an important role in the tsunami governing equations. 
+In our work, we perform a sensitivity analysis of the Manning’s n coefficient 
+and infer using water surface elevation data obtained during \tohoku tsunami. 
+We characterize the Manning’s friction coefficient by three different parameters 
+assumed constant in three regions: $n_1$ on-shore, $n_2$ near-shore and $n_3$ deep-water. 
+The efficiency of our approach stems from the use of polynomial chaos expansions to build an inexpensive surrogate for the numerical tsunami \geoclaw model that can be used to perform the sensitivity analysis.
+The surrogate also reduces the computational burden of the Markov Chain Monte-Carlo sampling needed for the Bayesian inference.
+Our objective is to sharpen three initial estimates of the three uncertain parameters.
+Our results indicate that Manning’s n friction coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n3=0.185$ while for $n_1$ no meaningful MAP value can be determined using the available data.
+
 \end{abstract}
@@
 \begin{keyword}
-tsunami \sep Manning's N friction \sep sensitivity analysis \sep polynomial chaos
+tsunami \sep Manning's n friction coefficient \sep sensitivity analysis \sep polynomial chaos \sep Bayesian inference
 \end{keyword}
@@
 \input{results}
+\clearpage
 
 \input{figures}
-\clearpage
 

### 2013-12-23 paper/tsunami.tex
modified intro and added one section on manning

@@
-%!TEX root = paper.tex
+\section{Manning’s Friction Coefficient}
+%We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
+
+Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
+water equations governing tsunamis and the roughness parameter $n$. 
+This Manning's $n$ coefficient is variable that is highly spatial 
+and dependent on the surface characteristics of the sea-floor. 
+The direct measurement of $n$ is therefore impossible, instead, values
+are often assigned from tables that are empirically estimated
+for common types of land that are determined from satellite images.
+As a result, the Manning’s n coefficients used in tsunami models
+often contain large amounts of uncertainty.
+
+The uncertainty in the Manning's $n$ coefficient leads to uncertainties
+in the predicted quantities of interest such as the water elevations
+and becomes a health risk during tsunami events. In addition, 
+the prediction of these quantities of interest are often highly sensitive to 
+the changes in these uncertain parameters. 
+In this work, we seek to quantify parametric uncertainty namely due 
+to the uncertainty in the roughness parameter $n$. To this end,
+we characterize the roughness $n$ by three different parameters
+that are varied in three different regions but are assumed
+constant in each region. The three regions are as follows: 
+on-shore whose coefficient is denoted by $n_1$, near-shore 
+denoted by $n_2$, and deep-water denoted by $n_3$.
+The three regions are shown in Figure~\ref{fig:ceofs}.
+To carry out the uncertainty quantification we have employed polynomial chaos expansions 
+as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
+to our problem is described in Section~\ref{sec:uqpce}.  
+To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $n \in [0.005-0.2]$.
 
@@
 The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
-\subsection{Parametric Uncertainty}
-In this work, we seek to quantify parametric uncertainty namely due 
-to the uncertainty in the roughness parameter $n$. To this end,
-we characterize the roughness $N$ by three different parameters
-that are varied in three different regions but are assume
-constant in each region. The three regions are as follows: 
-on-shore who coefficnet is denoted by $N_1$, near-shore 
-denoted by $N_2$, and deep-water denoted by $N_3$.
-The three regions are shown in Figure~\ref{fig:ceofs}.
-To carry out the uncertainty quantification we have employed polynomial chaos expansions 
-as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
-to our problem is described in Section~\ref{sec:uqpce}.  
-To represent the uncertainty in the Manning's $N$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $N \in [0.005-0.2]$.

### 2013-12-23 paper/paper.tex
title and abstract revisited

--- 2013-11-12 paper/paper.tex
@@
 
-\title{Polynomial Chaos for the Estimation of Manning's Friction}
+\title{Estimation of Manning's Friction Coefficient using DART Buoys Data during
+\tohoku Tsunami}
 \date{\today}
@@
 
+Tsunamis have been lately catastrophic due to the warnings
+issued late and to the inaccurate predictions of water elevations. 
+Tsunami computational models are often employed to explore multiple 
+different scenarios to predict water elevations due to the abundance 
+of uncertainty in the input data. However, accurate estimation of water elevations 
+requires accurate estimation of many model parameters,
+in particular the Manning's $n$ friction coefficient that plays an important role in the 
+tsunami governing equations. Our objective here is to present an efficient method for the 
+estimation of Manning's $n$ coefficient characterized by three different parameters 
+assumed constant in three regions: $n_1$ on-shore, $n_2$ near-shore and $n_3$ deep-water. 
+This method uses Polynomial Chaos (PC) to build an inexpensive surrogate for the 
+\geoclaw model and then employs the Bayesian inference that uses the DART buoys data
+obtained during \tohoku tsunami to estimate the three uncertain parameters. The efficiency of our approach stems from 
+using a surrogate model that reduces the computational burden of the Markov Chain Monte-Carlo 
+sampling needed in the Bayesian inference. The PC surrogate is also used to perform a sensitivity analysis. 
+Our results indicate that Manning’s $n$ coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined using the available data.
+
 \end{abstract}

### 2013-12-23 paper/intro.tex
Minor wording change

--- 2013-12-19 paper/intro.tex
@@
 \section{Introduction}
+\alert{We need to unify the name of the input parameters, Manning coefficint or friction or roughness.
+what is the best name ??}
 
 Tsunamis in the past decade have been responsible for some of the most deadly and costly natural disasters ever recorded.
-Coastal communities have faced this hazard by assessing the risk of these hazards, attempting to make informed decisions about the likelihood that such an event would occur with cost to life and property.
+Coastal communities have faced this hazard by assessing the risk they pose, attempting to make informed decisions about the likelihood that such an event would occur with cost to life and property.
 To do this, computational models of tsunami events are often employed to explore multiple different scenarios.

### 2013-12-23 paper/analysis.tex

@@
 of each of the uncertain parameters is shown in Figure~\ref{fig:sens} at the four gauges. 
-The Manning's roughness coefficient at the shore $N_2$ is clearly dominant and contributes
+The Manning's n coefficient at the shore $N_2$ is clearly dominant and contributes
 the most to the variance in the water surface elevation compared to the other two 
-Manning's roughness coefficients $N_1$ and $N_3$; this is true for almost the entire simulation time
-and at the four gauges. The Manning's roughness coefficient
+Manning's n coefficients $N_1$ and $N_3$; this is true for almost the entire simulation time
+and at the four gauges. The Manning's n coefficient
 in the bottom of ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
-during the second hour of simulation and Manning's roughness coefficient
+during the second hour of simulation and Manning's n coefficient
 at the land $N_1$ appears to be an insignificant contributor

### 2013-12-23 paper/intro.tex

--- 2013-12-23 paper/intro.tex
@@
 deadly and costly natural disasters ever recorded. Coastal communities 
-have faced this hazard by assessing the risk of these hazards, attempting 
+have faced this hazard by assessing the risk they pose, attempting 
 to make informed decisions about the likelihood that such an event would 

### 2013-12-23 paper/mcmc.tex

@@
 (DART) buoys observations  and their \geoclaw model counterparts
-for a reference simulation in which the  the Manning's N coefficients were set to their optimal values $N_1=N_2=N_3=0.015$. Figure~\ref{fig:scatter} 
+for a reference simulation in which the  the Manning's N coefficients were set to their optimal values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
 shows a scatter plot that compares the observed 
@@
 The difference between the observations and simulations is attributed to the uncertainty in the 
-Manning's N coefficient and to the measurement errors.
+Manning's n coefficients and to the measurement errors.
 \alert{An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
@@
 Bayesian inference is now used to estimate the Manning's 
-roughness coefficient ranges that minimize this scatter between 
+n coefficients ranges that minimize this scatter between 
 observed and modeled water surface elevation, and to estimate the variance of the noise in the measured data.
@@
 the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
-update the Manning's roughness coefficient distributions in light of the 
+update the Manning's n coefficients distributions in light of the 
 observed data. This sampling requires tens of thousands of 
@@
 The MCMC algorithm required $50,000$ iterations to guarantee convergence
-and to obtain the posteriors of the Manning roughness coefficients: 
-$N_1$,$N_2$ and $N_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
+and to obtain the posteriors of the Manning's n coefficients: 
+$n_1$,$n_2$ and $n_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
 shows the sample chains for the input parameters for different iterations of the MCMC algorithms. 
 The different panels shows well-mixed chains for all input parameters as well as the variance $\sigma^2$.
-The chain of  $N_{1}$ span the entire range of the prior meaning the observations are not informative 
-concerning this uncertain input.  On the contrary, the $N_{2}$ chain appears to be concentrated in the 
+The chain of  $n_{1}$ span the entire range of the prior meaning the observations are not informative 
+concerning this uncertain input.  On the contrary, the $n_{2}$ chain appears to be concentrated in the 
 lower end of the parameter range, with values between 0.005 and 0.1. 
-While for $N_{3}$, the chain appears to be concentrated in the 
+While for $n_{3}$, the chain appears to be concentrated in the 
 upper end of the parameter range. Finally, the chain for the variance 
@@
 ~\citep{Parzen1962,Silverman1986}.  The resulting posterior pdfs 
-of the three Manning's roughness coefficients $N_1,N_2,N_3$ are shown 
+of the three Manning's n coefficients $n_1,n_2,n_3$ are shown 
 in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
-Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
+Note that the first 10000 iterates, associated with the burn-in period, were discarded.  
 As expected from the chains shown in Figure
-~\ref{fig:mcmc}, the posterior pdfs of $N_1$ appear to be fairly flat, 
+~\ref{fig:mcmc}, the posterior pdfs of $n_1$ appear to be fairly flat, 
 and similar to the uniform prior; an indication that 
-the observed data were not useful to refine our prior knowledge for $N_1$.,
-In contrast, the posterior pdf of $N_2$ exhibits well-defined peak, 
+the observed data were not useful to refine our prior knowledge for $n_1$.,
+In contrast, the posterior pdf of $n_2$ exhibits well-defined peak, 
 with a Maximum A Posteriori (MAP) estimate of around 0.011
-and an extended tail towards the higher Manning's roughness coefficient values.
-While for $N_3$, we observed a posterior that has a well defined peak
-of 0.185 but no clear pdf shape. The posterior distribution of the variance 
+and an extended tail towards the higher Manning's n coefficient values.
+While for $n_3$, we observed a posterior that has a well defined peak
+of 0.18 but no clear pdf shape. The posterior distribution of the variance 
 appears to be well-defined of Gaussian-like shape. 
@@
 
-%MAPS 0.021308543478839   0.011083123405057   0.185744276109897   0.021045124706835
+%   0.180222380841571   0.011347632046398   0.185486684362890   0.021050903421753

### 2013-12-23 paper/intro.tex
Revision of intro, redid wrapping

@@
 
-Tsunamis in the past decade have been responsible for some of the most 
-deadly and costly natural disasters ever recorded. Coastal communities 
-have faced this hazard by assessing the risk they pose, attempting 
-to make informed decisions about the likelihood that such an event would 
-occur with cost to life and property.  To do this, computational models of 
-tsunami events are often employed to explore multiple different scenarios
-to predict the water elevation caused by the tsunamis. The accurate prediction 
-of water elevations requires accurate estimation of many model parameters that 
-are either measured directly, defined empirically, or inferred from
-sparsely collected data. Unfortunately, since tsunami events are relatively rare, 
-there is an abundance of uncertainty in the input data for these computational 
-models ranging from effects of the domain such as bathymetry and friction parameterizations, 
-to the earthquake source.  In this article, we aim to estimate the Manning's n friction
-coefficient that plays a key role in predicting the water elevation. 
-Our approach is to quantify the uncertainty in the predicted water elevation 
-and employ an inverse modeling approach to estimate the Manning's n coefficient 
-using real observations of water elevation measured during the \tohoku tsunami as a case study.
+Tsunamis in the past decade have been responsible for some of the most  deadly
+and costly natural disasters ever recorded. Coastal communities  have faced this
+hazard by assessing the risk they pose, attempting  to make informed decisions
+about the likelihood that such an event would  occur with cost to life and
+property.  To do this, computational models of  tsunami events are often
+employed to explore multiple different scenarios to predict the water elevation
+caused by the tsunamis. The accurate prediction  of water elevations requires
+accurate estimation of many model parameters that  are either measured directly,
+defined empirically, or inferred from sparsely collected data. Unfortunately,
+since tsunami events are relatively rare,  there is an abundance of uncertainty
+in the input data for these computational  models ranging from effects of the
+domain such as bathymetry and friction parameterizations,  to the earthquake
+source.  In this article, we aim to estimate the Manning's n friction
+coefficient. Our approach quantifies the uncertainty in the predicted water
+elevation  and employs an inverse modeling approach to estimate the Manning's
+$n$ coefficient using observations of water elevation measured during an event.
 
-Previous work looking into uncertainty in the input for tsunami simulations often has focused on the earthquake source but mainly as a way of constraining the various proposed earthquake scenarios only after the event (see for example \cite{MacInnes:2013cr}).
-For instance, Sarri \emph{et al.} recently built a statistical surrogate of a landslide-generated tsunami computer model by using an emulator~\cite{Sarri2012}.
-The implemented emulator was the Gaussian process that requires using a combination of prior knowledge about the simulator and
-appropriate choices of functions and parameters. Their work was presented as a proof-of-concept case study
-where they performed basic statistical and sensitivity analysis.
-In this work we focus on the uncertainty in the friction parameterization used in tsunami simulations.
-Changes to the friction parameterization has been shown in the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies were limited in the scope of the changes allowed.
-In a recent work, Mayo et el.
-reformulated a statistical data assimilation method generally used in the estimation of model states to estimate Manning's n coefficient~\cite{Mayo2013}.
-They used a low-dimensional representations of Manning's n coefficients and recovered it by assimilating water elevation data.
+Previous work looking into uncertainty in the input for tsunami simulations
+often has focused on the earthquake source but primarily as a way of
+constraining the various proposed earthquake scenarios only after the event (see
+for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
+tsunami was presented in Sarri \emph{et al.}, building a statistical surrogate
+using an emulator \cite{Sarri2012}.  The implemented emulator was the Gaussian
+process that requires using a combination of prior knowledge about the simulator
+and appropriate choices of functions and parameters. Their work was presented as
+a proof-of-concept case study where they performed basic statistical and
+sensitivity analysis. In this work we focus on the uncertainty in the friction
+parameterization used in tsunami simulations.  Changes to the friction
+parameterization has been shown in the past to lead to significant changes to
+inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies
+were limited in the scope of the changes to the parameterization allowed.  In
+recent work, Mayo et el. reformulated a statistical data assimilation method
+generally used in the estimation of model states to estimate the Manning's n
+coefficient~\cite{Mayo2013}. They used a low-dimensional representations of
+Manning's n coefficients and recovered it by assimilating water elevation data.
 
-In this work, we aim at making use of the recent developments in Uncertainty Quantification 
-(UQ) methods that allows probing the sensitivity of realistic tsunami models 
-to uncertain parameters, and inferring those parameters from specific observations. 
-In particular, we implement an inverse modeling approach that relied on the Bayesian inference approach.  
-We will also rely on Polynomial Chaos (PC) expansions to construct a 
-faithful surrogate of the response of the large and complex geophysical 
-model simulations.  The construction of the surrogate enables us to efficiently 
-implement the Bayesian inference formalism as it dramatically reduces the cost of 
-sampling the posterior distribution.
+\alert{Did the Talia et al. study use synthetic data?  Probably should mention
+this if that was the case.}
 
-PC expansions have been developed and applied in the engineering community to quantify uncertainties
-in numerical simulations; their principal use is in propagating input
-uncertainties through large, complex and non-linear models to compute the
-ensuing output uncertainties. Several researchers have been 
+The method proposed here utilizes recent developments in uncertainty
+quantification (UQ) methods that allow non-intrusive probing of the sensitivity
+of realistic tsunami  models to uncertain parameters, inferring those parameters
+from specific  observations.  In particular, we implement an inverse modeling
+approach that relies on the Bayesian inference approach.  We will also rely on
+polynomial chaos (PC) expansions to construct a faithful surrogate of the
+response of the large and complex geophysical model simulations.  THe
+construction of the surrogate enables us to efficiently implement the Bayesian
+inference formalism as it dramatically reduces the cost of sampling the
+posterior distribution.
+
+PC expansions have been developed and applied in the engineering community to
+quantify uncertainties in numerical simulations; their principal use is in
+propagating input uncertainties through large, complex and non-linear models to
+compute the ensuing output uncertainties. Several researchers have been
 investigating the applicability of PC expansions for UQ in oceanic simulations
-\citep{thacker2012,ashwanth2010,Alexanderian2012}. Recently,
-PC combined with Bayesian inference was used to estimate drag coefficient
-during Typhoon Fanapi 2010. The same problem was solved using
-the adjoint method and MCMC \cite{sraj:2013a,sraj:2013b}.
+\citep{thacker2012,ashwanth2010,Alexanderian2012}. Recently, PC combined with
+Bayesian inference was used to estimate drag coefficient during Typhoon Fanapi
+2010. The same problem was solved using the adjoint method and MCMC
+\cite{sraj:2013a,sraj:2013b}.
 
 In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
-to estimate Manning'n n coefficient of friction. 
-%We use the widely used Manning's n law to provide a relationship between the friction coefficient and the roughness parameter $n$.
-The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
-To carry out the uncertainty quantification we have employed polynomial chaos expansions.
-Here the quantity of interest is taken to be the surface elevation $\eta$ and will compared to available NOAA Deep-ocean Assessment and Reporting of Tsunamis (DART) buoys for the \tohoku tsunami.
-The forward model runs are performed using \geoclaw following the setup and using one of the source models explored in \cite{MacInnes:2013cr}.
+to estimate Manning's $n$ coefficient of friction using the \tohoku tsunami
+event as a case-study.  The quantity of interest is taken to be the surface
+elevation $\eta$ and is compared to the available NOAA Deep-ocean Assessment and
+Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
+\geoclaw package following the setup in MacInnes \emph{et al.} including the
+source models analyzed there \cite{MacInnes:2013cr}.
 

### 2013-12-23 paper/analysis.tex
minor

@@
 of each of the uncertain parameters is shown in Figure~\ref{fig:sens} at the four gauges. 
-The Manning's n coefficient at the shore $N_2$ is clearly dominant and contributes
+The Manning's $n$ coefficient at the shore $n_2$ is clearly dominant and contributes
 the most to the variance in the water surface elevation compared to the other two 
-Manning's n coefficients $N_1$ and $N_3$; this is true for almost the entire simulation time
-and at the four gauges. The Manning's n coefficient
-in the bottom of ocean $N_{3}$ at gauge number 21419 exhibits small sensitivity index 
-during the second hour of simulation and Manning's n coefficient
-at the land $N_1$ appears to be an insignificant contributor
+Manning's $n$ coefficients $n_1$ and $n_3$; this is true for almost the entire simulation time
+and at the four gauges. The Manning's $n$ coefficient
+in the bottom of ocean $n_{3}$ at gauge number 21419 exhibits small sensitivity index 
+during the second hour of simulation and Manning's $n$ coefficient
+at the land $n_1$ appears to be an insignificant contributor
 to the variance.
 
-In 2D, the sensitivity analysis shows also that $N_2$ is dominant
+In 2D, the sensitivity analysis shows also that $n_2$ is dominant
 at the same regions where we observe variance in water surface elevation. This is
 indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
-for $N_1$(top row) $N_2$(center row) and $N_3$ (bottom row)
+for $n_1$(top row) $n_2$(center row) and $n_3$ (bottom row)
 at different times as indicated in each panel.
@@
 This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
-range. Since $N_3$ shows insignificant contribution in the 
-uncertainty in the model output, we find the impact of $N_2$ and $N_3$ for 
-fixed value of $N_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
+range. Since $n_3$ shows insignificant contribution in the 
+uncertainty in the model output, we find the impact of $n_2$ and $n_3$ for 
+fixed value of $n_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
 and Figure~\ref{fig:response3} for different times as indicated. The most striking features are the relatively flat
-horizontal contours in the $N_3$ direction indicating that water surface elevation depends
-only mildly on $N_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,
-at gauge 21419, its the opposite case where the horizontal contours are in the $N_2 $ direction
-and there water surface elevation depends mildly on $N_2$.
+horizontal contours in the $n_3$ direction indicating that water surface elevation depends
+only mildly on $n_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,
+at gauge 21419, its the opposite case where the horizontal contours are in the $n_2 $ direction
+and there water surface elevation depends mildly on $n_2$.

### 2013-12-23 paper/coefs.tex
minor

@@
-\subsection{Inferring Manning Friction Coefficient}
+\subsection{Inferring Manning's Friction Coefficient}
  \label{sec:manning}
 
-We seek to infer the Manning's n coefficients from water surface elevation
+We seek to infer the Manning's $n$ coefficients from water surface elevation
 data measured at different gauges as shown in Figure~\ref{fig:setup}.
@@
 their model predicted counterparts $\vec G$ is predicted by \geoclaw.
-The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the
+The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
 three Manning's coefficients treated as the unknown parameters. 
@@
 \begin{equation} 
-p(N_1,N_2,N_3,\sigma^2 | \vec d) 
+p(n_1,n_2,n_3,\sigma^2 | \vec d) 
 \propto \frac{1}{\sqrt{2 \pi \sigma^2}} 
@@
 \exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
-\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
+\ p(\sigma^2)p(n_1)p(n_2) p(n_3)
 \label{eq:post_coef}
@@
 on some \emph{a priori} knowledge about the parameters. In our case, a uniform
-prior for the model parameters is assumed as indicated:
+prior for all model parameters is assumed where $n_i \in [0.005,0.2]$ and therefore:
 
 \begin{equation} 
-p(N_i) = \begin{cases}
-		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
-		0 &\text{otherwise}  , 
+p(n_i) = \begin{cases}
+		\displaystyle \frac{1}{0.2-0.005} &\text{for~} 0.005 <  N_i \leq 0.2 ,  \\
+		0 &\text{otherwise}  .
 \end{cases}
 \end{equation}
-where $ [a_i,b_i]$ denote the parameter ranges defined as.
 Regarding the variance, the only information we know 

### 2013-12-23 paper/formulation.tex
minor

--- 2013-11-13 paper/formulation.tex
@@
 First, the Bayesian approach to inverse problems is described in Section~\ref{sec:inference};
-its application to our problem of inferring the Manning Friction Coefficients in Section~\ref{sec:manning}.
+its application to our problem of inferring the Manning's $n$ coefficients in Section~\ref{sec:manning}.
 Finally, in Section~\ref{sec:uqpce}, we introduce

### 2013-12-23 paper/intro.tex
minor

@@
 domain such as bathymetry and friction parameterizations,  to the earthquake
-source.  In this article, we aim to estimate the Manning's n friction
+source.  In this article, we aim to estimate the Manning's $n$ friction
 coefficient. Our approach quantifies the uncertainty in the predicted water
@@
 recent work, Mayo et el. reformulated a statistical data assimilation method
-generally used in the estimation of model states to estimate the Manning's n
+generally used in the estimation of model states to estimate the Manning's $n$
 coefficient~\cite{Mayo2013}. They used a low-dimensional representations of
-Manning's n coefficients and recovered it by assimilating water elevation data.
+Manning's $n$ coefficients and recovered it by assimilating water elevation data.
 

### 2013-12-23 paper/mcmc.tex
minor

@@
 (DART) buoys observations  and their \geoclaw model counterparts
-for a reference simulation in which the  the Manning's N coefficients were set to their optimal values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
+for a reference simulation in which the  the Manning's $n$ coefficients were set to their optimal values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
 shows a scatter plot that compares the observed 
@@
 The difference between the observations and simulations is attributed to the uncertainty in the 
-Manning's n coefficients and to the measurement errors.
+Manning's $n$ coefficients and to the measurement errors.
 \alert{An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
@@
 Bayesian inference is now used to estimate the Manning's 
-n coefficients ranges that minimize this scatter between 
+$n$ coefficients ranges that minimize this scatter between 
 observed and modeled water surface elevation, and to estimate the variance of the noise in the measured data.
@@
 the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
-update the Manning's n coefficients distributions in light of the 
+update the Manning's $n$ coefficients distributions in light of the 
 observed data. This sampling requires tens of thousands of 
@@
 The MCMC algorithm required $50,000$ iterations to guarantee convergence
-and to obtain the posteriors of the Manning's n coefficients: 
+and to obtain the posteriors of the Manning's $n$ coefficients: 
 $n_1$,$n_2$ and $n_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
@@
 ~\citep{Parzen1962,Silverman1986}.  The resulting posterior pdfs 
-of the three Manning's n coefficients $n_1,n_2,n_3$ are shown 
+of the three Manning's $n$ coefficients $n_1,n_2,n_3$ are shown 
 in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
@@
 with a Maximum A Posteriori (MAP) estimate of around 0.011
-and an extended tail towards the higher Manning's n coefficient values.
+and an extended tail towards the higher Manning's $n$ coefficient values.
 While for $n_3$, we observed a posterior that has a well defined peak

### 2013-12-23 paper/paper.tex
minor

@@
 sampling needed in the Bayesian inference. The PC surrogate is also used to perform a sensitivity analysis. 
-Our results indicate that Manning’s $n$ coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined using the available data.
+Our results indicate that Manning's $n$ coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined using the available data.
 
@@
 \begin{keyword}
-tsunami \sep Manning's n friction coefficient \sep sensitivity analysis \sep polynomial chaos \sep Bayesian inference
+tsunami \sep Manning's $n$ friction coefficient \sep sensitivity analysis \sep polynomial chaos \sep Bayesian inference
 \end{keyword}

### 2013-12-23 paper/results.tex
minor

@@
 sensitivity analysis that enables us to rank the impact of the 
-different Manning's roughness coefficients on the uncertainty. Finally, 
+different Manning's $n$ coefficients on the uncertainty. Finally, 
 in Section~\ref{sec:inverse}, we present the results of the inverse problem 

### 2013-12-23 paper/tsunami.tex
minor

@@
-\section{Manning’s Friction Coefficient}
+\section{Manning's Friction Coefficient}
 %We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
@@
 for common types of land that are determined from satellite images.
-As a result, the Manning’s n coefficients used in tsunami models
+As a result, the Manning's $n$ coefficients used in tsunami models
 often contain large amounts of uncertainty.
@@
 to our problem is described in Section~\ref{sec:uqpce}.  
-To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $n \in [0.005-0.2]$.
+To represent the uncertainty in the Manning's $n$ coefficients a uniform 
+distribution was assumed as the prior where $n$ was sampled in the interval $n \in [0.005,0.2]$.
 

### 2013-12-23 paper/conc.tex
added outline

\section{Discussion and conclusions}
\label{sec:conc}

### 2013-12-24 paper/intro.tex
restructure tsunami section

@@
 \section{Introduction}
-
+\label{sec:intro}
 Tsunamis in the past decade have been responsible for some of the most  deadly
@@
 approach that relies on the Bayesian inference approach.  We will also rely on
-polynomial chaos (PC) expansions to construct a faithful surrogate of the
-response of the large and complex geophysical model simulations.  THe
+Polynomial Chaos (PC) expansions to construct a faithful surrogate of the
+response of the large and complex geophysical model simulations.  The
 construction of the surrogate enables us to efficiently implement the Bayesian
@@
 
-\alert{maybe an outline for this paper}
+The structure of this manuscript is as follows. First, we motivate the UQ problem 
+by discussing the uncertainties of the drag parameters in Section~\ref{sec:manning}. 
+Section~\ref{sec:simulation} summarizes the forward simulations setup
+and the DART buoys observational data. 
+Section~\ref{sec:formu} describes the two main numerical method employed:
+Bayesian inference formalism that brings together observations 
+and model results is briefly described in Section~\ref{sec:inference} 
+and PC expansions that were used to build the surrogate model are described
+in Section~\ref{sec:uqpce}. Section~\ref{sec:results} describes the main results in 
+three subsections: first, we present evidence that the surrogate is a faithful 
+representation of the \geoclaw model; second, we use the surrogate to explore the 
+response surfaces of simulated water elevations, and to compute their sensitivity 
+to the control parameters; and third, we present the posterior distribution obtained 
+via MCMC sampling and contrast it to the prior distributions. 
+A discussion of our findings and methodology is presented in Section~\ref{sec:conc}.

### 2013-12-24 paper/tsunami.tex
restructure tsunami section

@@
 \section{Manning's Friction Coefficient}
+\label{sec:manning}
 %We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
@@
 
-\section{Simulation Setup}
+\section{The \tohoku Tsunami} 
+\label{ssub:tohoku}
 
-In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following uncertainty analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
+The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
+The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
+In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.
+In this section we briefly describe the forward computational model used, \geoclaw, and the 
+simulation setup used in the uncertainty analysis and finally describe
+the observations from NOAA DART buoys.
+
+%\section{Simulation Setup}
+%\label{sec:simulation}
 
@@
 
-\subsection{The \tohoku Tsunami} \label{ssub:tohoku}
-
-The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
-The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
-In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.
-
-\subsubsection{Source Model}
+\subsection{Source Model and Bathymetry}
 
@@
 Access to the finest bathymetry located near the Japanese coastline was also not included as it primarily effected only the shore observations.
+\comment{The bathymetry is shown in Figure~\ref{fig:setup} in addition to the topography
+of the Pacific Ocean.}
 
-\subsubsection{Bathymetry}
-The bathymetry is shown in Figure~\ref{fig:setup} in addition to the topography
-of the Pacific Ocean.
-\subsubsection{Observations}
+\subsection{DART Buoys Observations}
 

### 2013-12-24 paper/tsunami.tex
more editing to the intro and other edits

@@
 
-Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
-water equations governing tsunamis and the roughness parameter $n$. 
+Manning's $n$ law provides a relationship between the friction coefficient 
+(Equation \ref{eq:coef}) appearing in the shallow
+water equations governing tsunamis (see Equation \ref{eq:swe})
+and the roughness parameter $n$. 
 This Manning's $n$ coefficient is variable that is highly spatial 
@@
 In this work, we seek to quantify parametric uncertainty namely due 
-to the uncertainty in the roughness parameter $n$. To this end,
-we characterize the roughness $n$ by three different parameters
+to the uncertainty in the friction parameter $n$. To this end,
+we characterize the Manning's $n$ coefficient by three different parameters
 that are varied in three different regions but are assumed
@@
 \section{The \tohoku Tsunami} 
-\label{ssub:tohoku}
+\label{sec:tohoku}
 
@@
     C_f = \frac{g n^2}{h^{5/3}}
+\label{eq:coef}
 \end{equation}
@@
 
-\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
-steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
-one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
-Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
-at the gauges ready.... Not sure if this explanation is worth including here ///}
+%\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
+%steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
+%one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
+%Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
+%at the gauges ready.... Not sure if this explanation is worth including here ///}
 
@@
 The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
+\comment{maybe more description of the data in the figure}.

### 2013-12-26 paper/tsunami.tex
Some tsunmai changes

--- 2013-12-23 paper/tsunami.tex
@@
-\section{Manning's Friction Coefficient}
+%!TEX root = paper.tex
+
+\section{Simulation Setup}
+
+In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following uncertainty analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
+
+\subsection{\geoclaw} \label{ssub:geoclaw}
+
+The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
+\begin{equation} \label{eq:swe}
+    \begin{aligned}
+    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
+    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
+    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
+    \end{aligned}
+\end{equation}
+where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
+longitudinal and latitudinal directions respectively, $g$ the acceleration due 
+to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
+friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
+given a Manning's $n$ parameterization such that:
+\begin{equation}
+    C_f = \frac{g n^2}{h^{5/3}}
+\end{equation}
+where $n$ is an empirically determined parameter known as Manning's $n$.
+
+The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.
+The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing, even when the momentum is non-zero, and entropy corrections \cite{George:2008aa}.
+
+\subsubsection{Adaptivity} \label{ssub:adaptivity}
+One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, the disturbance in the surface height $\eta$).
+\geoclaw implements these algorithms following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.
+
+For the simulations presented, refinement thresholds were used that matched what was presented in \cite{MacInnes:2013cr}.
+As was done there, four levels of refinement were used staring with a resolution of 1 degree in both the latitudinal and longitudinal directions down to 0.5' resolution (approximately 900 meters) located around the observation locations.
+The tolerance for the refinement criteria for sea-surface anomaly was 0.02 meters.
+
+\subsection{The \tohoku Tsunami} \label{ssub:tohoku}
+
+The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
+The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
+In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.
+
+\subsubsection{Source Model}
+
+The MacInnes \emph{et al.} study found that source models from tsunami
+inversions were
+
+Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.
+After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and .
+
+[4] Ammon used a seismic and GPS inversion technique
+[7] Saito et al. used a tsunami inversion
+
+Since the previous study found that the source model used did not impact the off-shore gauge stations, the ones used here from \cite{Ammon:2011dm}, was chose the source model that was commonly available.
+Access to the finest bathymetry located near the Japanese coastline was also not included as it primarily effected only the shore observations.
+
+\subsubsection{Bathymetry}
+The bathymetry used in the simulation is a combination of ETOPO 1' and 4' 
+accurate bathymetry for the region \cite{Amante:2009ud}.  The additional 
+bathymetry data used in the MacInnes \emph{et al.} was useful only near the
+shore and had a smaller effect on the regions near the observations used in
+this study.
+
+\subsubsection{Observations}
+The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
+The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
+The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
+
+\section{Manning’s $n$ Friction Parameterization}
 %We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
@@
 for common types of land that are determined from satellite images.
-As a result, the Manning's $n$ coefficients used in tsunami models
+As a result, the Manning’s n coefficients used in tsunami models
 often contain large amounts of uncertainty.
@@
 to our problem is described in Section~\ref{sec:uqpce}.  
-To represent the uncertainty in the Manning's $n$ coefficients a uniform 
-distribution was assumed as the prior where $n$ was sampled in the interval $n \in [0.005,0.2]$.
+To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $n \in [0.005-0.2]$.
 
-\section{Simulation Setup}
-
-In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following uncertainty analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
-
-\subsection{\geoclaw} \label{ssub:geoclaw}
-
-The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
-
-\begin{equation} \label{eq:swe}
-    \begin{aligned}
-    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
-    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
-    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
-    \end{aligned}
-\end{equation}
-
-where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
-longitudinal and latitudinal directions respectively, $g$ the acceleration due 
-to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
-friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
-given a Manning's $n$ parameterization such that:
-
-\begin{equation}
-    C_f = \frac{g n^2}{h^{5/3}}
-\end{equation}
-where $n$ is an empirically determined parameter.
-
-The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.
-The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing even when the momentum is non-zero, and entropy corrections \cite{George:2008aa}.
-
-One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, a tsunami wave).
-\geoclaw implements these algorithms following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.
-
-\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
-steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
-one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
-Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
-at the gauges ready.... Not sure if this explanation is worth including here ///}
-
-\subsection{The \tohoku Tsunami} \label{ssub:tohoku}
-
-The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
-The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
-In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.
-
-\subsubsection{Source Model}
-
-Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.
-After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and
-
-\begin{itemize}
-    \item Perhaps do one of the tsunami inversion fault models
-\end{itemize}
-
-Since the previous study found that the source model used did not impact the off-shore gauge stations, the ones used here from \cite{Ammon:2011dm}, was chose the source model that was commonly available.
-Access to the finest bathymetry located near the Japanese coastline was also not included as it primarily effected only the shore observations.
-
-\subsubsection{Bathymetry}
-The bathymetry is shown in Figure~\ref{fig:setup} in addition to the topography
-of the Pacific Ocean.
-\subsubsection{Observations}
-
-The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
-The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
-The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
+Note that on the Sendai plain a Manning's $n$ value of 0.035 was used to account for pasture, farmland and rice paddies.

### 2013-12-26 paper/formulation.tex

@@
 In this section, we introduce the methods employed in this work.
-First, the Bayesian approach to inverse problems is described in Section~\ref{sec:inference};
-its application to our problem of inferring the Manning's $n$ coefficients in Section~\ref{sec:manning}.
-Finally, in Section~\ref{sec:uqpce}, we introduce
+First, the Bayesian approach to inverse problems its application to our problem of inferring the Manning's $n$ coefficients is described in Section~\ref{sec:inference}.
+In Section~\ref{sec:uqpce}, we introduce
 a stochastic spectral method for forward uncertainty propagation and accelerating the Bayesian inference.

### 2013-12-26 paper/intro.tex

--- 2013-12-23 paper/intro.tex
@@
 employed to explore multiple different scenarios to predict the water elevation
-caused by the tsunamis. The accurate prediction  of water elevations requires
+caused by the tsunamis. The accurate prediction  of water elevations, however, requires
 accurate estimation of many model parameters that are either measured directly,
-defined empirically, or inferred from sparsely collected data. Unfortunately,
+defined empirically, or estimated from a collection of observed data. Unfortunately,
 since tsunami events are relatively rare,  there is an abundance of uncertainty
@@
 elevation  and employs an inverse modeling approach to estimate the Manning's
-$n$ coefficient using observations of water elevation measured during an event.
+$n$ coefficient using observations of water elevation measured during a tsunami event.
 
@@
 for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
-tsunami was presented in Sarri \emph{et al.}, building a statistical surrogate
+tsunami was presented in Sarri \emph{et al.}, by building a statistical surrogate
 using an emulator \cite{Sarri2012}. The implemented emulator was the Gaussian
@@
 and appropriate choices of functions and parameters. Their work was presented as
-a proof-of-concept case study where they performed basic statistical and
-sensitivity analysis. In this work we focus on the uncertainty in the friction
-parameterization used in tsunami simulations.  Changes to the friction
-parameterization has been shown in the past to lead to significant changes to
-inundation levels \cite{Myers:2001el,Jakeman:2010hk} but these previous studies
-were limited in the scope of the changes to the parameterization allowed.  In
-recent work, Mayo et el. reformulated a statistical data assimilation method
+a proof-of-concept case study where they only performed basic statistical and
+sensitivity analysis.  Changes to the friction parameterization has been shown in 
+the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} 
+but these previous studies were limited in the scope of the changes to the parameterization allowed.  
+In a recent work, Mayo et el. reformulated a statistical data assimilation method
 generally used in the estimation of model states to estimate the Manning's $n$
 coefficient~\cite{Mayo2013}. They used a low-dimensional representations of
-Manning's $n$ coefficients and recovered it by assimilating water elevation data.
+Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data.
 
-\alert{Did the Talia et al. study use synthetic data?  Probably should mention
-this if that was the case.}
-
-The method proposed here utilizes recent developments in uncertainty
-quantification (UQ) methods that allow non-intrusive probing of the sensitivity
-of realistic tsunami  models to uncertain parameters, inferring those parameters
-from specific  observations.  In particular, we implement an inverse modeling
-approach that relies on the Bayesian inference approach.  We will also rely on
-polynomial chaos (PC) expansions to construct a faithful surrogate of the
-response of the large and complex geophysical model simulations.  THe
+In this work, we focus our attention on the uncertainty in the friction
+parameterization used in tsunami simulations. We present a three-dimensional 
+representation of Manning's $n$ coefficient and an efficient method
+to estimate this representation using real water elevation data. The method proposed here utilizes recent 
+developments in Uncertainty Quantification (UQ) methods that allow 
+probing of the sensitivity of realistic tsunami  models to uncertain parameters
+without modifying the forward model and inferring those parameters
+from a number of observations.  In particular, we implement an inverse modeling
+approach that relies on the Bayesian inference technique.  We also employ
+Polynomial Chaos (PC) expansions to construct a faithful surrogate model of the
+response of the large and complex geophysical model simulations.  The
 construction of the surrogate enables us to efficiently implement the Bayesian
-inference formalism as it dramatically reduces the cost of sampling the
+inference technique as it dramatically reduces the cost of sampling the
 posterior distribution.
@@
 PC expansions have been developed and applied in the engineering community to
-quantify uncertainties in numerical simulations; their principal use is in
-propagating input uncertainties through large, complex and non-linear models to
-compute the ensuing output uncertainties. Several researchers have been
-investigating the applicability of PC expansions for UQ in oceanic simulations
-\citep{thacker2012,ashwanth2010,Alexanderian2012}. Recently, PC combined with
-Bayesian inference was used to estimate drag coefficient during Typhoon Fanapi
-2010. The same problem was solved using the adjoint method and MCMC
-\cite{sraj:2013a,sraj:2013b}.
+quantify uncertainties in numerical simulations.  The main advantage of
+using PC expansions is the ability of propagating input uncertainties through 
+large, complex and non-linear models to compute the corresponding output uncertainties. 
+Several researchers have been investigating the applicability of PC expansions for 
+UQ in oceanic simulations~\citep{thacker2012,ashwanth2010,Alexanderian2012}. 
+Recently, PC was combined with Bayesian inference to estimate the drag coefficient 
+using temperature data collected during Typhoon Fanapi 2010. The same problem was 
+solved using two different techniques: the adjoint method and Markov Chain Monte Carlo
+(MCMC)\cite{sraj:2013a,sraj:2013b}.
 
 In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
-to estimate Manning's $n$ coefficient of friction using the \tohoku tsunami
+to estimate Manning's $n$ coefficient using the \tohoku tsunami
 event as a case-study. The quantity of interest is taken to be the surface
@@
 
-\alert{maybe an outline for this paper}
+The structure of this manuscript is as follows. First, we discussing the uncertainties 
+of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
+and present a proposed three-dimensional representation of $n$. 
+In Section~\ref{sec:tohoku}, we describe our case-study the \tohoku tsunami,
+summarize the forward simulations setup and present the DART buoys observational data. 
+Section~\ref{sec:formu} describes the two main numerical methods employed in the UQ analysis:
+Bayesian inference formalism that brings together observations 
+and model results (briefly described in Section~\ref{sec:inference}) 
+and PC expansions that were used to build the surrogate model (described
+in Section~\ref{sec:uqpce}). Section~\ref{sec:results} describes the main results in 
+three subsections: first, we perform error and convergence studies
+as evidences that the PC surrogate is a faithful 
+representation of the \geoclaw model; second, we use the PC surrogate to 
+perform a statistical analysis and compute the  sensitivity of simulated water elevations
+to the uncertain input parameters; and third, we present the result of the
+posterior distributions obtained using Bayesian inference by MCMC. A discussion of our findings and methodology is finally presented in Section~\ref{sec:conc}.

### 2013-12-26 paper/paper.tex
restructure tsunami section

@@
 requires accurate estimation of many model parameters,
-in particular the Manning's $n$ friction coefficient that plays an important role in the 
-tsunami governing equations. Our objective here is to present an efficient method for the 
+in particular the Manning's $n$ friction coefficient. Our objective here is to 
+present an efficient method for the 
 estimation of Manning's $n$ coefficient characterized by three different parameters 
@@
 This method uses Polynomial Chaos (PC) to build an inexpensive surrogate for the 
-\geoclaw model and then employs the Bayesian inference that uses the DART buoys data
-obtained during \tohoku tsunami to estimate the three uncertain parameters. The efficiency of our approach stems from 
+\geoclaw model and then employs the Bayesian inference to estimate the three uncertain parameters
+using the DART buoys data obtained during \tohoku tsunami. The efficiency of our approach stems from 
 using a surrogate model that reduces the computational burden of the Markov Chain Monte-Carlo 
@@
 \input{results}
+\input{conc}
+
 \clearpage

### 2014-01-02 paper/coefs.tex
Cleaned up tsunami section and added some references

@@
-\subsection{Inferring Manning's Friction Coefficient}
- \label{sec:manning}
+\subsubsection{Inferring Manning's Friction Coefficient}
+\label{sec:infmanning}
 
 We seek to infer the Manning's $n$ coefficients from water surface elevation
-data measured at different gauges as shown in Figure~\ref{fig:setup}.
+data measured at different gauges as shown in Figure~\ref{fig:setup_buoy_data}.
 Bayesian inference can be applied directly to infer the uncertain parameters

### 2014-01-02 paper/pcstudy.tex
Cleaned up tsunami section and added some references

@@
 This is consistent with the distance from the gauge to the source of the earthquake
-(epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup}.
+(epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup_buoy_locations}.
 The uncertainty in the prediction of water surface  elevation continues till the end of the simulations

### 2014-01-02 paper/tsunami.tex
Cleaned up tsunami section and added some references

@@
 
-\section{Simulation Setup}
+\section{The \tohoku Tsunami} 
+\label{sec:tohoku}
 
-In this section we briefly describe the forward computational model used, \geoclaw, and the case study used in the following uncertainty analysis, the great \tohoku earthquake with observations from NOAA DART buoys.
+The \tohoku earthquake and tsunami of 2011 has been the subject of a number of
+studies due to the wealth of observational evidence and severity of the tsunami.
+The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and
+caused massive devastation across Japan.  In the case study presented, the
+simulation is based on work done in \cite{MacInnes:2013cr} and modified to use
+the specified variable friction field. In this section we briefly describe the
+forward computational model used, \geoclaw, the  simulation setup used in the
+uncertainty analysis and finally describe the observations from NOAA DART buoys.
 
 \subsection{\geoclaw} \label{ssub:geoclaw}
-
 The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
@@
 
-\subsection{The \tohoku Tsunami} \label{ssub:tohoku}
+\subsection{DART Buoys Observations}
 
-The \tohoku earthquake and tsunami of 2011 has been the subject of a number of studies due to the wealth of observational evidence and severity of the tsunami.
-The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and caused massive devastation across Japan.
-In the case study presented, the simulation is based on work done in \cite{MacInnes:2013cr} and modified to use the specified variable friction field.
+The observations used in this study are from the Deep-ocean Assessment and
+Reporting of Tsunamis (DART) buoy system developed and maintained by the
+National Oceanic and Atmospheric Administration (NOAA).  The purpose of the
+network is to provide early-warning detection and forecasting of tsunami
+propagation in the Pacific Ocean \cite{Milburn:1996wm}.  The DART buoys closest
+to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418,
+and 21419 whose locations are shown in Figure~\ref{fig:setup_buoy_locations} and
+whose de-tided data for the event are shown in Figure~\ref{fig:setup_buoy_data}.
+The time $t=0$ is set to be the start of the earthquake.  We can see that the
+primary tsunami waves generated reach the gauges between 40 minutes and 2 hours
+after the earthquake.  The large spikes later in time present in the data at
+gauge 21418 are data anomalies and not subsequent tsunami waves.
 
-\subsubsection{Source Model}
+\subsection{Bathymetry and Earthquake Source Model}
 
-The MacInnes \emph{et al.} study found that source models from tsunami
-inversions were
+The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
+accurate bathymetry for the region \cite{Amante:2009ud}.  The additional finer
+resolution bathymetry data used in MacInnes \emph{et al.} were not used in this
+study since the simulations were not sensitive to the finer bathymetry at the
+DART buoy locations in the deep-water.
 
-Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.
-After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and .
+Two earthquake source models were run to test whether the friction UQ analysis
+was sensitive to different rupture models, the first is from Saito et al.
+\cite{Saito:2011bh} and is based on an inversion analysis based on dispersive
+tsunami simulations.  The second is from Ammon \emph{et al.} \cite{Ammon:2011dm} and
+used inversion of the seismic waves to reconstruct the rupture.  These two
+models were chosen as they take very different approaches to the reconstruction
+of the original fault and were shown to obtain the best un-shifted match to the
+observations at the DART buoys for their respective types of reconstruction.  As
+discussed in MacInnes \emph{et al.}, the source models used do not include rupture
+timing and this could have an impact on the accuracy of the simulation.
 
-[4] Ammon used a seismic and GPS inversion technique
-[7] Saito et al. used a tsunami inversion
+\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}
 
-Since the previous study found that the source model used did not impact the off-shore gauge stations, the ones used here from \cite{Ammon:2011dm}, was chose the source model that was commonly available.
-Access to the finest bathymetry located near the Japanese coastline was also not included as it primarily effected only the shore observations.
+\section{Manning’s $n$ Friction Parameterization} \label{sec:manning}
 
-\subsubsection{Bathymetry}
-The bathymetry used in the simulation is a combination of ETOPO 1' and 4' 
-accurate bathymetry for the region \cite{Amante:2009ud}.  The additional 
-bathymetry data used in the MacInnes \emph{et al.} was useful only near the
-shore and had a smaller effect on the regions near the observations used in
-this study.
+Manning's $n$ law provides a relationship between the friction coefficient
+\eqref{eq:coef} appearing in the shallow water equations \eqref{eq:swe}
+governing tsunamis and the roughness parameter $n$.  This Manning's $n$
+coefficient is variable that is highly spatial  and dependent on the surface
+characteristics of the sea-floor.  The direct measurement of $n$ is therefore
+impossible, instead, values are often assigned from tables that are empirically
+estimated for common types of land that are determined from satellite images. As
+a result, the Manning’s n coefficients used in tsunami models often contain
+large amounts of uncertainty.
 
-\subsubsection{Observations}
-The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).
-The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.
-The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).
-
-\section{Manning’s $n$ Friction Parameterization}
-%We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.
-
-Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
-water equations governing tsunamis and the roughness parameter $n$. 
-This Manning's $n$ coefficient is variable that is highly spatial 
-and dependent on the surface characteristics of the sea-floor. 
-The direct measurement of $n$ is therefore impossible, instead, values
-are often assigned from tables that are empirically estimated
-for common types of land that are determined from satellite images.
-As a result, the Manning’s n coefficients used in tsunami models
-often contain large amounts of uncertainty.
-
-The uncertainty in the Manning's $n$ coefficient leads to uncertainties
-in the predicted quantities of interest such as the water elevations
-and becomes a health risk during tsunami events. In addition, 
-the prediction of these quantities of interest are often highly sensitive to 
-the changes in these uncertain parameters. 
-In this work, we seek to quantify parametric uncertainty namely due 
-to the uncertainty in the roughness parameter $n$. To this end,
-we characterize the roughness $n$ by three different parameters
-that are varied in three different regions but are assumed
-constant in each region. The three regions are as follows: 
-on-shore whose coefficient is denoted by $n_1$, near-shore 
-denoted by $n_2$, and deep-water denoted by $n_3$.
-The three regions are shown in Figure~\ref{fig:ceofs}.
-To carry out the uncertainty quantification we have employed polynomial chaos expansions 
-as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
-to our problem is described in Section~\ref{sec:uqpce}.  
-To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $n \in [0.005-0.2]$.
-
-Note that on the Sendai plain a Manning's $n$ value of 0.035 was used to account for pasture, farmland and rice paddies.
+The uncertainty in the Manning's $n$ coefficient leads to uncertainties in the
+predicted quantities of interest such as the water elevations and becomes a
+health risk during tsunami events. In addition,  the prediction of these
+quantities of interest are often highly sensitive to  the changes in these
+uncertain parameters.  In MacInnes \emph{et al.} a Manning's $n$ coefficient of 0.025
+was used except on the Sendai plane where 0.035 was used to account for pasture,
+farmland and rice paddies.  In this work, we seek to quantify parametric
+uncertainty namely due to the uncertainty in the roughness parameter $n$. To
+this end, we characterize the roughness $n$ by three different parameters that
+are varied in three different regions but are assumed constant in each region.
+The three regions are as follows:  on-shore whose coefficient is denoted by
+$n_1$, near- shore  denoted by $n_2$, and deep-water denoted by $n_3$. The three
+regions are shown in Figure~\ref{fig:ceofs}. To carry out the uncertainty
+quantification we have employed polynomial chaos expansions as used in
+\cite{sraj:2013a,sraj:2013b}.  The method and its application to our problem is
+described in Section~\ref{sec:uqpce}.  To represent the uncertainty in the
+Manning's $n$ coefficients a uniform distribution was assumed as the prior where
+$N$ was sampled in the interval $n \in [0.005-0.2]$.

### 2014-01-02 run_friction.py
Move over to new batch system

@@ -5,20 +5,23 @@ import sys
 
 import numpy
 
-import clawpack.clawutil.tests as clawtests
+import batch
 
-
-class FrictionTest(clawtests.Test):
+class FrictionJob(batch.Job):
     r""""""
 
     def __init__(self, index, values, 
-                                contours=(numpy.infty,0.0,-200.0,-numpy.infty)):
+                       source_path='$SRC/tohoku2011-paper1/sources/Ammon.txydz',
+                       contours=(numpy.infty,0.0,-200.0,-numpy.infty)):
         r""""""
 
-        super(FrictionTest, self).__init__()
+        super(FrictionJob, self).__init__()
+
+        self.source_path = os.path.expandvars(source_path)
 
         self.type = "tohoku"
-        self.name = "friction-sensitivity"
+        self.name = "friction-sensitivity-%s" % \
+                        os.path.split(os.path.splitext(self.source_path)[0])[-1]
         self.prefix = "fric_%s" % (str(index))
         self.executable = 'xgeoclaw'
 
@@ -34,11 +37,16 @@ class FrictionTest(clawtests.Test):
                        self.rundata.clawdata.lower, self.rundata.clawdata.upper,
                        contours, values])
 
+        # Set earthquake source model
+        dtopo_data = self.rundata.dtopo_data
+        dtopo_data.dtopofiles.append([1, 4, 4, self.source_path])
+
     def __str__(self):
-        output = super(FrictionTest, self).__str__()
+        output = super(FrictionJob, self).__str__()
+        output += "\n  Source model: %s" % self.source_path
         output += "\n  Friction Regions:"
         for (n,region) in enumerate(self.rundata.friction_data.friction_regions):
-            output += "\n  Region %s: %s" % (n, region)
+            output += "\n    Region %s: %s" % (n, region)
         return output
 
 
@@ -48,13 +56,16 @@ if __name__ == '__main__':
     if len(sys.argv) > 1:
         path = sys.argv[1]
 
+    source_path = '$SRC/tohoku2011-paper1/sources/Ammon.txydz'
+
     # Read in friction test values
     friction_values = numpy.loadtxt(path)
 
-    # Create all tests
-    tests = []
+    # Create all jobs
+    jobs = []
     for n in xrange(friction_values.shape[0]):
-        tests.append(FrictionTest(n, friction_values[n,:]))
+        jobs.append(FrictionJob(n, friction_values[n,:], source_path))
     
[…]

### 2014-01-02 plot_DART_data.py
Remove old plot script

@@ -1,16 +1,0 @@
-#!/usr/bin/env python
-
-import numpy
-
-import matplotlib.pyplot as plt
-
-def read_gauge(path, time_interval=[0.0,1e10]):
-
-    data = numpy.loadtxt(path, skiprows=2)
-
-
-def plot_gauge_data():
-    fig = plt.figure()
-
-def plot_gauge_locations():
-    fig = plt.figure()

### 2014-01-07 paper/intro.tex
changes based on ibrahm comments

@@
 about the likelihood that such an event would  occur with cost to life and
-property.  To do this, computational models of  tsunami events are often
+property.  Toward this goal, computational models of  tsunami events are often
 employed to explore multiple different scenarios to predict the water elevation
@@
 in the input data for these computational  models ranging from effects of the
-domain such as bathymetry and friction parameterizations,  to the earthquake
-source.  In this article, we aim to estimate the Manning's $n$ friction
-coefficient. Our approach quantifies the uncertainty in the predicted water
-elevation  and employs an inverse modeling approach to estimate the Manning's
+domain, such as bathymetry and friction parameterizations,  to the earthquake
+source.  In this study, we aim to estimate the Manning's $n$ friction
+coefficient \alert{add a sentence here about the Manning Coeff}. 
+We aim at quantifying the uncertainty in the predicted water
+elevation  and employs a Bayesian inverse modeling approach to estimate the Manning's
 $n$ coefficient using observations of water elevation measured during a tsunami event.
 
-Previous work looking into uncertainty in the input for tsunami simulations
-often has focused on the earthquake source but primarily as a way of
+Previous works looking into uncertainty in the input for tsunami simulations
+focused on the earthquake source but primarily as a way of
 constraining the various proposed earthquake scenarios only after the event (see
 for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
-tsunami was presented in Sarri \emph{et al.}, by building a statistical surrogate
-using an emulator \cite{Sarri2012}. The implemented emulator was the Gaussian
+tsunami was presented in Sarri \emph{et al.}~\cite{Sarri2012}, by building a statistical surrogate
+using an emulator. The implemented emulator was based on a Gaussian
 process that requires using a combination of prior knowledge about the simulator
@@
 sensitivity analysis.  Changes to the friction parameterization has been shown in 
-the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk} 
-but these previous studies were limited in the scope of the changes to the parameterization allowed.  
-In a recent work, Mayo et el. reformulated a statistical data assimilation method
+the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk}, 
+but these previous studies \comment{were limited in the scope of the changes to the parameterization allowed}.  
+In a recent work, Mayo \emph{et al.}~\cite{Mayo2013} reformulated a statistical data assimilation method
 generally used in the estimation of model states to estimate the Manning's $n$
-coefficient~\cite{Mayo2013}. They used a low-dimensional representations of
-Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data.
+coefficient. They used a low-dimensional representations of
+Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data
+in a storm surge setting.
 
@@
 representation of Manning's $n$ coefficient and an efficient method
-to estimate this representation using real water elevation data. The method proposed here utilizes recent 
-developments in Uncertainty Quantification (UQ) methods that allow 
+to estimate this representation using real water elevation data. The method proposed here 
+is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
 probing of the sensitivity of realistic tsunami  models to uncertain parameters
@@
 from a number of observations.  In particular, we implement an inverse modeling
-approach that relies on the Bayesian inference technique.  We also employ
+approach that relies on the Bayesian inference technique.  We resort to
 Polynomial Chaos (PC) expansions to construct a faithful surrogate model of the
-response of the large and complex geophysical model simulations.  The
-construction of the surrogate enables us to efficiently implement the Bayesian
+response of the large and complex geophysical model simulations \comment{ref on pcs}.
+The construction of the surrogate enables us to efficiently implement the Bayesian
 inference technique as it dramatically reduces the cost of sampling the
@@
 using PC expansions is the ability of propagating input uncertainties through 
-large, complex and non-linear models to compute the corresponding output uncertainties. 
-Several researchers have been investigating the applicability of PC expansions for 
+large, complex and non-linear models leading to the corresponding output uncertainties. 
+Several studies already investigated the efficiency of PC expansions for 
 UQ in oceanic simulations~\citep{thacker2012,ashwanth2010,Alexanderian2012}. 
-Recently, PC was combined with Bayesian inference to estimate the drag coefficient 
+Recently, PC was combined with Bayesian inference to estimate the wind drag coefficient 
 using temperature data collected during Typhoon Fanapi 2010. The same problem was 
@@
 Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
-\geoclaw package following the setup in MacInnes \emph{et al.} including the
-source models analyzed there \cite{MacInnes:2013cr}.
+\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} including the
+source models analyzed there.
 
-The structure of this manuscript is as follows. First, we discussing the uncertainties 
+The structure of this manuscript is as follows. First, we discuss the uncertainties 
 of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
-and present a proposed three-dimensional representation of $n$. 
-In Section~\ref{sec:tohoku}, we describe our case-study the \tohoku tsunami,
-summarize the forward simulations setup and present the DART buoys observational data. 
-Section~\ref{sec:formu} describes the two main numerical methods employed in the UQ analysis:
-Bayesian inference formalism that brings together observations 
-and model results (briefly described in Section~\ref{sec:inference}) 
-and PC expansions that were used to build the surrogate model (described
-in Section~\ref{sec:uqpce}). Section~\ref{sec:results} describes the main results in 
+and propose a three-dimensional representation of $n$. 
+Section~\ref{sec:tohoku} describes our case-study, the \tohoku tsunami,
+summarize the forward simulations setup, and present the DART buoys observational data. 
+Section~\ref{sec:formu} presents the two main methods employed in the UQ analysis:
+Bayesian inference that brings together observations 
+and model results (Section~\ref{sec:inference}) 
+and PC expansions that are used to build the surrogate model (Section~\ref{sec:uqpce}). Section~\ref{sec:results} describes the main results in 
 three subsections: first, we perform error and convergence studies
@@
 perform a statistical analysis and compute the  sensitivity of simulated water elevations
-to the uncertain input parameters; and third, we present the result of the
+to the uncertain input parameters; and third, we present the results of the
 posterior distributions obtained using Bayesian inference by MCMC. A discussion of our findings and methodology is finally presented in Section~\ref{sec:conc}.

### 2014-01-07 paper/pcstudy.tex
changes based on ibrahm comments

@@
 
-Figure~\ref{fig:rlzs} shows the evolution of the
+Figure~\ref{fig:rlzs} plots the evolution of the
 water surface elevation predicted at the four different gauges 
@@
 required to compute the PC expansions. We notice that the 
-variability in water surface elevation is insignificant for the first 
+variability in water surface elevation is insignificant in the first 
 hour at all gauges as the plots of the different realizations superimpose.
-Later in time, the variability starts to increase with time at the different gauges 
+Later in time, the variability starts to increase at the different gauges 
 as indicated from the thickness of the bands formed by the plots of the realizations.
 This variability appears to be significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
-for gauges 21401, 21413, 21418, 21419 respectively.
+for gauges 21401, 21413, 21418, 21419, respectively.
 This is consistent with the distance from the gauge to the source of the earthquake
 (epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup_buoy_locations}.
-The uncertainty in the prediction of water surface  elevation continues till the end of the simulations
+The uncertainty in the prediction of water surface  elevation persists till the end of the simulations
 at all gauges.
@@
 functions (pdfs) of water surface elevation at the different gauge locations
-converges with increased order of the PC expansion.  Sample
+converges with increased order of the PC expansion  \comment{add referece}.  Sample
 water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
@@
 sensitive to the refinement up to order $p=4$ but then becomes weakly 
-insensitive with additional refinement  ($p = 5$) as indicated in Figure~\ref{fig:pdfs3}. 
+insensitive with additional refinement  ($p = 5$) as shown in Figure~\ref{fig:pdfs3}. 
 We, therefore, used PC order $p = 5$ in all computations below.

### 2014-01-07 paper/tsunami.tex
Fix Manning's n formula coefficient
changes based on ibrahm comments

@@
 %!TEX root = paper.tex
+\section{Manning’s $n$ Friction Coefficient} \label{sec:manning}
+
+Manning's $n$ law provides a relationship between the friction coefficient
+\eqref{eq:coef} appearing in the shallow water equations \eqref{eq:swe}
+governing tsunamis and the roughness parameter $n$.  This Manning's $n$
+coefficient varies spatially  and is dependent on the surface
+characteristics of the sea-floor.  A direct measurement of $n$ is therefore
+impossible, instead, values are often assigned from tables that are empirically
+estimated for common types of land that are determined from satellite images. As
+a result, the Manning’s n coefficients used in tsunami models often contain
+large amounts of uncertainty.
+
+The uncertainty in the Manning's $n$ coefficient leads to uncertainties in the
+predicted quantities of interest such as the water elevations and becomes a
+health risk during tsunami events. In addition,  the prediction of these
+quantities of interest are often highly sensitive to changes in these
+uncertain parameters.  In MacInnes \emph{et al.} a Manning's $n$ coefficient of 0.025
+was used except on the Sendai plane where 0.035 was used to account for pasture,
+farmland and rice paddies.  In this work, we seek to quantify parametric
+uncertainty due to the uncertainty in the Manning's $n$ friction coefficient. To
+this end, we characterize $n$ as being contasnt over three different 
+regions. The three regions are:  on-shore whose coefficient is denoted by
+$n_1$, near- shore  denoted by $n_2$, and deep-water denoted by $n_3$. The three
+regions are shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
+with respect to those three parameters we employ Polynomial Chaos (PC) expansions as used in
+\cite{sraj:2013a,sraj:2013b}.  The method and its application to our problem is
+described in Section~\ref{sec:uqpce}.  To represent the uncertainty in the
+Manning's $n$ coefficients an uninformative uniform distribution was assumed as the prior with
+$n$  sampled from the interval $n \in [0.005-0.2]$.
 
@@
 The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and
-caused massive devastation across Japan.  In the case study presented, the
+caused massive devastation across Japan.  In our case study, the
 simulation is based on work done in \cite{MacInnes:2013cr} and modified to use
-the specified variable friction field. In this section we briefly describe the
-forward computational model used, \geoclaw, the  simulation setup used in the
-uncertainty analysis and finally describe the observations from NOAA DART buoys.
+the specified variable friction field. Hereafter, we briefly describe the
+forward computational model, \geoclaw, the  simulation setup used in the
+uncertainty analysis, and finally present the observations from NOAA DART buoys.
+\alert{describe a bit more the tsunami and its geophysical location}.
+\subsection{\geoclaw} \label{ssub:geoclaw}
+The \geoclaw software package uses a finite volume, wave-propagation approach as described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
 
-\subsection{\geoclaw} \label{ssub:geoclaw}
-The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
 \begin{equation} \label{eq:swe}
@@
     &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
-    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
-    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
+    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu, \\
+    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv,
     \end{aligned}
@@
 given a Manning's $n$ parameterization such that:
+
 \begin{equation}
-    C_f = \frac{g n^2}{h^{5/3}}
+    C_f = \frac{g n^2}{h^{4/3}},
 \label{eq:coef}
@@
 The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.
-The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing, even when the momentum is non-zero, and entropy corrections \cite{George:2008aa}.
+The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing, even when the momentum is non-zero, \alert{and entropy corrections} \cite{George:2008aa}.
 
-\subsubsection{Adaptivity} \label{ssub:adaptivity}
+%\subsubsection{Adaptivity} \label{ssub:adaptivity}
 One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, the disturbance in the surface height $\eta$).
-\geoclaw implements these algorithms following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.
+\geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.
 
@@
 and 21419 whose locations are shown in Figure~\ref{fig:setup_buoy_locations} and
-whose de-tided data for the event are shown in Figure~\ref{fig:setup_buoy_data}.
-The time $t=0$ is set to be the start of the earthquake.  We can see that the
+whose de-tided water surface elevation data for the event are shown in Figure~\ref{fig:setup_buoy_data}.
+The time $t=0$ is set to be the start of the earthquake \comment{frequency of data}.  We can see that the
 primary tsunami waves generated reach the gauges between 40 minutes and 2 hours
@@
 \comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}
-
-\section{Manning’s $n$ Friction Parameterization} \label{sec:manning}
-
-Manning's $n$ law provides a relationship between the friction coefficient
-\eqref{eq:coef} appearing in the shallow water equations \eqref{eq:swe}
-governing tsunamis and the roughness parameter $n$.  This Manning's $n$
-coefficient is variable that is highly spatial  and dependent on the surface
-characteristics of the sea-floor.  The direct measurement of $n$ is therefore
-impossible, instead, values are often assigned from tables that are empirically
-estimated for common types of land that are determined from satellite images. As
-a result, the Manning’s n coefficients used in tsunami models often contain
-large amounts of uncertainty.
-
-The uncertainty in the Manning's $n$ coefficient leads to uncertainties in the
-predicted quantities of interest such as the water elevations and becomes a
-health risk during tsunami events. In addition,  the prediction of these
-quantities of interest are often highly sensitive to  the changes in these
-uncertain parameters.  In MacInnes \emph{et al.} a Manning's $n$ coefficient of 0.025
-was used except on the Sendai plane where 0.035 was used to account for pasture,
-farmland and rice paddies.  In this work, we seek to quantify parametric
-uncertainty namely due to the uncertainty in the roughness parameter $n$. To
-this end, we characterize the roughness $n$ by three different parameters that
-are varied in three different regions but are assumed constant in each region.
-The three regions are as follows:  on-shore whose coefficient is denoted by
-$n_1$, near- shore  denoted by $n_2$, and deep-water denoted by $n_3$. The three
-regions are shown in Figure~\ref{fig:ceofs}. To carry out the uncertainty
-quantification we have employed polynomial chaos expansions as used in
-\cite{sraj:2013a,sraj:2013b}.  The method and its application to our problem is
-described in Section~\ref{sec:uqpce}.  To represent the uncertainty in the
-Manning's $n$ coefficients a uniform distribution was assumed as the prior where
-$N$ was sampled in the interval $n \in [0.005-0.2]$.

### 2014-01-08 run_friction.py
Add bathy transformation ability

@@ -56,7 +56,8 @@ if __name__ == '__main__':
     if len(sys.argv) > 1:
         path = sys.argv[1]
 
-    source_path = '$SRC/tohoku2011-paper1/sources/Ammon.txydz'
+    # source_path = '$SRC/tohoku2011-paper1/sources/Ammon.txydz'
+    source_path = os.path.abspath(os.path.join(os.getcwd(),'saito.xyzt'))
 
     # Read in friction test values
     friction_values = numpy.loadtxt(path)

### 2014-01-08 setplot.py

@@ -291,7 +291,7 @@ def setplot(plotdata):
     plotitem.colorbar_label = "Manning's-$n$ Coefficient"
     plotitem.amr_celledges_show = [0,0,0]
     plotitem.amr_patchedges_show = [0,0,0,0,0,0,0]
-    plotitem.colorbar_label = r"$n$")
+    plotitem.colorbar_label = r"$n$"
 
 
     #-----------------------------------------

### 2014-01-08 setrun.py

@@ -141,7 +141,7 @@ def setrun(claw_pkg='geoclaw'):
     clawdata.output_format == 'ascii'      # 'ascii' or 'netcdf' 
 
     clawdata.output_q_components = 'all'   # need all
-    clawdata.output_aux_components = [0,3]  # eta=h+B is in q
+    clawdata.output_aux_components = 'all'  # eta=h+B is in q
     clawdata.output_aux_onlyonce = False    # output aux arrays each frame
 
 
@@ -153,7 +153,7 @@ def setrun(claw_pkg='geoclaw'):
     # The current t, dt, and cfl will be printed every time step
     # at AMR levels <= verbosity.  Set verbosity = 0 for no printing.
     #   (E.g. verbosity == 2 means print only on levels 1 and 2.)
-    clawdata.verbosity = 1
+    clawdata.verbosity = 4
 
 
 
@@ -409,9 +409,12 @@ def setgeo(rundata):
     dtopo_data = rundata.dtopo_data
     # for moving topography, append lines of the form:  (<= 1 allowed for now!)
     #   [topotype, minlevel,maxlevel,fname]
-    dtopodir = os.path.expandvars('$SRC/tohoku2011-paper1/sources/')
+    # dtopodir = os.path.expandvars('$SRC/tohoku2011-paper1/sources/')
+    # dtopo_data.dtopofiles.append([1, 4, 4,
+    #                                    os.path.join(dtopodir, 'Ammon.txydz')])
+    dtopodir = os.getcwd()
     dtopo_data.dtopofiles.append([1, 4, 4,
-                                       os.path.join(dtopodir, 'Ammon.txydz')])
+                                       os.path.join(dtopodir, 'saito.xyzt')])
 
 
     # == setqinit.data values ==

### 2014-01-08 transform_dtopo.py

@@ -1,0 +1,187 @@
+#!/usr/bin/env python
+
+import sys
+import os
+
+import numpy
+import scipy.interpolate as interp
+import matplotlib.pyplot as plt
+
+import clawpack.visclaw.colormaps as colormaps
+
+def read_deformation(path):
+    r""""""
+
+    data = numpy.loadtxt(path)
+
+    x = data[:,0]
+    y = data[:,1]
+    z = data[:,2]
+
+    # Find maximum extents and minimum distance between points
+    extents = [numpy.min(x), numpy.min(y), numpy.max(x), numpy.max(y)]
+
+    return extents, x, y, z
+
+
+def project_deformation(extents, x, y, z):
+    r""""""
+
+    # Construct new grid
+    num_cells = [ int((extents[2] - extents[0]) / numpy.min(numpy.abs(numpy.diff(x)))),
+                  int((extents[3] - extents[1]) / numpy.min(numpy.abs(numpy.diff(y)))) ]
+    t = numpy.array([0.0,1.0])
+    X, Y = numpy.meshgrid(numpy.linspace(extents[0],extents[2],num_cells[0]), 
+                          numpy.linspace(extents[1],extents[3],num_cells[1]))
+    Z = numpy.zeros((num_cells[1], num_cells[0], 2))
+    Z[:,:,1] = interp.griddata((x, y), z, (X, Y), method='linear', fill_value=0.0)
+
+    return t, X, Y, Z
+
+
+def write_deformation_to_file(t, X, Y, Z, output_file, topo_type=1):
+    r"""Write out a dtopo file to *output_file*
+
+    input
+    -----
+     - *t* (numpy.ndarray[:]) - Array containing time points, note that 
+       *t.shape[0] == Z.shape[2]*.
+     - *X* (numpy.ndarray[:,:]) - Array containing x-coodinates (longitude), 
+       should be in the form given by *numpy.meshgrid*.
+     - *Y* (numpy.ndarray[:,:]) - Array containing y-coordinates (latitude),
+       should be in the form given by *numpy.meshgrid*.
+     - *Z* (numpy.ndarray[:,:,:]) - Array containing deformation from original
+       bathymetry.
+     - *output_file* (path) - Path to the output file to written to.
+     - *topo_type* (int) - Type of topography file to write out.  Default is 1.
+
+    """
+
+    # Temporary catch for non implemented topo_type input
+    if topo_type != 1:
[…]

### 2014-01-08 paper/analysis.tex

@@
 bounds at the four gauges as indicated in each panel.  
-Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant
+Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant,
 reflected in a significant standard deviation as seen earlier from the realizations shown 
@@
 (DART) buoys observations. Figure~\ref{fig:compare} 
-the PC mean water surface elevation with the observed 
+plots the PC mean water surface elevation with the observed 
 water surface elevation at the different gauges locations
@@
 The same statistical analysis can be performed for the
-entire domain in 2D. Figure~\ref{fig:mean2d}(top row) shows
+entire domain in 2D. Figure~\ref{fig:mean2d} (top row) shows
 the PC mean water surface elevation for the considered computational
 domain at three different times as indicated in the title of each panel.
-The standard deviation is also shown in Figure~\ref{fig:mean2d}(bottom row)
+The standard deviation is also shown in Figure~\ref{fig:mean2d} (bottom row)
 at different times. We clearly notice the propagation of the variance
@@
 indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
-for $n_1$(top row) $n_2$(center row) and $n_3$ (bottom row)
+for $n_1$ (top row) $n_2$ (center row) and $n_3$ (bottom row)
 at different times as indicated in each panel.
@@
 This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
-range. Since $n_3$ shows insignificant contribution in the 
+range. Since $n_3$ shows insignificant contribution to the 
 uncertainty in the model output, we find the impact of $n_2$ and $n_3$ for 
@@
 and Figure~\ref{fig:response3} for different times as indicated. The most striking features are the relatively flat
-horizontal contours in the $n_3$ direction indicating that water surface elevation depends
+horizontal contours in the $n_3$ direction suggesting that water surface elevation depends
 only mildly on $n_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,

### 2014-01-08 paper/coefs.tex

@@
 are the water surface elevation at the gauges;
-their model predicted counterparts $\vec G$ is predicted by \geoclaw.
+their model predicted counterparts $\vec G$ is represented by \geoclaw.
 The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
-three Manning's coefficients treated as the unknown parameters. 
+three Manning's coefficients treated as the unknown parameters and assumed independent. 
 In this case  Equation~\eqref{eq:post} can be written as:
@@
 \exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
-\ p(\sigma^2)p(n_1)p(n_2) p(n_3)
+\ p(\sigma^2)p(n_1)p(n_2) p(n_3).
 \label{eq:post_coef}
@@
 
-To complete the definition of the posterior, we need to choose a proper prior that should be based 
+To complete the definition of the posterior, we need to choose proper priors that should be based 
 on some \emph{a priori} knowledge about the parameters. In our case, a uniform
-prior for all model parameters is assumed where $n_i \in [0.005,0.2]$ and therefore:
+prior for all three parameters is assumed where $n_i \in [0.005,0.2]$ and therefore:
 
@@
 		\displaystyle \frac{1}{0.2-0.005} &\text{for~} 0.005 <  N_i \leq 0.2 ,  \\
-		0 &\text{otherwise}  .
+		0 &\text{otherwise}.
 \end{cases}
@@
 		0 &\text{otherwise}. 
-		\end{cases}
+		\end{cases}.
 \label{eq:var_pr}

### 2014-01-08 paper/formulation.tex

@@
 \label{sec:formu}
-In this section, we introduce the methods employed in this work.
-First, the Bayesian approach to inverse problems its application to our problem of inferring the Manning's $n$ coefficients is described in Section~\ref{sec:inference}.
-In Section~\ref{sec:uqpce}, we introduce
-a stochastic spectral method for forward uncertainty propagation and accelerating the Bayesian inference.
+In this section, we first, present the Bayesian approach to inverse problems 
+and its application to our problem of inferring the Manning's $n$ coefficients.
+We also introduce
+a stochastic Polynomial Chaos-based spectral method for forward uncertainty propagation and accelerating the Bayesian inference.
 

### 2014-01-08 paper/inference.tex
method section is completed for now
revisiting results section

@@
 Bayesian inference is a statistical approach to inverse problems
-that has recently gained much interest in different applications including
+that has recently gained great interest in different applications, including
 ocean~\citep{Alexanderian2011a,Zedler2012,sraj:2013a}
@@
 \citep{sivia}. We review this approach briefly below.
+\subsubsection{Formalism}
 
@@
 the parameters such that:
+
 \begin{equation}
-\vec d \approx \vec{G}( \vec \theta)
+\vec d \approx \vec{G}( \vec \theta).
 \end{equation}
+Applying the Bayes' rule yields:
 
-Applying the Bayes' rule yields:
 \begin{equation}
  p(\vec{\theta}| \vec d) \propto 
- p(\vec d | \vec{\theta}) \ p(\vec{\theta})  
+ p(\vec d | \vec{\theta}) \ p(\vec{\theta}), 
 \label{eq:bayes}
@@
 and finally $p(\vec{\theta}| \vec d)$ is the posterior,
-representing the probability that $\vec{\theta}$ is true given the data $\vec d $.
+representing the probability of occurrence of $\vec{\theta}$ given the data $\vec d $.
 
@@
 represents the discrepancy between the model and observations.
-Here, the components of $\vec \epsilon $ are i.i.d. random variables with density $p_{\epsilon}$.
+Here, the components of $\vec \epsilon $ are assumed to be i.i.d. random variables with density $p_{\epsilon}$.
 In this case the likelihood function can be written as:
+
 \begin{equation} 
@@
 \prod_{i=1}^m  
-p_\epsilon (d_i - G_i(\vec \theta))  	
+p_\epsilon (d_i - G_i(\vec \theta)).  	
 \label{eq:likelihood}
@@
 a simplified model and assume single $\sigma^2$ value.
-Consequently, Equation~\ref{eq:likelihood} becomes:
+Consequently, Equation~\eqref{eq:likelihood} becomes:
+
 \begin{equation} 
@@
 \frac{1}{\sqrt{2 \pi \sigma^2}}\prod_{i=1}^m   
-\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace 	
+\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace. 	
 \label{eq:likelihood2}
@@
 The joint posterior from Bayes' rule can be given as:
+
 \begin{equation} 
@@
 \exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace  
-\prod_{i=1}^n p(\theta_i)
+\prod_{i=1}^n p(\theta_i).
 \end{equation}
 The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
-In other words, $\sigma^2$ may become an additional object for Bayesian inference and may themselves be endowed with priors and estimated from data.
-In this case the joint posterior is finally written as follows:
+In other words, $\sigma^2$ may become an additional parameter for Bayesian inference and  
+therefore endowed with a prior and estimated from data. In this case the joint posterior is finally written as follows:
+
 \begin{equation} 
@@
 \exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace
-\ p(\sigma^2) \prod_{i=1}^n p(\theta_i)
+\ p(\sigma^2) \prod_{i=1}^n p(\theta_i).
 \label{eq:post}

### 2014-01-08 paper/mcmc.tex

@@
 (DART) buoys observations  and their \geoclaw model counterparts
-for a reference simulation in which the  the Manning's $n$ coefficients were set to their optimal values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
+for a reference simulation in which the  the Manning's $n$ coefficients were set to their default values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
 shows a scatter plot that compares the observed 
@@
 The different panels shows well-mixed chains for all input parameters as well as the variance $\sigma^2$.
-The chain of  $n_{1}$ span the entire range of the prior meaning the observations are not informative 
-concerning this uncertain input.  On the contrary, the $n_{2}$ chain appears to be concentrated in the 
+The chain of  $n_{1}$ spans the entire range of the prior, meaning that the observations are not informative 
+concerning this input.  In contrast, the $n_{2}$ chain appears to be concentrated in the 
 lower end of the parameter range, with values between 0.005 and 0.1. 
@@
 Note that the first 10000 iterates, associated with the burn-in period, were discarded.  
-As expected from the chains shown in Figure
-~\ref{fig:mcmc}, the posterior pdfs of $n_1$ appear to be fairly flat, 
+As expected from the chains shown in Figure~\ref{fig:mcmc}, the posterior pdfs of $n_1$ appear to be fairly flat, 
 and similar to the uniform prior; an indication that 
@@
 with a Maximum A Posteriori (MAP) estimate of around 0.011
-and an extended tail towards the higher Manning's $n$ coefficient values.
-While for $n_3$, we observed a posterior that has a well defined peak
+and an extended tail towards the larger Manning's $n$ coefficient values.
+While for $n_3$, we observe a posterior that has a well defined peak
 of 0.18 but no clear pdf shape. The posterior distribution of the variance 

### 2014-01-08 paper/paper.tex

@@
 
-\title{Estimation of Manning's Friction Coefficient using DART Buoys Data during
+\title{Uncertainty Quantification and Inference of Manning's Friction Coefficient using DART Buoys Data during
 \tohoku Tsunami}
@@
 \author[kaust]{Ihab Sraj}
-\ead{ihab.sraj@kaust.edu.sa}
-\ead{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
 \author[ut]{Kyle T. Mandli}
@@
 \author[kaust]{Ibrahim Hoteit}
-
+\ead{ibrahim.hoteit@kaust.edu.sa}
+\ead{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
 \address[kaust]{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
@@
 
-Tsunamis have been lately catastrophic due to the warnings
-issued late and to the inaccurate predictions of water elevations. 
-Tsunami computational models are often employed to explore multiple 
-different scenarios to predict water elevations due to the abundance 
-of uncertainty in the input data. However, accurate estimation of water elevations 
+%Tsunamis have been lately catastrophic due to the warnings
+%issued late and to the inaccurate predictions of water elevations. 
+Tsunami computational models are employed to explore multiple 
+different scenarios and to predict water elevations 
+%due to the abundance of uncertainty in the input data. 
+However, accurate estimation of water elevations 
 requires accurate estimation of many model parameters,
-in particular the Manning's $n$ friction coefficient. Our objective here is to 
-present an efficient method for the 
-estimation of Manning's $n$ coefficient characterized by three different parameters 
-assumed constant in three regions: $n_1$ on-shore, $n_2$ near-shore and $n_3$ deep-water. 
-This method uses Polynomial Chaos (PC) to build an inexpensive surrogate for the 
-\geoclaw model and then employs the Bayesian inference to estimate the three uncertain parameters
-using the DART buoys data obtained during \tohoku tsunami. The efficiency of our approach stems from 
-using a surrogate model that reduces the computational burden of the Markov Chain Monte-Carlo 
-sampling needed in the Bayesian inference. The PC surrogate is also used to perform a sensitivity analysis. 
-Our results indicate that Manning's $n$ coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined using the available data.
+in particular the Manning's $n$ friction coefficient. 
+Our objective here is to develop an efficient approach for the 
+uncertainty quantification and inference of Manning's $n$ coefficient 
+which we characterize here by three different parameters 
+as being constant over three different regions: $n_1$ on-shore, $n_2$ near-shore and $n_3$ deep-water. 
+We use Polynomial Chaos (PC) to build an inexpensive surrogate for the 
+\geoclaw model and then employs Bayesian inference to estimate 
+and quantify the uncertainties to those uncertain parameters
+using the DART buoys data collected during \tohoku tsunami. 
+The surrogate model significantly reduces the computational burden of the Markov Chain Monte-Carlo (MCMC) 
+sampling of the Bayesian inference. The PC surrogate is also used to perform a sensitivity analysis. 
+Our results indicate that Manning's $n$ coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined from the available data.
 

### 2014-01-08 paper/uqpce.tex
revisiting results section

@@
 this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
-once for every proposed set of parameters of the Markov chain. While a single \geoclaw simulation
+once for every proposed set of parameters of the Markov chain \comment{add reference}. While a single \geoclaw simulation
 takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
-to avoid forward model simulations altogether. This can be accomplished by constructing a 
+to avoid forward model simulations altogether. This is achieved by constructing a 
 surrogate model that requires a much
@@
 Polynomial Chaos expansions for accelerating Bayesian inference in this context 
-by building a surrogate, which, in addition can provide statistical properties, such as the mean, variance and sensitivities, efficiently.
+by building a surrogate model, which, in addition can provide statistical properties, such as the mean, variance and sensitivities, efficiently.
 
@@
 method below; for more details on this method, and other UQ methods
-refer to \citep{LeMaitreKnio2010}.
+the reader is referred to \citep{LeMaitreKnio2010}.
 
-Let $U=U(\bm{x},t,\xxi)$ denote a quantity of 
+Let $U=U(\bm{x},t,\xxi)$ denotes a quantity of 
 interest (QoI) that is the output of certain model.
-$U$ is normally function of space $\bm{x}$ and time $t$, and 
+$U$ is function of space $\bm{x}$ and time $t$, and 
 also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). 
 PC expresses $U$ in the form:
+
 \begin{equation}
-  U(\xbold,t,\xxi) \doteq \sum_{k = 0}^P U_k(\xbold,t) \Psi_k(\xxi)
+  U(\xbold,t,\xxi) \doteq \sum_{k = 0}^P U_k(\xbold,t) \Psi_k(\xxi),
 \label{eq:stochseries}
@@
 The choice of the basis is dictated by the probability density
-function of the stochastic variable $\xxi$, and which appears as a weight
+function of the stochastic variable $\xxi$, which appears as a weight
 function in the stochastic space's inner product:
+
 \begin{equation}
- \left<\Psi_i,\Psi_j\right> = \int \Psi_i(\xxi) \;\Psi_j(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi=\delta_{ij}\ave{\Psi_i^2}
+ \left<\Psi_i,\Psi_j\right> = \int \Psi_i(\xxi) \;\Psi_j(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi=\delta_{ij}\ave{\Psi_i^2},
 \label{eq:inner}
@@
 For multi-dimensional problems the basis functions are
-tensor products of 1D basis functions.
+tensor products of 1D basis functions \comment{add refer}.
 
@@
 of $U$ along the stochastic dimensions. It can be also seen as
-combining approximation and probabilistic frameworks, a combination
-that has proven extremely useful in solving UQ problems. The existence and convergence of this series
-is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
+combining approximation and probabilistic frameworks; a combination
+that has proven extremely useful in solving UQ problems\comment{add refer}.
+The existence and convergence of this series is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
 The series rate of convergence, and hence the number of terms to retain, depends on the smoothness of
@@
 In particular, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
+
 \begin{equation}
@@
  \frac{1}{\left< \Psi_k, \Psi_k \right>} 
- \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi
+ \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi.
 \end{equation}
@@
 \approx \left< U, \Psi_k \right>_Q
-= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q
+= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q,
 \end{equation}
@@
 quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
-respectively. The quadrature order should be commensurate with the
+respectively. The quadrature order should be commensurated with the
 truncation order, and should be high enough to avoid aliasing artifacts.
@@
 The computation of the ${U}_k$ can thus be expressed as a matrix-vector product of the form:
+
 \begin{equation} 
  U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
- \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>}
+ \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>},
 \end{equation} 
@@
 $\left<\Psi_0,\Psi_0\right>=1$, the expectation and variance of $U$ can be computed as:
-\begin{eqnarray}
- E[U]&=&\int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0  
- \label{eq:mean} \\
- E[(U-E[U])^2]&=&\int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
+
+\begin{equation}
+ E[U] = \int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0,  
+ \label{eq:mean}
+\end{equation}
+and \begin{equation}
+ E[(U-E[U])^2] = \int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
  \label{eq:sigma}
-\left<\Psi_k,\Psi_k\right>
-\end{eqnarray}
+\left<\Psi_k,\Psi_k\right>.
+\end{equation}
 
 PC representations of random variables also enable efficient
-approximation of the corresponding variance-based, global
-sensitivity indices. This can be done by computing the so-called {\it total} 
+approximation of the corresponding variance-based, \comment{global
+sensitivity indices}. This can be done by computing the so-called {\it total} 
 sensitivity indices of random variables from the PC representations or Sobol decomposition~\citep{Sobol:1993,Homma:1996,Sobol:2001}. To get the total sensitivity corresponding to the uncertain
 input $\xi_i$ we compute the total index~\citep{LeMaitreKnio2010,Crestaux,Sudret}:
+
 \begin{equation} \label{eq:T-hard}
@@
                \sum_{k \in K_i} U_k^2 \ave{\Psi_k^2}}
-              {\displaystyle\sum_{k = 1}^P U_k^2 \ave{\Psi_k^2}} .
+              {\displaystyle\sum_{k = 1}^P U_k^2 \ave{\Psi_k^2}},
 \end{equation}
@@
 computing the fraction of the total variance due to all the terms in the
-PC expansion which involve $\xi_i$.
+PC expansion that involve $\xi_i$.

### 2014-01-08 setplot.py
Output type modification

@@ -60,13 +60,14 @@ def setplot(plotdata):
     """ 
 
     plotdata.clearfigures()  # clear any old figures,axes,items data
+    plotdata.format = 'binary'
+
 
     # Import useful run data
     clawdata = data.ClawInputData(2)
     clawdata.read(os.path.join(plotdata.outdir, 'claw.data'))
     friction_data = surge.data.FrictionData()
     friction_data.read(os.path.join(plotdata.outdir, 'friction.data'))
-
 
     # To plot gauge locations on pcolor or contour plot, use this as
     # an afteraxis function:

### 2014-01-08 setrun.py

@@ -138,7 +138,7 @@ def setrun(claw_pkg='geoclaw'):
         clawdata.output_t0 = True
         
 
-    clawdata.output_format == 'ascii'      # 'ascii' or 'netcdf' 
+    clawdata.output_format = 'binary'      # 'ascii' or 'netcdf' 
 
     clawdata.output_q_components = 'all'   # need all
     clawdata.output_aux_components = 'all'  # eta=h+B is in q

### 2014-01-10 paper/intro.tex
Add some rewording and additional descriptions
Includes a review of the abstract, intro, and tsunami section along with
some comments and better worded descriptions of the friction section.

@@
 
-\section{Introduction}
-\label{sec:intro}
+\section{Introduction} \label{sec:intro}
+
 Tsunamis in the past decade have been responsible for some of the most  deadly
@@
 employed to explore multiple different scenarios to predict the water elevation
-caused by the tsunamis. The accurate prediction  of water elevations, however, requires
-accurate estimation of many model parameters that are either measured directly,
-defined empirically, or estimated from a collection of observed data. Unfortunately,
-since tsunami events are relatively rare,  there is an abundance of uncertainty
-in the input data for these computational  models ranging from effects of the
-domain, such as bathymetry and friction parameterizations,  to the earthquake
-source.  In this study, we aim to estimate the Manning's $n$ friction
-coefficient \alert{add a sentence here about the Manning Coeff}. 
-We aim at quantifying the uncertainty in the predicted water
-elevation  and employs a Bayesian inverse modeling approach to estimate the Manning's
-$n$ coefficient using observations of water elevation measured during a tsunami event.
+caused by the tsunamis. The accurate prediction  of water elevations, however,
+requires accurate estimation of many model parameters that are either measured
+directly, defined empirically, or estimated from a collection of observed data.
+Unfortunately, since tsunami events are relatively rare,  there is an abundance
+of uncertainty in the input data for these computational  models ranging from
+effects of the domain, such as bathymetry and friction parameterizations,  to
+the earthquake source.  In this study, we aim to estimate the Manning's $n$
+friction coefficient, a parameterization of the effect of bottom friction 
+commonly used in tsunami models.  We aim at quantifying the uncertainty in the 
+predicted water elevation  and employ
+a Bayesian inverse modeling approach to estimate the Manning's $n$ coefficient
+using observations of water elevation measured during a tsunami event.
 
-Previous works looking into uncertainty in the input for tsunami simulations
+Previous work looking into uncertainty in the input for tsunami simulations
 focused on the earthquake source but primarily as a way of
@@
 sensitivity analysis.  Changes to the friction parameterization has been shown in 
-the past to lead to significant changes to inundation levels \cite{Myers:2001el,Jakeman:2010hk}, 
-but these previous studies \comment{were limited in the scope of the changes to the parameterization allowed}.  
+the past to lead to significant changes to inundation levels \cite{Myers:2001el, 
+Jakeman:2010hk}, but these previous studies were limited in the range of the 
+values and locations used in the parameterization.  
 In a recent work, Mayo \emph{et al.}~\cite{Mayo2013} reformulated a statistical data assimilation method
 generally used in the estimation of model states to estimate the Manning's $n$
-coefficient. They used a low-dimensional representations of
+coefficient. They used a low-dimensional representation of
 Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data
@@
 In this work, we focus our attention on the uncertainty in the friction
-parameterization used in tsunami simulations. We present a three-dimensional 
+parameterization used in tsunami simulations. We present a \alert{three}-dimensional 
 representation of Manning's $n$ coefficient and an efficient method
@@
 Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
-\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} including the
-source models analyzed there.
+\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} 
+including the two source models analyzed there.
 
@@
 of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
-and propose a three-dimensional representation of $n$. 
+and propose a \alert{three}-dimensional representation of $n$. 
 Section~\ref{sec:tohoku} describes our case-study, the \tohoku tsunami,
@@
 posterior distributions obtained using Bayesian inference by MCMC. A discussion of our findings and methodology is finally presented in Section~\ref{sec:conc}.
+
+\comment{
+\begin{itemize}
+    \item There are two places where we refer to the parameterization as three-dimensional, I am not sure if this is really three-dimensional.
+I know that we are using depth contours but the parameterization is two-dimensional.
+I have a feeling that readers would take issue with this.
+    \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there.
+\end{itemize}
+}

### 2014-01-10 paper/paper.tex
Add some rewording and additional descriptions
Includes a review of the abstract, intro, and tsunami section along with
some comments and better worded descriptions of the friction section.

@@
 
-%Tsunamis have been lately catastrophic due to the warnings
-%issued late and to the inaccurate predictions of water elevations. 
-Tsunami computational models are employed to explore multiple 
-different scenarios and to predict water elevations 
-%due to the abundance of uncertainty in the input data. 
-However, accurate estimation of water elevations 
-requires accurate estimation of many model parameters,
-in particular the Manning's $n$ friction coefficient. 
-Our objective here is to develop an efficient approach for the 
-uncertainty quantification and inference of Manning's $n$ coefficient 
-which we characterize here by three different parameters 
-as being constant over three different regions: $n_1$ on-shore, $n_2$ near-shore and $n_3$ deep-water. 
-We use Polynomial Chaos (PC) to build an inexpensive surrogate for the 
-\geoclaw model and then employs Bayesian inference to estimate 
-and quantify the uncertainties to those uncertain parameters
-using the DART buoys data collected during \tohoku tsunami. 
-The surrogate model significantly reduces the computational burden of the Markov Chain Monte-Carlo (MCMC) 
-sampling of the Bayesian inference. The PC surrogate is also used to perform a sensitivity analysis. 
-Our results indicate that Manning's $n$ coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and $n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined from the available data.
+Tsunami computational models are employed to explore multiple  different
+scenarios and to predict water elevations.  However, accurate estimation of
+water elevations  requires accurate estimation of many model parameters, in
+particular the Manning's $n$ friction coefficient.  Our objective here is to
+develop an efficient approach for the  uncertainty quantification and inference
+of the Manning's $n$ coefficient  which we characterize here by three different
+parameters  as being constant over three different regions: $n_1$ on-shore,
+$n_2$ near-shore and $n_3$ deep-water.  We use Polynomial Chaos (PC) to build an
+inexpensive surrogate for the  \geoclaw model and then employ Bayesian inference
+to estimate  and quantify the uncertainties to those uncertain parameters using
+the DART buoy data collected during the \tohoku tsunami.  The surrogate model
+significantly reduces the computational burden of the Markov Chain Monte-Carlo
+(MCMC)  sampling of the Bayesian inference. The PC surrogate is also used to
+perform a sensitivity analysis.  Our results indicate that Manning's $n$
+coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and
+$n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined from the
+available data.
 

### 2014-01-10 paper/tsunami.tex
Add some rewording and additional descriptions
Includes a review of the abstract, intro, and tsunami section along with
some comments and better worded descriptions of the friction section.

@@
 \eqref{eq:coef} appearing in the shallow water equations \eqref{eq:swe}
-governing tsunamis and the roughness parameter $n$.  This Manning's $n$
-coefficient varies spatially  and is dependent on the surface
-characteristics of the sea-floor.  A direct measurement of $n$ is therefore
-impossible, instead, values are often assigned from tables that are empirically
-estimated for common types of land that are determined from satellite images. As
-a result, the Manning’s n coefficients used in tsunami models often contain
-large amounts of uncertainty.
+governing tsunamis and the empirically derived roughness parameter $n$.  The 
+value of $n$ can vary both spatially and temporally due to changes in the
+bottom roughness from wave action and other processes.  For tsunami modeling it
+is a common practice to assume a single value of Manning's $n$ for the areas
+that are initially below sea-level and sometimes use satellite images and 
+land-use data to determine the coefficient in areas above sea-level.  This
+process both above and below sea-level has several sources of uncertainty, all
+of which can have an impact on the quantities of interest in tsunami modeling.   
+In MacInnes \emph{et al.}, for
+instance, they found that adopting a uniform value of $n = 0.025$ everywhere
+lead to poor performance of the forward model's predicted inundation levels.  To
+address this they used $n = 0.035$ in the Sendai plane to account for land-uses 
+such as pasture, farmland, and rice paddies.  It is uncertainties such as these 
+that we wish to address in a systematic way so that accurate risk assessments 
+can be done.
 
-The uncertainty in the Manning's $n$ coefficient leads to uncertainties in the
-predicted quantities of interest such as the water elevations and becomes a
-health risk during tsunami events. In addition,  the prediction of these
-quantities of interest are often highly sensitive to changes in these
-uncertain parameters.  In MacInnes \emph{et al.} a Manning's $n$ coefficient of 0.025
-was used except on the Sendai plane where 0.035 was used to account for pasture,
-farmland and rice paddies.  In this work, we seek to quantify parametric
-uncertainty due to the uncertainty in the Manning's $n$ friction coefficient. To
-this end, we characterize $n$ as being contasnt over three different 
-regions. The three regions are:  on-shore whose coefficient is denoted by
-$n_1$, near- shore  denoted by $n_2$, and deep-water denoted by $n_3$. The three
-regions are shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
-with respect to those three parameters we employ Polynomial Chaos (PC) expansions as used in
-\cite{sraj:2013a,sraj:2013b}.  The method and its application to our problem is
+The direct result of the uncertainty in the Manning's $n$ coefficient lead to
+uncertainties in the quantities of interest such as the water elevation,
+especially in the most crucial area in risk assessments, the near-shore. In 
+addition,  the prediction of these quantities of interest are often highly 
+sensitive to changes in these uncertain parameters.  We seek to therefore 
+quantify the parametric uncertainty due to the uncertainty in the Manning's $n$ 
+friction coefficient by characterizing the spatial variability by using
+iso-baths to define distinct regions in the domain that are characterized by a
+single Manning's $n$ coefficient.  In this study three regions were used, an
+on-shore region (initially above sea-level), a near-shore region (between
+sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
+200 meters), denoted by $n_1$, $n_2$, and $n_3$ respectively and as shown in 
+Figure~\ref{fig:ceofs}. To quantify the uncertainties
+with respect to those three parameters we employ Polynomial Chaos (PC) 
+expansions as used in \cite{sraj:2013a,sraj:2013b}.  The method and its 
+application to our problem is
 described in Section~\ref{sec:uqpce}.  To represent the uncertainty in the
-Manning's $n$ coefficients an uninformative uniform distribution was assumed as the prior with
-$n$  sampled from the interval $n \in [0.005-0.2]$.
+Manning's $n$ coefficients an uninformative uniform distribution was assumed as 
+the prior with physically relevant values of $n$ sampled from the interval 
+$n \in [0.005-0.2]$.
+
+% This Manning's $n$
+% coefficient is allowed to vary spatially by defining a set of isobaths that 
+% partition the domain into disjoint regions where a single value of $n$ can be 
+% used.  The values used for $n$ can be chosen using satellite images and
+% land-use data but this is impractical for the values used in regions
+% below sea-level or in the case of highly variable land-use regions.  As a
+% result, values used for $n$ contain large amounts of uncertainty.
 
@@
 
-The \tohoku earthquake and tsunami of 2011 has been the subject of a number of
+The \tohoku earthquake and tsunami of 2011 have been the subject of a number of
 studies due to the wealth of observational evidence and severity of the tsunami.
@@
 caused massive devastation across Japan.  In our case study, the
-simulation is based on work done in \cite{MacInnes:2013cr} and modified to use
-the specified variable friction field. Hereafter, we briefly describe the
+simulation used is based on work done in \cite{MacInnes:2013cr} and modified to 
+use the specified variable friction field. Hereafter, we briefly describe the
 forward computational model, \geoclaw, the  simulation setup used in the
 uncertainty analysis, and finally present the observations from NOAA DART buoys.
+
 \alert{describe a bit more the tsunami and its geophysical location}.
+
+\comment{Kyle - I kind of figured the MacInnes paper goes into detail on this
+so I am not sure what pertinent information should be included here.  We could
+do some more info on the fault itself and some shore maximum inundations but we
+really aren't using any of that.}
+
 \subsection{\geoclaw} \label{ssub:geoclaw}
-The \geoclaw software package uses a finite volume, wave-propagation approach as described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
-
+The \geoclaw software package uses a finite volume, wave-propagation approach as
+described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water \
+equations:
 \begin{equation} \label{eq:swe}
@@
 given a Manning's $n$ parameterization such that:
-
 \begin{equation}
@@
 
-The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.
-The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing, even when the momentum is non-zero, \alert{and entropy corrections} \cite{George:2008aa}.
+The primary computational kernel in \geoclaw is the evaluation of the solution
+to the Riemann problem at each grid cell interface.  The Riemann solver used
+includes the ability to handle inundation (wet-dry interfaces), well-balancing,
+even when the momentum is non-zero, \alert{and entropy corrections}
+\cite{George:2008aa}.
 
-%\subsubsection{Adaptivity} \label{ssub:adaptivity}
-One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, the disturbance in the surface height $\eta$).
-\geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.
+One unique feature that \geoclaw has is the ability to adaptively refine the
+grids used for the computation essentially following a region of interest as
+time progresses (in this case, the disturbance in the surface height $\eta$).
+\geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa}
+and is described in detail in \cite{Berger:2011du} in the case of tsunami
+modeling.
 
-For the simulations presented, refinement thresholds were used that matched what was presented in \cite{MacInnes:2013cr}.
-As was done there, four levels of refinement were used staring with a resolution of 1 degree in both the latitudinal and longitudinal directions down to 0.5' resolution (approximately 900 meters) located around the observation locations.
-The tolerance for the refinement criteria for sea-surface anomaly was 0.02 meters.
+For the simulations presented, refinement thresholds were used that matched what
+was presented in \cite{MacInnes:2013cr}.  As was done there, four levels of
+refinement were used staring with a resolution of 1 degree in both the
+latitudinal and longitudinal directions down to 0.5' resolution (approximately
+900 meters) located around the observation locations.  The tolerance for the
+refinement criteria for sea-surface anomaly was 0.02 meters.
 
-\subsection{DART Buoys Observations}
+\subsection{DART Buoy Observations}
 
@@
 and 21419 whose locations are shown in Figure~\ref{fig:setup_buoy_locations} and
-whose de-tided water surface elevation data for the event are shown in Figure~\ref{fig:setup_buoy_data}.
-The time $t=0$ is set to be the start of the earthquake \comment{frequency of data}.  We can see that the
-primary tsunami waves generated reach the gauges between 40 minutes and 2 hours
-after the earthquake.  The large spikes later in time present in the data at
-gauge 21418 are data anomalies and not subsequent tsunami waves.
+whose de-tided water surface elevation data for the event are shown in
+Figure~\ref{fig:setup_buoy_data}. The time $t=0$ is set to be the start of the
+earthquake and is recorded every 15 seconds during the first few minutes of an
+event, otherwise recording every minute when triggered (a tsunami has occurred)
+or every 15 minutes when not.  We can see that the primary tsunami
+waves generated reach the gauges between 40 minutes and 2 hours after the
+earthquake.  The large spikes later in time present in the data at gauge 21418
+are data anomalies and not subsequent tsunami waves.
 
@@
 The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
-accurate bathymetry for the region \cite{Amante:2009ud}.  The additional finer
-resolution bathymetry data used in MacInnes \emph{et al.} were not used in this
-study since the simulations were not sensitive to the finer bathymetry at the
-DART buoy locations in the deep-water.
+accurate bathymetry for the region \cite{Amante:2009ud}.  While it is the case
+that the inundation region is the most sensitive to changes in the Manning's $n$
+coefficient, the finest bathymetry used in MacInnes \emph{et al.} were
+not included in the this study due to the availability of the data and the
+insensitivity of the simulation to the finer bathymetry at the DART buoy
+locations.
 
@@
 discussed in MacInnes \emph{et al.}, the source models used do not include rupture
-timing and this could have an impact on the accuracy of the simulation.
+timing which could have an impact on the accuracy of the simulation.
 
 \comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}
+
+\comment{We need to address not using the finest bathymetry better}
+
+% This issue
+% is mitigated somewhat by the decreasing sensitivity of water elevation levels
+% to friction with deepening water but the near shore region, where the setup for
+% inundation occurs, still can be highly sensitive to the friction values chosen.

### 2014-01-14 paper/conc.tex
few fixes and wrote a conclusion

@@
 \label{sec:conc}
+
+The present study aimed at estimating Manning's $n$ friction coefficient 
+which is crucial in the accurate prediction of water surface elevation in 
+tsunami events modeling. We proposed a three-dimensional representation
+of Manning's $n$ friction coefficient that was characterized using iso-baths 
+to define three distinct regions in the domain and represented each region
+with a single Manning's $n$ coefficient. The three regions were as follows:
+an on-shore region (initially above sea-level), a near-shore region (between
+sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
+200 meters) whose Manning's $n$ friction coefficient were denoted by $n_1$, $n_2$, 
+and $n_3$ respectively. The estimation relied on a Bayesian
+inference approach that sharpens the initial estimates of the three uncertain parameters
+using real observations. In our case, we used the DART buoys system data that provides water 
+surface elevation data. We specifically used the data collected at four different gauges 
+during \tohoku tsunami. To accelerate the Bayesian inference, we relied on the polynomial 
+chaos expansions to produce a faithful and efficient surrogate of the forward model \geoclaw. 
+The main results of this work were:
+\begin{itemize}
+
+\item We built a surrogate model using PC expansions that required 125 forward model runs of \geoclaw.
+The surrogate model was tested, validated and proved to be faithful.
+
+\item We used the PC surrogate to quantify the uncertainties in the predicted water surface elevation 
+due to the uncertainties in the Manning's $n$ coefficient.
+To this end, we estimated the mean and the standard deviation of water surface elevation and compared it with the measured data collected at found gauges.
+
+\item A global sensitivity analysis was performed to quantity the contribution of each uncertain parameter to the variance in water surface elevation.
+We found that the Manning’s $n$ coefficient at the shore $n_2$ is dominant and
+contributes the most to the variance in the water surface elevation compared
+to the other two Manning’s $n$ coefficients $n_1$ and $n_3$ for almost the
+entire tsunami event.
+
+\item We used Bayesian inference and identified the following MAP
+estimates for the three $n$ parameters using MCMC, namely $n_2=0.011$ and
+$n_3=0.180$, while for $n_1$ no meaningful MAP value could be determined from the
+available data.
+
+\end{itemize}
+
+Finally, we note that the present study focused on formulating and estimating a low-dimensional representation
+of the Manning's $n$ coefficient using UQ techniques namely Bayesian inference and PC expansions.
+A high-dimnensional representation of the Manning's $n$ coefficient would require a large number of forward runs that is computationally expensive.
+Alternatively, advanced techniques can be implemented to reduce the dimensionality of the problem
+that is the objective of a future study.

### 2014-01-14 paper/intro.tex
few fixes and wrote a conclusion

@@
 values and locations used in the parameterization.  
-In a recent work, Mayo \emph{et al.}~\cite{Mayo2013} reformulated a statistical data assimilation method
+In a recent work, Mayo \emph{et al.}~\cite{Mayo:2014} reformulated a statistical data assimilation method
 generally used in the estimation of model states to estimate the Manning's $n$
@@
 Polynomial Chaos (PC) expansions to construct a faithful surrogate model of the
-response of the large and complex geophysical model simulations \comment{ref on pcs}.
+response of the large and complex geophysical model simulations \cite{Najm:2009,Alexanderian2012,sraj:2013a,sraj:2013b}.
 The construction of the surrogate enables us to efficiently implement the Bayesian
@@
 I have a feeling that readers would take issue with this.
+    \item Ihab: the representation is 3D meaning we used three different values of n in three regions. The same concept was presented in Mayo paper where they did 2D representation.
     \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there. 
+    \item Ihab: Our main objective is to show the UQ techniques and how they work.
+The equations are known and we do not need to stress on them.
+We can even remove the whole equation or forward model section.
 \end{itemize}

### 2014-01-14 paper/mcmc.tex
few fixes and wrote a conclusion

@@
 Manning's $n$ coefficients and to the measurement errors.
-\alert{An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
-the epicenter of the earthquake and therefore receiving more impact by the tsunami.}
+An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
+the epicenter of the earthquake and therefore receiving more impact by the tsunami.
 

### 2014-01-14 paper/pcstudy.tex
few fixes and wrote a conclusion

@@
 functions (pdfs) of water surface elevation at the different gauge locations
-converges with increased order of the PC expansion  \comment{add referece}.  Sample
+converges with increased order of the PC expansion~\cite{Alexanderian2012,sraj:2013a}.  Sample
 water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}

### 2014-01-14 paper/tsunami.tex
few fixes and wrote a conclusion

@@
 Manning's $n$ law provides a relationship between the friction coefficient
-\eqref{eq:coef} appearing in the shallow water equations \eqref{eq:swe}
-governing tsunamis and the empirically derived roughness parameter $n$.  The 
-value of $n$ can vary both spatially and temporally due to changes in the
+$C_f$ appearing in the shallow water equations (SWE) governing tsunamis and 
+the empirically derived roughness parameter $n$:
+\begin{equation}
+    C_f = \frac{g n^2}{h^{4/3}},
+\label{eq:coef}
+\end{equation}
+The value of $n$ can vary both spatially and temporally due to changes in the
 bottom roughness from wave action and other processes.  For tsunami modeling it
@@
 quantify the parametric uncertainty due to the uncertainty in the Manning's $n$ 
-friction coefficient by characterizing the spatial variability by using
+friction coefficient by characterizing the spatial variability using
 iso-baths to define distinct regions in the domain that are characterized by a
@@
 sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
-200 meters), denoted by $n_1$, $n_2$, and $n_3$ respectively and as shown in 
-Figure~\ref{fig:ceofs}. To quantify the uncertainties
+200 meters), whose Manning's $n$ coefficient is denoted by $n_1$, $n_2$, and $n_3$ respectively 
+as shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
 with respect to those three parameters we employ Polynomial Chaos (PC) 
@@
 The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and
-caused massive devastation across Japan.  In our case study, the
+caused massive devastation across Japan. The source of the earthquake (epicenter)
+was located approximately 72 km east of \tohoku as indicated 
+in Figure~\ref{fig:ceofs}a. In our case study, the
 simulation used is based on work done in \cite{MacInnes:2013cr} and modified to 
@@
 uncertainty analysis, and finally present the observations from NOAA DART buoys.
-
-\alert{describe a bit more the tsunami and its geophysical location}.
 
@@
 to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
-friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
-given a Manning's $n$ parameterization such that:
-\begin{equation}
-    C_f = \frac{g n^2}{h^{4/3}},
-\label{eq:coef}
-\end{equation}
-where $n$ is an empirically determined parameter known as Manning's $n$.
+friction coefficient.  
+%As is common in tsunami modeling, $C_f$ is calculated 
+%given a Manning's $n$ parameterization such that:
+%\begin{equation}
+%    C_f = \frac{g n^2}{h^{4/3}},
+%\label{eq:coef}
+%\end{equation}
+%where $n$ is an empirically determined parameter known as Manning's $n$.
 
@@
 insensitivity of the simulation to the finer bathymetry at the DART buoy
-locations.
+locations. The bathymetry of the considered regions (and topography) is 
+shown in Figure~\ref{fig:setup}a.
 

### 2014-01-14 paper/uqpce.tex
few fixes and wrote a conclusion

@@
 this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
-once for every proposed set of parameters of the Markov chain \comment{add reference}. While a single \geoclaw simulation
+once for every proposed set of parameters of the Markov chain~\cite{MarzoukNajm2009,Malinverno2002}. While a single \geoclaw simulation
 takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
@@
 Polynomial Chaos expansions for accelerating Bayesian inference in this context 
-by building a surrogate model, which, in addition can provide statistical properties, such as the mean, variance and sensitivities, efficiently.
+by building a surrogate model, which, in addition can efficiently
+provide statistical properties, such as the mean, variance and sensitivities.
 
@@
 For multi-dimensional problems the basis functions are
-tensor products of 1D basis functions \comment{add refer}.
+tensor products of 1D basis functions~\cite{LeMaitreKnio2010}.
 
@@
 combining approximation and probabilistic frameworks; a combination
-that has proven extremely useful in solving UQ problems\comment{add refer}.
+that has proven extremely useful in solving UQ problems~\cite{Xiu:2003,Lin2009}.
 The existence and convergence of this series is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
@@
 
-PC representations of random variables also enable efficient
-approximation of the corresponding variance-based, \comment{global
-sensitivity indices}. This can be done by computing the so-called {\it total} 
-sensitivity indices of random variables from the PC representations or Sobol decomposition~\citep{Sobol:1993,Homma:1996,Sobol:2001}. To get the total sensitivity corresponding to the uncertain
-input $\xi_i$ we compute the total index~\citep{LeMaitreKnio2010,Crestaux,Sudret}:
+PC representations of random variables also enable conducting
+efficient global sensitivity analysis that aims to quantify the
+contribution of different random input parameters to the model variability.
+This can be done by computing the so-called {\it total} 
+sensitivity index $T_i$ that measures the contribution of
+the $i^{th}$ random input to total model variability by
+computing the fraction of the total variance due to all the terms in the
+PC expansion that involve $\xi_i$~\citep{LeMaitreKnio2010,Crestaux,Sudret}
+as follows:
+
+% of random variables from the PC representations or Sobol decomposition~\citep{Sobol:1993,Homma:1996,Sobol:2001}. The total sensitivity index  
+%
+%To get the total sensitivity corresponding to the uncertain
+%input $\xi_i$ we compute the total index:
 
@@
         \]
-        and $\vec{\alpha}^k$ is the
-multi-index associated with $k^{th}$ term in the
+        and $\vec{\alpha}^k$ is the multi-index associated with $k^{th}$ term in the
 PC expansion~\cite{LeMaitreKnio2010}.
 
-Using Equation~\eqref{eq:T-hard}, the computation of $T_i$ is straightforward.
-The total sensitivity index $T_i$ measures the contribution of
-the $i^{th}$ random input to total model variability by
-computing the fraction of the total variance due to all the terms in the
-PC expansion that involve $\xi_i$.
+%Using Equation~\eqref{eq:T-hard}, the computation of $T_i$ is straightforward.

### 2014-01-14 paper/conc.tex
fixed figures and removed comments from paper

@@
 which is crucial in the accurate prediction of water surface elevation in 
-tsunami events modeling. We proposed a three-dimensional representation
+tsunami events modeling. We proposed a three-parameter representation
 of Manning's $n$ friction coefficient that was characterized using iso-baths 

### 2014-01-14 paper/intro.tex
fixed figures and removed comments from paper

@@
 In this work, we focus our attention on the uncertainty in the friction
-parameterization used in tsunami simulations. We present a \alert{three}-dimensional 
+parameterization used in tsunami simulations. We present a three-parameter 
 representation of Manning's $n$ coefficient and an efficient method
@@
 to estimate Manning's $n$ coefficient using the \tohoku tsunami
-event as a case-study. The quantity of interest is taken to be the surface
-elevation $\eta$ and is compared to the available NOAA Deep-ocean Assessment and
+event as a case-study. The quantity of interest is taken to be the water surface
+elevation and is compared to the available NOAA Deep-ocean Assessment and
 Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
@@
 of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
-and propose a \alert{three}-dimensional representation of $n$. 
+and propose a three-parameter representation of $n$. 
 Section~\ref{sec:tohoku} describes our case-study, the \tohoku tsunami,
-summarize the forward simulations setup, and present the DART buoys observational data. 
+summarizes the forward simulations setup, and presents the DART buoys observational data. 
 Section~\ref{sec:formu} presents the two main methods employed in the UQ analysis:
@@
 
-\comment{
-\begin{itemize}
-    \item There are two places where we refer to the parameterization as three-dimensional, I am not sure if this is really three-dimensional.
+%\comment{
+%\begin{itemize}
+%    \item There are two places where we refer to the parameterization as three-dimensional, I am not sure if this is really three-dimensional.
 I know that we are using depth contours but the parameterization is two-dimensional.
 I have a feeling that readers would take issue with this.
-    \item Ihab: the representation is 3D meaning we used three different values of n in three regions. The same concept was presented in Mayo paper where they did 2D representation.
-    \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there. 
-    \item Ihab: Our main objective is to show the UQ techniques and how they work.
+%    \item Ihab: the representation is 3D meaning we used three different values of n in three regions. The same concept was presented in Mayo paper where they did 2D representation.
+%    \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there. 
+%    \item Ihab: Our main objective is to show the UQ techniques and how they work.
 The equations are known and we do not need to stress on them.
 We can even remove the whole equation or forward model section.
-\end{itemize}
-}
+%\end{itemize}
+%}

### 2014-01-14 paper/tsunami.tex
fixed figures and removed comments from paper

@@
 the empirically derived roughness parameter $n$:
+
 \begin{equation}
@@
 
-\comment{Kyle - I kind of figured the MacInnes paper goes into detail on this
-so I am not sure what pertinent information should be included here.  We could
-do some more info on the fault itself and some shore maximum inundations but we
-really aren't using any of that.}
+%\comment{Kyle - I kind of figured the MacInnes paper goes into detail on this
+%so I am not sure what pertinent information should be included here.  We could
+%do some more info on the fault itself and some shore maximum inundations but we
+%really aren't using any of that.}
 
@@
 includes the ability to handle inundation (wet-dry interfaces), well-balancing,
-even when the momentum is non-zero, \alert{and entropy corrections}
+even when the momentum is non-zero, and entropy corrections
 \cite{George:2008aa}.
@@
 grids used for the computation essentially following a region of interest as
-time progresses (in this case, the disturbance in the surface height $\eta$).
+time progresses (in this case, the disturbance in the surface height).
 \geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa}
@@
 
-\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}
+%\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}
 
-\comment{We need to address not using the finest bathymetry better}
+%\comment{We need to address not using the finest bathymetry better}
 

### 2014-01-16 paper/coefs.tex
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 three Manning's coefficients treated as the unknown parameters and assumed independent. 
-In this case  Equation~\eqref{eq:post} can be written as:
+This enables us to rewrite Equation~\eqref{eq:post} as:
 
@@
 on some \emph{a priori} knowledge about the parameters. In our case, a uniform
-prior for all three parameters is assumed where $n_i \in [0.005,0.2]$ and therefore:
+prior for all three Manning's $n$ parameters is assumed, with $n_i$ in the range  $[0.005,0.2]$; therefore:
 
@@
 p(n_i) = \begin{cases}
-		\displaystyle \frac{1}{0.2-0.005} &\text{for~} 0.005 <  N_i \leq 0.2 ,  \\
+		\displaystyle \frac{1}{0.2-0.005} &\text{for~} 0.005 <  n_i \leq 0.2 ,  \\
 		0 &\text{otherwise}.
@@
 is that $\sigma^2$ is always positive.
-We thus assume a Jeffreys prior \citep{sivia}, expressed as:
+We thus assume a Jeffreys prior \citep{sivia} for $\sigma^2$, expressed as:
 

### 2014-01-16 paper/formulation.tex
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 We also introduce
-a stochastic Polynomial Chaos-based spectral method for forward uncertainty propagation and accelerating the Bayesian inference.
+a Polynomial Chaos method used for forward uncertainty propagation, and 
+for accelerating the Bayesian inference based on the resulting surrogate.
 

### 2014-01-16 paper/inference.tex
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 To formulate the likelihood function, we let $\vec \epsilon = \vec d - \vec{G}$
-represents the discrepancy between the model and observations.
+represent the discrepancy between the model and observations.
 Here, the components of $\vec \epsilon $ are assumed to be i.i.d. random variables with density $p_{\epsilon}$.
-In this case the likelihood function can be written as:
+The likelihood function can thus be written as:
 
@@
 and normally distributed with mean
-zero and variance $\sigma^2$, i.e. 
-$\epsilon_i \sim N(0,\sigma^2)$. 
+zero and variance $\sigma^2$, i.e. $\epsilon_i \sim N(0,\sigma^2)$. 
 While in general $\sigma^2$ depends on the observations, in cases where the error
-amplitude is generally small and does not change throughout space and time, one may use
-a simplified model and assume single $\sigma^2$ value.
-Consequently, Equation~\eqref{eq:likelihood} becomes:
+amplitude is generally small and does not change throughout space and time.
+Consequently, one may use assume single $\sigma^2$ value in Equation~\eqref{eq:likelihood},
+resulting in:
 
@@
 \end{equation}
-The variance $\sigma^2$ is unknown \emph{a priori}; thus we treat it as a hyper-parameter.
-In other words, $\sigma^2$ may become an additional parameter for Bayesian inference and  
-therefore endowed with a prior and estimated from data. In this case the joint posterior is finally written as follows:
+Since the variance $\sigma^2$ is unknown \emph{a priori}; we treat it as a hyper-parameter.
+In other words, $\sigma^2$ become an additional parameter for Bayesian inference and  
+therefore endowed with a prior which is updated based on available observations. In this case the joint posterior is finally expressed as:
 

### 2014-01-16 paper/intro.tex
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 hazard by assessing the risk they pose, attempting  to make informed decisions
-about the likelihood that such an event would  occur with cost to life and
-property.  Toward this goal, computational models of  tsunami events are often
-employed to explore multiple different scenarios to predict the water elevation
+about the likelihood that such an event would  occur.
+Toward this goal, computational models of  tsunami events are often
+employed to explore multiple scenarios to predict the water elevation
 caused by the tsunamis. The accurate prediction  of water elevations, however,
@@
 directly, defined empirically, or estimated from a collection of observed data.
-Unfortunately, since tsunami events are relatively rare,  there is an abundance
-of uncertainty in the input data for these computational  models ranging from
+Unfortunately, since tsunami events are relatively rare,  there is a large
+uncertainty in the input data for these computational  models ranging from
 effects of the domain, such as bathymetry and friction parameterizations,  to
@@
 commonly used in tsunami models.  We aim at quantifying the uncertainty in the 
-predicted water elevation  and employ
-a Bayesian inverse modeling approach to estimate the Manning's $n$ coefficient
-using observations of water elevation measured during a tsunami event.
+predicted water elevation  and employ a Bayesian inverse modeling approach to 
+estimate the Manning's $n$ coefficient using observations of water elevation 
+measured during a tsunami event.
 
-Previous work looking into uncertainty in the input for tsunami simulations
-focused on the earthquake source but primarily as a way of
-constraining the various proposed earthquake scenarios only after the event (see
+\comment{Previous work looking into uncertainty in the input of tsunami models
+focused on the earthquake source models but primarily as a way of
+constraining the various proposed earthquake scenarios} (see
 for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
@@
 sensitivity analysis.  Changes to the friction parameterization has been shown in 
-the past to lead to significant changes to inundation levels \cite{Myers:2001el, 
+the past to lead to significant changes in predicted inundation levels \cite{Myers:2001el, 
 Jakeman:2010hk}, but these previous studies were limited in the range of the 
-values and locations used in the parameterization.  
+\comment{values and locations used in the parameterization}.  
 In a recent work, Mayo \emph{et al.}~\cite{Mayo:2014} reformulated a statistical data assimilation method
@@
 representation of Manning's $n$ coefficient and an efficient method
-to estimate this representation using real water elevation data. The method proposed here 
+to estimate this representation using measured water elevation data. The method proposed here 
 is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
@@
 from a number of observations.  In particular, we implement an inverse modeling
-approach that relies on the Bayesian inference technique.  We resort to
-Polynomial Chaos (PC) expansions to construct a faithful surrogate model of the
-response of the large and complex geophysical model simulations \cite{Najm:2009,Alexanderian2012,sraj:2013a,sraj:2013b}.
-The construction of the surrogate enables us to efficiently implement the Bayesian
-inference technique as it dramatically reduces the cost of sampling the
+approach that relies on a Bayesian inference technique.  We resort to
+Polynomial Chaos (PC) expansions to build a surrogate model ~\cite{Najm:2009,Alexanderian2012,Elsheikh2014,Young2013}
+of the response of the tsunami complex geophysical model \geoclaw used in this work.
+The PC surrogate enables us to avoid using the forward model in
+the Bayesian inference thus dramatically reducing the cost of sampling the
 posterior distribution.
@@
 Several studies already investigated the efficiency of PC expansions for 
-UQ in oceanic simulations~\citep{thacker2012,ashwanth2010,Alexanderian2012}. 
+UQ in oceanic simulations~\citep{thacker2012,ashwanth2010,Alexanderian2012,sraj:2013a}. 
 Recently, PC was combined with Bayesian inference to estimate the wind drag coefficient 
@@
 \geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} 
-including the two source models analyzed there.
+including the \comment{source models analyzed there}.
 

### 2014-01-16 paper/paper.tex
fixed figures and removed comments from paper
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 
-\author[kaust]{Ihab Sraj}
+\author[kaust1]{Ihab Sraj}
 \author[ut]{Kyle T. Mandli}
@@
 % \ead[http://users.ices.utexas.edu/~kyle]{http://users.ices.utexas.edu/~kyle}
-\author[duke]{Omar M. Knio}
+\author[duke,kaust2]{Omar M. Knio}
 \author[ut]{Clint N. Dawson}
@@
 \ead{ibrahim.hoteit@kaust.edu.sa}
-\ead{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
-\address[kaust]{Department of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+\ead{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+\address[kaust1]{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+
 \address[ut]{Institute for Computational Engineering and Science, University of Texas at Austin, 201 E 24th ST. Stop C0200, Austin, TX 78712-1229, USA}
@@
 Hudson Hall, Durham, North Carolina 27708, USA}
+\address[kaust2]{Division of Computer, Electrical and Mathematical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
 \date{\today}
@@
 
-Tsunami computational models are employed to explore multiple  different
+Tsunami computational models are employed to explore multiple
 scenarios and to predict water elevations.  However, accurate estimation of

### 2014-01-16 paper/tsunami.tex
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 is a common practice to assume a single value of Manning's $n$ for the areas
-that are initially below sea-level and sometimes use satellite images and 
-land-use data to determine the coefficient in areas above sea-level.  This
+that are initially below sea-level; while satellite images and 
+land-use data are used to determine the coefficient in areas above sea-level.  This
 process both above and below sea-level has several sources of uncertainty, all
 of which can have an impact on the quantities of interest in tsunami modeling.   
-In MacInnes \emph{et al.}, for
-instance, they found that adopting a uniform value of $n = 0.025$ everywhere
-lead to poor performance of the forward model's predicted inundation levels.  To
-address this they used $n = 0.035$ in the Sendai plane to account for land-uses 
-such as pasture, farmland, and rice paddies.  It is uncertainties such as these 
-that we wish to address in a systematic way so that accurate risk assessments 
+In MacInnes \emph{et al.}, for instance, they found that adopting a uniform value 
+of $n = 0.025$ everywhere leads to poor performance of the forward model's predicted 
+inundation levels.  To address this they used $n = 0.035$ in the Sendai plane to 
+account for land-uses such as pasture, farmland, and rice paddies.  It is 
+uncertainties such as these that we wish to address in a systematic way so that
+better risk assessments 
 can be done.
@@
 The direct result of the uncertainty in the Manning's $n$ coefficient lead to
-uncertainties in the quantities of interest such as the water elevation,
+uncertainties in water elevation,
 especially in the most crucial area in risk assessments, the near-shore. In 
-addition,  the prediction of these quantities of interest are often highly 
+addition,  the prediction of this and other quantities of interest are often highly 
 sensitive to changes in these uncertain parameters.  We seek to therefore 
@@
 as shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
-with respect to those three parameters we employ Polynomial Chaos (PC) 
+in these three parameters we employ Polynomial Chaos (PC) 
 expansions as used in \cite{sraj:2013a,sraj:2013b}.  The method and its 
@@
 The \tohoku earthquake and tsunami of 2011 have been the subject of a number of
-studies due to the wealth of observational evidence and severity of the tsunami.
+studies due to the wealth of observational evidence and severity of the tsunami~\cite{Kawase2014}.
 The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and
@@
 
-One unique feature that \geoclaw has is the ability to adaptively refine the
-grids used for the computation essentially following a region of interest as
+One of the unique features of \geoclaw is the ability to adaptively refine the
+grids used for the computation, essentially following a region of interest as
 time progresses (in this case, the disturbance in the surface height).
-\geoclaw implements these schemes following \cite{Berger:1984ui,Berger:1998aa}
-and is described in detail in \cite{Berger:2011du} in the case of tsunami
+To this end, \geoclaw implements the schemes in \cite{Berger:1984ui,Berger:1998aa}
+as described in detail in \cite{Berger:2011du} in the case of tsunami
 modeling.
@@
 was presented in \cite{MacInnes:2013cr}.  As was done there, four levels of
-refinement were used staring with a resolution of 1 degree in both the
+refinement were used starting with a resolution of 1 degree in both the
 latitudinal and longitudinal directions down to 0.5' resolution (approximately
@@
 Figure~\ref{fig:setup_buoy_data}. The time $t=0$ is set to be the start of the
-earthquake and is recorded every 15 seconds during the first few minutes of an
+earthquake and is recorded every 15 seconds during the first few minutes of the
 event, otherwise recording every minute when triggered (a tsunami has occurred)
@@
 The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
-accurate bathymetry for the region \cite{Amante:2009ud}.  While it is the case
+accurate bathymetry for the region \cite{Amante:2009ud}.  \comment{While it is the case
 that the inundation region is the most sensitive to changes in the Manning's $n$
@@
 locations. The bathymetry of the considered regions (and topography) is 
-shown in Figure~\ref{fig:setup}a.
-
+shown in Figure~\ref{fig:setup}a. }  
 Two earthquake source models were run to test whether the friction UQ analysis
-was sensitive to different rupture models, the first is from Saito et al.
+was sensitive to different rupture models. The first is from Saito et al.
 \cite{Saito:2011bh} and is based on an inversion analysis based on dispersive

### 2014-01-16 paper/uqpce.tex
changes based on omar comments, red sentences need rewording or are unclear, more to come

@@
 
-The four-dimensional posterior in Equation~\eqref{eq:post_coef} can be explored via MCMC;
-this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
+The four-dimensional posterior in Equation~\eqref{eq:post_coef} can be directly
+explored via MCMC; this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
 once for every proposed set of parameters of the Markov chain~\cite{MarzoukNajm2009,Malinverno2002}. While a single \geoclaw simulation
 takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
-to avoid forward model simulations altogether. This is achieved by constructing a 
+to avoid running the forward model at every realization of the MCMC. This is achieved by constructing a 
 surrogate model that requires a much
@@
 due to the uncertainties in a model's input. We briefly describe the PC
-method below; for more details on this method, and other UQ methods
+method below; for more details 
 the reader is referred to \citep{LeMaitreKnio2010}.
 
-Let $U=U(\bm{x},t,\xxi)$ denotes a quantity of 
-interest (QoI) that is the output of certain model.
-$U$ is function of space $\bm{x}$ and time $t$, and 
-also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). 
+Let $U=U(\bm{x},t,\xxi)$ denote a quantity of 
+interest (QoI) that is the output of a computational model.
+$U$ is function of space, $\bm{x}$, and time, $t$, and 
+also depends on the canonical vector of random variables $\xxi=(\xi_1,...,\xi_n)$
+that parameterize the uncertain inputs. 
 PC expresses $U$ in the form:
@@
 space. The total number of terms in the PC expansion is
-$P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimension and $p$ is the highest order
-of polynomials.
+$P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimensions and $p$ is the highest order
+polynomial order.
 
@@
 function of the stochastic variable $\xxi$, which appears as a weight
-function in the stochastic space's inner product:
+function in the probability space's inner product:
 
@@
 For uniform
-distributions the basis functions are hence scaled Legendre polynomials.
+distributions, the basis functions are scaled Legendre polynomials.
 For multi-dimensional problems the basis functions are
@@
 
-The series representation Equation~\eqref{eq:stochseries} can be viewed as a spectral expansion
-of $U$ along the stochastic dimensions. It can be also seen as
-combining approximation and probabilistic frameworks; a combination
-that has proven extremely useful in solving UQ problems~\cite{Xiu:2003,Lin2009}.
+The series representation ~\eqref{eq:stochseries} can be viewed as a spectral expansion
+of $U$ along the stochastic dimensions. It can also be seen as
+combination of approximation and probabilistic frameworks; this
+ has proven extremely useful in solving UQ problems~\cite{Xiu:2003,Lin2009}.
 The existence and convergence of this series is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
@@
 $U$ with respect to $\xxi$. The series converges spectrally fast with $P$
-when $U$ is infinitely smooth; the convergence rate becomes algebraic
+when $U$ is smooth; the convergence rate becomes algebraic
 when $U$ has finite smoothness \citep{Canuto:2006}. In practice the series convergence is monitored 
@@
 approach that allows the use of the forward model \geoclaw as a black box
-with no code modifications required. PC expansion coefficients are thus determined
+with no code modifications required. PC expansion coefficients are determined
 based on a set of response \geoclaw simulations at specified set of the uncertain parameters. 
-In particular, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
+Specifically, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
 
@@
 quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
-respectively. The quadrature order should be commensurated with the
+respectively. The quadrature order should be commensurate with the
 truncation order, and should be high enough to avoid aliasing artifacts.
 The choice of quadrature rule is hence critical to the performance
-of the PC (in its NISP version at least). In the current work, we employ the 
+of the PC (in its NISP version at least). \comment{In the current work, we employ the 
 tensorized Gaussian quadrature that yields 125 quadrature points for the set of three input parameters.
 The computational cost of this ensemble is low compared to the cost of the 
-tens of thousands of runs using the forward model.
+tens of thousands of runs using the forward model.}
 
@@
 The identification of the inner product weight function
-with the probability distribution of $\xxi$ simplifies the calculations of $U$'s statistical moments. 
-Noting that since $\Psi_0(\xxi)$ is a constant that can be normalized to satisfy 
+with the probability distribution of $\xxi$ simplifies the calculations of statistical moments of $U$. 
+Noting that since $\Psi_0(\xxi)$ is a constant that is normalized so that 
 $\left<\Psi_0,\Psi_0\right>=1$, the expectation and variance of $U$ can be computed as:
@@
 
-PC representations of random variables also enable conducting
-efficient global sensitivity analysis that aims to quantify the
-contribution of different random input parameters to the model variability.
+PC representations also enable conducting
+efficient global sensitivity analysis that quantify the
+contribution of different random input parameters to the variance in the output.
 This can be done by computing the so-called {\it total} 
@@
 \end{equation}
-where we define \[
+where \[
    K_i = \left\{ k \in \{1, \ldots, P\} :

### 2014-01-16 paper/analysis.tex
few fixes and wrote a conclusion
more changes and questions

@@
 The mean of the QoI and its standard deviation is computed
-from the PC coefficients as indicated in Equation~\eqref{eq:mean} and Equation~\eqref{eq:sigma}. 
+from the PC coefficients as indicated in Equations~\eqref{eq:mean} and \eqref{eq:sigma}. 
 In Figure~\ref{fig:ave} we plot the evolution of
-the calculated PC mean water surface elevation along with two standard deviations
-bounds at the four gauges as indicated in each panel.  
+the mean water surface elevation along with two standard deviations
+bounds at the four gauges.  
 Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant,
@@
 standard deviation in water surface elevations vanishes at few times instances
-during the tsunami. \alert{This can be explained as ...... }
+during the tsunami event. \comment{This can be attributed to the ...}
 
-These statistics can be used to compare with the 
-(DART) buoys observations. Figure~\ref{fig:compare} 
+The predicted mean water surface elevation  and its standard deviation 
+can be used to compare with the (DART) buoys observations collected during
+the tsunami event. Figure~\ref{fig:compare} 
 plots the PC mean water surface elevation with the observed 
@@
 \subsubsection{Sensitivity analysis}
-A global sensitivity analysis is next performed to quantity the contribution of each
+A global sensitivity analysis is performed to quantify the contribution of each
 uncertain parameter to the variance in water surface elevation. To this end, we calculate 
@@
 In 2D, the sensitivity analysis shows also that $n_2$ is dominant
-at the same regions where we observe variance in water surface elevation. This is
+at the same regions where we observe significant variance in water surface elevation. This is
 indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
 for $n_1$ (top row) $n_2$ (center row) and $n_3$ (bottom row)
-at different times as indicated in each panel.
+at different times.
 
@@
 range. Since $n_3$ shows insignificant contribution to the 
-uncertainty in the model output, we find the impact of $n_2$ and $n_3$ for 
-fixed value of $n_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
+uncertainty in the model output, we only consider the repsonse surface
+of $n_2$ versus $n_3$ for a fixed value of $n_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
 and Figure~\ref{fig:response3} for different times as indicated. The most striking features are the relatively flat
@@
 only mildly on $n_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,
-at gauge 21419, its the opposite case where the horizontal contours are in the $n_2 $ direction
-and there water surface elevation depends mildly on $n_2$.
+at gauge 21419, the horizontal contours are observed in the $n_2 $ direction
+and therefore water surface elevation depends mildly on $n_2$.

### 2014-01-16 paper/inference.tex
more changes and questions

@@
 Since the variance $\sigma^2$ is unknown \emph{a priori}; we treat it as a hyper-parameter.
-In other words, $\sigma^2$ become an additional parameter for Bayesian inference and  
-therefore endowed with a prior which is updated based on available observations. In this case the joint posterior is finally expressed as:
+In other words, $\sigma^2$ becomes an additional parameter for Bayesian inference and  
+therefore endowed with a prior which is updated based on available observations. In this 
+case the joint posterior is finally expressed as:
 

### 2014-01-16 paper/mcmc.tex
more changes and questions

@@
 water surface elevation at the different gauges locations and the  \geoclaw model counterparts. 
-The plots show a good agreement between the simulations and the 
+\comment{The plots show a good agreement between the simulations and the 
 observations at different locations and at different times. 
@@
 An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
-the epicenter of the earthquake and therefore receiving more impact by the tsunami.
+the epicenter of the earthquake and therefore receiving more impact by the tsunami.}
 
-Bayesian inference is now used to estimate the Manning's 
+\comment{Bayesian inference is now used to estimate the Manning's 
 $n$ coefficients ranges that minimize this scatter between 
@@
 efficient alternative that requires only evaluating the PC expansion
-for different values of the germ $\xxi$.
+for different values of the germ $\xxi$.}
 
@@
 lower end of the parameter range, with values between 0.005 and 0.1. 
-While for $n_{3}$, the chain appears to be concentrated in the 
+For $n_{3}$, the chain appears to be concentrated in the 
 upper end of the parameter range. Finally, the chain for the variance 
-appears to be well mixed with a well defined posterior range between 0.02 and 0.022.
+appears to be well mixed with a well defined posterior range between 0.02 and 0.022.\comment{discussion}.
 
@@
 and an extended tail towards the larger Manning's $n$ coefficient values.
-While for $n_3$, we observe a posterior that has a well defined peak
+For $n_3$, we observe a posterior that has a well defined peak
 of 0.18 but no clear pdf shape. The posterior distribution of the variance 
@@
 This value is a reflection of the mismatch between the model and 
-observed data.
+observed data. \comment{marginal "", correlation ''', joint}.
 

### 2014-01-16 paper/pcstudy.tex
more changes and questions

@@
 hour at all gauges as the plots of the different realizations superimpose.
-Later in time, the variability starts to increase at the different gauges 
-as indicated from the thickness of the bands formed by the plots of the realizations.
+Later in time, the variability starts to increase fro all gauges 
+as indicated by the thickness of the bands formed by the plots of curves for the selected realizations.
 This variability appears to be significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
@@
 where $\NISP$ is the 125-member ensemble obtained to construct the PC surrogate. 
-This error metric calculated at the different gauge locations is shown in Figure~\ref{fig:error};
+This error metric calculated at the different gauge locations, is shown in Figure~\ref{fig:error};
 the largest relative normalized error for 
 water surface elevation is about 0.1\%.
-
-%Contour maps of the relative normalized error for the entire simulation region are 
-%shown in Figure~\ref{fig:error2D} for various depths and dates to confirm the error trends of the
-%analysis box.  The error is largest after the wind
-%intensifies (bottom row) for all depths. For either day, the maximum
-%error is located at 50~m which coincides with the depth of the
-%original mixed layer measured by the AXBT. The maximum magnitude
-%recorded is about 1\% and occurs on Sep~18. The
-%elevated error region is located to the right of the storm and
-%extends from the surface down to 50~m, after which the impact of the
-%input uncertainty decreases substantially.  For the purpose of our
-%current study, and given that the majority of AXBT data are at
-%depth, the surrogate's errors are considered acceptable and small.
 

### 2014-01-16 paper/results.tex
more changes and questions

--- 2013-12-11 paper/results.tex
@@
 \label{sec:results}
-We begin the results section by performing
-an error and convergence analysis of the
-constructed surrogate using the PC expansions
-to establish its validity in Section~\ref{sec:analysis}.
-Then in Section~\ref{sec:forward}, we present
-statistical analysis to quantify the uncertainty in the predicted surface elevation
-as well as a sensitivity analysis to rank the impact of the different Manning's 
-roughness coefficients on this uncertainty. 
-Finally, in Section~\ref{sec:inverse}, we present the results of the inverse
-problem where we determined the posterior distributions of the input parameters 
-and analyzed them in light of the available gauge data.
+We begin by performing an error and convergence 
+analysis of the constructed surrogate using the PC expansions to 
+establish its validity. We then present a statistical analysis to quantify the 
+uncertainty in the predicted water surface elevation, as well as a 
+sensitivity analysis that enables us to rank the impact of the 
+different Manning's $n$ coefficients on the uncertainty. Finally, 
+we present the results of the inverse problem 
+where we determine the posterior distributions of the input uncertain
+parameters and analyze them in light of the available gauge data.
 \input{pcstudy}

### 2014-01-16 paper/intro.tex
Addressed comments and reworded

@@
 
-\comment{Previous work looking into uncertainty in the input of tsunami models
-focused on the earthquake source models but primarily as a way of
-constraining the various proposed earthquake scenarios} (see
+Previous work concerning uncertainty in tsunami modeling has been primarily been focused on constraining the likely earthquake source models (see
 for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
@@
 the past to lead to significant changes in predicted inundation levels \cite{Myers:2001el, 
-Jakeman:2010hk}, but these previous studies were limited in the range of the 
-\comment{values and locations used in the parameterization}.  
+Jakeman:2010hk}, but these previous studies were limited in range and spatial variation of the values used in the parameterization.  
 In a recent work, Mayo \emph{et al.}~\cite{Mayo:2014} reformulated a statistical data assimilation method
@@
 Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
-\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr} 
-including the \comment{source models analyzed there}.
+\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr}.
 
 The structure of this manuscript is as follows. First, we discuss the uncertainties 
-of the Manning'g $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
+of the Manning's $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
 and propose a three-parameter representation of $n$. 

### 2014-01-16 paper/tsunami.tex
Reworked all but bathymetry comment section

@@
 the empirically derived roughness parameter $n$:
-
 \begin{equation}
@@
 that are initially below sea-level; while satellite images and 
-land-use data are used to determine the coefficient in areas above sea-level.  This
+land-use data are sometimes used to determine the coefficient in areas above sea-level.  This
 process both above and below sea-level has several sources of uncertainty, all
@@
 uncertainties such as these that we wish to address in a systematic way so that
-better risk assessments 
-can be done.
+better risk assessments can be done.
 
@@
 was located approximately 72 km east of \tohoku as indicated 
-in Figure~\ref{fig:ceofs}a. In our case study, the
+in Figure~\ref{fig:ceofs}. In our case study, the
 simulation used is based on work done in \cite{MacInnes:2013cr} and modified to 
@@
 insensitivity of the simulation to the finer bathymetry at the DART buoy
-locations. The bathymetry of the considered regions (and topography) is 
-shown in Figure~\ref{fig:setup}a. }  
+locations.} The bathymetry and topography of the considered regions are 
+shown in Figure~\ref{fig:setup_buoy_locations}.
+
 Two earthquake source models were run to test whether the friction UQ analysis

### 2014-01-16 paper/inference.tex
Minor wording revisions

@@
 The prior distributions over the unknown parameters of the model
-are then formulated using our best belief. Lastly, given some observation data, Bayes rule 
+are then formulated using the best available knowledge of the parameters.  Lastly, given some observation data, Bayes rule 
 is used to obtain a posterior distribution for these unknown parameters
@@
 While in general $\sigma^2$ depends on the observations, in cases where the error
-amplitude is generally small and does not change throughout space and time.
-Consequently, one may use assume single $\sigma^2$ value in Equation~\eqref{eq:likelihood},
+amplitude is generally small \comment{and} does not change throughout space and time.
+Consequently, one may assume a single $\sigma^2$ value in Equation~\eqref{eq:likelihood},
 resulting in:
@@
 \end{equation}
-Since the variance $\sigma^2$ is unknown \emph{a priori}; we treat it as a hyper-parameter.
+Since the variance $\sigma^2$ is unknown \emph{a priori}\comment{;} we treat it as a hyper-parameter.
 In other words, $\sigma^2$ becomes an additional parameter for Bayesian inference and  

### 2014-01-16 paper/uqpce.tex
Minor wording revisions

@@
+%!TEX root = paper.tex
 \subsection{Accelerating Bayesian Inference}
@@
 
-The series representation ~\eqref{eq:stochseries} can be viewed as a spectral expansion
-of $U$ along the stochastic dimensions. It can also be seen as
+The series representation \eqref{eq:stochseries} can be viewed as a spectral expansion
+of $U$ along the stochastic dimensions. It can also be seen as a
 combination of approximation and probabilistic frameworks; this
  has proven extremely useful in solving UQ problems~\cite{Xiu:2003,Lin2009}.
-The existence and convergence of this series is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition of $U$ having a finite variance.
+The existence and convergence of this series is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition that $U$ has a finite variance.
 The series rate of convergence, and hence the number of terms to retain, depends on the smoothness of
@@
 with no code modifications required. PC expansion coefficients are determined
-based on a set of response \geoclaw simulations at specified set of the uncertain parameters. 
+based on a set of response \geoclaw simulations at a specified set of the uncertain parameters. 
 Specifically, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:
@@
 The choice of quadrature rule is hence critical to the performance
-of the PC (in its NISP version at least). \comment{In the current work, we employ the 
+of the PC (in its NISP version at least). \comment{In the current work, we employ a 
 tensorized Gaussian quadrature that yields 125 quadrature points for the set of three input parameters.
@@
 
-PC representations also enable conducting
+PC representations also enable
 efficient global sensitivity analysis that quantify the
-contribution of different random input parameters to the variance in the output.
+contributions of different random input parameters to the variance in the output.
 This can be done by computing the so-called {\it total} 

### 2014-01-16 paper/analysis.tex
Various rewording, typo fixes, and comments made

@@
 is now used as a surrogate to propagate prior input parameter uncertainty 
-through the forward model. We here exploit the surrogate 
+through the forward model.  Here we exploit the surrogate 
 to study the statistics of the water surface elevation and conduct 
@@
 in Figure~\ref{fig:rlzs}. An interesting observations is that the
-standard deviation in water surface elevations vanishes at few times instances
-during the tsunami event. \comment{This can be attributed to the ...}
+standard deviation in water surface elevations vanishes at a few time instances
+during the tsunami event.  \comment{Why is this?}
 
 The predicted mean water surface elevation  and its standard deviation 
-can be used to compare with the (DART) buoys observations collected during
+can be used to compare with the (DART) buoy observations collected during
 the tsunami event. Figure~\ref{fig:compare} 
@@
 Manning's $n$ coefficients $n_1$ and $n_3$; this is true for almost the entire simulation time
-and at the four gauges. The Manning's $n$ coefficient
-in the bottom of ocean $n_{3}$ at gauge number 21419 exhibits small sensitivity index 
-during the second hour of simulation and Manning's $n$ coefficient
-at the land $n_1$ appears to be an insignificant contributor
-to the variance.
+and at the four gauges.
+The Manning's $n$ coefficient in the deep ocean $n_{3}$ at gauge number 21419 exhibits a higher sensitivity index during the first hour of simulation however, as noted previously, none of the gauge locations see a significant variance between simulations until after the first hour of simulation.
 
@@
 can be used to construct a response surface for the uncertain input parameters.
-This is achieved by sampling the PC surrogate for different values of the germ $\xxi$ within the prior
+This is achieved by sampling the PC surrogate for different values of the term $\xxi$ within the prior
 range. Since $n_3$ shows insignificant contribution to the 
-uncertainty in the model output, we only consider the repsonse surface
+uncertainty in the model output, we only consider the response surface
 of $n_2$ versus $n_3$ for a fixed value of $n_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
@@
 and therefore water surface elevation depends mildly on $n_2$.
+
+\comment{Do you mean ``Since $n_1$ shows insignificant contribution...'' rather than $n_3$?  The value picked for $n_1$ also is quite high, can we pick something more reasonable such as 0.035?}

### 2014-01-16 paper/conc.tex
Various rewording, typo fixes, and comments made

--- 2013-12-23 paper/conc.tex
@@
+%!TEX root = paper.tex
 \section{Discussion and conclusions}
 \label{sec:conc}
+
+The present study aimed at estimating Manning's $n$ friction coefficient 
+which plays an important role in the accurate prediction of water surface elevation in 
+tsunami modeling. We proposed a three-parameter representation
+of Manning's $n$ friction coefficient that was characterized using iso-baths 
+to define three distinct regions, on-shore, near-shore, and deep-water, in the domain and represented each region with a single Manning's $n$ coefficient.
+The estimation relied on a Bayesian inference approach that sharpens the initial estimates of the three uncertain parameters
+using real observations.
+
+In our test case, the \tohoku tsunami, we used four DART buoy gauges that provide 
+surface elevation information.  To accelerate the Bayesian inference, we relied on the polynomial 
+chaos expansions to produce a faithful and efficient surrogate of the forward model \geoclaw.
+This PC surrogate model was then used to quantify the uncertainties in the predicted water surface elevations due to the uncertainties in the Manning's $n$ coefficient.
+This included the mean an standard deviation of water surface elevations which were also compared to the measured buoy data.
+A global sensitivity analysis was also performed in order to quantify the contribution of each uncertain parameter to the variance in surface elevation.
+It was found that the Manning's $n$ coefficient in the near-shore region $n_2$ contributed the most to the variance in the surface elevations compared to the other two Manning's $n$ coefficients in the on-shore and deep-water regions.
+Finally, using Bayesian inference, MAP estimates were found for the three region's $n$ parameters using MCMC.
+These values (excluding $n_1$ where no meaningful MAP value was found), $n_2=0.011$ and $n_3=0.180$, are \comment{unexpected and different from the expected values used commonly in tsunami modeling.
+This may be due to a number of other sources of uncertainty not taken into account in this analysis.}
+
+% The main results of this work were:
+% \begin{itemize}
+
+% \item We built a surrogate model using PC expansions that required 125 forward model runs of \geoclaw.
+% The surrogate model was tested, validated and proved to be faithful.
+
+% \item We used the PC surrogate to quantify the uncertainties in the predicted water surface elevation 
+% due to the uncertainties in the Manning's $n$ coefficient.
+To this end, we estimated the mean and the standard deviation of water surface elevation and compared it with the measured data collected at found gauges.
+
+% \item A global sensitivity analysis was performed to quantity the contribution of each uncertain parameter to the variance in water surface elevation.
+We found that the Manning’s $n$ coefficient at the shore $n_2$ is dominant and
+% contributes the most to the variance in the water surface elevation compared
+% to the other two Manning’s $n$ coefficients $n_1$ and $n_3$ for almost the
+% entire tsunami event.
+
+% \item We used Bayesian inference and identified the following MAP
+% estimates for the three $n$ parameters using MCMC, namely $n_2=0.011$ and
+% $n_3=0.180$, while for $n_1$ no meaningful MAP value could be determined from the
+% available data.
+
+% \end{itemize}
+
+Finally, we note that the present study focused on formulating and estimating a low-dimensional representation
+of the Manning's $n$ coefficient using UQ techniques namely Bayesian inference and PC expansions.
+A high-dimensional representation of the Manning's $n$ coefficient would require a large number of forward runs that is computationally expensive.
+Alternative methods can be implemented to reduce the dimensionality of the problem and will be the objective of a future study.

### 2014-01-16 paper/mcmc.tex
Various rewording, typo fixes, and comments made

@@
 
-Before we attempt solving the inverse problem of parameter estimation, 
+Before we attempt to solve the inverse problem of parameter estimation, 
 we first check the ability of \geoclaw to simulate water surface elevation
 realistically. To this end, we present a comparison between the 
-(DART) buoys observations  and their \geoclaw model counterparts
+(DART) buoy observations and their \geoclaw model counterparts
 for a reference simulation in which the the Manning's $n$ coefficients were set to their default values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
 shows a scatter plot that compares the observed 
-water surface elevation at the different gauges locations and the  \geoclaw model counterparts. 
-\comment{The plots show a good agreement between the simulations and the 
-observations at different locations and at different times. 
-The difference between the observations and simulations is attributed to the uncertainty in the 
-Manning's $n$ coefficients and to the measurement errors.
-An interesting observation is the more scatter in the data collection by gauge 21418 which is the closest to
-the epicenter of the earthquake and therefore receiving more impact by the tsunami.}
+water surface elevation at the different gauge locations and the \geoclaw model counterparts.
+The plots show a reasonable agreement between the simulation and the observations available.
+The differences between the simulation and observations can be attributed to uncertainty in the Manning's $n$ coefficients, errors in the earthquake rupture model, and insufficiently accurate bathymetry in the near-shore region.
+The high amount of scatter from gauge 21418 can probably be attributed to its proximity to the epicenter of the earthquake and shore region.
 
 \comment{Bayesian inference is now used to estimate the Manning's 
-$n$ coefficients ranges that minimize this scatter between 
-observed and modeled water surface elevation, and to estimate the variance of the noise in the measured data.
+$n$ coefficient's ranges that minimize this scatter between 
+observed and modeled water surface elevations and to estimate the variance of the noise in the measured data.
 To this end, an adaptive MCMC method is used to sample 
@@
 and to obtain the posteriors of the Manning's $n$ coefficients: 
-$n_1$,$n_2$ and $n_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
+$n_1$, $n_2$ and $n_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
 shows the sample chains for the input parameters for different iterations of the MCMC algorithms. 
-The different panels shows well-mixed chains for all input parameters as well as the variance $\sigma^2$.
+The different panels show well-mixed chains for all input parameters as well as the variance $\sigma^2$.
 The chain of  $n_{1}$ spans the entire range of the prior, meaning that the observations are not informative 
@@
 upper end of the parameter range. Finally, the chain for the variance 
-appears to be well mixed with a well defined posterior range between 0.02 and 0.022.\comment{discussion}.
+appears to be well mixed with a well defined posterior range between 0.02 and 0.022.  \comment{Add discussion}
 

### 2014-01-16 paper/pcstudy.tex
Various rewording, typo fixes, and comments made

@@
 hour at all gauges as the plots of the different realizations superimpose.
-Later in time, the variability starts to increase fro all gauges 
+Later in time, the variability starts to increase for all gauges 
 as indicated by the thickness of the bands formed by the plots of curves for the selected realizations.
@@
 for gauges 21401, 21413, 21418, 21419, respectively.
-This is consistent with the distance from the gauge to the source of the earthquake
-(epicenter located approximately 72 km east of \tohoku) where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup_buoy_locations}.
+This is consistent with the distance from the gauge to the epicenter of the earthquake located approximately 72 kilometer east of Japan where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup_buoy_locations}.
 The uncertainty in the prediction of water surface  elevation persists till the end of the simulations
@@
 with those obtained from the PC surrogate. The different curves (not shown) 
-reveal an excellent agreement for all times and gauges locations.
+reveal excellent agreement for all times and gauge locations.
 To quantify this agreement, we define
 an error metric that measures the relative normalized root mean-square error between the left hand side function 
-in Equation~(\ref{eq:stochseries}) and its PC representation at the sampling points:
+in Equation~\eqref{eq:stochseries} and its PC representation at the sampling points:
 

### 2014-01-19 paper/analysis.tex
fixed figures and text

@@
 standard deviation in water surface elevations vanishes at a few time instances
-during the tsunami event.  \comment{Why is this?}
+during the tsunami event.  \comment{The narrowing of the variance at these instances
+is possibility due to the waves that are arriving
+during these times and reflected from a single source or from a region
+where the Manning's $n$ values have little influence over these waves (as
+opposed to those being reflected off the Japanese coast).}
 
@@
 can be used to construct a response surface for the uncertain input parameters.
-This is achieved by sampling the PC surrogate for different values of the term $\xxi$ within the prior
-range. Since $n_3$ shows insignificant contribution to the 
+This is achieved by sampling the PC surrogate for different values of the 
+canonical vector of random variables  $\xxi$ within the prior
+range. Since $n_1$ shows insignificant contribution to the 
 uncertainty in the model output, we only consider the response surface
-of $n_2$ versus $n_3$ for a fixed value of $n_1=0.1025$ as illustrated in Figure~\ref{fig:response2}
-and Figure~\ref{fig:response3} for different times as indicated. The most striking features are the relatively flat
+of $n_2$ versus $n_3$ for a fixed value of $n_1=0.035$ as illustrated in Figure~\ref{fig:response2}
+at $t=7200~s$. The most striking features are the relatively flat
 horizontal contours in the $n_3$ direction suggesting that water surface elevation depends
@@
 and therefore water surface elevation depends mildly on $n_2$.
-
-\comment{Do you mean ``Since $n_1$ shows insignificant contribution...'' rather than $n_3$?  The value picked for $n_1$ also is quite high, can we pick something more reasonable such as 0.035?}

### 2014-01-19 paper/inference.tex
fixed figures and text

@@
 zero and variance $\sigma^2$, i.e. $\epsilon_i \sim N(0,\sigma^2)$. 
-While in general $\sigma^2$ depends on the observations, in cases where the error
-amplitude is generally small \comment{and} does not change throughout space and time.
+While in general $\sigma^2$ depends on the observations, in our case we assume that the error
+amplitude is generally small and does not change throughout space and time.
 Consequently, one may assume a single $\sigma^2$ value in Equation~\eqref{eq:likelihood},

### 2014-01-19 paper/mcmc.tex
fixed figures and text

@@
 
-\comment{Bayesian inference is now used to estimate the Manning's 
+Bayesian inference is now used to estimate the Manning's 
 $n$ coefficient's ranges that minimize this scatter between 
@@
 efficient alternative that requires only evaluating the PC expansion
-for different values of the germ $\xxi$.}
+for different values of the canonical vector of random variables $\xxi$.
 
@@
 The MAP value of $\sigma^2=0.021$ can be used to estimate the water surface elevation standard 
-deviation estimated to be 0.145~$m$. 
-This value is a reflection of the mismatch between the model and 
+deviation estimated to be 0.145~$m$. This value is a reflection of the mismatch between the model and 
 observed data. \comment{marginal "", correlation ''', joint}.
 
+With the MCMC chains computed, we can now con- struct the joint posterior distributions of the parameters
+using KDE.
+Since the posterior of m is close to the prior, we show in Fig.
+17 the joint posterior distribu- tions of a and Vmax (left panel), and of a and a2 on 17–18 September (right panel).
+These pdfs clearly show a single peak located at a 5 1.026 and Vmax 5 34 m s21, and that can be identified with the MAP estimates from the marginalized pdfs.
+The right panel also clearly shows a tight estimate for a with little spread around it.
+Setting m 5 0 (default HYCOM value), and substituting the MAP estimates for Vmax and a in Eq.
+(5), we obtain the ‘‘optimal’’ (most likely) curve describing the drag co- efficient CD’s dependence on wind speed; this curve is shown in Fig.
+18.
+In addition to decreasing the variance between simulations and observations, the impact of the MAP values on temperature can be assessed with the help of the surrogate.
+The results for the area-averaged SST and mixed layer depth (other quantities of interest can be treated similarly) are displayed in Fig.
+19.
+The default drag parameters underestimate the cooling by 0.58C and the mixed layer depth by as much as 15 m; note that the mixed layer depth feels the impact of the in- creased drag from 18 to 20 September, otherwise, the two curves overlap.
 %   0.180222380841571   0.011347632046398   0.185486684362890   0.021050903421753

### 2014-01-19 paper/pcstudy.tex
fixed figures and text

@@
 
-Figure~\ref{fig:rlzs} plots the evolution of the
+\comment{In the current work, we employ a tensorized Gaussian quadrature to construct the PC surrogate
+model. To this end, 125 \geoclaw simulations were needed to compute the PC coefficients using Equation~\ref{eq:nisp}.
+The computational cost of running this ensemble is low compared to the cost of the tens of thousands of 
+runs of the forward model.} Figure~\ref{fig:rlzs} plots the evolution of the
 water surface elevation predicted at the four different gauges 

### 2014-01-19 paper/tsunami.tex
fixed figures and text

@@
 The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
-accurate bathymetry for the region \cite{Amante:2009ud}.  \comment{While it is the case
+accurate bathymetry for the region \cite{Amante:2009ud}.  \comment{Needs rewording: While it is the case
 that the inundation region is the most sensitive to changes in the Manning's $n$
@@
 
-Two earthquake source models were run to test whether the friction UQ analysis
+\comment{Needs discussing in results section as well: Two earthquake source models were run to test whether the friction UQ analysis
 was sensitive to different rupture models. The first is from Saito et al.
@@
 discussed in MacInnes \emph{et al.}, the source models used do not include rupture
-timing which could have an impact on the accuracy of the simulation.
+timing which could have an impact on the accuracy of the simulation.}
 

### 2014-01-20 paper/mcmc.tex
new jpdfs and discussion of correlations

--- 2014-01-16 paper/mcmc.tex
@@
 
-Next, the computed MCMC chains are used to determine the posterior 
-distributions using kernel density estimation (KDE)
-~\citep{Parzen1962,Silverman1986}.  The resulting posterior pdfs 
-of the three Manning's $n$ coefficients $n_1,n_2,n_3$ are shown 
-in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
+Next, the computed MCMC chains are used to determine the marginalized posterior 
+distributions using kernel density estimation (KDE)~\citep{Parzen1962,Silverman1986}.  
+The resulting marginalized posterior pdfs of the three Manning's $n$ coefficients 
+$n_1,n_2,n_3$ are shown in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
 Note that the first 10000 iterates, associated with the burn-in period, were discarded.  
-As expected from the chains shown in Figure~\ref{fig:mcmc}, the posterior pdfs of $n_1$ appear to be fairly flat, 
-and similar to the uniform prior; an indication that 
-the observed data were not useful to refine our prior knowledge for $n_1$.,
-In contrast, the posterior pdf of $n_2$ exhibits well-defined peak, 
-with a Maximum A Posteriori (MAP) estimate of around 0.011
-and an extended tail towards the larger Manning's $n$ coefficient values.
-For $n_3$, we observe a posterior that has a well defined peak
-of 0.18 but no clear pdf shape. The posterior distribution of the variance 
-appears to be well-defined of Gaussian-like shape. 
-The MAP value of $\sigma^2=0.021$ can be used to estimate the water surface elevation standard 
-deviation estimated to be 0.145~$m$. 
-This value is a reflection of the mismatch between the model and 
-observed data. \comment{marginal "", correlation ''', joint}.
+As expected from the chains shown in Figure~\ref{fig:mcmc}, the marginalized posterior 
+pdfs of $n_1$ appear to be fairly flat, and similar to the uniform prior; an indication 
+that the observed data were not useful to refine our prior knowledge for $n_1$. In 
+contrast, the posterior pdf of $n_2$ exhibits well-defined peak, with a Maximum A 
+Posteriori (MAP) estimate of around 0.011 and an extended tail towards the larger 
+Manning's $n$ coefficient values. For $n_3$, we observe a posterior that has a well 
+defined peak of 0.18 but no clear pdf shape. The posterior distribution of the variance 
+appears to be well-defined of Gaussian-like shape. The MAP value of $\sigma^2=0.021$ 
+can be used to estimate the water surface elevation standard deviation estimated to 
+be 0.145~$m$. This value is a reflection of the mismatch between the model and 
+observed data.
+
+We finally investigate the existence of any correlation arising between the inferred parameters.
+This study is motivated by the fact that the Manning's $n$ coefficients have an important role 
+in characterizing the tsunami model predictions. Moreover, it is instructive to analyze the 
+correlation between the various $n$ parameters to gain some insight into their mutual dependence.
+To this end, we compute the two-dimensional marginalized joint posteriors for each 
+possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $p(n_1, n_2), p(n_1, n_3)$ and $p(n_2, n_3)$. 
+Figure~\ref{fig:joint}(left) shows the scatter plots of the MCMC samples of the combinations of 
+inferred parameters obtained from the chain samples of the marginalized posterior excluding 
+the burn-in iterations. The plots reveal no clear correlation in all three cases. 
+These samples can be readily exploited to reconstruct the joint density for each combination of parameters. 
+The right column of Figure~\ref{fig:joint} shows the contour plots of the marginalized joint posteriors 
+$p(n_1, n_2)$, $p(n_1, n_3)$ and $p(n_2, n_3)$. The contours are obtained from the
+distribution reconstructed via KDE and shown for the cases corresponding to greater than $10\%$
+of the maximum probability.
+
+To quantify the degree of no-correlation between each possible pair of parameters, we compute 
+the corresponding correlation coefficient which is a unitless scalar that falls in the range 
+$(-1, 1)$, with $1$ indicating a perfect positive correlation, $−1$ a perfect negative correlation and $0$ for
+a no-correlation case.
+% $R(n_2, n_3)$, $R(n_2, \sigma^2)$, and $R(n_3, \sigma^2)$.
+The correlation coefficient, $R(X, Y )$, between two random variables $X$ and $Y$ is defined as:
+
+\begin{equation}
+R(X,Y) = \frac{E[(X - \mu_X)(Y - \mu_Y )]}{\sigma_X \sigma_Y} 
+\end{equation}
+where $\sigma_X$ and  $\sigma_Y$ are the standard deviations of  $X$ and $Y$, $\mu_X$
+and $\mu_Y$  are the means of  $X$ and $Y$ respectively;
+and $E$ is the expectation operator.  In our case, the correlation coefficient
+was less than $0.06$ for all possible combinations of parameters; again indicating no correlation
+existing between the inferred parameters.
 

### 2014-01-23 paper/formulation.tex
Add Clint's edits

@@
 \label{sec:formu}
-In this section, we first, present the Bayesian approach to inverse problems 
+In this section, we first present the Bayesian approach to inverse problems 
 and its application to our problem of inferring the Manning's $n$ coefficients.

### 2014-01-23 paper/mcmc.tex
Add Clint's edits

@@
 that the observed data were not useful to refine our prior knowledge for $n_1$. In 
-contrast, the posterior pdf of $n_2$ exhibits well-defined peak, with a Maximum A 
+contrast, the posterior pdf of $n_2$ exhibits a well-defined peak, with a Maximum A 
 Posteriori (MAP) estimate of around 0.011 and an extended tail towards the larger 

### 2014-01-23 paper/tsunami.tex
Add Clint's edits

@@
 coefficient, the finest bathymetry used in MacInnes \emph{et al.} were
-not included in the this study due to the availability of the data and the
+not included in this study due to the availability of the data and the
 insensitivity of the simulation to the finer bathymetry at the DART buoy

### 2014-01-23 paper/analysis.tex
Add some more conclusions and revisions

@@
 standard deviation in water surface elevations vanishes at a few time instances
-during the tsunami event.  \comment{The narrowing of the variance at these instances
-is possibility due to the waves that are arriving
-during these times and reflected from a single source or from a region
-where the Manning's $n$ values have little influence over these waves (as
-opposed to those being reflected off the Japanese coast).}
+during the tsunami event.  The narrowing of the variance at these instances is
+possibly due to the waves the arrive due to reflections from a single source or
+from a region where Manning's $n$ values have little influence over these
+waves.
 

### 2014-01-23 paper/pcstudy.tex
Add Clint's edits
Add some more conclusions and revisions

--- 2014-01-16 paper/pcstudy.tex
@@
 
-Figure~\ref{fig:rlzs} plots the evolution of the
+In the current work, we employ a tensorized Gaussian quadrature to construct the PC surrogate
+model. To this end, 125 \geoclaw simulations were used to compute the PC coefficients using Equation~\ref{eq:nisp}.
+The computational cost of running this ensemble is low compared to the cost of the tens of thousands of 
+runs of the forward model.  Figure~\ref{fig:rlzs} plots the evolution of the
 water surface elevation predicted at the four different gauges 
@@
 insensitive with additional refinement  ($p = 5$) as shown in Figure~\ref{fig:pdfs3}. 
-We, therefore, used PC order $p = 5$ in all computations below.
+We therefore used PC order $p = 5$ in all computations below.
 This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 

### 2014-01-23 paper/tsunami.tex
Add some more conclusions and revisions

--- 2014-01-16 paper/tsunami.tex
@@
 The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
-accurate bathymetry for the region \cite{Amante:2009ud}.  \comment{While it is the case
-that the inundation region is the most sensitive to changes in the Manning's $n$
-coefficient, the finest bathymetry used in MacInnes \emph{et al.} were
-not included in the this study due to the availability of the data and the
-insensitivity of the simulation to the finer bathymetry at the DART buoy
-locations.} The bathymetry and topography of the considered regions are 
+accurate bathymetry for the region \cite{Amante:2009ud}.  Finer bathymetry was
+used in the study done in MacInnes \emph{et al.} in order to match the on-shore
+inundation levels.  In the current study this bathymetry was excluded since we
+are only concentrating on the data from the DART buoys and due to availability
+of the data.  The bathymetry and topography of the considered regions are 
 shown in Figure~\ref{fig:setup_buoy_locations}.

### 2014-01-29 paper/inference.tex
Add some more conclusions and revisions
changes some notations and figure captions

--- 2014-01-16 paper/inference.tex
@@
 \begin{equation}
- p(\vec{\theta}| \vec d) \propto 
- p(\vec d | \vec{\theta}) \ p(\vec{\theta}), 
+ \Pi(\vec{\theta}| \vec d) \propto 
+ L(\vec d | \vec{\theta}) \ q(\vec{\theta}), 
 \label{eq:bayes}
 \end{equation}
-where $p(\vec{\theta})$ is the prior of $\vec{\theta}$, representing the \emph{a priori} knowledge
+where $q(\vec{\theta})$ is the prior of $\vec{\theta}$, representing the \emph{a priori} knowledge
 about the parameters; 
-$p(\vec d| \vec{\theta})$ is the likelihood function representing
+$L(\vec d| \vec{\theta})$ is the likelihood function representing
 the probability of obtaining the data given the set of parameters $\vec{\theta}$;
-and finally $p(\vec{\theta}| \vec d)$ is the posterior,
+and finally $\Pi(\vec{\theta}| \vec d)$ is the posterior,
 representing the probability of occurrence of $\vec{\theta}$ given the data $\vec d $.
@@
 \begin{equation} 
-p(\vec d |  \vec{\theta}) 
+L(\vec d |  \vec{\theta}) 
 = 
@@
 zero and variance $\sigma^2$, i.e. $\epsilon_i \sim N(0,\sigma^2)$. 
-While in general $\sigma^2$ depends on the observations, in cases where the error
-amplitude is generally small \comment{and} does not change throughout space and time.
+While in general $\sigma^2$ (the variance of the noise in the measured data) depends on the observations, in our case we assume that the error
+amplitude is generally small and does not change throughout space and time.
 Consequently, one may assume a single $\sigma^2$ value in Equation~\eqref{eq:likelihood},
@@
 \begin{equation} 
-p(\vec{\theta}| \vec d)
+\Pi(\vec{\theta}| \vec d)
 \propto
@@
 \exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace  
-\prod_{i=1}^n p(\theta_i).
+\prod_{i=1}^n q(\theta_i).
 \end{equation}
-Since the variance $\sigma^2$ is unknown \emph{a priori}\comment{;} we treat it as a hyper-parameter.
+Since $\sigma^2$ is unknown \emph{a priori} we treat it as a hyper-parameter.
 In other words, $\sigma^2$ becomes an additional parameter for Bayesian inference and  
@@
 \begin{equation} 
-p(\vec{\theta},\sigma^2 | \vec d)
+\Pi(\vec{\theta},\sigma^2 | \vec d)
 \propto
@@
 \exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace
-\ p(\sigma^2) \prod_{i=1}^n p(\theta_i).
+\ q(\sigma^2) \prod_{i=1}^n q(\theta_i).
 \label{eq:post}

### 2014-01-29 paper/tsunami.tex
Remove mention of Saito rupture model

@@
 
-Two earthquake source models were run to test whether the friction UQ analysis
-was sensitive to different rupture models. The first is from Saito et al.
-\cite{Saito:2011bh} and is based on an inversion analysis based on dispersive
-tsunami simulations.  The second is from Ammon \emph{et al.} \cite{Ammon:2011dm} and
-used inversion of the seismic waves to reconstruct the rupture.  These two
-models were chosen as they take very different approaches to the reconstruction
-of the original fault and were shown to obtain the best un-shifted match to the
-observations at the DART buoys for their respective types of reconstruction.  As
-discussed in MacInnes \emph{et al.}, the source models used do not include rupture
-timing which could have an impact on the accuracy of the simulation.
-
-%\comment{Still working on getting the Saito \emph{et al.} source model, seems to do better at the DART buoys.}
-
-%\comment{We need to address not using the finest bathymetry better}
-
-% This issue
-% is mitigated somewhat by the decreasing sensitivity of water elevation levels
-% to friction with deepening water but the near shore region, where the setup for
-% inundation occurs, still can be highly sensitive to the friction values chosen.
+The earthquake source model that was used is due to Ammon \emph{et al.}
+\cite{Ammon:2011dm} who used inversion of the seismic waves to reconstruct the
+rupture.  This model was chosen as it provides a rupture reconstruction that was
+done using established techniques and did include tsunami inversion data.  The
+latter it was felt may introduce bias into the reported results of this study.
+As discussed in MacInnes \emph{et al.}, the source models used do not include
+rupture timing which could have an impact on the accuracy of the simulation.

### 2014-01-29 paper/mcmc.tex
Add some more conclusions and revisions

@@
 
-The MCMC algorithm required $50,000$ iterations to guarantee convergence
-and to obtain the posteriors of the Manning's $n$ coefficients: 
-$n_1$, $n_2$ and $n_3$ as well as for the variance $\sigma^2$. Figure \ref{fig:mcmc} 
-shows the sample chains for the input parameters for different iterations of the MCMC algorithms. 
-The different panels show well-mixed chains for all input parameters as well as the variance $\sigma^2$.
-The chain of  $n_{1}$ spans the entire range of the prior, meaning that the observations are not informative 
-concerning this input.  In contrast, the $n_{2}$ chain appears to be concentrated in the 
-lower end of the parameter range, with values between 0.005 and 0.1. 
-For $n_{3}$, the chain appears to be concentrated in the 
-upper end of the parameter range. Finally, the chain for the variance 
-appears to be well mixed with a well defined posterior range between 0.02 and 0.022.  \comment{Add discussion}
+The MCMC algorithm required $50,000$ iterations to guarantee convergence and to
+obtain the posteriors of the Manning's $n$ coefficients:  $n_1$, $n_2$ and $n_3$
+as well as for the noise variance $\sigma^2$. Figure \ref{fig:mcmc}  shows the sample
+chains for the input parameters for different iterations of the MCMC algorithms.
+The different panels show well-mixed chains for all input parameters as well as
+the noise variance $\sigma^2$. The chain of  $n_{1}$ spans the entire range of the
+prior, meaning that the observations are not informative  concerning this input.
+In contrast, the $n_{2}$ chain appears to be concentrated in the  lower end of
+the parameter range, with values between 0.005 and 0.1.  For $n_{3}$, the chain
+appears to be concentrated in the  upper end of the parameter range. Finally,
+the chain for the noise variance  appears to be well mixed with a well defined
+posterior range between 0.02 and 0.022.  These values align well with what is
+commonly considered physically relevant for near-shore and deep-water regions.
 
@@
 Manning's $n$ coefficient values. For $n_3$, we observe a posterior that has a well 
-defined peak of 0.18 but no clear pdf shape. The posterior distribution of the variance 
+defined peak of 0.18 but no clear pdf shape. The posterior distribution of the noise variance 
 appears to be well-defined of Gaussian-like shape. The MAP value of $\sigma^2=0.021$ 
@@
 To this end, we compute the two-dimensional marginalized joint posteriors for each 
-possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $p(n_1, n_2), p(n_1, n_3)$ and $p(n_2, n_3)$. 
+possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $\Pi(n_1, n_2), \Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. 
 Figure~\ref{fig:joint}(left) shows the scatter plots of the MCMC samples of the combinations of 
@@
 The right column of Figure~\ref{fig:joint} shows the contour plots of the marginalized joint posteriors 
-$p(n_1, n_2)$, $p(n_1, n_3)$ and $p(n_2, n_3)$. The contours are obtained from the
+$\Pi(n_1, n_2)$, $\Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. The contours are obtained from the
 distribution reconstructed via KDE and shown for the cases corresponding to greater than $10\%$

### 2014-01-29 paper/conc.tex
Add some more conclusions and revisions
Rewrite discussion/conclusions section
Include discussion about uncertainties from source and bathymetry and
more discussion about the resultant marginal PDFs and MAP values.

@@
 %!TEX root = paper.tex
-\section{Discussion and conclusions}
+\section{Discussion and Conclusions}
 \label{sec:conc}
 
-The present study aimed at estimating Manning's $n$ friction coefficient 
-which plays an important role in the accurate prediction of water surface elevation in 
-tsunami modeling. We proposed a three-parameter representation
-of Manning's $n$ friction coefficient that was characterized using iso-baths 
-to define three distinct regions, on-shore, near-shore, and deep-water, in the domain and represented each region with a single Manning's $n$ coefficient.
-The estimation relied on a Bayesian inference approach that sharpens the initial estimates of the three uncertain parameters
-using real observations.
+The present study aimed at estimating Manning's $n$ friction coefficient  which
+plays an important role in the accurate prediction of water surface elevation in
+tsunami modeling. We proposed a three-parameter representation of Manning's $n$
+friction coefficient that was characterized using iso-baths  to define three
+distinct regions, on-shore, near-shore, and deep-water, in the domain and
+represented each region with a single Manning's $n$ coefficient.  The estimation
+relied on a Bayesian inference approach that sharpens the initial estimates of
+the three uncertain parameters using real observations.
 
-In our test case, the \tohoku tsunami, we used four DART buoy gauges that provide 
-surface elevation information.  To accelerate the Bayesian inference, we relied on the polynomial 
-chaos expansions to produce a faithful and efficient surrogate of the forward model \geoclaw.
-This PC surrogate model was then used to quantify the uncertainties in the predicted water surface elevations due to the uncertainties in the Manning's $n$ coefficient.
-This included the mean an standard deviation of water surface elevations which were also compared to the measured buoy data.
-A global sensitivity analysis was also performed in order to quantify the contribution of each uncertain parameter to the variance in surface elevation.
-It was found that the Manning's $n$ coefficient in the near-shore region $n_2$ contributed the most to the variance in the surface elevations compared to the other two Manning's $n$ coefficients in the on-shore and deep-water regions.
-Finally, using Bayesian inference, MAP estimates were found for the three region's $n$ parameters using MCMC.
-These values (excluding $n_1$ where no meaningful MAP value was found), $n_2=0.011$ and $n_3=0.180$, are \comment{unexpected and different from the expected values used commonly in tsunami modeling.
-This may be due to a number of other sources of uncertainty not taken into account in this analysis.}
+In our test case, the \tohoku tsunami, we used four DART buoy gauges that
+provide  surface elevation information.  To accelerate the Bayesian inference,
+we relied on the polynomial  chaos expansions to produce a faithful and
+efficient surrogate of the forward model \geoclaw.  This PC surrogate model was
+then used to quantify the uncertainties in the predicted water surface
+elevations due to the uncertainties in the Manning's $n$ coefficient.  This
+included the mean an standard deviation of water surface elevations which were
+also compared to the measured buoy data.  A global sensitivity analysis was also
+performed in order to quantify the contribution of each uncertain parameter to
+the variance in surface elevation.  It was found that the Manning's $n$
+coefficient in the near-shore region $n_2$ contributed the most to the variance
+in the surface elevations compared to the other two Manning's $n$ coefficients
+in the on-shore and deep-water regions.  This is expected as the deep-water
+friction has little effect on the overall column.  On the other hand, the
+sensitivity of the analysis to the on-shore value being less than that in the
+deep-water was unexpected and we suspect is due primarily to the use of the
+DART buoy gauges used as they are far enough away from land regions that they
+would not have had as large of an impact as say if data from inundation zones
+was used.
 
-% The main results of this work were:
-% \begin{itemize}
+Using Bayesian inference, MAP estimates were found for the three region's $n$
+parameters using MCMC.  These values were (excluding $n_1$ where no meaningful
+MAP value was found), $n_2=0.011$ and $n_3=0.180$.  The values and their
+corresponding marginalized PDFs (Figure~\ref{fig:pdfs}) tell perhaps the most
+complete story in this analysis.  As was mentioned above, $n_2$ is the most
+sensitive and has the most clear MAP value.  Although the value was lower than
+expected, the tail includes the most commonly used values (0.022-0.025).  The
+deep-water value $n_3$ has a peak that is much higher than expected but a tail
+that does not taper off particularly quickly.  This is not altogether unexpected
+due to the low sensitivity of the simulations to this region's friction.  The
+on-shore value $n_1$ shows again how insensitive the simulations and
+observations used were to on-shore values.
 
-% \item We built a surrogate model using PC expansions that required 125 forward model runs of \geoclaw.
-% The surrogate model was tested, validated and proved to be faithful.
-
-% \item We used the PC surrogate to quantify the uncertainties in the predicted water surface elevation 
-% due to the uncertainties in the Manning's $n$ coefficient.
-To this end, we estimated the mean and the standard deviation of water surface elevation and compared it with the measured data collected at found gauges.
-
-% \item A global sensitivity analysis was performed to quantity the contribution of each uncertain parameter to the variance in water surface elevation.
-We found that the Manning’s $n$ coefficient at the shore $n_2$ is dominant and
-% contributes the most to the variance in the water surface elevation compared
-% to the other two Manning’s $n$ coefficients $n_1$ and $n_3$ for almost the
-% entire tsunami event.
-
-% \item We used Bayesian inference and identified the following MAP
-% estimates for the three $n$ parameters using MCMC, namely $n_2=0.011$ and
-% $n_3=0.180$, while for $n_1$ no meaningful MAP value could be determined from the
-% available data.
-
-% \end{itemize}
-
-Finally, we note that the present study focused on formulating and estimating a low-dimensional representation
-of the Manning's $n$ coefficient using UQ techniques namely Bayesian inference and PC expansions.
-A high-dimensional representation of the Manning's $n$ coefficient would require a large number of forward runs that is computationally expensive.
-Alternative methods can be implemented to reduce the dimensionality of the problem and will be the objective of a future study.
+Finally, it should be noted that the friction source term is only one of a
+number of other sources of uncertainty.  In particular, the rupture model and
+bathymetry used can often have a larger influence on the tsunami than friction
+does.  In the future, this type of analysis would be more illustrative if
+propagation of uncertainties in both the bathymetry and source models were
+included in each Manning's $n$ estimate.  We would also like to note that the
+present study focused on formulating and estimating a low-dimensional
+representation of the Manning's $n$ coefficient using UQ techniques namely
+Bayesian inference and PC expansions.  A high-dimensional representation of the
+Manning's $n$ coefficient would require a large number of forward runs that is
+computationally expensive.  Alternative methods can be implemented to reduce the
+dimensionality of the problem and will be the objective of a future study.

### 2014-01-30 setrun.py
Added sentence about insufficient info for inversion

@@ -122,7 +122,7 @@ def setrun(claw_pkg='geoclaw'):
 
     if clawdata.output_style == 1:
         # Output nout frames at equally spaced times up to tfinal:
-        clawdata.num_output_times = 4
+        clawdata.num_output_times = 16
         clawdata.tfinal = 4 * 3600.
         clawdata.output_t0 = True
 

### 2014-02-02 Dual_EnKF_Tsunami_old.py
Add new fault run script

@@ -1,0 +1,180 @@
+#!/usr/bin/env python
+
+"""
+The functions below implement the Dual Ensemble Kalman filter for 
+state-parameter estimation of the Tsunami model. The parameters 
+characterizes the earthquake (location, intensity, ...) that happens in 
+the sea. These parameters are only estimated at the initial time (t=0) 
+before the Tsunami waves start. Shallow water equations are then used to 
+propagate the resulting waves. 
+
+"""
+
+# Script originally adapted from matlab
+
+import numpy
+
+def swe_model(t, t_obs, fault_params=None):
+    r"""Evolve the shallow water model to t_observation from t_current
+
+    Input
+    -----
+     - *t* - (float) - Current time
+     - *t_obs* - (float) - Observation time, will run model to this time
+     - *fault_params* - (FaultData) - Data object containing fault specification,
+       if `fault_params = None` then continue from current time of model,
+       otherwise restart model with initial conditions using the 
+       `fault_params`.  Defaults to `None`.
+
+    Output
+    ------
+     - state - (ndarray) - Vector form of the state of the model at `t_obs`.
+    """
+
+    if fault_params is not None:
+        # Restart
+        raise NotImplemented("Restart not implemented yet.")
+    else:
+        # Continue to run GeoClaw forward
+        pass
+
+    return None
+
+def dual_ensemble_Kalman_filter(num_ensembles, num_obs, obs_times, t0, tf, t_observation, Xf, Pa, H, Y, R, sigma):
+    r"""
+
+    num_ensembles = Number of ensembles used
+    t0 = Initial time
+    t_observation = Time of observations
+    sigma = Standard error for perturbing the observations, size = [Ny x 1]
+
+
+    obs_times - (list) - List of observation times
+    Y - (ndarray(num_obs, len(obs_times)) - 
+
+    #
+    # In the Dual-EnKF, two parallel filters are simultaneously utilized for 
+    # the parameters and the state variables. First, the parameters might be
+    # propagated (here, kept unchanged) and then updated with available data of
+    # the state. The updated parameters are then used to integrate the state
+    # from the initial time to the current time. The data is then used again to
+    # update the state variables. 
+    #
[…]

### 2014-02-02 friction_paper/cover.tex
Move first paper to new directory
This is to make it easy to start on the tohoku-okada-pc study

--- 2013-10-30 paper/cover.tex
@@
 \bigskip
-Ihab Sraj$^1$, Kyle Mandli$^2$ \& Ibrahim Hoteit$^1$\\
+Ihab Sraj$^1$, Kyle Mandli$^2$, Ibrahim Hoteit$^1$, \& Clint Dawson$^2$\\
 \bigskip
@@
 King Abdullah University for Science and Technology, Thuwal, Saudi Arabia \\
-$^2$Department of Engineering, Austin, TX 
+$^2$Institute for Computational Engineering and Science\\
+University of Texas at Austin\\
+Austin, TX 
 \end{center}

### 2014-02-02 friction_paper/intro.tex
Move first paper to new directory
This is to make it easy to start on the tohoku-okada-pc study

--- 2014-01-16 paper/intro.tex
@@
 coefficient. They used a low-dimensional representation of
-Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data
-in a storm surge setting.
+Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data using tidal simulations.
 

### 2014-02-02 friction_paper/coefs.tex

--- 2014-01-16 paper/coefs.tex
@@
 \begin{equation} 
-p(n_1,n_2,n_3,\sigma^2 | \vec d) 
+\Pi(n_1,n_2,n_3,\sigma^2 | \vec d) 
 \propto \frac{1}{\sqrt{2 \pi \sigma^2}} 
@@
 \exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
-\ p(\sigma^2)p(n_1)p(n_2) p(n_3).
+\ q(\sigma^2)q(n_1)q(n_2) q(n_3).
 \label{eq:post_coef}
@@
 on some \emph{a priori} knowledge about the parameters. In our case, a uniform
-prior for all three Manning's $n$ parameters is assumed, with $n_i$ in the range  $[0.005,0.2]$; therefore:
+prior for all three Manning's $n$ parameters is assumed, with $n_i$ in the range  $[0.005 - 0.2]$; therefore:
 
 \begin{equation} 
-p(n_i) = \begin{cases}
+q(n_i) = \begin{cases}
 		\displaystyle \frac{1}{0.2-0.005} &\text{for~} 0.005 <  n_i \leq 0.2 ,  \\
@@
 \end{equation}
-Regarding the variance, the only information we know 
+Regarding the noise variance, the only information we know 
 is that $\sigma^2$ is always positive.
@@
 \begin{equation} 
-p(\sigma^2) =  \begin{cases}
+q(\sigma^2) =  \begin{cases}
 		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\

### 2014-02-02 friction_paper/uqpce.tex

--- 2014-01-16 paper/uqpce.tex
@@
 The choice of quadrature rule is hence critical to the performance
-of the PC (in its NISP version at least). \comment{In the current work, we employ a 
-tensorized Gaussian quadrature that yields 125 quadrature points for the set of three input parameters.
-The computational cost of this ensemble is low compared to the cost of the 
-tens of thousands of runs using the forward model.}
+of the PC (in its NISP version at least).
 
@@
  \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>},
+\label{eq:nisp}
 \end{equation} 
@@
         \]
-        and $\vec{\alpha}^k$ is the multi-index associated with $k^{th}$ term in the
+        and $\vec{\alpha}^k$ is the multi-index associated with the $k^{th}$ term in the
 PC expansion~\cite{LeMaitreKnio2010}.

### 2014-02-03 friction_paper/analysis.tex
implemented some final corrections by Ibrahim

--- 2014-01-23 paper/analysis.tex
@@
 in Figure~\ref{fig:rlzs}. An interesting observations is that the
-standard deviation in water surface elevations vanishes at a few time instances
+standard deviation in water surface elevations narrows at a few time instances
 during the tsunami event.  The narrowing of the variance at these instances is
-possibly due to the waves the arrive due to reflections from a single source or
+possibly due to the waves that arrive due to reflections from a single source or
 from a region where Manning's $n$ values have little influence over these
@@
 The predicted mean water surface elevation  and its standard deviation 
-can be used to compare with the (DART) buoy observations collected during
+can be used for comparison with the (DART) buoy observations collected during
 the tsunami event. Figure~\ref{fig:compare} 
@@
 The Manning's $n$ coefficient in the deep ocean $n_{3}$ at gauge number 21419 exhibits a higher sensitivity index during the first hour of simulation however, as noted previously, none of the gauge locations see a significant variance between simulations until after the first hour of simulation.
+The on-shore Manning's $n$ coefficient, $n_{1 }$, has no effect on the variance in
+the water surface elevation for the entire simulation time at all gauges.
 
@@
 at gauge 21419, the horizontal contours are observed in the $n_2 $ direction
-and therefore water surface elevation depends mildly on $n_2$.
+meaning that water surface elevation depends only mildly on $n_2$.

### 2014-02-03 friction_paper/conc.tex
implemented some final corrections by Ibrahim

--- 2014-01-29 paper/conc.tex
@@
 elevations due to the uncertainties in the Manning's $n$ coefficient.  This
-included the mean an standard deviation of water surface elevations which were
+included the mean and standard deviation of water surface elevations which were
 also compared to the measured buoy data.  A global sensitivity analysis was also
@@
 sensitivity of the analysis to the on-shore value being less than that in the
-deep-water was unexpected and we suspect is due primarily to the use of the
-DART buoy gauges used as they are far enough away from land regions that they
-would not have had as large of an impact as say if data from inundation zones
-was used.
+deep-water was unexpected. We suspect that this is primarily due to the use of the
+DART buoy gauges, being far enough from land regions to have a significant impact. 
+This might not be the case if data from inundation zones (that
+were not available to us) were used.
 
@@
 due to the low sensitivity of the simulations to this region's friction.  The
-on-shore value $n_1$ shows again how insensitive the simulations and
-observations used were to on-shore values.
+on-shore value $n_1$ shows again how insensitive the used simulations and
+observations were to on-shore values in our particular setting.  
+In conclusion, it is clear that the use of only the DART buoy observations and the 
+bathymetry resolution used here is not sufficiently informative to invert for the 
+Manning's $n$ values with any strong confidence, especially outside of the near-shore region.
 
@@
 number of other sources of uncertainty.  In particular, the rupture model and
-bathymetry used can often have a larger influence on the tsunami than friction
-does.  In the future, this type of analysis would be more illustrative if
+bathymetry used can often have an important influence on the tsunami.  
+In the future, this type of analysis would be more illustrative if
 propagation of uncertainties in both the bathymetry and source models were
-included in each Manning's $n$ estimate.  We would also like to note that the
+included in each Manning's $n$ estimate.  The
 present study focused on formulating and estimating a low-dimensional
-representation of the Manning's $n$ coefficient using UQ techniques namely
+representation of the Manning's $n$ coefficient using UQ techniques, namely
 Bayesian inference and PC expansions.  A high-dimensional representation of the
-Manning's $n$ coefficient would require a large number of forward runs that is
-computationally expensive.  Alternative methods can be implemented to reduce the
-dimensionality of the problem and will be the objective of a future study.
+Manning's $n$ coefficient would require a large number of forward runs and this remains
+computationally expensive.  We will work on alternative methods to reduce the
+dimensionality of the problem and this will be the objective of a future study.

### 2014-02-03 friction_paper/inference.tex
implemented some final corrections by Ibrahim

--- 2014-01-29 paper/inference.tex
@@
 that has recently gained great interest in different applications, including
-ocean~\citep{Alexanderian2011a,Zedler2012,sraj:2013a}
+ocean~\citep{Alexagnderian2011a,Zedler2012,sraj:2013a}
 climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
@@
 The prior distributions over the unknown parameters of the model
-are then formulated using the best available knowledge of the parameters.  Lastly, given some observation data, Bayes rule 
-is used to obtain a posterior distribution for these unknown parameters
-\citep{sivia}. We review this approach briefly below.
+are then formulated using the best available knowledge of the parameters.  Given some observation data, Bayes rule 
+is then used to computer the posterior distribution for these unknown parameters
+\citep{sivia}. We briefly review this approach below.
 \subsubsection{Formalism}

### 2014-02-03 friction_paper/intro.tex
implemented some final corrections by Ibrahim

@@
 Toward this goal, computational models of  tsunami events are often
-employed to explore multiple scenarios to predict the water elevation
+employed to explore multiple scenarios and to predict the water elevation
 caused by the tsunamis. The accurate prediction  of water elevations, however,
@@
 
-Previous work concerning uncertainty in tsunami modeling has been primarily been focused on constraining the likely earthquake source models (see
+Previous work concerning uncertainty quantification in tsunami modeling has been primarily 
+focused on constraining the likely earthquake source models (see
 for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
@@
 The PC surrogate enables us to avoid using the forward model in
-the Bayesian inference thus dramatically reducing the cost of sampling the
+the Bayesian inference, thus dramatically reducing the cost of sampling the
 posterior distribution.
@@
 quantify uncertainties in numerical simulations.  The main advantage of
-using PC expansions is the ability of propagating input uncertainties through 
+using PC expansions is the ability of efficiently propagating input uncertainties through 
 large, complex and non-linear models leading to the corresponding output uncertainties. 
@@
 
-The structure of this manuscript is as follows. First, we discuss the uncertainties 
-of the Manning's $n$ coefficient to motivate the UQ problem in Section~\ref{sec:manning}
-and propose a three-parameter representation of $n$. 
-Section~\ref{sec:tohoku} describes our case-study, the \tohoku tsunami,
-summarizes the forward simulations setup, and presents the DART buoys observational data. 
+The structure of this manuscript is as follows. First, we describe our case-study, the \tohoku tsunami,
+summarize the forward simulations setup, and present the DART buoys observational data
+in Section~\ref{sec:tohoku}. Section~\ref{sec:manning} discusses the uncertainties 
+of the Manning's $n$ coefficient to motivate the UQ problem and propose a three-parameter representation of $n$. 
 Section~\ref{sec:formu} presents the two main methods employed in the UQ analysis:

### 2014-02-03 friction_paper/mcmc.tex
Move first paper to new directory
This is to make it easy to start on the tohoku-okada-pc study
implemented some final corrections by Ibrahim

--- 2014-01-29 paper/mcmc.tex
@@
 Before we attempt to solve the inverse problem of parameter estimation, 
-we first check the ability of \geoclaw to simulate water surface elevation
-realistically. To this end, we present a comparison between the 
+we first check the ability of \geoclaw to realistically simulate water 
+surface elevation. To this end, we present a comparison between the 
 (DART) buoy observations and their \geoclaw model counterparts
@@
 observed and modeled water surface elevations and to estimate the variance of the noise in the measured data.
-To this end, an adaptive MCMC method is used to sample 
+To this end, an adaptive MCMC method is implemented to sample 
 the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
@@
 the parameter range, with values between 0.005 and 0.1. For $n_{3}$, the chain
-appears to be concentrated in the  upper end of the parameter range. Finally,
-the chain for the noise variance  appears to be well mixed with a well defined
-posterior range between 0.02 and 0.022.  These values align well with what is
+appears to be concentrated in the  upper end of the parameter range. These values align well with what is
 commonly considered physically relevant for near-shore and deep-water regions.
+Finally, the chain for the noise variance  appears to be well mixed with a well defined
+posterior range between 0.02 and 0.022.
 
@@
 $n_1,n_2,n_3$ are shown in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
-Note that the first 10000 iterates, associated with the burn-in period, were discarded.  
+Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
 As expected from the chains shown in Figure~\ref{fig:mcmc}, the marginalized posterior 
@@
 We finally investigate the existence of any correlation arising between the inferred parameters.
-This study is motivated by the fact that the Manning's $n$ coefficients have an important role 
-in characterizing the tsunami model predictions. Moreover, it is instructive to analyze the 
+%This study is motivated by the fact that the Manning's $n$ coefficients have an important role 
+%in characterizing the tsunami model predictions. 
+It is indeed instructive to analyze the 
 correlation between the various $n$ parameters to gain some insight into their mutual dependence.
@@
 possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $\Pi(n_1, n_2), \Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$.
-Figure~\ref{fig:joint}(left) shows the scatter plots of the MCMC samples of the combinations of 
+Figure~\ref{fig:joint} (left) shows the scatter plots of the MCMC samples of the combinations of
 inferred parameters obtained from the chain samples of the marginalized posterior excluding 
@@
 was less than $0.06$ for all possible combinations of parameters; again indicating no correlation
-existing between the inferred parameters.
+between the inferred parameters.
 

### 2014-02-03 friction_paper/pcstudy.tex
implemented some final corrections by Ibrahim

--- 2014-01-23 paper/pcstudy.tex
@@
 model. To this end, 125 \geoclaw simulations were used to compute the PC coefficients using Equation~\ref{eq:nisp}.
-The computational cost of running this ensemble is low compared to the cost of the tens of thousands of 
-runs of the forward model.  Figure~\ref{fig:rlzs} plots the evolution of the
+The computational cost of running this ensemble is low compared to the cost of the thousands of 
+runs of the forward model that would have been required if a standard Monte-Carlo uncertainty propagation technique was to be adopted.  Figure~\ref{fig:rlzs} plots the evolution of the
 water surface elevation predicted at the four different gauges 

### 2014-02-03 friction_paper/tsunami.tex
implemented some final corrections by Ibrahim

--- 2014-01-29 paper/tsunami.tex
@@
 %!TEX root = paper.tex
-\section{Manning’s $n$ Friction Coefficient} \label{sec:manning}
-
-Manning's $n$ law provides a relationship between the friction coefficient
-$C_f$ appearing in the shallow water equations (SWE) governing tsunamis and 
-the empirically derived roughness parameter $n$:
-\begin{equation}
-    C_f = \frac{g n^2}{h^{4/3}},
-\label{eq:coef}
-\end{equation}
-The value of $n$ can vary both spatially and temporally due to changes in the
-bottom roughness from wave action and other processes.  For tsunami modeling it
-is a common practice to assume a single value of Manning's $n$ for the areas
-that are initially below sea-level; while satellite images and 
-land-use data are sometimes used to determine the coefficient in areas above sea-level.  This
-process both above and below sea-level has several sources of uncertainty, all
-of which can have an impact on the quantities of interest in tsunami modeling.   
-In MacInnes \emph{et al.}, for instance, they found that adopting a uniform value 
-of $n = 0.025$ everywhere leads to poor performance of the forward model's predicted 
-inundation levels.  To address this they used $n = 0.035$ in the Sendai plane to 
-account for land-uses such as pasture, farmland, and rice paddies.  It is 
-uncertainties such as these that we wish to address in a systematic way so that
-better risk assessments can be done.
-
-The direct result of the uncertainty in the Manning's $n$ coefficient lead to
-uncertainties in water elevation,
-especially in the most crucial area in risk assessments, the near-shore. In 
-addition,  the prediction of this and other quantities of interest are often highly 
-sensitive to changes in these uncertain parameters.  We seek to therefore 
-quantify the parametric uncertainty due to the uncertainty in the Manning's $n$ 
-friction coefficient by characterizing the spatial variability using
-iso-baths to define distinct regions in the domain that are characterized by a
-single Manning's $n$ coefficient.  In this study three regions were used, an
-on-shore region (initially above sea-level), a near-shore region (between
-sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
-200 meters), whose Manning's $n$ coefficient is denoted by $n_1$, $n_2$, and $n_3$ respectively 
-as shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
-in these three parameters we employ Polynomial Chaos (PC) 
-expansions as used in \cite{sraj:2013a,sraj:2013b}.  The method and its 
-application to our problem is
-described in Section~\ref{sec:uqpce}.  To represent the uncertainty in the
-Manning's $n$ coefficients an uninformative uniform distribution was assumed as 
-the prior with physically relevant values of $n$ sampled from the interval 
-$n \in [0.005-0.2]$.
-
-% This Manning's $n$
-% coefficient is allowed to vary spatially by defining a set of isobaths that 
-% partition the domain into disjoint regions where a single value of $n$ can be 
-% used.  The values used for $n$ can be chosen using satellite images and
-% land-use data but this is impractical for the values used in regions
-% below sea-level or in the case of highly variable land-use regions.  As a
-% result, values used for $n$ contain large amounts of uncertainty.
 
@@
 The \geoclaw software package uses a finite volume, wave-propagation approach as
-described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water \
-equations:
+described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water
+equations (SWE):
 \begin{equation} \label{eq:swe}
@@
 earthquake.  The large spikes later in time present in the data at gauge 21418
-are data anomalies and not subsequent tsunami waves.
+are data anomalies and not subsequent tsunami waves \textbf{any refrs}.
 
@@
 rupture timing which could have an impact on the accuracy of the simulation.
+
+\section{Manning's $n$ Friction Coefficient} \label{sec:manning}
+
+Manning's $n$ law provides a relationship between the friction coefficient
+$C_f$ appearing in the SWE governing tsunamis (Equation \ref{eq:swe}) and 
+the empirically derived roughness parameter $n$:
+\begin{equation}
+    C_f = \frac{g n^2}{h^{4/3}},
+\label{eq:coef}
+\end{equation}
+The value of $n$ may vary both spatially and temporally due to changes in the
+bottom roughness from wave action and other processes.  For tsunami modeling it
+is a common practice to assume a single value of Manning's $n$ for the areas
+that are initially below sea-level; while satellite images and 
+land-use data are sometimes used to determine the coefficient in areas above sea-level.  
+Both above and below sea-level, this process has several sources of uncertainty, all
+of which can have an impact on the quantities of interest in tsunami modeling.   
+MacInnes \emph{et al.}, for instance, found that adopting a uniform value 
+of $n = 0.025$ everywhere leads to poor performance of the forward model's predicted 
+inundation levels.  To address this they used $n = 0.035$ in the Sendai plane to 
+account for land-uses such as pasture, farmland, and rice paddies.  We aim to address such
+uncertainties in a systematic way in order to better study risk assessments can be done.
+
+The direct result of the uncertainty in the Manning's $n$ coefficient leads to
+uncertainties in water elevation,
+especially in the most crucial area in risk assessments, the near-shore. In 
+addition,  the prediction of this and other quantities of interest are often highly 
+sensitive to changes in these uncertain parameters.  We therefore seek to 
+quantify the parametric uncertainty due to the uncertainty in the Manning's $n$ 
+friction coefficient by characterizing the spatial variability using
+iso-baths to define distinct regions in the domain that are characterized by a
+single Manning's $n$ coefficient.  In this study three regions were used, an
+on-shore region (initially above sea-level), a near-shore region (between
+sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
+200 meters), whose Manning's $n$ coefficient is denoted by $n_1$, $n_2$, and $n_3$ respectively 
+as shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
+in these three parameters we employ Polynomial Chaos (PC) 
+expansions as used in \cite{sraj:2013a,sraj:2013b}.  The method and its 
+application to our problem is
+described in Section~\ref{sec:uqpce}.  To represent the initial uncertainty in the
+Manning's $n$ coefficients an uninformative uniform distribution was assumed as 
+the prior with physically relevant values of $n$ sampled from the interval 
+$n \in [0.005-0.2]$.
+
+% This Manning's $n$
+% coefficient is allowed to vary spatially by defining a set of isobaths that 
+% partition the domain into disjoint regions where a single value of $n$ can be 
+% used.  The values used for $n$ can be chosen using satellite images and
+% land-use data but this is impractical for the values used in regions
+% below sea-level or in the case of highly variable land-use regions.  As a
+% result, values used for $n$ contain large amounts of uncertainty.

### 2014-02-03 friction_paper/uqpce.tex
implemented some final corrections by Ibrahim

@@
 explored via MCMC; this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
-once for every proposed set of parameters of the Markov chain~\cite{MarzoukNajm2009,Malinverno2002}. While a single \geoclaw simulation
-takes $\sim 15~$mins, depending on the details of the MCMC algorithm used, it is desirable 
-to avoid running the forward model at every realization of the MCMC. This is achieved by constructing a 
-surrogate model that requires a much
+once for every proposed set of parameters of the Markov chain~\cite{MarzoukNajm2009,Malinverno2002}. 
+While a single \geoclaw simulation takes about $15~$mins, depending on the details of the MCMC algorithm used, 
+it is desirable to avoid running the forward model at every realization of the MCMC. 
+This is achieved by constructing a surrogate model that requires a much
 smaller ensemble of \geoclaw runs, and that can be used instead
@@
 dependencies of model outputs on the uncertain model inputs
-as a polynomial truncated expansion. This method has been developed in 
-the engineering community to represent uncertainties in the output of 
-numerical simulations~\citep{Villegas2012,Lin2009,Xiu2004}
-due to the uncertainties in a model's input. We briefly describe the PC
-method below; for more details 
+as a polynomial truncated expansion~\citep{Villegas2012,Lin2009,Xiu2004}.
+We briefly describe the PC method below; for more details 
 the reader is referred to \citep{LeMaitreKnio2010}.
@@
 The identification of the inner product weight function
-with the probability distribution of $\xxi$ simplifies the calculations of statistical moments of $U$. 
+with the probability distribution of $\xxi$ simplifies the calculations of the statistical moments of $U$. 
 Noting that since $\Psi_0(\xxi)$ is a constant that is normalized so that 
@@
    K_i = \left\{ k \in \{1, \ldots, P\} :
-           \vec{\alpha}^k_i > 0 \right\}
+           \vec{\alpha}^k_i > 0 \right\},
         \]

### 2014-02-03 friction_paper/analysis.tex
minor and changes spacing

@@
 during the tsunami event.  The narrowing of the variance at these instances is
-possibly due to the waves that arrive due to reflections from a single source or
-from a region where Manning's $n$ values have little influence over these
-waves.
+possibly due to the waves that arrive due to reflections from a single source
+and then move away from the gauge location, and thus no variance in the water
+surface elevation.
+
+% or from a region where Manning's $n$ values have little influence over these
+%waves.
 

### 2014-02-03 friction_paper/tsunami.tex
minor and changes spacing

@@
 %!TEX root = paper.tex
-
 \section{The \tohoku Tsunami} 
@@
 uncertainty analysis, and finally present the observations from NOAA DART buoys.
-
-%\comment{Kyle - I kind of figured the MacInnes paper goes into detail on this
-%so I am not sure what pertinent information should be included here.  We could
-%do some more info on the fault itself and some shore maximum inundations but we
-%really aren't using any of that.}
 
@@
 earthquake.  The large spikes later in time present in the data at gauge 21418
-are data anomalies and not subsequent tsunami waves \textbf{any refrs}.
+are data anomalies and not subsequent tsunami waves as observed in \cite{MacInnes:2013cr}.
 
@@
 account for land-uses such as pasture, farmland, and rice paddies.  We aim to address such
-uncertainties in a systematic way in order to better study risk assessments can be done.
+uncertainties in a systematic way in order to better study risk assessments.
 
@@
 $n \in [0.005-0.2]$.
-
-% This Manning's $n$
-% coefficient is allowed to vary spatially by defining a set of isobaths that 
-% partition the domain into disjoint regions where a single value of $n$ can be 
-% used.  The values used for $n$ can be chosen using satellite images and
-% land-use data but this is impractical for the values used in regions
-% below sea-level or in the case of highly variable land-use regions.  As a
-% result, values used for $n$ contain large amounts of uncertainty.

### 2014-02-03 friction_paper/analysis.tex
omars corrections

@@
 through the forward model.  Here we exploit the surrogate 
-to study the statistics of the water surface elevation and conduct 
+to study the statistics of the water surface elevation, and to conduct 
 a global sensitivity analysis of the impact of the uncertain input parameters.
@@
 reflected in a significant standard deviation as seen earlier from the realizations shown 
-in Figure~\ref{fig:rlzs}. An interesting observations is that the
+in Figure~\ref{fig:rlzs}. An interesting observation is that the
 standard deviation in water surface elevations narrows at a few time instances
@@
 water surface elevation at the different gauges locations
-with error bars indicated at few time instances. 
+with variances indicated at few time instances. 
 The plots show a good agreement between the PC simulations and the 

### 2014-02-03 friction_paper/coefs.tex
implemented some final corrections by Ibrahim
omars corrections

@@
 Bayesian inference can be applied directly to infer the uncertain parameters
-using Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
+based on Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
 are the water surface elevation at the gauges;
@@
 
-To complete the definition of the posterior, we need to choose proper priors that should be based 
+To determine the posterior, we need to choose proper priors that should be based 
 on some \emph{a priori} knowledge about the parameters. In our case, a uniform
@@
 
-Inferring the coefficients requires 
+Inferring the coefficients amounts to 
 sampling the posterior. In general, when the space of the unknown 
@@
 the Markov Chain Monte  Carlo (MCMC) method. 
-We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
-sample the posterior distribution accurately and efficiently.
+We rely on an adaptive Metropolis MCMC algorithm~\citep{Gareth2009,Haario2001} to
+accurately and efficiently sample the posterior distribution.

### 2014-02-03 friction_paper/conc.tex
minor and changes spacing
omars corrections

@@
 MAP value was found), $n_2=0.011$ and $n_3=0.180$.  The values and their
-corresponding marginalized PDFs (Figure~\ref{fig:pdfs}) tell perhaps the most
-complete story in this analysis.  As was mentioned above, $n_2$ is the most
+corresponding marginalized PDFs tell perhaps the most
+complete story in this analysis.  As mentioned earlier, $n_2$ is the most
 sensitive and has the most clear MAP value.  Although the value was lower than
@@
 that does not taper off particularly quickly.  This is not altogether unexpected
-due to the low sensitivity of the simulations to this region's friction.  The
+due to the low sensitivity of the QoI to this region's friction.  The
 on-shore value $n_1$ shows again how insensitive the used simulations and
@@
 number of other sources of uncertainty.  In particular, the rupture model and
-bathymetry used can often have an important influence on the tsunami.  
+bathymetry might often have an important influence on the tsunami.  
 In the future, this type of analysis would be more illustrative if

### 2014-02-03 friction_paper/inference.tex
omars corrections

@@
 The key idea in this approach is to express all forms of uncertainty
-in terms of probability. The first step in Bayesian inference 
+in terms of canonical random variables. The first step in Bayesian inference 
 is to formulate the forward problem (response surface/model) using 
@@
 Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs.
-We consider a forward model $\vec G$ that predicts the data function of
+We consider a forward model $\vec G$ that predicts the data as function of
 the parameters such that:
@@
 \prod_{i=1}^m  
-p_\epsilon (d_i - G_i(\vec \theta)).  	
+p_\epsilon (d_i - G_i(\vec \theta)), 	
 \label{eq:likelihood}
 \end{equation}
+where $m$ is the number of observations (dimension of $\vec{d}$).
 
@@
 \end{equation}
-The joint posterior from Bayes' rule can be given as:
+The joint posterior from Bayes' rule can thus be given as:
 

### 2014-02-03 friction_paper/intro.tex
omars corrections

@@
 Polynomial Chaos (PC) expansions to build a surrogate model ~\cite{Najm:2009,Alexanderian2012,Elsheikh2014,Young2013}
-of the response of the tsunami complex geophysical model \geoclaw used in this work.
+of the response of the tsunami predicted using the \geoclaw model.
 The PC surrogate enables us to avoid using the forward model in

### 2014-02-03 friction_paper/mcmc.tex
omars corrections

@@
 water surface elevation at the different gauge locations and the \geoclaw model counterparts.
-The plots show a reasonable agreement between the simulation and the observations available.
+The plots show a reasonable agreement between the simulation and the observations.
 The differences between the simulation and observations can be attributed to uncertainty in the Manning's $n$ coefficients, errors in the earthquake rupture model, and insufficiently accurate bathymetry in the near-shore region.
@@
 appears to be well-defined of Gaussian-like shape. The MAP value of $\sigma^2=0.021$ 
-can be used to estimate the water surface elevation standard deviation estimated to 
+can be used to estimate the water surface elevation standard deviation that was found to 
 be 0.145~$m$. This value is a reflection of the mismatch between the model and 
@@
 $\Pi(n_1, n_2)$, $\Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. The contours are obtained from the
-distribution reconstructed via KDE and shown for the cases corresponding to greater than $10\%$
+distribution reconstructed via KDE, and depict values that are greater than $10\%$
 of the maximum probability.
@@
 and $E$ is the expectation operator.  In our case, the correlation coefficient
-was less than $0.06$ for all possible combinations of parameters; again indicating no correlation
+was in the range $(-0.002 , 0.06)$ for all possible combinations of parameters; again indicating no correlation
 between the inferred parameters.

### 2014-02-03 friction_paper/paper.tex
Move first paper to new directory
This is to make it easy to start on the tohoku-okada-pc study
minor and changes spacing
omars corrections

--- 2014-01-16 paper/paper.tex
@@
 % \documentclass[12pt]{article}
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass[review,12pt]{elsarticle}
 
@@
 \usepackage{latexsym,amssymb,amsmath,color}
-\newcommand{\revised}[1]{\textcolor{blue}{{#1}}}
-\newcommand{\comment}[1]{\textcolor{red}{{#1}}}
-\newcommand{\alert}[1]{\textbf{\textcolor{red}{{#1}}}}
+% \newcommand{\revised}[1]{\textcolor{blue}{{#1}}}
+% \newcommand{\comment}[1]{\textcolor{red}{{#1}}}
+% \newcommand{\alert}[1]{\textbf{\textcolor{red}{{#1}}}}
 \newcommand{\be}{\begin{equation}}
@@
 \author[ut]{Kyle T. Mandli}
-% \ead{kyle@ices.utexas.edu}
-% \ead[http://users.ices.utexas.edu/~kyle]{http://users.ices.utexas.edu/~kyle}
 \author[duke,kaust2]{Omar M. Knio}
 \author[ut]{Clint N. Dawson}
-\author[kaust]{Ibrahim Hoteit}
+\author[kaust1,kaust2]{Ibrahim Hoteit\corref{cor1}}
 \ead{ibrahim.hoteit@kaust.edu.sa}
-\ead{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+%\ead{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
 \address[kaust1]{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
@@
 \address[kaust2]{Division of Computer, Electrical and Mathematical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
+\cortext[cor1]{Corresponding author}
 \date{\today}
@@
 
-Tsunami computational models are employed to explore multiple
+Tsunami computational models are employed to explore multiple flooding
 scenarios and to predict water elevations.  However, accurate estimation of
 water elevations  requires accurate estimation of many model parameters, in
-particular the Manning's $n$ friction coefficient.  Our objective here is to
+particular the Manning's $n$ friction coefficient.  Our objective is to
 develop an efficient approach for the  uncertainty quantification and inference
@@
 
-\clearpage
-
-\input{figures}
-
-\clearpage
 \bibliographystyle{elsarticle-num}
 \bibliography{biblio}
+\clearpage
+\input{figures} 
 \end{document}

### 2014-02-03 friction_paper/pcstudy.tex
omars corrections

@@
 
-A final check for the validity of the PC approximation 
+Another check for the validity of the PC approximation 
 consists of verifying whether the probability density

### 2014-02-03 friction_paper/tsunami.tex
omars corrections

@@
 inundation levels.  In the current study this bathymetry was excluded since we
-are only concentrating on the data from the DART buoys and due to availability
-of the data.  The bathymetry and topography of the considered regions are 
-shown in Figure~\ref{fig:setup_buoy_locations}.
+are only concentrating on the data from the DART buoys. The bathymetry and 
+topography of the considered regions are shown in Figure~\ref{fig:setup_buoy_locations}.
 
@@
 \cite{Ammon:2011dm} who used inversion of the seismic waves to reconstruct the
-rupture.  This model was chosen as it provides a rupture reconstruction that was
-done using established techniques and did include tsunami inversion data.  The
-latter it was felt may introduce bias into the reported results of this study.
+rupture.  This model was chosen as it provides a rupture reconstruction based on
+established techniques and did include tsunami inversion data.  The
+latter may introduce bias into the reported results of this study and was consequently ignored.
 As discussed in MacInnes \emph{et al.}, the source models used do not include
@@
 especially in the most crucial area in risk assessments, the near-shore. In 
-addition,  the prediction of this and other quantities of interest are often highly 
-sensitive to changes in these uncertain parameters.  We therefore seek to 
+addition,  the prediction of this and other quantities of interest is often highly 
+sensitive to changes in these uncertain parameters.  Therefore, we seek to 
 quantify the parametric uncertainty due to the uncertainty in the Manning's $n$ 
@@
 iso-baths to define distinct regions in the domain that are characterized by a
-single Manning's $n$ coefficient.  In this study three regions were used, an
+single Manning's $n$ coefficient.  In this study three regions are used, an
 on-shore region (initially above sea-level), a near-shore region (between

### 2014-02-03 friction_paper/uqpce.tex
omars corrections

@@
 dependencies of model outputs on the uncertain model inputs
-as a polynomial truncated expansion~\citep{Villegas2012,Lin2009,Xiu2004}.
+as a truncated polynomial expansion~\citep{Villegas2012,Lin2009,Xiu2004}.
 We briefly describe the PC method below; for more details 
@@
 where $U_k(\xbold,t)$ are the polynomial coefficients, and
-$\Psi_k(\xxi)$ are functions that form an orthogonal basis of an underlying probability
-space. The total number of terms in the PC expansion is
+$\Psi_k(\xxi)$ are elements of an orthogonal basis of an underlying probability
+space. The total number of terms in the truncated PC expansion is
 $P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimensions and $p$ is the highest order
-polynomial order.
+polynomial retained.
 
 The choice of the basis is dictated by the probability density
-function of the stochastic variable $\xxi$, which appears as a weight
+function of the stochastic vector $\xxi$, which appears as a weight
 function in the probability space's inner product:
@@
   \left< U, \Psi_k \right> 
-\approx \left< U, \Psi_k \right>_Q
-= \sum_{q=1}^Q U(\xxi_q) \Psi_k(\xxi_q) \omega_q,
+\approx \left< U, \Psi_k \right>_{\cal Q}
+= \sum_{q=1}^{Q} U(\xxi_q) \Psi_k(\xxi_q) \omega_q,
 \end{equation}
-where the subscript $Q$ refers to approximating the inner product integral with
+where the subscript $\cal Q$ refers to approximating the inner product integral with
 quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
-respectively. The quadrature order should be commensurate with the
+respectively and $Q$ is the number of nodes in the multi-dimensional quadrature. The quadrature order should be commensurate with the
 truncation order, and should be high enough to avoid aliasing artifacts.
@@
  U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
- \Pi_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>},
+ {\cal P}_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>},
 \label{eq:nisp}
 \end{equation} 
-where $\Pi_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
+where ${\cal P}_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
 from an ensemble of the deterministic model realizations with the uncertain parameters set at
@@
 \begin{equation}
- E[U] = \int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0,  
+ E[U] = \int U \, \rho(\xxi) \, \mbox{d}\xxi \approx \left< U,\Psi_0\right>_{\cal Q} = U_0,  
  \label{eq:mean}
@@
 and \begin{equation}
- E[(U-E[U])^2] = \int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
+ E[(U-E[U])^2] = \int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi \approx \sum_{k=1}^P U_k^2
  \label{eq:sigma}

### 2014-04-06 friction_paper/response_1.tex
Compile all reviewer feedback and reformat
Corrects for LaTeX idiosynchocies and enumerates some of the items.

%
%  Reviewer Responses Revision 1
%  6 April, 2014
%
\documentclass[]{article}

% Use utf-8 encoding for foreign characters
\usepackage[utf8]{inputenc}

% Setup for fullpage use
\usepackage{fullpage}

% Uncomment some of the following if you use the features
%
% Running Headers and footers
%\usepackage{fancyhdr}

% Multipart figures
%\usepackage{subfigure}

% More symbols
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{latexsym}

\usepackage{url}

% Surround parts of graphics with box
\usepackage{boxedminipage}

% Package for including code in the document
\usepackage{listings}

% If you want to generate a toc for each chapter (use with book)
\usepackage{minitoc}

% This is now the recommended way for checking for PDFLaTeX:
\usepackage{ifpdf}

\ifpdf
\usepackage[pdftex]{graphicx}
\else
\usepackage{graphicx}
\fi

% Commands 
\include{commands}

% Graphics
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{epstopdf}

\title{Response to Reviewers of ``Uncertainty Quantication and Inference of Manning's Friction Coefficient using DART Buoys Data during Tohoku Tsunami''}
\author{I. Sraj, K.T. Mandli, O.M. Knio, Dawson, C.N., Hoteit, I.}
\date{\today}

\begin{document}

\ifpdf
\DeclareGraphicsExtensions{.pdf, .jpg, .tif}
\else
\DeclareGraphicsExtensions{.eps, .jpg}
\fi

\makesmalltitle{Response to Reviewers of ``Adaptive Mesh Refinement for Storm Surge''}{Revision 1 - \today}

\section*{Reviewer 1}
\begin{quote}
The article presents an uncertainty analysis for a tsunami propagation model (GeoClaw) using the method of polynomial chaos. The focus is on the uncertainty in specifying the roughness parameter (aka Manning coefficient) controlling the amount of momentum lost to bottom drag, and on its ensuing effect on sea surface height prediction. The uncertainty analysis involves the forward propagation of the uncertainty via an ensemble type calculation, the validation of the uncertainty representation, the analysis of the statistical output, and the use of observational data to infer the likely distribution of the roughness parameter. The work presented highlight the benefits of incorporating newly developed tools for uncertainty quantification within the framework of risk assessment and disaster preparation.

The major concern of inundation model is to predict the tsunami run-up once it hits the shore, as the authors point out correctly in the introduction. Yet most of the figures concern the off-shore buoy data; only figure 11 show on-shore result. I would urge the authors to present the forward uncertainty propagation (and the uncertainty analysis) for an on-shore location, even in the absence of verification data, because inundation is the major practical concern. This would highlight the practical side of PC methods for risk assessment.

It is somewhat surprising to me that the near-shore roughness is so important to off-shore SSH, particularly given its small surface extent compared to deep-water roughness. It this unique to the Tohoku tsunami since it was generated in the near shore? Also are the high sensitivities for $n_1$ and $n_3$ (figure 11 and compare to the front location in figure 9) reliable? They seem to be beyond the tsunami reach after 1 hour of propagation (assuming an average gravity wave speed of 200 m/s would give it a radius of 720 km, and these regions seem twice as far ~1500 km). Likewise for figure 10, is the sensitivity index meaningful at the buoys prior to the tsunami arrival time? Some clarification of this point is needed.

I would like to see some more physical interpretation of the mathematical results and assumptions. For example a rational for using 3 roughness parameters, a dynamical explanation for the non-monotonic behavior of the SSH standard deviation in time (the explanation given are somewhat speculative and it would be nice to strengthen them with simple analysis, I am referring to figure 7 and lines 393-395 in the text).

I am a little confused about figure 16 and what it represents. I was expecting somewhat of a maximum around the 2D MAP values, but instead the contours are spread out. Could the authors elaborate on this point some more? Is it a problem with the colormap with only blue appearing? It could be useful to normalize the scales of these distribution to simplify their interpretation as joint pdfs.
\end{quote}

\subsection*{Detailed Comments}
\begin{enumerate}
\item Use either Manning’s coefficient or simply n, not “Manning’s n coefficient”
\item In abstract fix “T ohoku” to “Tohoku”
\item Lines 3-8: Suggestion: Coastal communities have faced these hazard by assessing the risk they pose, and by attempting to make informed decisions about their likely impacts. Computational models of tsunami events are often employed to explore various tsunami disaster scenarios and their ensuing inundation levels.
\item Line 24: change “The implemented emulator” to “This emulator”
\item Line 31: remove the comma after “In a recent work”
\item Line 33 and 34: Use either “Manning’s coefficient” or n.
\item Lines 51-54: Change to: Their main advantage is the efficient propagation of input uncertainties through large, complex and non-linear models to calculate the corresponding output uncertainties.
\item Line 81: The section title should be changed as it does not reflect the content to something like:  “Simulating the Tohoku Tsunami”
\item Line 85: Change “The source of the earthquake (epicenter)” to “The epicenter”
\item Lines 95-99: the sea surface anomaly does not appear anywhere in the equations or the explanation even though it is the primary quantity of interest here.
\item Lines 100-115 should be a single paragraph describing GEOCLAW. I suggest the following :
\begin{quote}
GeoClaw was originally designed to solve systems of hyperbolic equations where the primary computational kernel is the solution to the Riemann problem at each grid cell interface. The model’s main features which are relevant for tsunami modeling include: the ability to handle inundation (wet-dry interfaces), implementing a well-balanced scheme to suppress spurious motion induced by topographic variations even when the momentum is non-zero, and entropy corrections [16]. One of the unique features of GeoClaw is the ability to adaptively refine the computational grid to resolve particular dynamical features as they evolve in time (the surface height disturbance in the present case). GeoClaw implements the adaptive schemes in [17, 18] and described in detail in [19] for the case of tsunami modeling. The refinement thresholds of [1] were adopted: four levels of refinement were used starting with a resolution of 1 degree in both the latitudinal and longitudinal directions down to 0.5 resolution (approximately 900 meters) located around the observation locations. The tolerance for the refinement criteria for sea-surface anomaly was 0.02 meters.  
\end{quote}
It would be useful if more details are given about the adaptive procedure since it tracks the inundation front.

\item Lines 124-127:
\begin{enumerate}
    \item move the time-reference and the description of the buoy recording protocol to the figure caption as they detract from the main narrative.
    \item rewrite the buoy recording protocol, e.g.  The buoys record SSH every 15 minutes in normal operating conditions; once an earthquake is reported they record SSH every 15 seconds during the first few minutes of the even, and every minute after that (how long...).
\end{enumerate}
\item Lines 128 - remove the word ``generated''
\item Lines 129–131 - awkward, suggested:  The large spikes reported by gauge 21418 at later time are data anomalies and not subsequent tsunami waves as observed in [1].  The text as written is confusing, it is not clear whether reference [1] reported the spikes as data anomalies or though they were tsunamis. If it is the former the ``as observed in [1]'' should be moved to the beginning of the sentence.  If these are indeed data anomalies were they included in the inversion, and if not would the PC machinery have flagged them as anomalies?.
\item Line 142–145 - This sentence seems to contradict the previous one where the Ammon model was praised for including tsunami data. Please clarify.
\item Section 3 the end of the 1st paragraph and the beginning of the second paragraphs have several interdependent sentences. I would reorganize them to flow more naturally and logically. For example, and starting at line 155–167:  In either case this process introduces uncertainty in model output, most importantly in the forecasted water elevation in the most important at-risk zones, namely the near-shore. MacInnes et al., for instance, found that adopting a uniform value of n = 0.025 everywhere leads to poor performance of the forward models predicted inundation levels. To address this they used n = 0.035 in the Sendai plane to account for land-uses such as pasture, farmland, and rice paddies. We aim to address such uncertainties systematically by quantifying their impact on water elevations.  We seek to quantify the parametric uncertainty in Mannings friction coefficient by characterizing its spatial variability and its depth-dependence; we thus define distinct regions in the domain that are characterized by a single coefficient.
\item Line 188 - missing reference.
\item Line 192 - I thought the response surface was connected with the PC expansion and not with the Bayesian inference. I would remove the parenthetical remark as it raises more question than conveys information.
\item Line 196 - Change ``computer'' to ``compute''
\item Line 192 - delete ``the''
\item subsection Formalism: I would avoid introducing more notation in the paper when explaining Bayesian inference. So instead of a 0 use directly $n = (n_1 , n_2 , n_3 )$ and instead of $d$ use whatever notation is appropriate for surface elevation. I would also urge the authors to compress this section as much as possible and to strengthen its links to other sections (by avoiding extra notations for example).
\item Line 153 - delete "accelerating Bayesian inference in this context by” and replace the ``for'' by ``to''.
\item Line 154 - misplaced commas ``, which, in addition'' should be ``which, in addition,''.
\item polynomial chaos subsection The authors should either not use $\xi$ or explain how it is related to n.
\item Line 296 ``Using NISP the stochastic integrals are solved '' should read ``The stochastic integrals are then approximated''
\item Line 332: ``input uncertain parameters'' should be ``uncertain input parameters''
\item Line 335: The polynomial degree was never mentioned.
\item line 347: ``the plots of curves for the selected realizations'' should be ``the curves of different realizations''
\item line 349: It is not clear in which way the onset of variability is consistent with distance.
\item Figures 2 and 3: It is hard to distinguish data for gauges 21413 and 21419 as they both appear green.
Please use a different color for one of the gauges.
\item Figure 4: Can the spikeness in the error curves be explained?
\item Figures 5 and 6: since the different curves are distinguished by color it would be easier on the readers
eyes if a solid line is used instead of a dashed line.
\item Line 385 There is no reason to use QoI since the only model output here is sea surface elevation.
\item Line 389 change the ``where'' to a ``when'', the ``is'' to ``becomes'' and delete the rest of the sentence.
\item Line 391–392 The description ``narrows at a few time instances during the tsunami event'' is a little
awkward, how about ``waxes and wanes as the tsunami evolves''?
\item Line 393–395 Sentence does not really explain the dynamical reason behind the non-monotonic behavior of the standard deviation. The sentence needs to be rewritten, e.g. the waxing is probably caused by waves reflecting off a single source and arriving at the gauge simultaneously while the waning is associated with waves propagating away from the gauges.
\end{enumerate}

\section*{Reviewer 2}
\begin{quote}      
Overview

This paper uses Bayesian inference and polynomial chaos modelling of the Tohoku tsunami to attempt inferences of nearshore, offshore, and onshore Manning's n friction coefficients. The GEOCLAW model is used as training set for the polynomial chaos modelling. The writing quality is quite high and I found few grammatical mistakes.

Results of the study show that the nearshore Manning's n has a MAP estimate of 0.011, and the offshore MAP value is 0.018. No good estimate could be made for Manning's n over land. The nearshore Manning's n value is lower than is used by most models but is not impossible.

The quality of work in this paper shows a good technical level and is clearly presented. Nonetheless, as I was reading it I kept asking: ``how can the authors neglect the effects of source uncertainty and other factors in favor of Manning's n''? It is not until the last paragraph of the conclusions that the authors address this point which to me seems central to the entire inference. Source models are known differ significantly between various inversions. Why do the authors believe that Manning's n is the controlling parameter? The world is full of models that where coefficients are calibrated to account for other processes. Is this happening here? This point needs to be addressed early and rigorously.

Other Points

A comparison is needed between error levels before and after Bayesian inference. It is not possible to determine how much the simulation was improved. This comparison should be clearly made.

Bad reference line 188.

Figure 16 does not have contour plots as stated in the text and caption.

For Figures 3 and 10, it would be better to plot the lines with different styles, not just different colours. Now they are impossible to distinguish when printed on a BW laser printer.
\end{quote}

\section*{Reviewer 3}
\begin{quote}
This paper uses a polynomial chaos (PC) approximation of a tsunami model in order to perform Bayesian estimation of the friction coefficient used in the model. Although the tsunami model is a distributed parameter system, the dimensionality of the estimation problem is reduced by allowing the friction coefficient to take on separate values in an on-shore region, a near-shore region, and in the deep water. The results of the investigation are an analysis of the probability distribution of the a posteriori friction coefficients, and an analysis of sensitivity.

It seems that the main innovation of this paper is the use of a PC expansion for parameter estimation in a water level model. I like that the Results section begins with an error and convergence study of the PC expansion. But I think the over-all approach of the paper may be mis-guided, and, at a minimum, it needs more commentary to justify the approach taken and significance of results.
\end{quote}

\subsection*{Major comments:}
\begin{enumerate}
\item Model justification and setup - Eqn 1: The authors should provide some comments about the applicability of the Manning formula, vs., say, the Chezy formula, for bottom stress in a tsunami model. Since the tsunami in deep water (or on the continental shelf) is not in a state of fully-developed turbulence, neither friction formula is physically justified. Perhaps this fact can be used to develop more appropriate priors for the friction coefficients. At lines, 443-456, the ability of GeoClaw to realistically simulate the water surface elevation is discussed, but I think this should appear earlier in the paper. Lines 451-454 mention important factors, topography and the source model, which should be compared to the influence of the friction coefficients. Lines 454-456 mention the high scatter for gauge 21418 which might justify the use of a different sigma values for this gauge if a representation error is included in sigma. The literature review in the introduction should probably mention older work of the
Dutch groups to infer friction coefficients and bottom depth in water elevation models (e.g., see Lardner et al papers in early 1990's, Heemink et al 2002, etc.).

\item Treatment of the coastline - It seems that the high model resolution is used around the measurement sites, rather than at the coastline where it might be dynamically significant. Could this be related to why $n_1$ was not influential in the results? Although the model includes inundation effects, it seems that the resolution at the coastline was too coarse to include this process.

\item It seems that the Metropolis algorithm is over-used - The response surfaces illustrated in Figure 12 are very smooth. It would seem that a gradient-based search method would be very efficient at finding optimal values of the unknown parameters ($n_1$, $n_2$, $n-3$, and $\sigma$). Cannot the a posteriori pdf's be approximated (Fig 15) using a PC-like quadrature formula? It is because an importance sampling technique is used here that I disagree with the first point concerning ``efficiency'' in the ``Highlights''.

\item Analysis of significance - Perhaps I missed it, but I did not see any evaluation of the a posteriori goodness-of-fit to the DART buoy data. Is the improvement significant compared to the estimated measurement error, sigma? Is the two-parameter fit (using $n_1$ and $n_2$) a statistically significant improvement over a one-parameter fit? Can you incorporate a low-dimensional representation of the source model parameters or topography in your approach? Even a cursory evaluation of the size of these effects would do a lot to justify your emphasis on identification of the friction coefficient.
\end{enumerate}

\subsection*{Minor comments:}
\begin{enumerate}
\item title: ``Buoys Data'' --> ``Buoys'' or ``Buoy Data''
\item Highlights: [see Major comment 3] I do not agree with the first point, that they ``present an efficient approach''.
\item line 99: The units of $C_f$ should be stated and it should be distinguished from the non-dimensional friction coefficient used in other water level model (e.g., POM).
\item lines 113-114: Why was the model resolution refined at the observation locations? Shouldn't it be refined at the coastline or in regions of complex and shallow topography?
\item line 129: Why are data anomalies shown? The spurious data should be omitted.
\item line 150: Manning's n is presented in dimensional form, but the units are missing. I would suggest that you use a non-dimensional presentation; although, I realize that dimensional values are commonly used in the engineering literature.
\item line 188: missing citation
\item line 236: I disagree with the use of a Jeffrey's prior. The accuracy of of DART instruments is known or could be inferred from the harmonic analysis used to de-tide the data.
\item Figure 16, right: caption mentions ``contour plots'' but I see only dots in the figure panels.
\item References: [13] missing date; [14] capitalization; etc.
\end{enumerate}

\bibliographystyle{plain}
\bibliography{database}
\end{document}

### 2014-04-06 friction_paper/response_1.tex
Add response lines, clean up some other stuff

@@
 
-% Commands 
-\include{commands}
+\newcommand{\makesmalltitle}[3]
+{
+\noindent
+\begin{minipage}{1.0\linewidth}
+  {\LARGE \bf 
+    #1}\\[-1.5mm]
+  \noindent\rule{\linewidth}{1pt}\\
+  #2 \hfill \hbox{#3}
+  \\ %[8mm]
+\end{minipage}
+}
+
+\usepackage{color}
+\newcommand{\alert}[1]{{\color{red} #1}}
 
@@
 
-\makesmalltitle{Response to Reviewers of ``Adaptive Mesh Refinement for Storm Surge''}{Revision 1 - \today}
+\makesmalltitle{Response to Reviewers of ``Uncertainty Quantification and Inference of Manning's Friction Coefficient using DART Buoys Data during Tohoku Tsunami''}{Revision 1 - \today}
 
@@
 
+\alert{Needs response.} \\
+
 \subsection*{Detailed Comments}
 \begin{enumerate}
-\item Use either Manning’s coefficient or simply n, not “Manning’s n coefficient”
-\item In abstract fix “T ohoku” to “Tohoku”
+\item Use either Manning’s coefficient or simply n, not “Manning’s n coefficient” \\
+
+\alert{Needs response.} \\
+
+\item In abstract fix “T ohoku” to “Tohoku” \\
+
+\alert{Needs response.} \\
+
 \item Lines 3-8: Suggestion: Coastal communities have faced these hazard by assessing the risk they pose, and by attempting to make informed decisions about their likely impacts.
 Computational models of tsunami events are often employed to explore various tsunami disaster scenarios and their ensuing inundation levels.
-\item Line 24: change “The implemented emulator” to “This emulator”
-\item Line 31: remove the comma after “In a recent work”
-\item Line 33 and 34: Use either “Manning’s coefficient” or n.
+\\
+
+\alert{Needs response.} \\
+
+\item Line 24: change “The implemented emulator” to “This emulator” \\
+
+\alert{Needs response.} \\
+
+\item Line 31: remove the comma after “In a recent work” \\
+
+\alert{Needs response.} \\
+
+\item Line 33 and 34: Use either “Manning’s coefficient” or n. \\
+
+\alert{Needs response.} \\
+
 \item Lines 51-54: Change to: Their main advantage is the efficient propagation of input uncertainties through large, complex and non-linear models to calculate the corresponding output uncertainties.
-\item Line 81: The section title should be changed as it does not reflect the content to something like:  “Simulating the Tohoku Tsunami”
-\item Line 85: Change “The source of the earthquake (epicenter)” to “The epicenter”
-\item Lines 95-99: the sea surface anomaly does not appear anywhere in the equations or the explanation even though it is the primary quantity of interest here.
+\\
+
+\alert{Needs response.} \\
+
+\item Line 81: The section title should be changed as it does not reflect the content to something like:  “Simulating the Tohoku Tsunami” \\
+
+\alert{Needs response.} \\
+
+\item Line 85: Change “The source of the earthquake (epicenter)” to “The epicenter” \\
+
+\alert{Needs response.} \\
+
+\item Lines 95-99: the sea surface anomaly does not appear anywhere in the equations or the explanation even though it is the primary quantity of interest here. \\
+
+\alert{Needs response.} \\
+
 \item Lines 100-115 should be a single paragraph describing GEOCLAW. I suggest the following :
@@
 \end{quote}
-It would be useful if more details are given about the adaptive procedure since it tracks the inundation front.
+It would be useful if more details are given about the adaptive procedure since it tracks the inundation front. \\
+
+\alert{Needs response.} \\
 
@@
 \end{enumerate}
-\item Lines 128 - remove the word ``generated''
+
+\alert{Needs response.} \\
+
+\item Lines 128 - remove the word ``generated'' \\
+
+\alert{Needs response.} \\
+
 \item Lines 129–131 - awkward, suggested:  The large spikes reported by gauge 21418 at later time are data anomalies and not subsequent tsunami waves as observed in [1].
@@
 If these are indeed data anomalies were they included in the inversion, and if not would the PC machinery have flagged them as anomalies?.
-\item Line 142–145 - This sentence seems to contradict the previous one where the Ammon model was praised for including tsunami data. Please clarify.
+\\
+
+\alert{Needs response.} \\
+
+\item Line 142–145 - This sentence seems to contradict the previous one where the Ammon model was praised for including tsunami data. Please clarify. \\
+
+\alert{Needs response.} \\
+
 \item Section 3 the end of the 1st paragraph and the beginning of the second paragraphs have several interdependent sentences.
@@
 We seek to quantify the parametric uncertainty in Mannings friction coefficient by characterizing its spatial variability and its depth-dependence; we thus define distinct regions in the domain that are characterized by a single coefficient.
-\item Line 188 - missing reference.
+\\
+
+\alert{Needs response.} \\
+
+\item Line 188 - missing reference. \\
+
+\alert{Needs response.} \\
+
 \item Line 192 - I thought the response surface was connected with the PC expansion and not with the Bayesian inference.
 I would remove the parenthetical remark as it raises more question than conveys information.
-\item Line 196 - Change ``computer'' to ``compute''
-\item Line 192 - delete ``the''
+\\
+
+\alert{Needs response.} \\
+
+\item Line 196 - Change ``computer'' to ``compute'' \\
+
+\alert{Needs response.} \\
+
+\item Line 192 - delete ``the'' \\
+
+\alert{Needs response.} \\
+
 \item subsection Formalism: I would avoid introducing more notation in the paper when explaining Bayesian inference.
@@
 I would also urge the authors to compress this section as much as possible and to strengthen its links to other sections (by avoiding extra notations for example).
-\item Line 153 - delete "accelerating Bayesian inference in this context by” and replace the ``for'' by ``to''.
-\item Line 154 - misplaced commas ``, which, in addition'' should be ``which, in addition,''.
-\item polynomial chaos subsection The authors should either not use $\xi$ or explain how it is related to n.
-\item Line 296 ``Using NISP the stochastic integrals are solved '' should read ``The stochastic integrals are then approximated''
-\item Line 332: ``input uncertain parameters'' should be ``uncertain input parameters''
-\item Line 335: The polynomial degree was never mentioned.
-\item line 347: ``the plots of curves for the selected realizations'' should be ``the curves of different realizations''
-\item line 349: It is not clear in which way the onset of variability is consistent with distance.
-\item Figures 2 and 3: It is hard to distinguish data for gauges 21413 and 21419 as they both appear green.
-Please use a different color for one of the gauges.
-\item Figure 4: Can the spikeness in the error curves be explained?
-\item Figures 5 and 6: since the different curves are distinguished by color it would be easier on the readers
-eyes if a solid line is used instead of a dashed line.
-\item Line 385 There is no reason to use QoI since the only model output here is sea surface elevation.
-\item Line 389 change the ``where'' to a ``when'', the ``is'' to ``becomes'' and delete the rest of the sentence.
-\item Line 391–392 The description ``narrows at a few time instances during the tsunami event'' is a little
-awkward, how about ``waxes and wanes as the tsunami evolves''?
+\\
+
+\alert{Needs response.} \\
+
+\item Line 153 - delete "accelerating Bayesian inference in this context by” and replace the ``for'' by ``to''. \\
+
+\alert{Needs response.} \\
+
+\item Line 154 - misplaced commas ``, which, in addition'' should be ``which, in addition,''. \\
+
+\alert{Needs response.} \\
+
+\item polynomial chaos subsection The authors should either not use $\xi$ or explain how it is related to n. \\
+
+\alert{Needs response.} \\
+
+\item Line 296 ``Using NISP the stochastic integrals are solved '' should read ``The stochastic integrals are then approximated'' \\
+
+\alert{Needs response.} \\
+
+\item Line 332: ``input uncertain parameters'' should be ``uncertain input parameters'' \\
+
+\alert{Needs response.} \\
+
+\item Line 335: The polynomial degree was never mentioned. \\
+
+\alert{Needs response.} \\
+
+\item line 347: ``the plots of curves for the selected realizations'' should be ``the curves of different realizations'' \\
+
+\alert{Needs response.} \\
+
+\item line 349: It is not clear in which way the onset of variability is consistent with distance. \\
+
+\alert{Needs response.} \\
+
+\item Figures 2 and 3: It is hard to distinguish data for gauges 21413 and 21419 as they both appear green. \\
+
+\alert{Needs response.} \\
+
+Please use a different color for one of the gauges. \\
+
+\alert{Needs response.} \\
+
+\item Figure 4: Can the spikeness in the error curves be explained? \\
+
+\alert{Needs response.} \\
+
+\item Figures 5 and 6: since the different curves are distinguished by color it would be easier on the readers eyes if a solid line is used instead of a dashed line. \\
+
+\alert{Needs response.} \\
+
+\item Line 385 There is no reason to use QoI since the only model output here is sea surface elevation.\\
+
+\alert{Needs response.} \\
+
+\item Line 389 change the ``where'' to a ``when'', the ``is'' to ``becomes'' and delete the rest of the sentence.\\
+
+\alert{Needs response.} \\
+
+\item Line 391–392 The description ``narrows at a few time instances during the tsunami event'' is a little awkward, how about ``waxes and wanes as the tsunami evolves''?\\
+
+\alert{Needs response.} \\
+
 \item Line 393–395 Sentence does not really explain the dynamical reason behind the non-monotonic behavior of the standard deviation.
 The sentence needs to be rewritten, e.g.
-the waxing is probably caused by waves reflecting off a single source and arriving at the gauge simultaneously while the waning is associated with waves propagating away from the gauges.
+the waxing is probably caused by waves reflecting off a single source and arriving at the gauge simultaneously while the waning is associated with waves propagating away from the gauges.\\
+
+\alert{Needs response.}
+
 \end{enumerate}
@@
 \section*{Reviewer 2}
+\subsection*{Overview}
 \begin{quote}      
-Overview
-
 This paper uses Bayesian inference and polynomial chaos modelling of the Tohoku tsunami to attempt inferences of nearshore, offshore, and onshore Manning's n friction coefficients.
@@
 This point needs to be addressed early and rigorously.
+\end{quote}
 
-Other Points
+\alert{Needs response.} \\
 
-A comparison is needed between error levels before and after Bayesian inference. It is not possible to determine how much the simulation was improved. This comparison should be clearly made.
+\subsection*{Other Points}
+\begin{enumerate}
+\item A comparison is needed between error levels before and after Bayesian inference. It is not possible to determine how much the simulation was improved. This comparison should be clearly made.\\
 
-Bad reference line 188.
+\alert{Needs response.} \\
 
-Figure 16 does not have contour plots as stated in the text and caption.
+\item Bad reference line 188.\\
 
-For Figures 3 and 10, it would be better to plot the lines with different styles, not just different colours. Now they are impossible to distinguish when printed on a BW laser printer.
-\end{quote}
+\alert{Needs response.} \\
+
+\item Figure 16 does not have contour plots as stated in the text and caption. \\
+
+\alert{Needs response.} \\
+
+\item For Figures 3 and 10, it would be better to plot the lines with different styles, not just different colors. Now they are impossible to distinguish when printed on a BW laser printer.\\
+
+\alert{Needs response.} \\
+
+\end{enumerate}
 
@@
 
+\alert{Needs response.} \\
+
 \subsection*{Major comments:}
@@
 The literature review in the introduction should probably mention older work of the
-Dutch groups to infer friction coefficients and bottom depth in water elevation models (e.g., see Lardner et al papers in early 1990's, Heemink et al 2002, etc.).
+Dutch groups to infer friction coefficients and bottom depth in water elevation models (e.g., see Lardner et al papers in early 1990's, Heemink et al 2002, etc.).\\
+
+\alert{Needs response.} \\
 
@@
 Could this be related to why $n_1$ was not influential in the results?
-Although the model includes inundation effects, it seems that the resolution at the coastline was too coarse to include this process.
+Although the model includes inundation effects, it seems that the resolution at the coastline was too coarse to include this process.\\
+
+\alert{Needs response.} \\
 
@@
 Cannot the a posteriori pdf's be approximated (Fig 15) using a PC-like quadrature formula?
-It is because an importance sampling technique is used here that I disagree with the first point concerning ``efficiency'' in the ``Highlights''.
+It is because an importance sampling technique is used here that I disagree with the first point concerning ``efficiency'' in the ``Highlights''.\\
+
+\alert{Needs response.} \\
 
@@
 Can you incorporate a low-dimensional representation of the source model parameters or topography in your approach?
-Even a cursory evaluation of the size of these effects would do a lot to justify your emphasis on identification of the friction coefficient.
+Even a cursory evaluation of the size of these effects would do a lot to justify your emphasis on identification of the friction coefficient.\\
+
+\alert{Needs response.} \\
+
 \end{enumerate}
@@
 \begin{enumerate}
-\item title: ``Buoys Data'' --> ``Buoys'' or ``Buoy Data''
-\item Highlights: [see Major comment 3] I do not agree with the first point, that they ``present an efficient approach''.
-\item line 99: The units of $C_f$ should be stated and it should be distinguished from the non-dimensional friction coefficient used in other water level model (e.g., POM).
-\item lines 113-114: Why was the model resolution refined at the observation locations? Shouldn't it be refined at the coastline or in regions of complex and shallow topography?
-\item line 129: Why are data anomalies shown? The spurious data should be omitted.
+\item title: ``Buoys Data'' --> ``Buoys'' or ``Buoy Data''\\
+
+\alert{Needs response.} \\
+
+\item Highlights: [see Major comment 3] I do not agree with the first point, that they ``present an efficient approach''.\\
+
+\alert{Needs response.} \\
+
+\item line 99: The units of $C_f$ should be stated and it should be distinguished from the non-dimensional friction coefficient used in other water level model (e.g., POM).\\
+
+\alert{Needs response.} \\
+
+\item lines 113-114: Why was the model resolution refined at the observation locations? Shouldn't it be refined at the coastline or in regions of complex and shallow topography?\\
+
+\alert{Needs response.} \\
+
+\item line 129: Why are data anomalies shown? The spurious data should be omitted.\\
+
+\alert{Needs response.} \\
+
 \item line 150: Manning's n is presented in dimensional form, but the units are missing.
-I would suggest that you use a non-dimensional presentation; although, I realize that dimensional values are commonly used in the engineering literature.
-\item line 188: missing citation
-\item line 236: I disagree with the use of a Jeffrey's prior. The accuracy of of DART instruments is known or could be inferred from the harmonic analysis used to de-tide the data.
-\item Figure 16, right: caption mentions ``contour plots'' but I see only dots in the figure panels.
-\item References: [13] missing date; [14] capitalization; etc.
+I would suggest that you use a non-dimensional presentation; although, I realize that dimensional values are commonly used in the engineering literature.\\
+
+\alert{Needs response.} \\
+
+\item line 188: missing citation\\
+
+\alert{Needs response.} \\
+
+\item line 236: I disagree with the use of a Jeffrey's prior. The accuracy of of DART instruments is known or could be inferred from the harmonic analysis used to de-tide the data.\\
+
+\alert{Needs response.} \\
+
+\item Figure 16, right: caption mentions ``contour plots'' but I see only dots in the figure panels.\\
+
+\alert{Needs response.} \\
+
+\item References: [13] missing date; [14] capitalization; etc.\\
+
+\alert{Needs response.} \\
+
 \end{enumerate}

### 2014-04-07 friction_revised/highlights.tex
created folder for revised version of the paper

 \documentclass[12pt]{article}
\usepackage{xspace} % Incorporates correct optional space after commands
\newcommand{\geoclaw}{{\sc GeoClaw}\xspace}
\newcommand{\tohoku}{T\={o}hoku\xspace}

\begin{document}

\begin{itemize}
\item  We present an efficient approach for the inference
of the Manning's $n$ coefficient.
\item  We propose a three-parameter representation of Manning's n coefficient.
\item We use Polynomial Chaos to build an inexpensive surrogate for the \geoclaw model.
\item We use Bayesian inference to infer the parameters using data from \tohoku tsunami.
\item We report \textit{pdfs} and Maximum-A-Posteriori values of the uncertain parameters.
\end{itemize}
\end{document}

### 2014-04-07 setrun.py
Revert to Ammon source

@@ -409,12 +409,12 @@ def setgeo(rundata):
     dtopo_data = rundata.dtopo_data
     # for moving topography, append lines of the form:  (<= 1 allowed for now!)
     #   [topotype, minlevel,maxlevel,fname]
-    # dtopodir = os.path.expandvars('$SRC/tohoku2011-paper1/sources/')
-    # dtopo_data.dtopofiles.append([1, 4, 4,
-    #                                    os.path.join(dtopodir, 'Ammon.txydz')])
-    dtopodir = os.getcwd()
+    dtopodir = os.path.expandvars('$SRC/tohoku2011-paper1/sources/')
     dtopo_data.dtopofiles.append([1, 4, 4,
-                                       os.path.join(dtopodir, 'saito.xyzt')])
+                                       os.path.join(dtopodir, 'Ammon.txydz')])
+    # dtopodir = os.getcwd()
+    # dtopo_data.dtopofiles.append([1, 4, 4,
+    #                                    os.path.join(dtopodir, 'saito.xyzt')])
 
 
     # == setqinit.data values ==

### 2014-04-07 friction_paper/revised/paper.tex
Move friction paper related dirs into paper dir

%&pdflatex
%
%
%
%

% \documentclass[12pt]{article}
\documentclass[review,12pt]{elsarticle}

\usepackage{graphicx}
\usepackage{bm}
\usepackage{natbib}
\usepackage{xspace} % Incorporates correct optional space after commands
\usepackage{hyperref}

% Add line numbers
\usepackage[mathlines]{lineno} % Control line numbers

% Multipart figures
\usepackage{subcaption}  % "Newest" subfigure like package

% \oddsidemargin 0in
% \evensidemargin 0in
% \textwidth= 6.5in
% \topmargin -0.50in
% \textheight 9in

\usepackage{ifpdf} % For declaring rules for graphics below

\usepackage{latexsym,amssymb,amsmath,color}
% \newcommand{\revised}[1]{\textcolor{blue}{{#1}}}
% \newcommand{\comment}[1]{\textcolor{red}{{#1}}}
% \newcommand{\alert}[1]{\textbf{\textcolor{red}{{#1}}}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\xxi}{\vec{\xi}}
\newcommand{\xxii}{\xxi^{(\scriptscriptstyle{-}\scriptstyle{i})}}
\newcommand{\var}[1]{{\mathrm{Var}}\left[ {#1} \right]}
\newcommand{\ip}[2]{\left( {#1}, {#2} \right)}
\newcommand{\U}{\mathcal{U}}
\newcommand{\normim}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega^{*})}}
\newcommand{\avemu}[1]{\mathrm{E}\left({#1}\right)}
\newcommand{\ave}[1]{\left\langle {#1} \right\rangle}
\newcommand{\prob}[1]{\mathrm{Prob}\left\{ {#1} \right\}}
\newcommand{\ind}[1]{\mathrm{\chi}_{\scriptscriptstyle {#1} }}
\newcommand{\NISP}{\mathcal{S}}
\newcommand{\ipmu}[2]{\left( {#1}, {#2} \right)_\mu}
\newcommand{\norm}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega)}}
\newcommand{\pard}[2]{\frac{\partial{#1}}{\partial{#2}}}
\def\xbold{{\vec{x}}}
\renewcommand{\vec}[1]{{\mathchoice
                     {\mbox{\boldmath$\displaystyle{#1}$}}
                     {\mbox{\boldmath$\textstyle{#1}$}}
                     {\mbox{\boldmath$\scriptstyle{#1}$}}
                     {\mbox{\boldmath$\scriptscriptstyle{#1}$}}}}

\newcommand{\pd}{\pard}
\newcommand{\geoclaw}{{\sc GeoClaw}\xspace}
\newcommand{\tohoku}{T\={o}hoku\xspace}
\newcommand{\dx}{\ensuremath{\Delta x}}   % Delta x
\newcommand{\dy}{\ensuremath{\Delta y}}   % Delta y
\newcommand{\dt}{\ensuremath{\Delta t}}   % Delta t

\journal{Ocean Modelling}

\begin{document}

\ifpdf
\DeclareGraphicsExtensions{.pdf, .png, .jpg, .tif}
\else
\DeclareGraphicsExtensions{.png, .jpg, .tif, .eps}
\fi

\begin{frontmatter}

\title{Uncertainty Quantification and Inference of Manning's Friction Coefficient using DART Buoys Data during
\tohoku Tsunami}
\date{\today}

\author[kaust1]{Ihab Sraj}
\author[ut]{Kyle T. Mandli}
\author[duke,kaust2]{Omar M. Knio}
\author[ut]{Clint N. Dawson}
\author[kaust1,kaust2]{Ibrahim Hoteit\corref{cor1}}
\ead{ibrahim.hoteit@kaust.edu.sa}
%\ead{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
\address[kaust1]{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}

\address[ut]{Institute for Computational Engineering and Science, University of Texas at Austin, 201 E 24th ST. Stop C0200, Austin, TX 78712-1229, USA}
\address[duke]{Department of Mechanical Engineering and Materials Science, Duke University, 144
Hudson Hall, Durham, North Carolina 27708, USA}
\address[kaust2]{Division of Computer, Electrical and Mathematical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
\cortext[cor1]{Corresponding author}
\date{\today}

\begin{abstract}

Tsunami computational models are employed to explore multiple flooding
scenarios and to predict water elevations.  However, accurate estimation of
water elevations  requires accurate estimation of many model parameters, in
particular the Manning's $n$ friction coefficient.  Our objective is to
develop an efficient approach for the  uncertainty quantification and inference
of the Manning's $n$ coefficient which we characterize here by three different
parameters as being constant over three different regions: $n_1$ on-shore,
$n_2$ near-shore and $n_3$ deep-water.  We use Polynomial Chaos (PC) to build an
inexpensive surrogate for the  \geoclaw model and then employ Bayesian inference
to estimate  and quantify the uncertainties to those uncertain parameters using
the DART buoy data collected during the \tohoku tsunami.  The surrogate model
significantly reduces the computational burden of the Markov Chain Monte-Carlo
(MCMC)  sampling of the Bayesian inference. The PC surrogate is also used to
perform a sensitivity analysis.  Our results indicate that Manning's $n$
coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and
$n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined from the
available data.

\end{abstract}

\begin{keyword}
tsunami \sep Manning's $n$ friction coefficient \sep sensitivity analysis \sep polynomial chaos \sep Bayesian inference
\end{keyword}

\end{frontmatter}
\linenumbers

% \input{cover}
% \clearpage

%!TEX root = paper.tex

\section{Introduction} \label{sec:intro}

Tsunamis in the past decade have been responsible for some of the most  deadly
and costly natural disasters ever recorded. Coastal communities have faced this
hazard by assessing the risk they pose, attempting  to make informed decisions
about the likelihood that such an event would  occur.
Toward this goal, computational models of  tsunami events are often
employed to explore multiple scenarios and to predict the water elevation
caused by the tsunamis. The accurate prediction  of water elevations, however,
requires accurate estimation of many model parameters that are either measured
directly, defined empirically, or estimated from a collection of observed data.
Unfortunately, since tsunami events are relatively rare,  there is a large
uncertainty in the input data for these computational  models ranging from
effects of the domain, such as bathymetry and friction parameterizations,  to
the earthquake source.  In this study, we aim to estimate the Manning's $n$
friction coefficient, a parameterization of the effect of bottom friction 
commonly used in tsunami models.  We aim at quantifying the uncertainty in the 
predicted water elevation  and employ a Bayesian inverse modeling approach to 
estimate the Manning's $n$ coefficient using observations of water elevation 
measured during a tsunami event.

Previous work concerning uncertainty quantification in tsunami modeling has been primarily 
focused on constraining the likely earthquake source models (see
for example \cite{MacInnes:2013cr}).  A similar approach for landslide-generated
tsunami was presented in Sarri \emph{et al.}~\cite{Sarri2012}, by building a statistical surrogate
using an emulator. The implemented emulator was based on a Gaussian
process that requires using a combination of prior knowledge about the simulator
and appropriate choices of functions and parameters. Their work was presented as
a proof-of-concept case study where they only performed basic statistical and
sensitivity analysis.  Changes to the friction parameterization has been shown in 
the past to lead to significant changes in predicted inundation levels \cite{Myers:2001el, 
Jakeman:2010hk}, but these previous studies were limited in range and spatial variation of the values used in the parameterization.  
In a recent work, Mayo \emph{et al.}~\cite{Mayo:2014} reformulated a statistical data assimilation method
generally used in the estimation of model states to estimate the Manning's $n$
coefficient. They used a low-dimensional representation of
Manning's $n$ coefficients and recovered it by assimilating synthetic water elevation data using tidal simulations.

In this work, we focus our attention on the uncertainty in the friction
parameterization used in tsunami simulations. We present a three-parameter 
representation of Manning's $n$ coefficient and an efficient method
to estimate this representation using measured water elevation data. The method proposed here 
is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
probing of the sensitivity of realistic tsunami  models to uncertain parameters
without modifying the forward model and inferring those parameters
from a number of observations.  In particular, we implement an inverse modeling
approach that relies on a Bayesian inference technique.  We resort to
Polynomial Chaos (PC) expansions to build a surrogate model ~\cite{Najm:2009,Alexanderian2012,Elsheikh2014,Young2013}
of the response of the tsunami predicted using the \geoclaw model.
The PC surrogate enables us to avoid using the forward model in
the Bayesian inference, thus dramatically reducing the cost of sampling the
posterior distribution.

PC expansions have been developed and applied in the engineering community to
quantify uncertainties in numerical simulations.  The main advantage of
using PC expansions is the ability of efficiently propagating input uncertainties through 
large, complex and non-linear models leading to the corresponding output uncertainties. 
Several studies already investigated the efficiency of PC expansions for 
UQ in oceanic simulations~\citep{thacker2012,ashwanth2010,Alexanderian2012,sraj:2013a}. 
Recently, PC was combined with Bayesian inference to estimate the wind drag coefficient 
using temperature data collected during Typhoon Fanapi 2010. The same problem was 
solved using two different techniques: the adjoint method and Markov Chain Monte Carlo
(MCMC)\cite{sraj:2013a,sraj:2013b}.

In this work, we follow a similar procedure as in \cite{sraj:2013a,sraj:2013b}
to estimate Manning's $n$ coefficient using the \tohoku tsunami
event as a case-study. The quantity of interest is taken to be the water surface
elevation and is compared to the available NOAA Deep-ocean Assessment and
Reporting Tsunamis (DART) buoys.  The forward model runs are performed using the
\geoclaw package following the setup in MacInnes \emph{et al.}~\cite{MacInnes:2013cr}.

The structure of this manuscript is as follows. First, we describe our case-study, the \tohoku tsunami,
summarize the forward simulations setup, and present the DART buoys observational data
in Section~\ref{sec:tohoku}. Section~\ref{sec:manning} discusses the uncertainties 
of the Manning's $n$ coefficient to motivate the UQ problem and propose a three-parameter representation of $n$. 
Section~\ref{sec:formu} presents the two main methods employed in the UQ analysis:
Bayesian inference that brings together observations 
and model results (Section~\ref{sec:inference}) 
and PC expansions that are used to build the surrogate model (Section~\ref{sec:uqpce}). Section~\ref{sec:results} describes the main results in 
three subsections: first, we perform error and convergence studies
as evidences that the PC surrogate is a faithful 
representation of the \geoclaw model; second, we use the PC surrogate to 
perform a statistical analysis and compute the  sensitivity of simulated water elevations
to the uncertain input parameters; and third, we present the results of the
posterior distributions obtained using Bayesian inference by MCMC. A discussion of our findings and methodology is finally presented in Section~\ref{sec:conc}.

%\comment{
%\begin{itemize}
%    \item There are two places where we refer to the parameterization as three-dimensional, I am not sure if this is really three-dimensional.  I know that we are using depth contours but the parameterization is two-dimensional.  I have a feeling that readers would take issue with this.
%    \item Ihab: the representation is 3D meaning we used three different values of n in three regions. The same concept was presented in Mayo paper where they did 2D representation.
%    \item I think the order here needs to be revised to talk about the forward model and the equations we are solving first since the Manning's $n$ discussion refers to equations there. 
%    \item Ihab: Our main objective is to show the UQ techniques and how they work. The equations are known and we do not need to stress on them. We can even remove the whole equation or forward model section.
%\end{itemize}
%}

%!TEX root = paper.tex
\section{The \tohoku Tsunami} 
\label{sec:tohoku}

The \tohoku earthquake and tsunami of 2011 have been the subject of a number of
studies due to the wealth of observational evidence and severity of the tsunami~\cite{Kawase2014}.
The earthquake itself had an estimated magnitude of $\text{M}_\text{w}$ 9.0 and
caused massive devastation across Japan. The source of the earthquake (epicenter)
was located approximately 72 km east of \tohoku as indicated 
in Figure~\ref{fig:ceofs}. In our case study, the
simulation used is based on work done in \cite{MacInnes:2013cr} and modified to 
use the specified variable friction field. Hereafter, we briefly describe the
forward computational model, \geoclaw, the  simulation setup used in the
uncertainty analysis, and finally present the observations from NOAA DART buoys.

\subsection{\geoclaw} \label{ssub:geoclaw}
The \geoclaw software package uses a finite volume, wave-propagation approach as
described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water
equations (SWE):

\begin{equation} \label{eq:swe}
    \begin{aligned}
    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu, \\
    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv,
    \end{aligned}
\end{equation} 
where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
longitudinal and latitudinal directions respectively, $g$ the acceleration due 
to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
friction coefficient.  
%As is common in tsunami modeling, $C_f$ is calculated 
%given a Manning's $n$ parameterization such that:
%\begin{equation}
%    C_f = \frac{g n^2}{h^{4/3}},
%\label{eq:coef}
%\end{equation}
%where $n$ is an empirically determined parameter known as Manning's $n$.

The primary computational kernel in \geoclaw is the evaluation of the solution
to the Riemann problem at each grid cell interface.  The Riemann solver used
includes the ability to handle inundation (wet-dry interfaces), well-balancing,
even when the momentum is non-zero, and entropy corrections
\cite{George:2008aa}.

One of the unique features of \geoclaw is the ability to adaptively refine the
grids used for the computation, essentially following a region of interest as
time progresses (in this case, the disturbance in the surface height).
To this end, \geoclaw implements the schemes in \cite{Berger:1984ui,Berger:1998aa}
as described in detail in \cite{Berger:2011du} in the case of tsunami
modeling.

For the simulations presented, refinement thresholds were used that matched what
was presented in \cite{MacInnes:2013cr}.  As was done there, four levels of
refinement were used starting with a resolution of 1 degree in both the
latitudinal and longitudinal directions down to 0.5' resolution (approximately
900 meters) located around the observation locations.  The tolerance for the
refinement criteria for sea-surface anomaly was 0.02 meters.

\subsection{DART Buoy Observations}

The observations used in this study are from the Deep-ocean Assessment and
Reporting of Tsunamis (DART) buoy system developed and maintained by the
National Oceanic and Atmospheric Administration (NOAA).  The purpose of the
network is to provide early-warning detection and forecasting of tsunami
propagation in the Pacific Ocean \cite{Milburn:1996wm}.  The DART buoys closest
to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418,
and 21419 whose locations are shown in Figure~\ref{fig:setup_buoy_locations} and
whose de-tided water surface elevation data for the event are shown in
Figure~\ref{fig:setup_buoy_data}. The time $t=0$ is set to be the start of the
earthquake and is recorded every 15 seconds during the first few minutes of the
event, otherwise recording every minute when triggered (a tsunami has occurred)
or every 15 minutes when not.  We can see that the primary tsunami
waves generated reach the gauges between 40 minutes and 2 hours after the
earthquake.  The large spikes later in time present in the data at gauge 21418
are data anomalies and not subsequent tsunami waves as observed in \cite{MacInnes:2013cr}.

\subsection{Bathymetry and Earthquake Source Model}

The bathymetry used in the simulation is a combination of ETOPO 1' and 4'
accurate bathymetry for the region \cite{Amante:2009ud}.  Finer bathymetry was
used in the study done in MacInnes \emph{et al.} in order to match the on-shore
inundation levels.  In the current study this bathymetry was excluded since we
are only concentrating on the data from the DART buoys. The bathymetry and 
topography of the considered regions are shown in Figure~\ref{fig:setup_buoy_locations}.

The earthquake source model that was used is due to Ammon \emph{et al.}
\cite{Ammon:2011dm} who used inversion of the seismic waves to reconstruct the
rupture.  This model was chosen as it provides a rupture reconstruction based on
established techniques and did include tsunami inversion data.  The
latter may introduce bias into the reported results of this study and was consequently ignored.
As discussed in MacInnes \emph{et al.}, the source models used do not include
rupture timing which could have an impact on the accuracy of the simulation.

\section{Manning's $n$ Friction Coefficient} \label{sec:manning}

Manning's $n$ law provides a relationship between the friction coefficient
$C_f$ appearing in the SWE governing tsunamis (Equation \ref{eq:swe}) and 
the empirically derived roughness parameter $n$:

\begin{equation}
    C_f = \frac{g n^2}{h^{4/3}},
\label{eq:coef}
\end{equation}
The value of $n$ may vary both spatially and temporally due to changes in the
bottom roughness from wave action and other processes.  For tsunami modeling it
is a common practice to assume a single value of Manning's $n$ for the areas
that are initially below sea-level; while satellite images and 
land-use data are sometimes used to determine the coefficient in areas above sea-level.  
Both above and below sea-level, this process has several sources of uncertainty, all
of which can have an impact on the quantities of interest in tsunami modeling.   
MacInnes \emph{et al.}, for instance, found that adopting a uniform value 
of $n = 0.025$ everywhere leads to poor performance of the forward model's predicted 
inundation levels.  To address this they used $n = 0.035$ in the Sendai plane to 
account for land-uses such as pasture, farmland, and rice paddies.  We aim to address such
uncertainties in a systematic way in order to better study risk assessments.

The direct result of the uncertainty in the Manning's $n$ coefficient leads to
uncertainties in water elevation,
especially in the most crucial area in risk assessments, the near-shore. In 
addition,  the prediction of this and other quantities of interest is often highly 
sensitive to changes in these uncertain parameters.  Therefore, we seek to 
quantify the parametric uncertainty due to the uncertainty in the Manning's $n$ 
friction coefficient by characterizing the spatial variability using
iso-baths to define distinct regions in the domain that are characterized by a
single Manning's $n$ coefficient.  In this study three regions are used, an
on-shore region (initially above sea-level), a near-shore region (between
sea-level and the 200 meter iso-bath), and the deep-water region (deeper than
200 meters), whose Manning's $n$ coefficient is denoted by $n_1$, $n_2$, and $n_3$ respectively 
as shown in Figure~\ref{fig:ceofs}. To quantify the uncertainties
in these three parameters we employ Polynomial Chaos (PC) 
expansions as used in \cite{sraj:2013a,sraj:2013b}.  The method and its 
application to our problem is
described in Section~\ref{sec:uqpce}.  To represent the initial uncertainty in the
Manning's $n$ coefficients an uninformative uniform distribution was assumed as 
the prior with physically relevant values of $n$ sampled from the interval 
$n \in [0.005-0.2]$.
\section{Formulation}
\label{sec:formu}
In this section, we first present the Bayesian approach to inverse problems 
and its application to our problem of inferring the Manning's $n$ coefficients.
We also introduce
a Polynomial Chaos method used for forward uncertainty propagation, and 
for accelerating the Bayesian inference.
\subsection{Bayesian Inference}
 \label{sec:inference}

Bayesian inference is a statistical approach to inverse problems
that has recently gained great interest in different applications, including
ocean~\citep{Alexagnderian2011a,Zedler2012,sraj:2013a}
climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
The key idea in this approach is to express all forms of uncertainty
in terms of canonical random variables. The first step in Bayesian inference 
is to formulate the forward problem (response surface/model) using 
a suitable likelihood function and a product of conditional probability densities. 
The prior distributions over the unknown parameters of the model
are then formulated using the best available knowledge of the parameters.  Given some observation data, Bayes rule 
is then used to computer the posterior distribution for these unknown parameters
\citep{sivia}. We briefly review this approach below.
\subsubsection{Formalism}

Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs. We consider a forward model $\vec G$ that predicts the data as function of 
the parameters such that:

\begin{equation}
\vec d \approx \vec{G}( \vec \theta).
\end{equation}
Applying the Bayes' rule yields:

\begin{equation}
 \Pi(\vec{\theta}| \vec d) \propto 
 L(\vec d | \vec{\theta}) \ q(\vec{\theta}), 
\label{eq:bayes}
\end{equation}
where $q(\vec{\theta})$ is the prior of $\vec{\theta}$, representing the \emph{a priori} knowledge
about the parameters; 
$L(\vec d| \vec{\theta})$ is the likelihood function representing
the probability of obtaining the data given the set of parameters $\vec{\theta}$;
and finally $\Pi(\vec{\theta}| \vec d)$ is the posterior,
representing the probability of occurrence of $\vec{\theta}$ given the data $\vec d $.

To formulate the likelihood function, we let $\vec \epsilon = \vec d - \vec{G}$
represent the discrepancy between the model and observations.
Here, the components of $\vec \epsilon $ are assumed to be i.i.d. random variables with density $p_{\epsilon}$.
The likelihood function can thus be written as:

\begin{equation} 
L(\vec d |  \vec{\theta}) 
= 
\prod_{i=1}^m  
p_\epsilon (d_i - G_i(\vec \theta)), 	
\label{eq:likelihood}
\end{equation}
where $m$ is the number of observations (dimension of $\vec{d}$).

In our application, we assume that the errors $\epsilon_i$ are independent
and normally distributed with mean
zero and variance $\sigma^2$, i.e. $\epsilon_i \sim N(0,\sigma^2)$. 
While in general $\sigma^2$ (the variance of the noise in the measured data) depends on the observations, in our case we assume that the error
amplitude is generally small and does not change throughout space and time.
Consequently, one may assume a single $\sigma^2$ value in Equation~\eqref{eq:likelihood},
resulting in:

\begin{equation} 
L(\vec d |  \vec{\theta}) 
= 
\frac{1}{\sqrt{2 \pi \sigma^2}}\prod_{i=1}^m   
\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace. 	
\label{eq:likelihood2}
\end{equation}
The joint posterior from Bayes' rule can thus be given as:

\begin{equation} 
\Pi(\vec{\theta}| \vec d)
\propto
\frac{1}{\sqrt{2 \pi \sigma^2}}   \prod_{i=1}^m  
\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace  
\prod_{i=1}^n q(\theta_i).
\end{equation}
Since $\sigma^2$ is unknown \emph{a priori} we treat it as a hyper-parameter.
In other words, $\sigma^2$ becomes an additional parameter for Bayesian inference and  
therefore endowed with a prior which is updated based on available observations. In this 
case the joint posterior is finally expressed as:

\begin{equation} 
\Pi(\vec{\theta},\sigma^2 | \vec d)
\propto
\frac{1}{\sqrt{2 \pi \sigma^2}}   \prod_{i=1}^m  
\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace
\ q(\sigma^2) \prod_{i=1}^n q(\theta_i).
\label{eq:post}
\end{equation}

\subsubsection{Inferring Manning's Friction Coefficient}
\label{sec:infmanning}

We seek to infer the Manning's $n$ coefficients from water surface elevation
data measured at different gauges as shown in Figure~\ref{fig:setup_buoy_data}.
Bayesian inference can be applied directly to infer the uncertain parameters
based on Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their model predicted counterparts $\vec G$ is represented by \geoclaw.
The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
three Manning's coefficients treated as the unknown parameters and assumed independent. 
This enables us to rewrite Equation~\eqref{eq:post} as:

\begin{equation} 
\Pi(n_1,n_2,n_3,\sigma^2 | \vec d) 
\propto \frac{1}{\sqrt{2 \pi \sigma^2}} 
 \prod_{i=1}^m  
\exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
\ q(\sigma^2)q(n_1)q(n_2) q(n_3).
\label{eq:post_coef}
\end{equation}

To determine the posterior, we need to choose proper priors that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for all three Manning's $n$ parameters is assumed, with $n_i$ in the range  $[0.005 - 0.2]$; therefore:

\begin{equation} 
q(n_i) = \begin{cases}
		\displaystyle \frac{1}{0.2-0.005} &\text{for~} 0.005 <  n_i \leq 0.2 ,  \\
		0 &\text{otherwise}.
\end{cases}
\end{equation}
Regarding the noise variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia} for $\sigma^2$, expressed as:

\begin{equation} 
q(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}.
\label{eq:var_pr}
\end{equation}

Inferring the coefficients amounts to 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC algorithm~\citep{Gareth2009,Haario2001} to
accurately and efficiently sample the posterior distribution.

%!TEX root = paper.tex
\subsection{Accelerating Bayesian Inference}
\label{sec:uqpce}

The four-dimensional posterior in Equation~\eqref{eq:post_coef} can be directly
explored via MCMC; this requires repeated simulations (tens of thousands) of the forward \geoclaw model, 
once for every proposed set of parameters of the Markov chain~\cite{MarzoukNajm2009,Malinverno2002}. 
While a single \geoclaw simulation takes about $15~$mins, depending on the details of the MCMC algorithm used, 
it is desirable to avoid running the forward model at every realization of the MCMC. 
This is achieved by constructing a surrogate model that requires a much
smaller ensemble of \geoclaw runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
Polynomial Chaos expansions for accelerating Bayesian inference in this context 
by building a surrogate model, which, in addition can efficiently
provide statistical properties, such as the mean, variance and sensitivities.

\subsubsection{Polynomial Chaos}

Polynomial Chaos (PC) is a probabilistic methodology that expresses the 
dependencies of model outputs on the uncertain model inputs
as a truncated polynomial expansion~\citep{Villegas2012,Lin2009,Xiu2004}.
We briefly describe the PC method below; for more details 
the reader is referred to \citep{LeMaitreKnio2010}.

Let $U=U(\bm{x},t,\xxi)$ denote a quantity of 
interest (QoI) that is the output of a computational model.
$U$ is function of space, $\bm{x}$, and time, $t$, and 
also depends on the canonical vector of random variables $\xxi=(\xi_1,...,\xi_n)$
that parameterize the uncertain inputs. 
PC expresses $U$ in the form:

\begin{equation}
  U(\xbold,t,\xxi) \doteq \sum_{k = 0}^P U_k(\xbold,t) \Psi_k(\xxi),
\label{eq:stochseries}
\end{equation} 
where $U_k(\xbold,t)$ are the polynomial coefficients, and
$\Psi_k(\xxi)$ are elements of an orthogonal basis of an underlying probability
space. The total number of terms in the truncated PC expansion is
$P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimensions and $p$ is the highest order
polynomial retained.

The choice of the basis is dictated by the probability density
function of the stochastic vector $\xxi$, which appears as a weight
function in the probability space's inner product:

\begin{equation}
 \left<\Psi_i,\Psi_j\right> = \int \Psi_i(\xxi) \;\Psi_j(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi=\delta_{ij}\ave{\Psi_i^2},
\label{eq:inner}
\end{equation}
where $\delta_{ij}$ is the Kronecker delta.
For uniform
distributions, the basis functions are scaled Legendre polynomials.
For multi-dimensional problems the basis functions are
tensor products of 1D basis functions~\cite{LeMaitreKnio2010}.

The series representation \eqref{eq:stochseries} can be viewed as a spectral expansion
of $U$ along the stochastic dimensions. It can also be seen as a
combination of approximation and probabilistic frameworks; this
 has proven extremely useful in solving UQ problems~\cite{Xiu:2003,Lin2009}. The existence and convergence of this series is asserted by the Cameron-Martin theorem \citep{Cameron:1947} with the condition that $U$ has a finite variance.
The series rate of convergence, and hence the number of terms to retain, depends on the smoothness of
$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
when $U$ is smooth; the convergence rate becomes algebraic
when $U$ has finite smoothness \citep{Canuto:2006}. In practice the series convergence is monitored 
via various error metrics as discussed in the results section.

\subsubsection{Non Intrusive Spectral Projection (NISP)}
The computation of the coefficients of the PC expansions $U_k$
can be done using a number of procedures. Here we adopt a non-intrusive
approach that allows the use of the forward model \geoclaw as a black box
with no code modifications required. PC expansion coefficients are determined
based on a set of response \geoclaw simulations at a specified set of the uncertain parameters. 
Specifically, we rely on the Non-Intrusive Spectral Projection (NISP) method that exploits the orthogonality of the basis and applies the Galerkin projection to find the PC expansion coefficients as follows:

\begin{equation}
 U_k(\bm{x},t) = \frac{\left< U, \Psi_k \right>}{\left< \Psi_k, \Psi_k \right>} = 
 \frac{1}{\left< \Psi_k, \Psi_k \right>} 
 \int U(\bm{x},t,\xxi) \Psi_k(\xxi) \rho(\xxi) \mbox{ d}\xxi.
\end{equation}
This orthogonal projection minimizes the $L_2$ error on the space spanned by the basis.
Using NISP the stochastic integrals are solved using a numerical quadrature to obtain:

\begin{equation}
  \left< U, \Psi_k \right> 
\approx \left< U, \Psi_k \right>_{\cal Q}
= \sum_{q=1}^{Q} U(\xxi_q) \Psi_k(\xxi_q) \omega_q,
\end{equation}
where the subscript $\cal Q$ refers to approximating the inner product integral with
quadrature, and $\xxi_q$ and $\omega_q$ are multi-dimensional quadrature points and weights,
respectively and $Q$ is the number of nodes in the multi-dimensional quadrature. The quadrature order should be commensurate with the
truncation order, and should be high enough to avoid aliasing artifacts.
The choice of quadrature rule is hence critical to the performance
of the PC (in its NISP version at least).

The computation of the ${U}_k$ can thus be expressed as a matrix-vector product of the form:

\begin{equation} 
 U_k(\bm{x},t)=\sum_q \Pi_{kq} U(\bm{x},t,\xxi_q),\;\;\;
 {\cal P}_{kq}=\frac{\Psi_k(\xxi_q)\omega_q}{\left< \Psi_k, \Psi_k \right>},
\label{eq:nisp}
\end{equation} 
where ${\cal P}_{kq}$ is the projection matrix and $U(\bm{x},t,\xxi_q)$ is obtained
from an ensemble of the deterministic model realizations with the uncertain parameters set at
the quadrature value $\xxi_q$.

\subsubsection{Statistical moments and sensitivity analysis}
The identification of the inner product weight function
with the probability distribution of $\xxi$ simplifies the calculations of the statistical moments of $U$. 
Noting that since $\Psi_0(\xxi)$ is a constant that is normalized so that 
$\left<\Psi_0,\Psi_0\right>=1$, the expectation and variance of $U$ can be computed as:

\begin{equation}
 E[U] = \int U \, \rho(\xxi) \, \mbox{d}\xxi \approx \left< U,\Psi_0\right>_{\cal Q} = U_0,  
 \label{eq:mean}
\end{equation}
and

\begin{equation}
 E[(U-E[U])^2] = \int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi \approx \sum_{k=1}^P U_k^2
 \label{eq:sigma}
\left<\Psi_k,\Psi_k\right>.
\end{equation}

PC representations also enable
efficient global sensitivity analysis that quantify the
contributions of different random input parameters to the variance in the output.
This can be done by computing the so-called {\it total} 
sensitivity index $T_i$ that measures the contribution of
the $i^{th}$ random input to total model variability by
computing the fraction of the total variance due to all the terms in the
PC expansion that involve $\xi_i$~\citep{LeMaitreKnio2010,Crestaux,Sudret}
as follows:

% of random variables from the PC representations or Sobol decomposition~\citep{Sobol:1993,Homma:1996,Sobol:2001}. The total sensitivity index  
%
%To get the total sensitivity corresponding to the uncertain
%input $\xi_i$ we compute the total index:

\begin{equation} \label{eq:T-hard}
   T_i =
         \frac{\displaystyle
               \sum_{k \in K_i} U_k^2 \ave{\Psi_k^2}}
              {\displaystyle\sum_{k = 1}^P U_k^2 \ave{\Psi_k^2}},
\end{equation}
where

\[
   K_i = \left\{ k \in \{1, \ldots, P\} :
           \vec{\alpha}^k_i > 0 \right\},
        \]
and $\vec{\alpha}^k$ is the multi-index associated with the $k^{th}$ term in the
PC expansion~\cite{LeMaitreKnio2010}.

%Using Equation~\eqref{eq:T-hard}, the computation of $T_i$ is straightforward.

\section{Results}
\label{sec:results}
We begin by performing an error and convergence 
analysis of the constructed surrogate using the PC expansions to 
establish its validity. We then present a statistical analysis to quantify the 
uncertainty in the predicted water surface elevation, as well as a 
sensitivity analysis that enables us to rank the impact of the 
different Manning's $n$ coefficients on the uncertainty. Finally, 
we present the results of the inverse problem 
where we determine the posterior distributions of the input uncertain
parameters and analyze them in light of the available gauge data.
\subsection{Error and convergence study}
\label{sec:analysis}

In the current work, we employ a tensorized Gaussian quadrature to construct the PC surrogate
model. To this end, 125 \geoclaw simulations were used to compute the PC coefficients using Equation~\ref{eq:nisp}.
The computational cost of running this ensemble is low compared to the cost of the thousands of 
runs of the forward model that would have been required if a standard Monte-Carlo uncertainty propagation technique was to be adopted.  Figure~\ref{fig:rlzs} plots the evolution of the
water surface elevation predicted at the four different gauges 
using \geoclaw for the 125 different realizations 
required to compute the PC expansions. We notice that the 
variability in water surface elevation is insignificant in the first 
hour at all gauges as the plots of the different realizations superimpose.
Later in time, the variability starts to increase for all gauges 
as indicated by the thickness of the bands formed by the plots of curves for the selected realizations.
This variability appears to be significant at $t\sim$ 5400~s, 6000~s, 3600~s and 7200~s
for gauges 21401, 21413, 21418, 21419, respectively.
This is consistent with the distance from the gauge to the epicenter of the earthquake located approximately 72 kilometer east of Japan where gauge 21418 is the closest and gauge 21419 is the farthest as shown in Figure~\ref{fig:setup_buoy_locations}.
The uncertainty in the prediction of water surface  elevation persists till the end of the simulations
at all gauges.

In order to check the consistency of the PC approximation, we compare
water surface elevation from the realizations 
with those obtained from the PC surrogate. The different curves (not shown) 
reveal excellent agreement for all times and gauge locations.
To quantify this agreement, we define
an error metric that measures the relative normalized root mean-square error between the left hand side function 
in Equation~\eqref{eq:stochseries} and its PC representation at the sampling points:

\begin{equation} 
   E = \frac{\displaystyle
         \left(\sum_{\xxi \in \NISP} \left|U(\xxi) - \sum_{k = 0}^{P}
U_k\Psi_k(\xxi)\right|^2
         \right)^{1/2}}
        {\displaystyle
          \left(\sum_{\xxi \in \NISP} \left|U(\xxi)\right|^2\right)^{1/2} 
          },
\label{eq:error}
\end{equation}
where $\NISP$ is the 125-member ensemble obtained to construct the PC surrogate. 
This error metric calculated at the different gauge locations, is shown in Figure~\ref{fig:error};
the largest relative normalized error for 
water surface elevation is about 0.1\%.

Another check for the validity of the PC approximation 
consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC expansion~\cite{Alexanderian2012,sraj:2013a}.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3} where
the different curves correspond to increased order of PC ($p= 1-5$).
The plots indicate double peaked distributions that are
well-resolved with PC order $p=3$ at t = 7200 s as shown in Figure~\ref{fig:pdfs2}.
However, at t = 10800 s further refinement is needed as the pdfs are
sensitive to the refinement up to order $p=4$ but then becomes weakly 
insensitive with additional refinement  ($p = 5$) as shown in Figure~\ref{fig:pdfs3}. 
We therefore used PC order $p = 5$ in all computations below.
This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 
model surrogate that can be used in both the forward and inverse problems.

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
The PC expansion created using an ensemble of 125 \geoclaw simulations
is now used as a surrogate to propagate prior input parameter uncertainty 
through the forward model.  Here we exploit the surrogate 
to study the statistics of the water surface elevation, and to conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.

\subsubsection{Statistical analysis}
The mean of the QoI and its standard deviation is computed
from the PC coefficients as indicated in Equations~\eqref{eq:mean} and \eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the mean water surface elevation along with two standard deviations
bounds at the four gauges.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant,
reflected in a significant standard deviation as seen earlier from the realizations shown 
in Figure~\ref{fig:rlzs}. An interesting observation is that the
standard deviation in water surface elevations narrows at a few time instances
during the tsunami event.  The narrowing of the variance at these instances is
possibly due to the waves that arrive due to reflections from a single source
and then move away from the gauge location, and thus no variance in the water
surface elevation.

% or from a region where Manning's $n$ values have little influence over these
%waves.

The predicted mean water surface elevation  and its standard deviation 
can be used for comparison with the (DART) buoy observations collected during
the tsunami event. Figure~\ref{fig:compare} 
plots the PC mean water surface elevation with the observed 
water surface elevation at the different gauges locations
with variances indicated at few time instances. 
The plots show a good agreement between the PC simulations and the 
observations at different locations and at different times.

The same statistical analysis can be performed for the
entire domain in 2D. Figure~\ref{fig:mean2d} (top row) shows
the PC mean water surface elevation for the considered computational
domain at three different times as indicated in the title of each panel.
The standard deviation is also shown in Figure~\ref{fig:mean2d} (bottom row)
at different times. We clearly notice the propagation of the variance
due to the parametric uncertainty with time
and along the tsunami's reflections.

\subsubsection{Sensitivity analysis}
A global sensitivity analysis is performed to quantify the contribution of each
uncertain parameter to the variance in water surface elevation. To this end, we calculate 
the total sensitivity index using the PC coefficients as shown in Equation~\eqref{eq:T-hard}~\citep{Alexanderian2012,Sudret,Crestaux}. The evolution of the total sensitivity index
of each of the uncertain parameters is shown in Figure~\ref{fig:sens} at the four gauges. 
The Manning's $n$ coefficient at the shore $n_2$ is clearly dominant and contributes
the most to the variance in the water surface elevation compared to the other two 
Manning's $n$ coefficients $n_1$ and $n_3$; this is true for almost the entire simulation time
and at the four gauges. The Manning's $n$ coefficient in the deep ocean $n_{3}$ at gauge number 21419 exhibits a higher sensitivity index during the first hour of simulation however, as noted previously, none of the gauge locations see a significant variance between simulations until after the first hour of simulation.
The on-shore Manning's $n$ coefficient, $n_{1 }$, has no effect on the variance in
the water surface elevation for the entire simulation time at all gauges.

In 2D, the sensitivity analysis shows also that $n_2$ is dominant
at the same regions where we observe significant variance in water surface elevation. This is
indicated from Figure~\ref{fig:sens2d} that shows the total sensitivity index
for $n_1$ (top row) $n_2$ (center row) and $n_3$ (bottom row)
at different times.

\subsubsection{Response surface}
In addition to the statistical and sensitivity analysis, the PC surrogate 
can be used to construct a response surface for the uncertain input parameters.
This is achieved by sampling the PC surrogate for different values of the 
canonical vector of random variables  $\xxi$ within the prior
range. Since $n_1$ shows insignificant contribution to the 
uncertainty in the model output, we only consider the response surface
of $n_2$ versus $n_3$ for a fixed value of $n_1=0.035$ as illustrated in Figure~\ref{fig:response2}
at $t=7200~s$. The most striking features are the relatively flat
horizontal contours in the $n_3$ direction suggesting that water surface elevation depends
only mildly on $n_3$ even during peak tsunami events. This is true for gauges 21401, 21413 and 21418. However,
at gauge 21419, the horizontal contours are observed in the $n_2 $ direction
meaning that water surface elevation depends only mildly on $n_2$.
\subsection{Inverse Problem} 
\label{sec:inverse}

Before we attempt to solve the inverse problem of parameter estimation, 
we first check the ability of \geoclaw to realistically simulate water 
surface elevation. To this end, we present a comparison between the 
(DART) buoy observations and their \geoclaw model counterparts
for a reference simulation in which the the Manning's $n$ coefficients were set to their default values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
shows a scatter plot that compares the observed 
water surface elevation at the different gauge locations and the \geoclaw model counterparts. The plots show a reasonable agreement between the simulation and the observations.  The differences between the simulation and observations can be attributed to uncertainty in the Manning's $n$ coefficients, errors in the earthquake rupture model, and insufficiently accurate bathymetry in the near-shore region.  The high amount of scatter from gauge 21418 can probably be attributed to its proximity to the epicenter of the earthquake and shore region.

Bayesian inference is now used to estimate the Manning's 
$n$ coefficient's ranges that minimize this scatter between 
observed and modeled water surface elevations and to estimate the variance of the noise in the measured data.
To this end, an adaptive MCMC method is implemented to sample 
the posterior distributions \citep{Gareth2009,Haario2001} and consequently 
update the Manning's $n$ coefficients distributions in light of the 
observed data. This sampling requires tens of thousands of 
\geoclaw model runs that are prohibitively expensive as each MCMC 
sample requires an independent \geoclaw realization. Instead,
the surrogate model created using PC expansions provides a computationally
efficient alternative that requires only evaluating the PC expansion
for different values of the canonical vector of random variables $\xxi$.

The MCMC algorithm required $50,000$ iterations to guarantee convergence and to
obtain the posteriors of the Manning's $n$ coefficients:  $n_1$, $n_2$ and $n_3$
as well as for the noise variance $\sigma^2$. Figure \ref{fig:mcmc}  shows the sample
chains for the input parameters for different iterations of the MCMC algorithms.
The different panels show well-mixed chains for all input parameters as well as
the noise variance $\sigma^2$. The chain of  $n_{1}$ spans the entire range of the
prior, meaning that the observations are not informative  concerning this input.
In contrast, the $n_{2}$ chain appears to be concentrated in the  lower end of
the parameter range, with values between 0.005 and 0.1. For $n_{3}$, the chain
appears to be concentrated in the  upper end of the parameter range. These values align well with what is
commonly considered physically relevant for near-shore and deep-water regions.
Finally, the chain for the noise variance  appears to be well mixed with a well defined
posterior range between 0.02 and 0.022.

Next, the computed MCMC chains are used to determine the marginalized posterior 
distributions using kernel density estimation (KDE)~\citep{Parzen1962,Silverman1986}.  
The resulting marginalized posterior pdfs of the three Manning's $n$ coefficients 
$n_1,n_2,n_3$ are shown in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
As expected from the chains shown in Figure~\ref{fig:mcmc}, the marginalized posterior 
pdfs of $n_1$ appear to be fairly flat, and similar to the uniform prior; an indication 
that the observed data were not useful to refine our prior knowledge for $n_1$. In 
contrast, the posterior pdf of $n_2$ exhibits a well-defined peak, with a Maximum A 
Posteriori (MAP) estimate of around 0.011 and an extended tail towards the larger 
Manning's $n$ coefficient values. For $n_3$, we observe a posterior that has a well 
defined peak of 0.18 but no clear pdf shape. The posterior distribution of the noise variance 
appears to be well-defined of Gaussian-like shape. The MAP value of $\sigma^2=0.021$ 
can be used to estimate the water surface elevation standard deviation that was found to 
be 0.145~$m$. This value is a reflection of the mismatch between the model and 
observed data.

We finally investigate the existence of any correlation arising between the inferred parameters.
%This study is motivated by the fact that the Manning's $n$ coefficients have an important role 
%in characterizing the tsunami model predictions. 
It is indeed instructive to analyze the 
correlation between the various $n$ parameters to gain some insight into their mutual dependence.
To this end, we compute the two-dimensional marginalized joint posteriors for each 
possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $\Pi(n_1, n_2), \Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. Figure~\ref{fig:joint} (left) shows the scatter plots of the MCMC samples of the combinations of 
inferred parameters obtained from the chain samples of the marginalized posterior excluding 
the burn-in iterations. The plots reveal no clear correlation in all three cases. 
These samples can be readily exploited to reconstruct the joint density for each combination of parameters. 
The right column of Figure~\ref{fig:joint} shows the contour plots of the marginalized joint posteriors 
$\Pi(n_1, n_2)$, $\Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. The contours are obtained from the
distribution reconstructed via KDE, and depict values that are greater than $10\%$
of the maximum probability.

To quantify the degree of no-correlation between each possible pair of parameters, we compute 
the corresponding correlation coefficient which is a unitless scalar that falls in the range 
$(-1, 1)$, with $1$ indicating a perfect positive correlation, $−1$ a perfect negative correlation and $0$ for
a no-correlation case.
% $R(n_2, n_3)$, $R(n_2, \sigma^2)$, and $R(n_3, \sigma^2)$.
The correlation coefficient, $R(X, Y )$, between two random variables $X$ and $Y$ is defined as:

\begin{equation}
R(X,Y) = \frac{E[(X - \mu_X)(Y - \mu_Y )]}{\sigma_X \sigma_Y} 
\end{equation}
where $\sigma_X$ and  $\sigma_Y$ are the standard deviations of  $X$ and $Y$, $\mu_X$
and $\mu_Y$  are the means of  $X$ and $Y$ respectively;
and $E$ is the expectation operator.  In our case, the correlation coefficient
was in the range $(-0.002 , 0.06)$ for all possible combinations of parameters; again indicating no correlation
between the inferred parameters.

%!TEX root = paper.tex
\section{Discussion and Conclusions}
\label{sec:conc}

The present study aimed at estimating Manning's $n$ friction coefficient  which
plays an important role in the accurate prediction of water surface elevation in
tsunami modeling. We proposed a three-parameter representation of Manning's $n$
friction coefficient that was characterized using iso-baths  to define three
distinct regions, on-shore, near-shore, and deep-water, in the domain and
represented each region with a single Manning's $n$ coefficient.  The estimation
relied on a Bayesian inference approach that sharpens the initial estimates of
the three uncertain parameters using real observations.

In our test case, the \tohoku tsunami, we used four DART buoy gauges that
provide  surface elevation information.  To accelerate the Bayesian inference,
we relied on the polynomial  chaos expansions to produce a faithful and
efficient surrogate of the forward model \geoclaw.  This PC surrogate model was
then used to quantify the uncertainties in the predicted water surface
elevations due to the uncertainties in the Manning's $n$ coefficient.  This
included the mean and standard deviation of water surface elevations which were
also compared to the measured buoy data.  A global sensitivity analysis was also
performed in order to quantify the contribution of each uncertain parameter to
the variance in surface elevation.  It was found that the Manning's $n$
coefficient in the near-shore region $n_2$ contributed the most to the variance
in the surface elevations compared to the other two Manning's $n$ coefficients
in the on-shore and deep-water regions.  This is expected as the deep-water
friction has little effect on the overall column.  On the other hand, the
sensitivity of the analysis to the on-shore value being less than that in the
deep-water was unexpected. We suspect that this is primarily due to the use of the
DART buoy gauges, being far enough from land regions to have a significant impact. 
This might not be the case if data from inundation zones (that
were not available to us) were used.

Using Bayesian inference, MAP estimates were found for the three region's $n$
parameters using MCMC.  These values were (excluding $n_1$ where no meaningful
MAP value was found), $n_2=0.011$ and $n_3=0.180$.  The values and their
corresponding marginalized PDFs tell perhaps the most
complete story in this analysis.  As mentioned earlier, $n_2$ is the most
sensitive and has the most clear MAP value.  Although the value was lower than
expected, the tail includes the most commonly used values (0.022-0.025).  The
deep-water value $n_3$ has a peak that is much higher than expected but a tail
that does not taper off particularly quickly.  This is not altogether unexpected
due to the low sensitivity of the QoI to this region's friction.  The
on-shore value $n_1$ shows again how insensitive the used simulations and
observations were to on-shore values in our particular setting.  
In conclusion, it is clear that the use of only the DART buoy observations and the 
bathymetry resolution used here is not sufficiently informative to invert for the 
Manning's $n$ values with any strong confidence, especially outside of the near-shore region.

Finally, it should be noted that the friction source term is only one of a
number of other sources of uncertainty.  In particular, the rupture model and
bathymetry might often have an important influence on the tsunami.  
In the future, this type of analysis would be more illustrative if
propagation of uncertainties in both the bathymetry and source models were
included in each Manning's $n$ estimate.  The
present study focused on formulating and estimating a low-dimensional
representation of the Manning's $n$ coefficient using UQ techniques, namely
Bayesian inference and PC expansions.  A high-dimensional representation of the
Manning's $n$ coefficient would require a large number of forward runs and this remains
computationally expensive.  We will work on alternative methods to reduce the
dimensionality of the problem and this will be the objective of a future study.

\bibliographystyle{elsarticle-num}
\begin{thebibliography}{10}
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
\expandafter\ifx\csname href\endcsname\relax
  \def\href#1#2{#2} \def\path#1{#1}\fi

\bibitem{MacInnes:2013cr}
B.~T. MacInnes, A.~R. Gusman, R.~J. LeVeque, Y.~Tanioka, {Comparison of
  Earthquake Source Models for the 2011 Tohoku Event Using Tsunami Simulations
  and Near-Field Observations}, Bulletin of the Seismological Society of
  America 103~(2B) (2013) 1256--1274.

\bibitem{Sarri2012}
A.~Sarri, S.~Guillas, F.~Dias,
  \href{http://www.nat-hazards-earth-syst-sci.net/12/2003/2012/}{Statistical
  emulation of a tsunami model for sensitivity analysis and uncertainty
  quantification}, Natural Hazards and Earth System Science 12~(6) (2012)
  2003--2018.
\newblock \href {http://dx.doi.org/10.5194/nhess-12-2003-2012}
  {\path{doi:10.5194/nhess-12-2003-2012}}.
\newline\urlprefix\url{http://www.nat-hazards-earth-syst-sci.net/12/2003/2012/}

\bibitem{Myers:2001el}
E.~P. Myers, A.~M. Baptista, {Analysis of Factors Influencing Simulations of
  the 1993 Hokkaido Nansei-Oki and 1964 Alaska Tsunamis}, Natural Hazards
  23~(1) (2001) 1--28.

\bibitem{Jakeman:2010hk}
J.~D. Jakeman, O.~M. Nielsen, K.~V. Putten, R.~Mleczko, D.~Burbidge,
  N.~Horspool, {Towards spatially distributed quantitative assessment of
  tsunami inundation models}, Ocean Dynamics 60~(5) (2010) 1115--1138.

\bibitem{Mayo:2014}
T.~Mayo, T.~Butler, C.~Dawson, I.~Hoteit, Data assimilation within the advanced
  circulation (adcirc) modeling framework for the estimation of manning’s
  friction coefficient, accepted to Ocean Modelling.

\bibitem{Najm:2009}
H.~Najm, Uncertainty quantification and polynomial chaos techniques in
  computational fluid dynamics, Ann. Rev. Fluid Mech. 41 (2009) 35--52.

\bibitem{Alexanderian2012}
A.~Alexanderian, J.~Winokur, I.~Sraj, A.~Srinivasan, M.~Iskandarani,
  W.~Thacker, O.~Knio,
  \href{http://dx.doi.org/10.1007/s10596-012-9286-2}{Global sensitivity
  analysis in an ocean general circulation model: a sparse spectral projection
  approach}, Computational Geosciences 16 (2012) 757--778,
  10.1007/s10596-012-9286-2.
\newline\urlprefix\url{http://dx.doi.org/10.1007/s10596-012-9286-2}

\bibitem{Elsheikh2014}
A.~H. Elsheikh, I.~Hoteit, M.~F. Wheeler,
  \href{http://www.sciencedirect.com/science/article/pii/S004578251300296X}{Efficient
  bayesian inference of subsurface flow models using nested sampling and sparse
  polynomial chaos surrogates}, Computer Methods in Applied Mechanics and
  Engineering 269~(0) (2014) 515 -- 537.
\newblock \href {http://dx.doi.org/http://dx.doi.org/10.1016/j.cma.2013.11.001}
  {\path{doi:http://dx.doi.org/10.1016/j.cma.2013.11.001}}.
\newline\urlprefix\url{http://www.sciencedirect.com/science/article/pii/S004578251300296X}

\bibitem{Young2013}
K.~C. Young, M.~D. Grace,
  \href{http://link.aps.org/doi/10.1103/PhysRevLett.110.110402}{Simulation of
  stochastic quantum systems using polynomial chaos expansions}, Phys. Rev.
  Lett. 110 (2013) 110402.
\newblock \href {http://dx.doi.org/10.1103/PhysRevLett.110.110402}
  {\path{doi:10.1103/PhysRevLett.110.110402}}.
\newline\urlprefix\url{http://link.aps.org/doi/10.1103/PhysRevLett.110.110402}

\bibitem{thacker2012}
W.~C. Thacker, A.~Srinivasan, M.~Iskandarani, O.~M. Knio, M.~L. Henaff,
  Propagating boundary uncertainties using polynomial expansions, Ocean
  Modelling 43--44 (2012) 52--63.
\newblock \href {http://dx.doi.org/10.1016/j.ocemod.2011.11.011}
  {\path{doi:10.1016/j.ocemod.2011.11.011}}.

\bibitem{ashwanth2010}
A.~Srinivasan, J.~Helgers, C.~B. Paris, M.~LeHenaff, H.~Kang, V.~Kourafalou,
  M.~Iskandarani, W.~C. Thacker, J.~P. Zysman, N.~F. Tsinoremas, O.~M. Knio,
  \href{http://dx.doi.org/10.1109/MTAGS.2010.5699424}{Many task computing for
  modeling the fate of oil discharged from the deep water horizon well
  blowout}, in: Many-Task Computing on Grids and Supercomputers (MTAGS), 2010
  IEEE Workshop on, IEEE, November, 2010, pp. 1--7.
\newblock \href {http://dx.doi.org/10.1109/MTAGS.2010.5699424}
  {\path{doi:10.1109/MTAGS.2010.5699424}}.
\newline\urlprefix\url{http://dx.doi.org/10.1109/MTAGS.2010.5699424}

\bibitem{sraj:2013a}
I.~Sraj, M.~Iskandarani, A.~Srinivasan, W.~C. Thacker, J.~Winokur,
  A.~Alexanderian, C.-Y. Lee, S.~S. Chen, O.~M. Knio, Bayesian inference of
  drag parameters using fanapi axbt data, Monthly Weather Review 141 (2013)
  2347--2367.
\newblock \href {http://dx.doi.org/10.1175/MWR-D-12-00228.1}
  {\path{doi:10.1175/MWR-D-12-00228.1}}.

\bibitem{sraj:2013b}
I.~Sraj, M.~Iskandarani, A.~Srinivasan, W.~C. Thacker, O.~M. Knio, Drag
  parameter estimation using gradients and hessian from a polynomial chaos
  model surrogate, Monthly Weather Review.

\bibitem{Kawase2014}
T.~Hiraishi, N.~Yoneyama, Y.~Baba, R.~Azuma,
  \href{http://dx.doi.org/10.1007/978-4-431-54418-0_4}{Field survey of the
  damage caused by the 2011 off the pacific coast of tohoku earthquake
  tsunami}, in: H.~Kawase (Ed.), Studies on the 2011 Off the Pacific Coast of
  Tohoku Earthquake, Natural Disaster Science and Mitigation Engineering: DPRI
  reports, Springer Japan, 2014, pp. 37--48.
\newblock \href {http://dx.doi.org/10.1007/978-4-431-54418-0_4}
  {\path{doi:10.1007/978-4-431-54418-0_4}}.
\newline\urlprefix\url{http://dx.doi.org/10.1007/978-4-431-54418-0_4}

\bibitem{LeVeque:1997eg}
R.~J. LeVeque, {Wave Propagation Algorithms for Multidimensional Hyperbolic
  Systems}, Journal of Computational Physics 131~(2) (1997) 327--353.

\bibitem{George:2008aa}
D.~L. George, {Augmented Riemann solvers for the shallow water equations over
  variable topography with steady states and inundation}, Journal of
  Computational Physics 227~(6) (2008) 3089--3113.

\bibitem{Berger:1984ui}
M.~J. Berger, J.~Oliger, {Adaptive mesh refinement for hyperbolic partial
  differential equations}, Journal of Computational Physics 52 (1984) 484--512.

\bibitem{Berger:1998aa}
M.~J. Berger, R.~J. LeVeque, {Adaptive Mesh Refinement using Wave-Propagation
  Algorithms for Hyperbolic Systems}, SIAM Journal on Numerical Analysis 35
  (1998) 2298--2316.

\bibitem{Berger:2011du}
R.~J. LeVeque, D.~L. George, M.~J. Berger, {Tsunami Propagation and inundation
  with adaptively refined finite volume methods}, Acta Numerica (2011)
  211--289.

\bibitem{Milburn:1996wm}
H.~B. Milburn, A.~I. Nakamura, F.~I. Gonz{\'a}lez, {Real-time tsunami reporting
  from the deep ocean}, in: Oceans 96 MTS/IEEE Conference, Fort Lauderdale, FL,
  1996, pp. 390--394.

\bibitem{Amante:2009ud}
C.~Amante, B.~W. Eakins, {ETOPO1 1 Arc-Minute Global Relief Model: Procedures,
  Data Sources and Analysis}, Tech. Rep. NGDC-24, NOAA (Mar. 2009).

\bibitem{Ammon:2011dm}
Ammon, {A rupture model of the 2011 off the Pacific coast of Tohoku
  Earthquake}, Earth, Planets and Space 63~(7) (2011) 693--696.

\bibitem{Zedler2012}
S.~Zedler, G.~Kanschat, R.~Korty, I.~Hoteit, A new approach for the
  determination of the drag coefficient from the upper ocean response to a
  tropical cyclone: a feasibility study., Journal of Oceanography 68~(2) (2012)
  227 -- 241.

\bibitem{OlsonEtAl2012}
R.~Olson, R.~Sriver, M.~Goes, N.~M. Urban, H.~D. Matthews, M.~Haran, K.~Keller,
  A climate sensitivity estimate using bayesian fusion of instrumental
  observations and an earth system model, Journal of Geophysical Research 117
  (2012) D04103.

\bibitem{Malinverno2002}
A.~Malinverno,
  \href{http://dx.doi.org/10.1046/j.1365-246X.2002.01847.x}{Parsimonious
  bayesian markov chain monte carlo inversion in a nonlinear geophysical
  problem}, Geophysical Journal International 151~(3) (2002) 675--688.
\newblock \href {http://dx.doi.org/10.1046/j.1365-246X.2002.01847.x}
  {\path{doi:10.1046/j.1365-246X.2002.01847.x}}.
\newline\urlprefix\url{http://dx.doi.org/10.1046/j.1365-246X.2002.01847.x}

\bibitem{sivia}
D.~S. Sivia, Data Analysis - A Bayesian Tutorial, Oxford Science Publications,
  Oxford, 2006.

\bibitem{Gareth2009}
G.~O. Roberts, J.~S. Rosenthal, Examples of adaptive {MCMC}, Journal of
  Computational and Graphical Statistics 18~(2) (2009) 349--367.
\newblock \href
  {http://arxiv.org/abs/http://amstat.tandfonline.com/doi/pdf/10.1198/jcgs.2009.06134}
  {\path{arXiv:http://amstat.tandfonline.com/doi/pdf/10.1198/jcgs.2009.06134}},
  \href {http://dx.doi.org/10.1198/jcgs.2009.06134}
  {\path{doi:10.1198/jcgs.2009.06134}}.

\bibitem{Haario2001}
H.~Haario, E.~Saksman, J.~Tamminen,
  \href{http://www.jstor.org/stable/3318737}{An adaptive metropolis algorithm},
  Bernoulli 7~(2) (2001) 223--242.
\newline\urlprefix\url{http://www.jstor.org/stable/3318737}

\bibitem{MarzoukNajm2009}
Y.~M. Marzouk, H.~N. Najm, Dimensionality reduction and polynomial chaos
  acceleration of bayesian inference in inverse problems, Journal of
  Computational Phyics 228~(6) (2009) 1862--1902.
\newblock \href {http://dx.doi.org/10.1016/j.jcp.2008.11.024}
  {\path{doi:10.1016/j.jcp.2008.11.024}}.

\bibitem{Villegas2012}
M.~Villegas, F.~Augustin, A.~Gilg, A.~Hmaidi, U.~Wever, Application of the
  polynomial chaos expansion to the simulation of chemical reactors with
  uncertainties, Mathematics and Computers in Simulation 82~(5) (2012)
  805--817.
\newblock \href {http://dx.doi.org/10.1016/j.matcom.2011.12.001}
  {\path{doi:10.1016/j.matcom.2011.12.001}}.

\bibitem{Lin2009}
G.~Lin, G.~E. Karniadakis, Sensitivity analysis and stochastic simulations of
  non-equilibrium plasma flow, International Journal for Numerical Methods in
  Engineering 80~(6-7) (2009) 738--766.
\newblock \href {http://dx.doi.org/10.1002/nme.2582}
  {\path{doi:10.1002/nme.2582}}.

\bibitem{Xiu2004}
D.~Xiu, D.~Tartakovsky, Uncertainty quantification for flow in highly
  heterogeneous porous media, in: W.~G.~G. Cass T.~Miller, Matthew W.~Farthing,
  G.~F. Pinder (Eds.), Computational Methods in Water Resources: Volume 1, Vol.
  55, Part 1 of Developments in Water Science, Elsevier, 2004, pp. 695 -- 703.

\bibitem{LeMaitreKnio2010}
O.~P. {Le Ma\^{i}tre}, O.~M. Knio, Spectral Methods for Uncertainty
  Quantification, Springer-Verlag, 2010.

\bibitem{Xiu:2003}
D.~Xiu, G.~Karniadakis, Modeling uncertainty in flow simulations via
  generalized {Polynomial Chaos}, J. Comput. Phys. 187 (2003) 137--167.

\bibitem{Cameron:1947}
R.~H. Cameron, W.~T. Martin, The orthogonal development of non-linear
  functionals in series of fourier-hermite functionals, Ann. Math. 48 (1947)
  385--392.

\bibitem{Canuto:2006}
C.~G. Canuto, M.~Y. Hussaini, A.~Quarteroni, T.~A. Zang, Spectral
  Methods:Fundamentals in Single Domains, Scientific Computation,
  Springer-Verlag, 2006.

\bibitem{Crestaux}
T.~Crestaux, O.~L. Maitre, J.-M. Martinez, Polynomial chaos expansion for
  sensitivity analysis, Reliability Engineering \& System Safety 94~(7) (2009)
  1161 -- 1172, special Issue on Sensitivity Analysis.

\bibitem{Sudret}
B.~Sudret, Global sensitivity analysis using polynomial chaos expansions,
  Reliability Engineering \& System Safety 93~(7) (2008) 964 -- 979, bayesian
  Networks in Dependability.
\newblock \href {http://dx.doi.org/DOI: 10.1016/j.ress.2007.04.002}
  {\path{doi:DOI: 10.1016/j.ress.2007.04.002}}.

\bibitem{Parzen1962}
E.~Parzen, \href{http://www.jstor.org/stable/2237880}{On estimation of a
  probability density function and mode}, The Annals of Mathematical Statistics
  33~(3) (1962) 1065--1076.
\newline\urlprefix\url{http://www.jstor.org/stable/2237880}

\bibitem{Silverman1986}
B.~W. Silverman, {Density estimation: for statistics and data analysis},
  Chapman and Hall, London, 1986.

\end{thebibliography}
\clearpage

\begin{figure}[ht]      
\centering
\includegraphics[width=0.6\textwidth]{Figure1.pdf}
\caption{Manning's $n$ coefficients at three regions: $n_1$ on-shore, $n_2$ near shore, $n_3$ deep-water.}
\label{fig:ceofs}
\end{figure}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.45\textwidth}
    \includegraphics[width=\textwidth]{Figure2a.pdf}
    \caption{} \label{fig:setup_buoy_locations}
\end{subfigure}
\begin{subfigure}[c]{0.45\textwidth}
    \includegraphics[width=\textwidth]{Figure2b.pdf} 
    \caption{} \label{fig:setup_buoy_data}
\end{subfigure}
\caption{(a) The topography, bathymetry and gauge locations used in the simulation. (b) Observed de-tided water surface height at all the DART buoys used.}
\label{fig:setup}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\centering
\begin{tabular}{clc}        
\includegraphics[width=0.6\textwidth]{Figure3.pdf} 
\end{tabular}
\caption{125 \geoclaw realizations at different gauge locations.}
\label{fig:rlzs}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[ht]
\centering
\begin{tabular}{clc}        
\includegraphics[width=0.5\textwidth]{Figure4a.pdf} &
\includegraphics[width=0.5\textwidth]{Figure4b.pdf} \\
\includegraphics[width=0.5\textwidth]{Figure4c.pdf} &
\includegraphics[width=0.5\textwidth]{Figure4d.pdf}

\end{tabular}
\caption{Relative normalized error between realizations and 
the corresponding PC surrogates at different gauge locations
calculated using Equation~(\ref{eq:error}).}
\label{fig:error}
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\centering

\begin{tabular}{clcl}
\includegraphics[width=0.5\textwidth]{Figure5a.pdf} &
\includegraphics[width=0.5\textwidth]{Figure5b.pdf} \\
\includegraphics[width=0.5\textwidth]{Figure5c.pdf} &
\includegraphics[width=0.5\textwidth]{Figure5d.pdf}
\end{tabular}
\caption{pdf of water surface elevation at the different gauge locations at t = 7200 s.}
\label{fig:pdfs2}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\centering
\begin{tabular}{clc}

\includegraphics[width=0.5\textwidth]{Figure6a.pdf} &
\includegraphics[width=0.5\textwidth]{Figure6b.pdf} \\
\includegraphics[width=0.5\textwidth]{Figure6c.pdf} &
\includegraphics[width=0.5\textwidth]{Figure6d.pdf}
\end{tabular}
\caption{pdf of water surface elevation at the different gauge locations at t = 10800 s.}
\label{fig:pdfs3}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\begin{tabular}{clc}

\includegraphics[width=0.475\textwidth]{Figure7a.pdf} &
\includegraphics[width=0.475\textwidth]{Figure7b.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure7c.pdf} &
\includegraphics[width=0.475\textwidth]{Figure7d.pdf}
\end{tabular}
\caption{Evolution of PC mean water surface elevation at different gauge locations.}
\label{fig:ave}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\begin{tabular}{clc}
%        
\includegraphics[width=0.475\textwidth]{Figure8a.pdf} &
\includegraphics[width=0.475\textwidth]{Figure8b.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure8c.pdf} &
\includegraphics[width=0.475\textwidth]{Figure8d.pdf}
\end{tabular}
\caption{Comparison of PC mean 
and observed data of water surface elevation with time at the four gauges.}
\label{fig:compare}
\end{figure}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
        \begin{tabular}{ccc}
\hspace*{-55pt}
\includegraphics[width=0.45\textwidth]{Figure9a.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure9b.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure9c.pdf} \\
\hspace*{-55pt}
\includegraphics[width=0.45\textwidth]{Figure9d.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure9e.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure9f.pdf}
\end{tabular}
\caption{PC mean (top row) and standard deviation (bottom row) of the water surface elevation at different times.}
\label{fig:mean2d}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[ht]
\begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{Figure10a.pdf} &
\includegraphics[width=0.475\textwidth]{Figure10b.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure10c.pdf} &
\includegraphics[width=0.475\textwidth]{Figure10d.pdf}
\end{tabular}
\caption{Total sensitivity index of different input parameters.}
\label{fig:sens}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[ht]
\begin{tabular}{clc}
 \hspace*{-55pt}
\includegraphics[width=0.45\textwidth]{Figure11a.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure11b.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure11c.pdf} \\
\hspace*{-55pt}
\includegraphics[width=0.45\textwidth]{Figure11d.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure11e.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure11f.pdf} \\
\hspace*{-55pt}
\includegraphics[width=0.45\textwidth]{Figure11g.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure11h.pdf} &
\hspace*{-45pt}
\includegraphics[width=0.45\textwidth]{Figure11i.pdf}
\end{tabular}
\caption{Total sensitivity index for $n_1$ (top row) $n_2$ (center row) and $n_3$ (bottom row)
 at different times as indicated.}
\label{fig:sens2d}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\centering

\begin{tabular}{clcl}
\includegraphics[width=0.5\textwidth]{Figure12a.pdf} &
\includegraphics[width=0.5\textwidth]{Figure12b.pdf} \\
\includegraphics[width=0.5\textwidth]{Figure12c.pdf} &
\includegraphics[width=0.5\textwidth]{Figure12d.pdf}
\end{tabular}
\caption{Response surface of water surface elevation at the different gauge locations at t = 7200 s
as function of $n_2$ and $n_3$, with fixed $n_1=0.035$}.
\label{fig:response2}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[ht]
\centering
\includegraphics[width=0.575\textwidth]{Figure13.pdf} 
\caption{Scatter plot of measured water surface elevation against their \geoclaw
model counterparts during Tsunami \tohoku. The data points are colored 
differently for the different gauges as indicated in the legend. }
\label{fig:scatter}

\end{figure}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[ht]
\begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{Figure14a.pdf} &
\includegraphics[width=0.475\textwidth]{Figure14b.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure14c.pdf} &
\includegraphics[width=0.475\textwidth]{Figure14d.pdf}
\end{tabular}
\caption{Chain samples for the three Manning's $n$ coefficients $n_1,n_2,n_3$ and $\sigma^2$
the noise variance. The vertical dotted lines corresponds to the
burn-in iterations.}
\label{fig:mcmc} 
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[ht]
        \begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{Figure15a.pdf} &
\includegraphics[width=0.475\textwidth]{Figure15b.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure15c.pdf} &
\includegraphics[width=0.475\textwidth]{Figure15d.pdf}
        \end{tabular}
        \caption{KDE of the marginalized posterior distributions for the three Manning's $n$ coefficients 
        $\Pi(n_1)$, $\Pi(n_2)$ and $\Pi(n_3)$, 
and the marginalized posterior distribution of the noise variance 
$\Pi(\sigma^2)$.}
\label{fig:pdfs} 
        \end{figure}

%        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[ht]
        \begin{tabular}{clc}
\includegraphics[width=0.475\textwidth]{Figure16a.pdf} &
\includegraphics[width=0.475\textwidth]{Figure16b.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure16c.pdf} &
\includegraphics[width=0.475\textwidth]{Figure16d.pdf} \\
\includegraphics[width=0.475\textwidth]{Figure16e.pdf} &
\includegraphics[width=0.475\textwidth]{Figure16f.pdf}
        \end{tabular}
\caption{Left: Scatter plots of the pairs $(n_1,n_2)$,$(n_1,n_3)$ and $(n_2,n_3)$ obtained from the
chain samples (Figure~\ref{fig:mcmc}). 
Right: Contour plots of the marginalized joint posterior distribution $\Pi(n_1,n_2)$, $\Pi(n_1,n_3)$ and $\Pi(n_2,n_3)$.  Contours of greater than $10\%$ of the maximum probability are shown.}
\label{fig:joint} 
        \end{figure}

\end{document}
